### 2021-07-09 index.Rmd
init commit

--- 
title: "R for Clinical Study Reports and Submission"
author: ["Yilong Zhang"]
knit: "bookdown::render_book"
description: >
  This book will teach you how to prepare clinical statistiacal report and submit to regulatory agency, 
  the essential part of clinical trial development.

site: bookdown::bookdown_site

documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes

---

# Welcome! {-}

<img src="images/cover.jpg" width="100%" align="right" alt="" class="cover" />

Welcome to R for clinical study report and submission. 
Clinical study reports (CSR) are the essential part of clinical trials development. 
a CSR is an "integrated" full scientific report of an individual clinical trial.

The [ICH E3: strcuture and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) 
provide guidance to assist sponsors in the development of a CSR. 
In this book, you will learn how to use R to prepare CSR and submit to regulatory agency.

This is the work-in-progress draft of the book.

### 2021-07-09 intro.Rmd
init commit

# Introduction {#intro}

## Folder structure

In clinical trial development, developer needs to organize source code to deliver SDTM or ADaM datasets and table, listing and figures (TLFs).
For example, a Phase 3 clinical trial typically require hundreds of TLFs. 
A consistent and well-defined folder structure is crucial to manage a clinical trial development project.

We recommended to use R package folder structure to organize all the analysis and reporting related source code and documentation
for a clinical trial development project. 
By using the R package folder structure you can follow the convention to organize your code as other contributors on CRAN.
The consistent approach simplify the communication for developers within and across organizations.

- For a new R developer, it is an essential step to learn R package when you want to share your work with others. 
You will learn one folder structure that is widely used in R community with outstanding tutorial and tools for free.

- For an experienced R developer, there is minimal learning curve.

- For an organization, it simplify process, tool, template, and training development, because a unified folder structure is used to develop and maintain standard tool and analysis project.

The workflow around an R package can also improve the tractability and reproducibility for an analysis project [@marwick2018packaging].

In addition, R package folder structure are also recommended to develop Shiny app as discussed in [Chapter 20 of Master Shiny](https://mastering-shiny.org/scaling-packaging.html) book and [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/golem.html) book.

## In this book

This book is an intermediate level book by assuming reader has equipped some R programming and clinical development knowledge.
The assumption of each part is as below:

- Part 1, "Delivering TLFs in CSR" provides general information with examples to create table, listing and figures. 
In this part, we assume readers are an individual contributor of a clinical project with some experience in R programming. 
Some good reference includes [Hands-On Programming with R](https://rstudio-education.github.io/hopr/) book and 
[R for Data Science](https://r4ds.had.co.nz/) book.

- Part 2, "Clinical Trial Project" provides general information with examples to manage a clinical trial analysis and reporting 
project. In this part, we assume readers are project lead of a clinical project with experience in R package development. 
Some good reference includes [R packages](https://r-pkgs.org/) and [The tidyverse style guide](https://style.tidyverse.org/).

- Part 3, "Submission package in eCTD" provides general information in preparing submission package related to CSR in 
electronic Common Technical Document (eCTD). In this part, we assume readers are project lead of a clinical project 
with experience in R package development as well.

## Philosophy

We follow the same philosophy described in [Section 1.1 of R Packages book](https://r-pkgs.org/intro.html?q=Marwick#intro-phil) and quote here.

- Anything that can be automated, should be automated. 
- Do as little as possible by hand. Do as much as possible with functions. 
- The goal is to spend your time thinking about what you want to do rather than thinking about the minutiae of package structure.

### 2021-07-09 intro.Rmd
update part3 title

@@
 
-- Part 3, "Submission package in eCTD" provides general information in preparing submission package related to CSR in 
+- Part 3, "eCTD Submission package" provides general information in preparing submission package related to CSR in 
 electronic Common Technical Document (eCTD). In this part, we assume readers are project lead of a clinical project 

### 2021-07-09 tlf.Rmd
update template

# (PART) Part Delivering TLFs in CSR {-}

# Reporting

ToDo:

add r2rtf related inforamtion

### 2021-07-30 intro.Rmd
first draft of tlf

@@
 For example, a Phase 3 clinical trial typically require hundreds of TLFs. 
-A consistent and well-defined folder structure is crucial to manage a clinical trial development project.
+A consistent and well-defined folder structure is crucial to manage a clinical trial analysis and reporting (A&R) project.
 
-We recommended to use R package folder structure to organize all the analysis and reporting related source code and documentation
-for a clinical trial development project. 
+We recommend to use R package folder structure to organize all the analysis and reporting related source code and documentation
+for a clinical trial A&R project. 
 By using the R package folder structure you can follow the convention to organize your code as other contributors on CRAN.
@@
 In this part, we assume readers are an individual contributor of a clinical project with some experience in R programming. 
-Some good reference includes [Hands-On Programming with R](https://rstudio-education.github.io/hopr/) book and 
-[R for Data Science](https://r4ds.had.co.nz/) book.
+We expect readers are familiar with data manipulation in R. Some good reference includes [Hands-On Programming with R](https://rstudio-education.github.io/hopr/), 
+[R for Data Science](https://r4ds.had.co.nz/) and [Data Manipulation with R](https://www.amazon.com/Data-Manipulation-R-Use/dp/0387747303).
 
-- Part 2, "Clinical Trial Project" provides general information with examples to manage a clinical trial analysis and reporting 
-project. In this part, we assume readers are project lead of a clinical project with experience in R package development. 
+- Part 2, "Clinical Trial Project" provides general information with examples to manage a clinical trial A&R 
+project. In this part, we assume a reader is a project lead with experience in R package development. 
 Some good reference includes [R packages](https://r-pkgs.org/) and [The tidyverse style guide](https://style.tidyverse.org/).
 
-- Part 3, "eCTD Submission package" provides general information in preparing submission package related to CSR in 
-electronic Common Technical Document (eCTD). In this part, we assume readers are project lead of a clinical project 
+- Part 3, "eCTD Submission package" provides general information in preparing 
+submission package related to clinical study report (CSR) in 
+electronic Common Technical Document (eCTD). 
+In this part, we assume a reader is a project lead of a clinical project 
 with experience in R package development as well.
@@
 
-We follow the same philosophy described in [Section 1.1 of R Packages book](https://r-pkgs.org/intro.html?q=Marwick#intro-phil) and quote here.
+We share the same philosophy described in [Section 1.1 of R Packages book](https://r-pkgs.org/intro.html?q=Marwick#intro-phil) and quote here.
 
-- Anything that can be automated, should be automated. 
-- Do as little as possible by hand. Do as much as possible with functions. 
-- The goal is to spend your time thinking about what you want to do rather than thinking about the minutiae of package structure.
+- "Anything that can be automated, should be automated." 
+- "Do as little as possible by hand. Do as much as possible with functions." 
+- "The goal is to spend your time thinking about what you want to do 
+rather than thinking about the minutiae of package structure."

### 2021-07-30 tlf.Rmd
update title
first draft of tlf

@@
-# (PART) Part Delivering TLFs in CSR {-}
+# (PART) Delivering TLFs in CSR {-}
 
@@
 
-ToDo:
+## Background
 
-add r2rtf related inforamtion
+In clinical trial development, it is a critical step to submit information on medicinal products to regulatory agencies.
+[Electronic Common Technical Document (eCTD)](https://en.wikipedia.org/wiki/Electronic_common_technical_document)
+has become a worldwide regulatory submission standard format. 
+For example, [the United States Food and Drug Administration (US FDA) required new drug applications, biologics license applications must be submitted using eCTD format](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd).
+Clinical Data Interchange Standards Consortium (CDISC) provided a
+[pilot project
+following ICH E3 guidance](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse).
+
+Within eCTD, clinical study reports (CSR) are located at module 5. 
+[ICH E3 guidance](https://www.ich.org/page/efficacy-guidelines) provide 
+compilation of structure and content of clinical study reports.
+
+A typical CSR contain full details of an individual clinical study 
+including clinical and statistical description, presentations, and analyses. 
+A large number of tables and figures are incorporated into the main text and appendices 
+of a CSR.
+In CDISC pilot project, an [example CSR](file:///C:/Users/zhanyilo/OneDrive%20-%20Merck%20Sharp%20&%20Dohme,%20Corp/Desktop/cdiscpilot01.pdf) is also provided.
+If you are interested in more examples of clinical study reports,
+the European Medicines Agency (EMA) publishes clinical data including 
+clinical study reports submitted by pharmaceutical companies 
+on [EMA clincal data website](https://clinicaldata.ema.europa.eu/web/cdp/home).
+
+Building CSR is a team work with clinicians, medical writers, statisticians and statistical programmers.
+Here, We focus on the work and deliverables completed by statisticians and statistical programmers. 
+In an organization, A group of statisticians and statistical programmers commonly work together to 
+define, develop, validate and deliver tables, listings and figures (TLFs) required for a CSR.
+Microsoft Word is widely used to prepare CSR in pharmaceutical industry. 
+Therefore, `rtf`, `doc`, `docx` are commonly used format to deliver TLFs.
+
+In this chapter, our focus is to illustrate how to create tables, listings and figures (TLFs) in RTF format 
+that is commonly used in a CSR. The examples are in compliance of 
+[FDA's Portable Document Format (PDF) Specifications](https://www.fda.gov/media/76797/download)
+
+> FDA's PDF specification is a general reference, each organization may define 
+> more specific TLF format requirement that can be different from the examples in this book.
+
+These TLFs are typically used to summarize efficacy and safety statistical analysis results of a drug (or treatment).
+
+## Table for Baseline Characteristics
+
+```{r}
+library(table1)
+library(r2rtf)
+data(r2rtf_adsl)
+
+tbl <- table1(~ SEX + AGE + RACE | TRT01P, data = r2rtf_adsl)
+tbl
+```
+
+```{r}
+tbl %>% as.data.frame() %>% 
+        rtf_body() %>% 
+        rtf_encode() %>% 
+        write_rtf("tlf/bs-1.rtf")
+```
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+r2rtf:::rtf_convert_format("tlf/bs-1.rtf", output_dir = "tlf")
+knitr::include_graphics("tlf/bs-1.pdf")
+```

### 2021-07-30 tlf.Rmd
update code to transfer PDF

@@
 ```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-r2rtf:::rtf_convert_format("tlf/bs-1.rtf", output_dir = "tlf")
 knitr::include_graphics("tlf/bs-1.pdf")

### 2021-08-05 index.Rmd
update json
update code to transfer PDF
disposition table

@@
 description: >
-  This book will teach you how to prepare clinical statistiacal report and submit to regulatory agency, 
+  This book will teach you how to prepare tables, listings and figures for clinical study report and submit to regulatory agency, 
   the essential part of clinical trial development.
@@
 ---
+
+```{r, include = FALSE}
+library(bslib)
+library(xml2)
+library(downlit)
+```
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE, message = FALSE)
+```
+
+```{r, eval = FALSE, include=FALSE}
+r2rtf:::rtf_convert_format(list.files("tlf", pattern = ".rtf", full.names = TRUE), output_dir = "tlf")
+```
+
+```{r, eval = FALSE, include=FALSE}
+rsconnect::writeManifest(appPrimaryDoc = "index.Rmd", contentCategory = "site")
+```
 

### 2021-08-05 tlf-baseline.Rmd
disposition table

--- 2021-07-09 tlf.Rmd
@@
-# (PART) Part Delivering TLFs in CSR {-}
+## Table for Baseline Characteristics
 
-# Reporting
+```{r}
+library(table1)
+library(r2rtf)
+data(r2rtf_adsl)
 
-ToDo:
+tbl <- table1(~ SEX + AGE + RACE | TRT01P, data = r2rtf_adsl)
+tbl
+```
 
-add r2rtf related inforamtion
+```{r}
+tbl <- tbl %>% as.data.frame() %>% 
+               as_tibble() %>% 
+               mutate( across(everything(), 
+                        ~ stringr::str_replace_all(.x, intToUtf8(160), " ") ) 
+                      )
+```
+
+```{r}
+tbl %>% 
+      rtf_body() %>% 
+      rtf_encode() %>% 
+      write_rtf("tlf/bs-1.rtf")
+```
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/bs-1.pdf")
+```

### 2021-08-05 tlf-disposition.Rmd
disposition table

## Table for Study Patients

In section 10, study patients of a CSR, we need to summarize accounting 
of all patients who entered the study.

```{r}
library(haven)     # Read SAS data
library(dplyr)     # Manipulate data
library(tidyr)     # Manipulate data
library(r2rtf)     # Reporting in RTF format
```

The first step is to read relevant datasets into R. 
For disposition table, all the information is saved in the ADSL dataset.
We can use `haven` package to read the dataset.

```{r}
adsl <- read_sas("adam_data/adsl.sas7bdat")
```

We illustrate how to prepare a report data for a simple disposition of patients table using variables below:

- USUBJID: Unique subject identifier 
- TRT01P: Planed treatment
- TRT01PN: Planned treatment encoding
- DISCONFL: Discontinued from study flag
- DCREASCD: Discontinued from study reason

```{r}
adsl %>% select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>% 
         head()
```
```{r}
library(tidyr)
```

```{r}
n_rand <- adsl %>% group_by(TRT01PN) %>%
                   summarize(n = n()) %>% 
                   pivot_wider(names_from = TRT01PN, names_prefix = "n_", values_from = n) %>% 
                   mutate(row = "Patients in population")
n_rand
```

```{r}
n_disc <- adsl %>% group_by(TRT01PN) %>%
                   summarize(n = sum(DISCONFL == "Y"), 
                             pct = formatC(n / n() * 100, digits = 1, format = "f", width = 5)) %>% 
                   pivot_wider(names_from = TRT01PN, values_from = c(n, pct) ) %>% 
                   mutate(row = "Discontinued")
n_disc
```

```{r}
n_reason <- adsl %>% group_by(TRT01PN) %>%
                     mutate(n_total = n()) %>%
                     group_by(TRT01PN, DCREASCD) %>%
                     summarize(n = n(),
                               pct = formatC(n / unique(n_total) * 100, digits = 1, format = "f", width = 5)) %>% 
                     pivot_wider(id_cols = DCREASCD, names_from = TRT01PN, 
                                 values_from = c(n, pct), values_fill = list(n = 0, pct = "  0.0") ) %>% 
                     rename(row = DCREASCD)
n_reason
```

```{r}
n_complete <- n_reason %>% filter(row == "Completed")
n_complete
```

```{r}
n_reason   <- n_reason %>% filter(row != "Completed") %>% 
                  mutate(row = paste0("    ", row))
n_reason
```

```{r}
tbl_disp <- bind_rows(n_rand, n_complete, n_disc, n_reason) %>% 
            select(row, ends_with(c("_0", "_54", "_81")))

tbl_disp
```

```{r}
tbl_disp %>% rtf_body() %>% 
             rtf_encode() %>% 
             write_rtf("tlf/tbl_disp1.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tbl_disp1.pdf")
```

```{r}
adsl %>% select(TRT01P, TRT01PN) %>% 
         unique() %>% 
         arrange(TRT01PN)
```

```{r}
tbl_disp %>% rtf_title("Disposition of Patients")

             rtf_colheader(" | Placebo | Xanomeline Low Dose| Xanomeline High Dose", 
                           col_rel_width = c(3, rep(2,3)) ) %>%

             rtf_colheader(" | n | (%) | n | (%) | n | (%)",
                           col_rel_width = c(3, rep(c(0.7, 1.3),3)),
                           border_top = c("", rep("single", 6)),
                           border_left = c("single", rep(c("single",""), 3))) %>%

             rtf_body(col_rel_width      = c(3, rep(c(0.7, 1.3),3)),
                      text_justification = c("l", rep("c",6)),
                      border_left        = c("single", rep(c("single",""), 3))) %>%

             rtf_encode() %>%

             write_rtf("tlf/tbl_disp2.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tbl_disp2.pdf")
```

### 2021-08-05 tlf.Rmd
disposition table

@@
 
-## Table for Baseline Characteristics
+## Structure and Content
 
-```{r}
-library(table1)
-library(r2rtf)
-data(r2rtf_adsl)
+In the rest of this chapter, we are following the [ICH E3 guidance](https://www.ich.org/page/efficacy-guidelines) on 
+structure and content of clincial study reports.
 
-tbl <- table1(~ SEX + AGE + RACE | TRT01P, data = r2rtf_adsl)
-tbl
-```
+In a CSR, most of TLFs are located in
 
-```{r}
-tbl %>% as.data.frame() %>% 
-        rtf_body() %>% 
-        rtf_encode() %>% 
-        write_rtf("tlf/bs-1.rtf")
-```
+- Section 10: Study patients 
+- Section 11: Efficacy evaluation
+- Section 12: Safety evaluation 
+- Section 14: Tables, Figures and Graphs referred to but not included in the text
+- Section 16: Appendices
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/bs-1.pdf")
-```
+## Datasets
+
+We used public available CDISC pilot [study data located at CDISC Bitbucket repository](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).
+
+For simplicity, we have downloaded all these datasets into `adam_data` folder of 
+this project and transfer them from `.xpt` format to `sas7bdat` format.
+
+The dataset structure is following [CDISC Analysis Data Model (ADaM)](https://www.cdisc.org/standards/foundational/adam).

### 2021-08-11 project-folder.Rmd
initiate project folder info

# (PART) Clinical Trial Project {-}

# Project Folder

In a late-stage clinical trial, the number of A&R deliverables can easily be in the hundreds. 
For an organization, it is possible to have multiple ongoing clinical trials in a clincial program.

A clearly defined project folder structure can have many benefits to 
clinical development teams in an organization. 
Specifically a well defined project folder can achieve:

- Consistency
- Reproducibility
- Automation
- Compliance

## Consistency

For consistency, a well defined folder structure with potential templates
ensures project teams organize A&R work consistently across multiple projects. 
Consistent folder structure also reduce communication cost between study team members 
and enhances the transparency of projects.

For R user, you already benefit from a well defined and consistent folder structure. 
That is the R package folder structure. 
Every R package developer is required to follow the same convention to organize their R
functions, before the function package can be disseminated through the Comprehensive R Archive Network
(CRAN). As a user, you can easily install and user those packages after installing from CRAN.

We recommend to also use R package folder structure to organize analysis scripts 
for clincial trial development. 
Using the R package folder structure to streamline data analysis work has been proposed 
(@marwick2018packaging, @wuanalysis).

In this book, we refer an R package as **project specific R package** 
if the purpose of an R package is to organize analysis scripts for a clinical project.
We refer an R package as **standard R package** if the purpose of an R package is to 
share commonly used R functions to be hosted in an code repository such as CRAN.

Below is a minimal sufficient folders and files based on the R package folder structure.

- `*.Rproj`: RStudio project file used to open RStudio project.
- `DESCRIPTION`: Metadata for a package including authors, license, dependency etc.
- `vignettes/`: Analysis scripts using Rmarkdown.
- `R/`: Project specific R functions.
- `man/`: Manual of project specific R functions.

We demonstrate the the idea using the [`esubdemo`](https://github.com/elong0527/esubdemo) project.

A general discussion of the R package folder structure can be found in Chapter 4 of “R packages:
organize, test, document, and share your code” book (@wickham2015r).

## Reproducibility

Reproducibility of analysis can be challenge. 
By using R package folder structure and proper tools (e.g. `renv`, `packrat`), 
we can easily achieve reproducibility for R and R package versions.

> This is the same level of reproducibility in most SAS environment.
> https://support.sas.com/techsup/pcn/altopsys.html

In `esubdemo` project, a reproducible environment is created when you open the 
`esubdemo.Rproj` from RStudio. 
While RStudio open `esubdemo` project, RStudio will automatically execute R code in `.Rprofile`.
So we are using `.Rprofile` to setup a reproducible environment. 
More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html).

> `.Rprofile` is only for project specific R package. 
> A standard R package should not use `.Rprofile`.

After the `esubdemo` project is open, the code in `.Rprofile` will automatically check 
current R version is the same as we defined in `.Rprofile`

```{r, eval = FALSE}
R_version <- "4.0.1"	# set up project R version
```

If the R version mismatch, an error message is displayed.

```
Error: The current R version is not the same as the current project in R4.0.1
```

If we are running under the same R version, 
you shall be able to see similar messages to inform how we control R package version.

```
* Project '~/esubdemo' loaded. [renv 0.14.0]
Current oroject R package repository:
    https://mran.microsoft.com/snapshot/2021-08-06
```

The R package version is controlled in two layers. 
Firstly, we define a snapshot date in `.Rprofile`. 
The snapshot date allows to freeze the repository. 
In other words, all R packages installed in this RStudio project 
are based on the frozen R version at the snapshot date. 
Here it is `2021-08-06` by using Microsoft R Application Network (MRAN) server.

```{r, eval = FALSE}
snapshot  <- "2021-08-06" 									                           # set up snapshot date
repos     <- paste0("https://mran.microsoft.com/snapshot/", snapshot)  # set up repository based on snapshot
```

> RStudio Package Manager (RSPM) provided a solution to host both 
> public avaiable and internally developed R packages. 
> However, public RSPM server (https://packagemanager.rstudio.com/client/#/repos/2/overview) 
> did not provide daily snapshot as MRAN had.

Secondly, we use `renv` to lock R package versions and saved in `renv.lock` file. 
The exact same R package versions can be restored with `renv::restore()`.
An introduction of `renv` can be found in its [package website](https://rstudio.github.io/renv/articles/renv.html).

```
* Project '~/github-repo/esubdemo' loaded. [renv 0.14.0]
```

> `renv.lock` file and `renv/` folder are only for project specific R package. 
> A standard R package should not use `renv`.

Once R packages have been properly installed, 
the system will use the R packages located in the search path defined based on the order of `.libPaths()`.
The startup message also provided the R package search path.

```
Below R package path are searching in order to find installed R pacakges in this R session:
    /home/esubdemo/renv/library/R-4.0/x86_64-pc-linux-gnu
    /rtmp/Rtmp8d5mcL/renv-system-library
```

For an organization, a cloud based R environment (e.g. RStudio Workbench) 
can enhance the development and operation of a clinical development team by 
centralize managing R and R package version. 
More details can be found in [https://environments.rstudio.com/](https://environments.rstudio.com/).

In summary, to achieve reproducibility for a project specific R package, 
a clinical project team can work under a controlled R environment 
in the same R version and R package versions defined by a repository snapshot date.

## Automation

- ToDo

There are additional folders and files required in this demo for an analysis project

> people may use different folder name as it is not a standard R package folder.
> These folders and files needs to be added in `.Rbuildignore` to pass
> R package check

- `adam`: ADaM datasets in `.sas7bdat` format.
  + one may also put it in `inst/exdata` following [R package convention](https://r-pkgs.org/data.html)
  + in reality, we suggest to have real data saved outside of this project. (e.g. in a database)
- `output`: TLFs output 
- `renv.lock` and `renv`: R package management using `renv` package. ([Introduction](https://rstudio.github.io/renv/articles/renv.html)) 
- `_pkgdown.yml`: [pkgdown](https://pkgdown.r-lib.org/articles/pkgdown.html) configuration file
- `.Rprofile`: Project startup file to setup running environment including R version, repository, folder path etc. 
  - We further use `inst/startup.R` and `R/zzz.R` to allow the startup file is executed while running. `devtools::load_all()` and RStudio build panel.

## Compliance

- ToDo

### 2021-08-11 project-management.Rmd
initiate project folder info

# Project Management

- ToDo

A clearly defined software development lifecycle (SDLC) for the programming of A&R deliverables is crucial to manage all
the outputs and to facilitate the development of the deliverables.
A typical SDLC for the A&R deliverables is comprised of 4 stages: define, develop, validate and operations.
The define stage is the most critical stage in the lifecycle as the requirements for all A&R deliverables are
gathered and documented. Validation plan for the deliverables is also planned and documented at this
stage. After the define stage is complete, the programming development occurs in the develop stage. The
programs completed in the develop stage are then validated according to the validation plan in the validate
stage. Test runs of the completed programs are conducted to ensure quality of the deliverables.
The development of programs and their associated validation is usually carried out in a dedicated test area
in the computing platform. After the validation stage is completed, the operations stage commences with
the promotion of validated programs to the production area in the computing platform. The promoted
programs are then executed in the production area with production data to generate the final version of the
deliverables. The deliverables are then handed over to the end customer for review. Any changes in
requirements of the deliverables at this stage based on end customer review, results in the deliverable
being processed through the 4 stages of SDLC again, starting with the changes to the specifications in the
define stage.
Several different software and programming languages have used in analyzing clinical trial data, with R
being among them. R is a popular programming language and an open-source software environment. R
has been widely used in various fields of data analytics for statistical analyses. There is a growing interest
within the pharmaceutical industry in analyzing clinical trial data using R and its associated software under
a regulatory environment. However, there is a gap in defining the appropriate SDLC to complete typical
A&R deliverables and ensure regulatory compliance. 
ToDo:

- add mkstdpkg/mkdev/mkuse related information.

### 2021-08-20 project-folder.Rmd
update project information

@@
 
-# Project Folder
+# Overview
 
 In a late-stage clinical trial, the number of A&R deliverables can easily be in the hundreds. 
-For an organization, it is possible to have multiple ongoing clinical trials in a clincial program.
+For an organization, it is also common to have multiple ongoing clinical trials in a clinical program.
+
+In this part, we would like to put you as an A&R leader for a clinical project.
+
+In Chapter \@ref(folder), we will discuss how to organize source code, documents, 
+deliverables in an A&R clinical project. 
+We recommend to use the R package folder structure.
+
+In Chapter \@ref(manage), we will discuss a process or system development lifecycle 
+to manage A&R of a clinical project. 
+We recommend to follow an agile management approach to define, develop, validate and deliver work.
+
+# Project Folder {#folder}
 
@@
 
-## Reproducibility
+## Reproducibility {#reproduce}
 
@@
 
-If the R version mismatch, an error message is displayed.
+If the R version mismatch, an error message is displayed as below.
 

### 2021-08-20 project-management.Rmd
update project information

@@
-# Project Management
+# Project Management {#manage}
 
-- ToDo
+For A&R deliverables in a clinical project development, 
+a clearly defined process or system development lifecycle (SDLC) 
+is crucial to ensure regulatory compliant.
 
-A clearly defined software development lifecycle (SDLC) for the programming of A&R deliverables is crucial to manage all
-the outputs and to facilitate the development of the deliverables.
-A typical SDLC for the A&R deliverables is comprised of 4 stages: define, develop, validate and operations.
-The define stage is the most critical stage in the lifecycle as the requirements for all A&R deliverables are
-gathered and documented. Validation plan for the deliverables is also planned and documented at this
-stage. After the define stage is complete, the programming development occurs in the develop stage. The
-programs completed in the develop stage are then validated according to the validation plan in the validate
-stage. Test runs of the completed programs are conducted to ensure quality of the deliverables.
-The development of programs and their associated validation is usually carried out in a dedicated test area
-in the computing platform. After the validation stage is completed, the operations stage commences with
-the promotion of validated programs to the production area in the computing platform. The promoted
-programs are then executed in the production area with production data to generate the final version of the
-deliverables. The deliverables are then handed over to the end customer for review. Any changes in
-requirements of the deliverables at this stage based on end customer review, results in the deliverable
-being processed through the 4 stages of SDLC again, starting with the changes to the specifications in the
-define stage.
-Several different software and programming languages have used in analyzing clinical trial data, with R
-being among them. R is a popular programming language and an open-source software environment. R
-has been widely used in various fields of data analytics for statistical analyses. There is a growing interest
-within the pharmaceutical industry in analyzing clinical trial data using R and its associated software under
-a regulatory environment. However, there is a gap in defining the appropriate SDLC to complete typical
-A&R deliverables and ensure regulatory compliance. 
-ToDo:
+SDLC for A&R deliverables can be defined in stages:
 
-- add mkstdpkg/mkdev/mkuse related information.
+- planning: a planning stage to define the scope of a project.  
+- development: a development stage to implement target deliverables. 
+- validation: a validation stage to verify target deliverables. 
+- operation: an operation stage to deliver work to stakeholders.
+
+The define stage is an important stage in the SDLC lifecycle 
+as the requirements for all A&R deliverables are gathered and documented.
+
+Importantly, we should not consider SDLC as a linear process. 
+For example, if study team identify new requirement in a development or validation stage, 
+team shall return to planning stage to discuss and align the scope. 
+An [agile project management](https://www.atlassian.com/agile/project-management) approach is suitable 
+and recommended to an A&R clinical development project.
+The goal is to embrace an iterative approach that continue improve target deliverables based on
+frequent customer feedback.
+
+There are many good tools to implement agile project management strategy, for example:
+
+- [github project board](https://docs.github.com/en/issues/organizing-your-work-with-project-boards/managing-project-boards/about-project-boards)
+- [Jira](https://www.atlassian.com/software/jira)
+
+## Planning
+
+In the planning stage, a project leader shall identifies all the deliverables, 
+e.g. a list of table, listing, and figures (TLFs). 
+For each TLFs, team shall prepare necessary specifications:
+
+- mock-up
+- validation level (e.g. independent review or double programming)
+- etc.
+
+ToDo: add a simplified mock-up and A&R grid.
+
+project leader shall also align work assignment with team members.
+The purpose is to answer the question on "who is doing what?".
+
+ToDo: add a simplified validation tracker
+
+Project lead shall also setup a project folder as discussed in Chapter \@ref(folder).
+The project initiation can be simplified by creating an 
+[RStudio project template](https://rstudio.github.io/rstudio-extensions/rstudio_project_templates.html).
+
+To enable reproducibility, project leader shall also review the startup file 
+(i.e. `.Rprofile` discussed in Section \@ref(manage)) and define:
+
+- R version 
+- Repository of R packages with a snapshot date 
+- Project package library path 
+- etc.
+
+> After project initiate, modifying `.Rprofile` will be a risk of reproducibility 
+> and shall be handled carefully if necessary.
+
+## Development
+
+After a project is initiated, the study team start to develop TLFs based on 
+pre-defined mock-up tables assigned to each team member.
+
+The analysis code and relevant description can be saved in R markdown files 
+in the `vignettes` folder. 
+The use of R markdown allow developer to assemble narrative text, code and its
+comments in one place to simplify documentation. 
+It would be helpful to define a name convention for all TLFs deliverables. 
+For example, we can use `tlf_` prefix in the filename to indicate that the Rmarkdown is to deliver TLFs. 
+For example, `tlf_001_ds.Rmd` is an Rmarkdown file for disposition table. 
+Multiple TLFs with similar design can be included in one Rmarkdown file.
+
+ToDo: add a file path for `esubdemo`
+
+To validate the results, 
+developer shall also save the last dataset before creating TLFs in RTF format to `.Rdata` files.
+
+## Validation
+
+## Operation

### 2021-08-23 tlf-overview.Rmd
update content structure
proofread of chap 1,2.

--- 2021-08-05 tlf.Rmd
@@
 
-# Reporting
+# Overview
 
@@
 following ICH E3 guidance](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse).
+(??? What is the correlation between CDISC and eCTD? Are they to standard format of CSR? Besides, as a beginner, I feel it is better to explain "pilot project" in details.)
 
@@
 A large number of tables and figures are incorporated into the main text and appendices 
-of a CSR.
-In CDISC pilot project, an [example CSR](file:///C:/Users/zhanyilo/OneDrive%20-%20Merck%20Sharp%20&%20Dohme,%20Corp/Desktop/cdiscpilot01.pdf) is also provided.
-If you are interested in more examples of clinical study reports,
+of a CSR. In CDISC pilot project, an [example CSR](file:///C:/Users/zhanyilo/OneDrive%20-%20Merck%20Sharp%20&%20Dohme,%20Corp/Desktop/cdiscpilot01.pdf) (??? This link is not workable)
+is also provided. If you are interested in more examples of clinical study reports, 
 the European Medicines Agency (EMA) publishes clinical data including 
@@
 Building CSR is a team work with clinicians, medical writers, statisticians and statistical programmers.
-Here, We focus on the work and deliverables completed by statisticians and statistical programmers. 
+Here, we focus on the work and deliverables completed by statisticians and statistical programmers. 
 In an organization, A group of statisticians and statistical programmers commonly work together to 
@@
 In the rest of this chapter, we are following the [ICH E3 guidance](https://www.ich.org/page/efficacy-guidelines) on 
-structure and content of clincial study reports.
+structure and content of clinical study reports.
 

### 2021-08-26 tlf-overview.Rmd
tlf-overview.Rmd edited online with Bitbucket

@@
 A large number of tables and figures are incorporated into the main text and appendices 
-of a CSR. In CDISC pilot project, an [example CSR](file:///C:/Users/zhanyilo/OneDrive%20-%20Merck%20Sharp%20&%20Dohme,%20Corp/Desktop/cdiscpilot01.pdf) (??? This link is not workable)
+of a CSR.
+In CDISC pilot project, an [example CSR](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/53-clin-stud-rep/535-rep-effic-safety-stud/5351-stud-rep-contr/cdiscpilot01/cdiscpilot01.pdf)
 is also provided. If you are interested in more examples of clinical study reports, 

### 2021-08-27 tlf-overview.Rmd
tlf-overview.Rmd edited online with Bitbucket
update tlf-overview page

--- 2021-08-05 tlf.Rmd
@@
 The dataset structure is following [CDISC Analysis Data Model (ADaM)](https://www.cdisc.org/standards/foundational/adam).
+
+## Tools
+
+In this part, we mainly use R packages below to illustrate 
+how to deliver TLFs in CSR.
+
+- [tidyverse](https://www.tidyverse.org/): prepare reporting ready datasets. 
+- [r2rtf](https://merck.github.io/r2rtf/): create RTF outputs
+
+### tidyverse
+
+`tidyverse` is a collection of R packages to simplify the workflow to manipulate, 
+visualize and analyze data in R. 
+Those R packages share [the tidy tools manifesto](https://cran.r-project.org/web/packages/tidyverse/vignettes/manifesto.html) 
+and easy to use for interactive data analysis.
+
+RStudio provided outstanding [cheatsheets](https://www.rstudio.com/resources/cheatsheets/), 
+and [tutorials](https://github.com/rstudio-education/remaster-the-tidyverse) for `tidyverse`.
+
+There are also books to introduce tidyverse. 
+We assume reader to have experience in using `tidyverse` in this book.
+
+- [The tidyverse cookbook](https://rstudio-education.github.io/tidyverse-cookbook/)
+- [R for Data Science](https://r4ds.had.co.nz/)
+
+### r2rtf
+
+`r2rtf` is an R package to create production ready tables and figures in RTF format.
+The R package is designed to
+
+- provide simple "verb" functions that correspond to each component of a table, 
+  to help you translate data frame to table in RTF file.
+- enable pipes (`%>%`). 
+- only focus on **table format**. 
+  Data manipulation and analysis shall be handled by other R packages. (e.g. `tidyverse`)
+
+Before creating an RTF table we need to:
+
+- Figure out table layout.
+
+- Split the layout into small tasks in the form of a computer program.
+
+- Execute the program.
+
+We provided a brief introduction of `r2rtf` and show how to transfer 
+data frames into Table, Listing, and Figure (TLFs).
+
+Other extended examples and features are covered in 
+[`r2rtf` package website](https://merck.github.io/r2rtf/articles/index.html).
+
+To explore the basic RTF generation verbs in `r2rtf`, 
+we'll use the dataset `r2rtf_adae` saved in the `r2rtf` package. 
+This dataset contain adverse events (AE) information from a clinical trial.
+
+Below is the meaning of relevant variables. 
+More information can be found in help page of the dataset (`?r2rtf_adae`)
+
+In this example we consider three variables:
+
+- USUBJID: Unique Subject Identifier
+- TRTA: Actual Treatment 
+- AEDECOD: Dictionary-Derived Term
+
+```{r}
+library(r2rtf)
+r2rtf_adae %>%
+  select(USUBJID, TRTA, AEDECOD) %>%
+  head(4)
+```
+`dplyr` and `tidyr` packages within `tidyverse` are used 
+for data manipulation to create a data frame 
+that contains all the information we want to add in an RTF table.
+
+```{r}
+library(tidyr)
+library(dplyr)
+tbl <- r2rtf_adae %>%
+  count(TRTA, AEDECOD) %>%
+  pivot_wider(names_from = TRTA, values_from = n, values_fill = 0)
+
+tbl %>% head(4)
+```
+
+Now we have a dataset `tbl` in preparing the final RTF table.
+
+`r2rtf` aims to provide one function for each type of table layout.
+Commonly used verbs includes:
+
+- `rtf_page`: RTF page information
+- `rtf_title`: RTF title information
+- `rtf_colheader`: RTF column header information
+- `rtf_body`: RTF table body information
+- `rtf_footnote`: RTF footnote information
+- `rtf_source`: RTF data source information
+
+All these verbs is designed to enables usage of pipes (`%>%`).
+A full list of all functions can be found in [package reference](https://merck.github.io/r2rtf/reference/index.html)
+
+A minimal example below illustrates how to combine verbs using pipes to create an RTF table.
+
+- `rtf_body` is used to define table body layout. 
+- `rtf_encode` transfers table layout information into RTF syntax.
+- `write_rtf` save RTF encoding into a file with file extension `.rtf`
+
+```{r}
+head(tbl) %>%
+  rtf_body() %>%                        # Step 1 Add table  attributes
+  rtf_encode() %>%                      # Step 2 Convert attributes to RTF encode
+  write_rtf("tlf/intro-ae1.rtf")        # Step 3 Write to a .rtf file
+```
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/intro-ae1.pdf")
+```
+
+If we want to adjust the width of each column to 
+provide more space to the first column, 
+this can be achieved by updating `col_rel_width` argument in `rtf_body`.
+
+In this example, the input of `col_rel_width` is a vector with same length for number of columns. 
+This argument defines the relative width of each column 
+within a pre-defined total column width.
+
+In this example, the defined relative width is `3:2:2:2`. 
+Only the ratio of `col_rel_width` is used. 
+Therefore it is equivalent to use `col_rel_width = c(6,4,4,4)` 
+or `col_rel_width = c(1.5,1,1,1)`.
+
+```{r}
+head(tbl) %>%
+  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%    # define relative width
+  rtf_encode() %>%
+  write_rtf("tlf/intro-ae2.rtf")
+```
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/intro-ae2.pdf")
+```
+
+In previous example, we found issue of misaligned column header. 
+We can fix the issue by using `rtf_colheader` function.
+
+In `rtf_colheader`, `colheader` argument is used to provide content of column header. 
+We use `"|"` to separate the columns.
+
+In the example below, `"Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose"`
+define a column header with 4 columns.
+
+```{r}
+head(tbl) %>%
+  rtf_colheader(
+    colheader = "Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose",
+    col_rel_width = c(3, 2, 2, 2)
+  ) %>%
+  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
+  rtf_encode() %>%
+  write_rtf("tlf/intro-ae3.rtf")
+```
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/intro-ae3.pdf")
+```
+
+In `rtf_*` functions such as `rtf_body`, `rtf_footnote`, 
+`text_justification` argument is used to align text. 
+Default is `"c"` for center justification. 
+To vary text justification by column, use character vector with length of vector equal to
+number of columns displayed (e.g., `c("c","l","r")`).
+
+All possible inputs can be found in the table below.
+
+```{r}
+r2rtf:::justification()
+```
+
+Below is an example to make first column left aligned and center aligned for the rest.
+
+```{r}
+head(tbl) %>%
+  rtf_body(text_justification = c("l", "c", "c", "c")) %>%
+  rtf_encode() %>%
+  write_rtf("tlf/intro-ae5.rtf")
+```
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/intro-ae5.pdf")
+```
+
+In `rtf_*` functions such as `rtf_body`, `rtf_footnote`, etc.,
+`border_left`, `border_right`, `border_top`, and `border_bottom` control cell borders.
+
+If we want to remove the top border of `"Adverse Events"` in header, 
+we can change default value `"single"` to `""` in `border_top` argument as showed below.
+
+`r2rtf` support 26 different border types. The details can be found in 
+the [r2rtf package website](https://merck.github.io/r2rtf/articles/rtf-row.html#border-type).
+
+In this example, we also demonstrate the possibility to add multiple column headers.
+
+```{r}
+head(tbl) %>%
+  rtf_colheader(
+    colheader = " | Treatment",
+    col_rel_width = c(3, 6)
+  ) %>%
+  rtf_colheader(
+    colheader = "Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose",
+    border_top = c("", "single", "single", "single"),
+    col_rel_width = c(3, 2, 2, 2)
+  ) %>%
+  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
+  rtf_encode() %>%
+  write_rtf("tlf/intro-ae7.rtf")
+```
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/intro-ae7.pdf")
+```
+
+In the r2rtf [get started](https://merck.github.io/r2rtf/articles/r2rtf.html) page
+there are more examples to illustrate how to customize
+
+- title, subtitle
+- footnote, data source
+- special character
+- etc
+
+Those features will be introduced when we first use them in the rest of chapters.
+
+There are other R packages that can achieve similar goals.
+Interested reader can refer "Microsoft/LibreOffice Formats" section in 
+[CRAN Task View: Reproducible Research](https://cran.r-project.org/web/views/ReproducibleResearch.html)
+for an overview.

### 2021-08-27 intro.Rmd

@@
 
-In clinical trial development, developer needs to organize source code to deliver SDTM or ADaM datasets and table, listing and figures (TLFs).
+In clinical trial development, developers needs to organize source code to deliver Study Data Tabulation Model (SDTM) or Analysis Dataset Model (ADaM) datasets and table, listing and figures (TLFs).
 For example, a Phase 3 clinical trial typically require hundreds of TLFs. 
@@
 
-We recommend to use R package folder structure to organize all the analysis and reporting related source code and documentation
-for a clinical trial A&R project. 
-By using the R package folder structure you can follow the convention to organize your code as other contributors on CRAN.
-The consistent approach simplify the communication for developers within and across organizations.
+We recommend to use R package folder structure to organize all the A&R related source code and documentation for a clinical trial A&R project. 
+By using the R package folder structure, you can follow the convention to organize your code as other contributors on CRAN.
+This consistent approach simplifies the communication for all developers within and across organizations.
 

### 2021-08-27 tlf-overview.Rmd
add clarification for tables

@@
 of a CSR.
-In CDISC pilot project, an [example CSR](file:///C:/Users/zhanyilo/OneDrive%20-%20Merck%20Sharp%20&%20Dohme,%20Corp/Desktop/cdiscpilot01.pdf) is also provided.
-If you are interested in more examples of clinical study reports,
+In CDISC pilot project, an [example CSR](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/53-clin-stud-rep/535-rep-effic-safety-stud/5351-stud-rep-contr/cdiscpilot01/cdiscpilot01.pdf)
+is also provided. If you are interested in more examples of clinical study reports, 
 the European Medicines Agency (EMA) publishes clinical data including 
@@
 - special character
-- etc
+- etc.
 
@@
 
-There are other R packages that can achieve similar goals.
+There are other R packages that can create TLFs in `.rtf` or `.docx` format.
 Interested reader can refer "Microsoft/LibreOffice Formats" section in 

### 2021-08-27 tlf-disposition.Rmd
update output
update content structure
update dispostion table

@@
-## Table for Study Patients
+# Table for Study Patients
 
-In section 10, study patients of a CSR, we need to summarize accounting 
-of all patients who entered the study.
+Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
+we need to summarize accounting of all patients who entered the study, 
+in the section 10.1, disposition of patients of a CSR.
 
@@
 
+In this chapter, we illustrate how to create a disposition table as below.
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/tbl_disp.pdf")
+```
+
 The first step is to read relevant datasets into R. 
-For disposition table, all the information is saved in the ADSL dataset.
+For disposition table, all the required information is saved in the ADSL dataset.
 We can use `haven` package to read the dataset.
@@
 
-We illustrate how to prepare a report data for a simple disposition of patients table using variables below:
+We illustrate how to prepare a report data for a simplified disposition 
+of patients table using variables below:
 
@@
 ```{r}
-adsl %>% select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>% 
-         head()
+adsl %>%
+  select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>%
+  head(4)
 ```
-```{r}
-library(tidyr)
-```
+In the code below, we calculated patients in the analysis population.
 
 ```{r}
-n_rand <- adsl %>% group_by(TRT01PN) %>%
+n_rand <- adsl %>%
+  group_by(TRT01PN) %>%
   summarize(n = n()) %>%
-                   pivot_wider(names_from = TRT01PN, names_prefix = "n_", values_from = n) %>% 
+  pivot_wider(
+    names_from = TRT01PN,
+    names_prefix = "n_",
+    values_from = n
+  ) %>%
   mutate(row = "Patients in population")
@@
 ```
+In the code below, we calculate number and percentage of patients who discontinued the study.
+
+Here we use `formatC()` to customize the numeric value to be 1 decimal with fixed width of 5.
 
 ```{r}
-n_disc <- adsl %>% group_by(TRT01PN) %>%
-                   summarize(n = sum(DISCONFL == "Y"), 
-                             pct = formatC(n / n() * 100, digits = 1, format = "f", width = 5)) %>% 
-                   pivot_wider(names_from = TRT01PN, values_from = c(n, pct) ) %>% 
+n_disc <- adsl %>%
+  group_by(TRT01PN) %>%
+  summarize(
+    n = sum(DISCONFL == "Y"),
+    pct = formatC(n / n() * 100,
+      digits = 1, format = "f", width = 5
+    )
+  ) %>%
+  pivot_wider(
+    names_from = TRT01PN,
+    values_from = c(n, pct)
+  ) %>%
   mutate(row = "Discontinued")
+
 n_disc
 ```
+In the code below, we calculate number and percentage of patients 
+who completed/discontinued the study in different reasons.
 
 ```{r}
-n_reason <- adsl %>% group_by(TRT01PN) %>%
+n_reason <- adsl %>%
+  group_by(TRT01PN) %>%
   mutate(n_total = n()) %>%
   group_by(TRT01PN, DCREASCD) %>%
-                     summarize(n = n(),
-                               pct = formatC(n / unique(n_total) * 100, digits = 1, format = "f", width = 5)) %>% 
-                     pivot_wider(id_cols = DCREASCD, names_from = TRT01PN, 
-                                 values_from = c(n, pct), values_fill = list(n = 0, pct = "  0.0") ) %>% 
+  summarize(
+    n = n(),
+    pct = formatC(n / unique(n_total) * 100,
+      digits = 1, format = "f", width = 5
+    )
+  ) %>%
+  pivot_wider(
+    id_cols = DCREASCD,
+    names_from = TRT01PN,
+    values_from = c(n, pct),
+    values_fill = list(n = 0, pct = "  0.0")
+  ) %>%
   rename(row = DCREASCD)
+
 n_reason
@@
 
+In the code below, we calculate number and percentage of patients 
+who complete the study. We split `n_reason` because we need to customize 
+the row order of the table.
+
 ```{r}
 n_complete <- n_reason %>% filter(row == "Completed")
+
 n_complete
 ```
+In the code below, we calculate number and percentage of patients 
+who discontinued the study in different reasons. 
+Here we use `paste0("    ", row)` to add some leading space for individual reasons for study discontinuation.
 
 ```{r}
-n_reason   <- n_reason %>% filter(row != "Completed") %>% 
+n_reason <- n_reason %>%
+  filter(row != "Completed") %>%
   mutate(row = paste0("    ", row))
+
 n_reason
 ```
+
+Now we combined individual rows into the whole table for reporting purpose. 
+`tbl_disp` is used as input for `r2rtf` to create final report.
 
@@
 
+We start to define the format of the output. We highlighted items that 
+is not discussed in previous discussion.
+
+`rtf_title` define table title. We can provide a vector for the title argument. 
+Each value is a separate line. 
+The format can also be controlled by providing a vector input in text format.
+
 ```{r}
-tbl_disp %>% rtf_body() %>% 
+tbl_disp %>%
+
+  # Table title
+  rtf_title("Disposition of Patients") %>%
+
+  # First row of column header
+  rtf_colheader(" | Placebo | Xanomeline Low Dose| Xanomeline High Dose",
+    col_rel_width = c(3, rep(2, 3))
+  ) %>%
+
+  # Second row of column header
+  rtf_colheader(" | n | (%) | n | (%) | n | (%)",
+    col_rel_width = c(3, rep(c(0.7, 1.3), 3)),
+    border_top = c("", rep("single", 6)),
+    border_left = c("single", rep(c("single", ""), 3))
+  ) %>%
+
+  # Table body
+  rtf_body(
+    col_rel_width = c(3, rep(c(0.7, 1.3), 3)),
+    text_justification = c("l", rep("c", 6)),
+    border_left = c("single", rep(c("single", ""), 3))
+  ) %>%
+
+  # Encoding RTF syntax
   rtf_encode() %>%
-             write_rtf("tlf/tbl_disp1.rtf")
+
+  # Save to a file
+  write_rtf("tlf/tbl_disp.rtf")
 ```
@@
 ```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/tbl_disp1.pdf")
+knitr::include_graphics("tlf/tbl_disp.pdf")
 ```
-
-```{r}
-adsl %>% select(TRT01P, TRT01PN) %>% 
-         unique() %>% 
-         arrange(TRT01PN)
-```
-
-```{r}
-tbl_disp %>% rtf_title("Disposition of Patients")
-
-             rtf_colheader(" | Placebo | Xanomeline Low Dose| Xanomeline High Dose", 
-                           col_rel_width = c(3, rep(2,3)) ) %>%
-
-             rtf_colheader(" | n | (%) | n | (%) | n | (%)",
-                           col_rel_width = c(3, rep(c(0.7, 1.3),3)),
-                           border_top = c("", rep("single", 6)),
-                           border_left = c("single", rep(c("single",""), 3))) %>%
-
-             rtf_body(col_rel_width      = c(3, rep(c(0.7, 1.3),3)),
-                      text_justification = c("l", rep("c",6)),
-                      border_left        = c("single", rep(c("single",""), 3))) %>%
-
-             rtf_encode() %>%
-
-             write_rtf("tlf/tbl_disp2.rtf")
-```
-
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/tbl_disp2.pdf")
-```

### 2021-08-27 tlf-overview.Rmd
update dispostion table

@@
 
-In the rest of this chapter, we are following the [ICH E3 guidance](https://www.ich.org/page/efficacy-guidelines) on 
-structure and content of clinical study reports.
+In the rest of this chapter, we are following the 
+[ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf) 
+on structure and content of clinical study reports.
 
@@
 library(r2rtf)
+
 r2rtf_adae %>%
@@
 library(dplyr)
+
 tbl <- r2rtf_adae %>%
@@
 head(tbl) %>%
-  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%    # define relative width
+  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
+
+  # define relative width
   rtf_encode() %>%
@@
 head(tbl) %>%
+
   rtf_colheader(
@@
   ) %>%
+
   rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
+
   rtf_encode() %>%
+
   write_rtf("tlf/intro-ae3.rtf")
@@
 head(tbl) %>%
+
   rtf_body(text_justification = c("l", "c", "c", "c")) %>%
+
   rtf_encode() %>%
+
   write_rtf("tlf/intro-ae5.rtf")
@@
   ) %>%
+
   rtf_colheader(
@@
   ) %>%
+
   rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
+
   rtf_encode() %>%
+
   write_rtf("tlf/intro-ae7.rtf")
@@
 
-In the r2rtf [get started](https://merck.github.io/r2rtf/articles/r2rtf.html) page
+In the `r2rtf` R package [get started](https://merck.github.io/r2rtf/articles/r2rtf.html) page
 there are more examples to illustrate how to customize

### 2021-08-27 tlf-planning.Rmd
update dispostion table

# Table for Planning

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we need to summarize important deviations in the section 10.2, protocol deviations of a CSR.

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we need to summarize which patients were included in each efficacy analysis 
in the section 11.1, data sets analysed of a CSR.

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we need to summarize critical demographic and baseline characteristics of the patients
in the section 11.2, demographic and other baseline characteristics of a CSR.

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we need to summarize measurements of compliance of the patients
in the section 11.3, measurements of treatment compliance of a CSR.

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we need to summarize primary and secondary efficacy endpoints. 
in the section 11.4, efficacy results and tabulations of individual patient data of a CSR.

### 2021-08-28 tlf-population.Rmd
add analysis population example

# Table for Study Patients {#population}

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we need to summarize which patients were included in each efficacy analysis 
in the section 11.1, data sets analysed of a CSR.

```{r}
library(haven) # Read SAS data
library(dplyr) # Manipulate data
library(tidyr) # Manipulate data
library(r2rtf) # Reporting in RTF format
```

In this chapter, we illustrate how to create a summary table for
analysis population in a study.

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tbl_pop.pdf")
```

The first step is to read relevant datasets into R. 
For analysis population table, 
all the required information is saved in the ADSL dataset.
We can use `haven` package to read the dataset.

```{r}
adsl <- read_sas("adam_data/adsl.sas7bdat")
```

We illustrate how to prepare a report data for a simplified 
analysis population table using variables below:

- USUBJID: Unique subject identifier 
- ITTFL: Intent-to-treat population flag
- EFFFL: Efficacy population flag
- SAFFL: Safty population flag

```{r}
adsl %>%
  select(USUBJID, ITTFL, EFFFL, SAFFL) %>%
  head(4)
```
## Helper Functions

Before we write analysis code, let's discuss the possibility to
reuse R code by writing some toy helper function.

As discussed in 
[R for data science](https://r4ds.had.co.nz/functions.html#when-should-you-write-a-function),
"You should consider writing a function 
whenever you’ve copied and pasted a block of code more than twice"

In Chapter \@ref(dispositon), there are few repeating steps:

- format percentage using `formatC` 
- calculate number and percentage by groups

We create two ad-hoc functions and apply them to all rest of tables.

To format number and percentage, we created a function called `fmt_num`.
It is a very simple function that is a wrapper of `formatC`.

```{r}
fmt_num <- function(x, digits, width = digits + 4){
  formatC(x,
          digits = digits, 
          format = "f", 
          width = width
    )
}
```

The main reason to create `fmt_num` is to enhance readability of the analysis code.

For example, we can compare the two versions of code to format percentage 
used in \@ref(dispositon).

```{r, eval = FALSE}
formatC(n / n() * 100,
      digits = 1, format = "f", width = 5
    )
```

```{r, eval = FALSE}
fmt_pct(n / n() * 100)
```

To calculate number and percentage of patients by groups, 
we provide a simple (but not robust) wrapper function, `count_by` using `dplyr` and `tidyr`.

The function can be enhanced in multiple ways, but here we only focus on simplificy and readability.
More details about writting R functions can be found in [Stat454 course](https://stat545.com/functions-part1.html).

```{r}
count_by <- function(data,               # Input data set
                     grp,                # Group variable
                     var,                # Analysis variable
                     var_label = var,    # Analysis variable label
                     id = "USUBJID") {   # Subject ID variable 
  data <- data %>% rename(grp = !!grp, var = !!var, id = !!id)

  left_join(
    count(data, grp, var),
    count(data, grp, name = "tot"), 
    by = "grp",
  ) %>%
    mutate(
      pct = fmt_num(100 * n / tot, digits = 1),
      n   = fmt_num(n, digits = 0), 
      npct = paste0(n , " (", pct, ")")
    ) %>%
    pivot_wider(
      id_cols = var,
      names_from = grp,
      values_from = c(n, pct, npct),
      values_fill = list(n = "0", pct = fmt_pct(0))
    ) %>% 
    mutate(var_label = var_label)
}
```

By using `count_by`, we can simplify the analysis code as below.

```{r}
count_by(adsl, "TRT01PN", "EFFFL")
```

### Analysis Code

With the helper functions `count_by`, we can easily prepare report dataset as

```{r}
# Derive a randomization flag 
adsl <- adsl %>% mutate(RANDFL = "Y") 
pop <- count_by(adsl, "TRT01PN", "RANDFL", var_label = "Participants in Population") %>% 
          select(var_label,starts_with("n_"))
```

```{r}
pop1 <- bind_rows(
  count_by(adsl, "TRT01PN", "ITTFL", 
           var_label = "Participants included in ITT population"), 
  count_by(adsl, "TRT01PN", "EFFFL", 
           var_label = "Participants included in efficacy population"),
  count_by(adsl, "TRT01PN", "SAFFL", 
           var_label = "Participants included in safety population")
) %>% 
  filter(var == "Y") %>% 
  select(var_label, starts_with("npct_"))
```

```{r}
names(pop) <- gsub("n_", "npct_", names(pop))
tbl_pop <- bind_rows(pop, pop1)
tbl_pop
```

```{r}
rel_width <- c(2, rep(1, 3))
colheader <- " | \\line Placebo | Xanomeline \\line Low Dose| Xanomeline \\line High Dose"
tbl_pop %>%

  # Table title
  rtf_title("Participants Accounting in Analysis Population", 
            "(All Participants Randomized)") %>%

  # First row of column header
  rtf_colheader(colheader,
    col_rel_width = rel_width
  ) %>%

  # Second row of column header
  rtf_colheader(" | n (%) | n (%) | n (%)",
    border_top = "",
    col_rel_width = rel_width) %>%

  # Table body
  rtf_body(
    col_rel_width = rel_width,
    text_justification = c("l", rep("c", 3))
  ) %>%

  # Encoding RTF syntax
  rtf_encode() %>%

  # Save to a file
  write_rtf("tlf/tbl_pop.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tbl_pop.pdf")
```

### 2021-08-28 tlf-population.Rmd
tlf-population: fix typo

@@
-# Table for Study Patients {#population}
+# Table for Analysis Population {#population}
 
@@
 
-In Chapter \@ref(dispositon), there are few repeating steps:
+In Chapter \@ref(disposition), there are few repeating steps to:
 
@@
 ```{r}
-fmt_num <- function(x, digits, width = digits + 4){
+fmt_num <- function(x, digits, width = digits + 4) {
   formatC(x,
@@
 For example, we can compare the two versions of code to format percentage 
-used in \@ref(dispositon).
+used in \@ref(disposition) and `fmt_num`.
 
@@
 ```{r, eval = FALSE}
-fmt_pct(n / n() * 100)
+fmt_num(n / n() * 100, digits = 1)
 ```
@@
       n = fmt_num(n, digits = 0),
-      npct = paste0(n , " (", pct, ")")
+      npct = paste0(n, " (", pct, ")")
     ) %>%
@@
 ```{r}
-count_by(adsl, "TRT01PN", "EFFFL")
+count_by(adsl, "TRT01PN", "EFFFL") %>% 
+  select(- ends_with(c("_54", "_81")))
 ```
 
-### Analysis Code
+## Analysis Code
 
@@
 adsl <- adsl %>% mutate(RANDFL = "Y")
-pop <- count_by(adsl, "TRT01PN", "RANDFL", var_label = "Participants in Population") %>% 
-          select(var_label,starts_with("n_"))
+
+pop <- count_by(adsl, "TRT01PN", "RANDFL", 
+                var_label = "Participants in Population") %>%
+  select(var_label, starts_with("n_"))
 ```
@@
   count_by(adsl, "TRT01PN", "ITTFL",
-           var_label = "Participants included in ITT population"), 
+    var_label = "Participants included in ITT population"
+  ),
   count_by(adsl, "TRT01PN", "EFFFL",
-           var_label = "Participants included in efficacy population"),
+    var_label = "Participants included in efficacy population"
+  ),
   count_by(adsl, "TRT01PN", "SAFFL",
-           var_label = "Participants included in safety population")
+    var_label = "Participants included in safety population"
+  )
 ) %>%
@@
 
+Now we combined individual rows into the whole table for reporting purpose. 
+`tbl_pop` is used as input for `r2rtf` to create final report.
+
 ```{r}
@@
 tbl_pop <- bind_rows(pop, pop1)
-tbl_pop
+
+tbl_pop %>% select(var_label, npct_0)
 ```
+
+We start to define the format of the output.
 
@@
 rel_width <- c(2, rep(1, 3))
-colheader <- " | \\line Placebo | Xanomeline \\line Low Dose| Xanomeline \\line High Dose"
+colheader <- " | Placebo | Xanomeline line Low Dose| Xanomeline line High Dose"
 tbl_pop %>%
-
   # Table title
-  rtf_title("Participants Accounting in Analysis Population", 
-            "(All Participants Randomized)") %>%
-
+  rtf_title(
+    "Participants Accounting in Analysis Population",
+    "(All Participants Randomized)"
+  ) %>%
   # First row of column header
@@
   ) %>%
-
   # Second row of column header
@@
     border_top = "",
-    col_rel_width = rel_width) %>%
-
+    col_rel_width = rel_width
+  ) %>%
   # Table body
@@
   ) %>%
-
   # Encoding RTF syntax
   rtf_encode() %>%
-
   # Save to a file

### 2021-08-28 index.Rmd
update tlf-overview page
add analysis population example
update package dependency

@@
 library(downlit)
+library(haven) 
+library(dplyr) 
+library(tidyr) 
+library(r2rtf) 
 ```
@@
 knitr::opts_chunk$set(echo = TRUE, message = FALSE)
+options(dplyr.summarise.inform = FALSE)
 ```

### 2021-08-28 tlf-population.Rmd
update fmt_pct to fmt_num

@@
       values_from = c(n, pct, npct),
-      values_fill = list(n = "0", pct = fmt_pct(0))
+      values_fill = list(n = "0", pct = fmt_num(0))
     ) %>%

### 2021-08-28 tlf-population.Rmd
update digits

@@
       values_from = c(n, pct, npct),
-      values_fill = list(n = "0", pct = fmt_num(0))
+      values_fill = list(n = "0", pct = fmt_num(0, digits = 0))
     ) %>%

### 2021-08-29 tlf-baseline.Rmd
update content structure
update tlf-baseline table

@@
-## Table for Baseline Characteristics
+# Table for Baseline Characteristics
+
+Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
+we need to summarize critical demographic and baseline characteristics of the patients
+in the section 11.2, demographic and other baseline characteristics of a CSR.
+
+In this chapter, we illustrate how to create a simplified 
+baseline characteristics table in a study.
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/tlf_base.pdf")
+```
+
+There is many R packages that can effciently summarize baseline information.
+The [`table1`](https://github.com/benjaminrich/table1) R package is one of them.
 
@@
 library(r2rtf)
-data(r2rtf_adsl)
+library(haven)
+library(stringr)
+```
 
-tbl <- table1(~ SEX + AGE + RACE | TRT01P, data = r2rtf_adsl)
+As in previous chapters, we first read `adsl` dataset that contain all 
+required information for baseline characteristics table.
+
+```{r}
+adsl <- read_sas("adam_data/adsl.sas7bdat")
+```
+
+For simplicity, we only analyze `SEX`, `AGE` and `RACE` in this example
+using the `table1` R package. 
+More details of the `table` R package can be found in the package
+[vignettes](https://benjaminrich.github.io/table1/vignettes/table1-examples.html)
+
+The `table1` R package directly create an HTML report.
+
+```{r}
+tbl <- table1(~ SEX + AGE + RACE | TRT01P, data = adsl)
 tbl
@@
 
+The code below transfer the output into a dataframe 
+that only contain ASCII character 
+recommended by regulatory agencies. 
+`tbl_base` is used as input for `r2rtf` to create final report.
+
 ```{r}
-tbl <- tbl %>% as.data.frame() %>% 
-               as_tibble() %>% 
+tbl_base <- tbl %>% 
+              as.data.frame() %>% 
               mutate( across(everything(), 
-                        ~ stringr::str_replace_all(.x, intToUtf8(160), " ") ) 
+                      ~ str_replace_all(.x, intToUtf8(160), " ") ) 
                     )
+
+names(tbl_base) <- str_replace_all(names(tbl_base), intToUtf8(160), " ")
+tbl_base
 ```
 
+We start to define the format of the output. We highlighted items that 
+is not discussed in previous discussion.
+
+`text_indent_first` and `test_indent_left` are used to control the indent
+space of text. They are helpful when you need to control the white space 
+of a long sentence. For example, "AMERICAN INDIAN OR ALASKA NATIVE" 
+in the table.
+
 ```{r}
-tbl %>% 
-      rtf_body() %>% 
+colheader1 <- paste(names(tbl_base), collapse = "|")
+colheader2 <- paste(tbl_base[1, ], collapse = "|")
+rel_width <- c(2.5, rep(1, 4))
+
+tbl_base[-1, ] %>% 
+      rtf_title("Participant Baseline Characteristics", 
+                "(All Participants Randomized)") %>%
+      rtf_colheader(colheader1, 
+                    col_rel_width = rel_width) %>% 
+      rtf_colheader(colheader2, 
+                    border_top = "",
+                    col_rel_width = rel_width) %>%
+      rtf_body(col_rel_width = rel_width,
+               text_justification = c("l", rep("c", 4)), 
+               text_indent_first = -240,
+               text_indent_left = 180) %>% 
       rtf_encode() %>% 
-      write_rtf("tlf/bs-1.rtf")
+      write_rtf("tlf/tlf_base.rtf")
 ```
@@
 ```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/bs-1.pdf")
+knitr::include_graphics("tlf/tlf_base.pdf")
 ```

### 2021-08-29 tlf-efficacy.Rmd
add efficacy ananylsis example

# Efficacy

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we need to summarize primary and secondary efficacy endpoints. 
in the section 11.4, efficacy results and tabulations of individual patient data of a CSR.

```{r}
library(haven)   # Read SAS data
library(dplyr)   # Manipulate data
library(tidyr)   # Manipulate data
library(r2rtf)   # Reporting in RTF format
library(emmeans) # LS mean estimation
```

For efficacy analysis, we only analyze change from baseline glucose data at Week 24.

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tlf_eff.pdf")
```

## Analysis Dataset

To prepare analysis dataset, 
we need both `adsl` and `adlbc` datasets for this analysis.

```{r}
adsl <- read_sas("adam_data/adsl.sas7bdat")
adlb <- read_sas("adam_data/adlbc.sas7bdat")
```

We first define the analysis dataset using 
efficacy population flag `EFFFL` and 
all records post baseline (`AVISITN > 1`) and on or before Week 24 (`AVISITN <= 12`).

```{r}
gluc <- adcibic %>%
  left_join(adsl %>% select(USUBJID, EFFFL), by = "USUBJID") %>%
  filter(EFFFL == "Y" & PARAMCD == "GLUC") %>% 
  arrange(TRTPN) %>% 
  mutate(TRTP = factor(TRTP, levels = unique(TRTP)))

ana <- gluc %>%
  filter(AVISITN > 0 & AVISITN <= 24) %>%
  arrange(AVISITN) %>%
  mutate(AVISIT = factor(AVISIT, levels = unique(AVISIT)))
```

Below is the first few records of the analysis dataset.

```{r}
ana %>%
  select(USUBJID, TRTPN, AVISIT, AVAL, BASE, CHG) %>%
  head(4)
```
## Helper Functions

To prepare report dataframe, we created a few helper functions 
by using the `fmt_num` function defined in Chapter \@ref(population).

- Format Estimators

```{r}
fmt_est <- function(.mean,
                    .sd,
                    digits = c(1, 2)) {
  .mean <- fmt_num(.mean, digits[1], width = digits[1] + 4)
  .sd <- fmt_num(.sd, digits[2], width = digits[2] + 3)
  paste0(.mean, " (", .sd, ")")
}
```

- Format Confidence Interval

```{r}
fmt_ci <- function(.est,
                   .lower,
                   .upper,
                   digits = 2,
                   width = digits + 3) {
  .est <- fmt_num(.est, digits, width)
  .lower <- fmt_num(.lower, digits, width)
  .upper <- fmt_num(.upper, digits, width)
  paste0(.est, " (", .lower, ",", .upper, ")")
}
```

- Format P-Value

```{r}
fmt_pval <- function(.p, digits = 3) {
  scale <- 10^(-1 * digits)
  p_scale <- paste0("<", digits)
  if_else(.p < scale, p_scale, fmt_num(.p, digits = digits))
}
```

## Summary of Observed Data

We first summarize observed data at Baseline and Week 24

```{r}
t11 <- gluc %>%
  filter(AVISITN %in% c(0, 24)) %>%
  group_by(TRTPN, TRTP, AVISITN) %>%
  summarise(
    n = n(),
    mean_sd = fmt_est(mean(AVAL), sd(AVAL))
  ) %>%
  pivot_wider(
    id_cols = c(TRTP, TRTPN),
    names_from = AVISITN,
    values_from = c(n, mean_sd)
  )

t11
```

We also summarize observed change from baseline glucose at Week 24.

```{r}
t12 <- gluc %>%
  filter(AVISITN %in% 24) %>%
  group_by(TRTPN, AVISITN) %>%
  summarise(
    n_chg = n(),
    mean_chg = fmt_est(
      mean(CHG, na.rm = TRUE),
      sd(CHG, na.rm = TRUE)
    )
  )

t12
```

## Missing Data Imputation

In clinical trials, missing data is inevitable. 
In this study, there are also missing values in glucose data.

```{r}
count(ana, AVISIT)
```

For simplicity and illustration purpose, 
we use the last observation carried forward (LOCF) approach to handle missing data.
LOCF approach is a single imputation approach that is **not recommended** 
in real studies. 
Interested readers can find more discussion on missing data approaches in the book:
[The Prevention and Treatment of Missing Data in Clinical Trials](https://www.ncbi.nlm.nih.gov/books/NBK209904/pdf/Bookshelf_NBK209904.pdf).

```{r}
ana_locf <- ana %>%
  group_by(USUBJID) %>%
  mutate(locf = AVISITN == max(AVISITN)) %>%
  filter(locf)
```

## ANCOVA model

We start to analyze the imputed data using ANCOVA model with treatment and baseline glucose as covariates.

```{r}
fit <- lm(CHG ~ BASE + TRTP, data = ana_locf)
summary(fit)
```

Then we use the `emmeans` R package to obtain 
within and between group least square mean (LS mean)

```{r}
fit_within <- emmeans(fit, "TRTP")
fit_within
```

```{r}
t13 <- fit_within %>%
  as_tibble() %>%
  mutate(ls = fmt_ci(emmean, lower.CL, upper.CL)) %>%
  select(TRTP, ls)
t13
```

```{r}
fit_between <- pairs(fit_within, reverse = TRUE)
fit_between
```

```{r}
t2 <- fit_between %>%
  as_tibble() %>%
  mutate(
    ls = fmt_ci(
      estimate,
      estimate - 1.96 * SE,
      estimate + 1.96 * SE
    ),
    p = fmt_pval(p.value)
  ) %>%
  filter(str_detect(contrast, "- Placebo")) %>%
  select(contrast, ls, p)

t2
```

## Reporting

We combine `t11`, `t12` and `t13` to get the first part of the report data

```{r}
t1 <- cbind(
  t11 %>% ungroup() %>% select(TRTP, ends_with("0"), ends_with("24")),
  t12 %>% ungroup() %>% select(ends_with("chg")),
  t13 %>% ungroup() %>% select(ls)
)
t1
```

Then we use `r2rtf` to prepare the table format for `t1`.
We also highlight how to handle special character in this example.

Special characters `^` and `_` are used to define superscript and subscript of text. And `{}` is to define the part that will be impacted. 
For example, `{^a}` provide a superscript `a` for footnote notation.   
`r2rtf` also support most LaTeX characters. 
Examples can be found in
[`r2rtf` Get Started Page](https://merck.github.io/r2rtf/articles/r2rtf.html#special-character). 
The `text_convert` argument in `r2rtf_xxx` functions control whether convert special characters.

```{r}
t1_rtf <- t1 %>% 
  data.frame() %>%
  rtf_title(c(
    "ANCOVA of Change from Baseline Glucose (mmol/L) at Week 24",
    "LOCF",
    "Efficacly Analysis Population"
  )) %>%
  rtf_colheader("| Baseline | Week 24 | Change from Baseline",
    col_rel_width = c(2.5, 2, 2, 4)
  ) %>%
  rtf_colheader(paste(
    "Treatment |",
    paste0(rep("N | Mean (SD) | ", 3), collapse = ""),
    "LS Mean (95% CI){^a}"
  ),
  col_rel_width = c(2.5, rep(c(0.5, 1.5), 3), 2)
  ) %>%
  rtf_body(text_justification = c("l", rep("c", 7)), 
           col_rel_width = c(2.5, rep(c(0.5, 1.5), 3), 2)) %>%
  rtf_footnote(c(
    "{^a}Based on an ANCOVA model after adjusting baseline value. LOCF approach is used to impute missing values.",
    "ANCOVA = Analysis of Covariance, LOCF = Last Observation Carried Forward",
    "CI = Confidence Interval, LS = Least Squares, SD = Standard Deviation"
  ))

t1_rtf %>%
  rtf_encode() %>%
  write_rtf("tlf/tlf_eff1.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tlf_eff1.pdf")
```

we also use `r2rtf` to prepare the table format for `t2`

```{r}
t2_rtf <- t2 %>% 
  data.frame() %>%
  rtf_colheader("Pairwise Comparison | Difference in LS Mean (95% CI){^a} | p-Value",
    col_rel_width = c(4.5, 4, 2)
  ) %>%
  rtf_body(text_justification = c("l", "c", "c"), 
           col_rel_width = c(4.5, 4, 2))

t2_rtf %>%
  rtf_encode() %>%
  write_rtf("tlf/tlf_eff2.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tlf_eff2.pdf")
```

Finally we combine the two parts to get the final table using `r2rtf`.
This is achieved by providing a list of `t1_rtf` and `t2_rtf` as input for
`rtf_encode`.

```{r}
list(t1_rtf, t2_rtf) %>%
  rtf_encode() %>%
  write_rtf("tlf/tlf_eff.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tlf_eff.pdf")
```

### 2021-08-29 tlf-baseline.Rmd
update baseline table

@@
-# Table for Baseline Characteristics
+# Baseline Characteristics
 
@@
 library(haven)
+library(dplyr)
+library(tidyr)
 library(stringr)
+library(tools)
 ```
@@
 ```{r}
-tbl <- table1(~ SEX + AGE + RACE | TRT01P, data = adsl)
+ana <- adsl %>% 
+  mutate(SEX = factor(SEX, c("F", "M"), c("Female", "Male")), 
+         RACE = toTitleCase(tolower(RACE)))
+
+tbl <- table1(~ SEX + AGE + RACE | TRT01P, data = ana)
 tbl
@@
               as.data.frame() %>% 
+              as_tibble() %>%
               mutate( across(everything(), 

### 2021-08-29 tlf-overview.Rmd
update minor formating item

@@
 ```{r}
-library(r2rtf)
-
 r2rtf_adae %>%
@@
 ```{r}
-library(tidyr)
-library(dplyr)
-
 tbl <- r2rtf_adae %>%

### 2021-08-29 tlf-ae-summary.Rmd
place holder for ae tables

--- 2021-08-28 tlf-population.Rmd
@@
-# Table for Analysis Population {#population}
+# AE Summary
 
@@
 ```
-
-In this chapter, we illustrate how to create a summary table for
-analysis population in a study.
-
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/tbl_pop.pdf")
-```
-
-The first step is to read relevant datasets into R. 
-For analysis population table, 
-all the required information is saved in the ADSL dataset.
-We can use `haven` package to read the dataset.
-
-```{r}
-adsl <- read_sas("adam_data/adsl.sas7bdat")
-```
-
-We illustrate how to prepare a report data for a simplified 
-analysis population table using variables below:
-
-- USUBJID: Unique subject identifier 
-- ITTFL: Intent-to-treat population flag
-- EFFFL: Efficacy population flag
-- SAFFL: Safty population flag
-
-```{r}
-adsl %>%
-  select(USUBJID, ITTFL, EFFFL, SAFFL) %>%
-  head(4)
-```
-## Helper Functions
-
-Before we write analysis code, let's discuss the possibility to
-reuse R code by writing some toy helper function.
-
-As discussed in 
-[R for data science](https://r4ds.had.co.nz/functions.html#when-should-you-write-a-function),
-"You should consider writing a function 
-whenever you’ve copied and pasted a block of code more than twice"
-
-In Chapter \@ref(disposition), there are few repeating steps to:
-
-- format percentage using `formatC` 
-- calculate number and percentage by groups
-
-We create two ad-hoc functions and apply them to all rest of tables.
-
-To format number and percentage, we created a function called `fmt_num`.
-It is a very simple function that is a wrapper of `formatC`.
-
-```{r}
-fmt_num <- function(x, digits, width = digits + 4) {
-  formatC(x,
-    digits = digits,
-    format = "f",
-    width = width
-  )
-}
-```
-
-The main reason to create `fmt_num` is to enhance readability of the analysis code.
-
-For example, we can compare the two versions of code to format percentage 
-used in \@ref(disposition) and `fmt_num`.
-
-```{r, eval = FALSE}
-formatC(n / n() * 100,
-  digits = 1, format = "f", width = 5
-)
-```
-
-```{r, eval = FALSE}
-fmt_num(n / n() * 100, digits = 1)
-```
-
-To calculate number and percentage of patients by groups, 
-we provide a simple (but not robust) wrapper function, `count_by` using `dplyr` and `tidyr`.
-
-The function can be enhanced in multiple ways, but here we only focus on simplificy and readability.
-More details about writting R functions can be found in [Stat454 course](https://stat545.com/functions-part1.html).
-
-```{r}
-count_by <- function(data, # Input data set
-                     grp, # Group variable
-                     var, # Analysis variable
-                     var_label = var, # Analysis variable label
-                     id = "USUBJID") { # Subject ID variable
-  data <- data %>% rename(grp = !!grp, var = !!var, id = !!id)
-
-  left_join(
-    count(data, grp, var),
-    count(data, grp, name = "tot"),
-    by = "grp",
-  ) %>%
-    mutate(
-      pct = fmt_num(100 * n / tot, digits = 1),
-      n = fmt_num(n, digits = 0),
-      npct = paste0(n, " (", pct, ")")
-    ) %>%
-    pivot_wider(
-      id_cols = var,
-      names_from = grp,
-      values_from = c(n, pct, npct),
-      values_fill = list(n = "0", pct = fmt_num(0))
-    ) %>%
-    mutate(var_label = var_label)
-}
-```
-
-By using `count_by`, we can simplify the analysis code as below.
-
-```{r}
-count_by(adsl, "TRT01PN", "EFFFL") %>% 
-  select(- ends_with(c("_54", "_81")))
-```
-
-## Analysis Code
-
-With the helper functions `count_by`, we can easily prepare report dataset as
-
-```{r}
-# Derive a randomization flag
-adsl <- adsl %>% mutate(RANDFL = "Y")
-
-pop <- count_by(adsl, "TRT01PN", "RANDFL", 
-                var_label = "Participants in Population") %>%
-  select(var_label, starts_with("n_"))
-```
-
-```{r}
-pop1 <- bind_rows(
-  count_by(adsl, "TRT01PN", "ITTFL",
-    var_label = "Participants included in ITT population"
-  ),
-  count_by(adsl, "TRT01PN", "EFFFL",
-    var_label = "Participants included in efficacy population"
-  ),
-  count_by(adsl, "TRT01PN", "SAFFL",
-    var_label = "Participants included in safety population"
-  )
-) %>%
-  filter(var == "Y") %>%
-  select(var_label, starts_with("npct_"))
-```
-
-Now we combined individual rows into the whole table for reporting purpose. 
-`tbl_pop` is used as input for `r2rtf` to create final report.
-
-```{r}
-names(pop) <- gsub("n_", "npct_", names(pop))
-tbl_pop <- bind_rows(pop, pop1)
-
-tbl_pop %>% select(var_label, npct_0)
-```
-
-We start to define the format of the output.
-
-```{r}
-rel_width <- c(2, rep(1, 3))
-colheader <- " | Placebo | Xanomeline line Low Dose| Xanomeline line High Dose"
-tbl_pop %>%
-  # Table title
-  rtf_title(
-    "Participants Accounting in Analysis Population",
-    "(All Participants Randomized)"
-  ) %>%
-  # First row of column header
-  rtf_colheader(colheader,
-    col_rel_width = rel_width
-  ) %>%
-  # Second row of column header
-  rtf_colheader(" | n (%) | n (%) | n (%)",
-    border_top = "",
-    col_rel_width = rel_width
-  ) %>%
-  # Table body
-  rtf_body(
-    col_rel_width = rel_width,
-    text_justification = c("l", rep("c", 3))
-  ) %>%
-  # Encoding RTF syntax
-  rtf_encode() %>%
-  # Save to a file
-  write_rtf("tlf/tbl_pop.rtf")
-```
-
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/tbl_pop.pdf")
-```

### 2021-08-29 tlf-efficacy.Rmd
bug fixed: using adlb data

@@
 ```{r}
-gluc <- adcibic %>%
+gluc <- adlb %>%
   left_join(adsl %>% select(USUBJID, EFFFL), by = "USUBJID") %>%

### 2021-08-29 tlf-ae-summary.Rmd
add ae summary table

@@
 we need to summarize which patients were included in each efficacy analysis 
-in the section 11.1, data sets analysed of a CSR.
+in the section 12.2, adverse events of a CSR.
 
@@
 ```
+
+In this chapter, we illustrate how to summarize AE information in a study.
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/tlf_ae_summary.pdf")
+```
+
+The data used to summarize AE information is in `adsl` and `adae` datasets.
+
+```{r}
+adsl <- read_sas("adam_data/adsl.sas7bdat")
+adae <- read_sas("adam_data/adae.sas7bdat")
+```
+
+We first summarize participants in population by treatment group.
+
+```{r}
+pop <- adsl %>%
+  filter(SAFFL == "Y") %>%
+  rename(TRTAN = TRT01AN) %>%
+  count(TRTAN, name = "tot")
+
+pop
+```
+
+We transform the data to simplify the analysis of each required AE criteria of interest.
+
+- With one or more adverse events
+- With drug-related adverse events
+- With serious adverse events
+- With serious drug-related adverse events
+- Who died"
+
+```{r}
+tidy_ae <- adae %>%
+  mutate(
+    all = SAFFL == "Y",
+    drug = AEREL %in% c("POSSIBLE", "PROBABLE"),
+    ser = AESER == "Y",
+    drug_ser = drug & ser,
+    die = AEOUT == "FATAL"
+  ) %>%
+  select(USUBJID, TRTAN, all, drug, ser, drug_ser, die) %>%
+  pivot_longer(cols = c(all, drug, ser, drug_ser, die))
+
+tidy_ae %>% head(4)
+```
+We summarize the number and percentage of patients meet each AE criteria.
+
+```{r}
+ana <- tidy_ae %>%
+  filter(value == TRUE) %>%
+  group_by(TRTAN, name) %>%
+  summarise(n = n_distinct(USUBJID)) %>%
+  left_join(pop, by = "TRTAN") %>%
+  mutate(
+    pct = fmt_num(n / tot * 100, digits = 1),
+    n = fmt_num(n, digits = 0),
+    pct = paste0("(", pct, ")")
+  )
+
+ana %>% head(4)
+```
+We prepare reporting ready dataset for each AE criteria.
+
+```{r}
+t_ae <- ana %>%
+  pivot_wider(
+    id_cols = "name",
+    names_from = TRTAN,
+    values_from = c(n, pct),
+    values_fill = list(
+      n = "   0",
+      pct = "(  0.0)"
+    )
+  )
+
+t_ae <- t_ae %>%
+  mutate(name = factor(
+    name,
+    c("all", "drug", "ser", "drug_ser", "die"),
+    c(
+      "With one or more adverse events",
+      "With drug-related adverse events",
+      "With serious adverse events",
+      "With serious drug-related adverse events",
+      "Who died"
+    )
+  )) %>%
+  arrange(name)
+```
+
+We prepare reporting ready dataset for analysis population.
+
+```{r}
+t_pop <- adsl %>%
+  filter(SAFFL == "Y") %>%
+  count_by("TRT01AN", "SAFFL",
+    var_label = "Participants in population"
+  ) %>%
+  rename(name = var_label) %>%
+  select(name, starts_with("n_"))
+
+t_pop
+```
+The final report data is saved in `tbl_ae_summary`
+
+```{r}
+tbl_ae_summary <- bind_rows(t_pop, t_ae) %>%
+  select(name, ends_with("_0"), ends_with("_54"), ends_with("_81"))
+
+tbl_ae_summary
+```
+
+We start to define the format of the output.
+
+```{r}
+tbl_ae_summary %>% 
+  rtf_title(
+    "Analysis of Adverse Event Summary",
+    "(Safety Analysis Population)"
+  ) %>%
+  rtf_colheader(" | Placebo | Xanomeline Low Dose| Xanomeline High Dose",
+    col_rel_width = c(3.5, rep(2, 3))
+  ) %>%
+  rtf_colheader(" | n | (%) | n | (%) | n | (%)",
+    col_rel_width = c(3.5, rep(c(0.7, 1.3), 3)),
+    border_top = c("", rep("single", 6)),
+    border_left = c("single", rep(c("single", ""), 3))
+  ) %>%
+  rtf_body(
+    col_rel_width = c(3.5, rep(c(0.7, 1.3), 3)),
+    text_justification = c("l", rep("c", 6)),
+    border_left = c("single", rep(c("single", ""), 3))
+  ) %>%
+  rtf_footnote("Every subject is counted a single time for each applicable row and column.") %>%
+  rtf_encode() %>%
+  write_rtf("tlf/tlf_ae_summary.rtf")
+```
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/tlf_ae_summary.pdf")
+```

### 2021-08-29 tlf-specific-ae.Rmd
place holder for ae tables
add specific AE example

--- 2021-08-29 tlf-ae-summary.Rmd
@@
-# AE Summary
+# Specific AE
 
@@
 ```
+
+In this chapter, we illustrate how to summarize simplified specific AE information in a study.
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/tlf_spec_ae.pdf")
+```
+
+The data used to summarize AE information is in `adsl` and `adae` datasets.
+
+```{r}
+adsl <- read_sas("adam_data/adsl.sas7bdat")
+adae <- read_sas("adam_data/adae.sas7bdat")
+```
+
+For illustration purpose, we only provide count in the simplified table. 
+The percentage of participants 
+for each AE criteria can be calculated as in Chapter \@ref(aesummary).
+
+In this way, let's focus on the analysis script for advanced feature for table layout.
+
+- group content: AE can be summarized in multiple nested layer. 
+(e.g. by system organ class (SOC, `AESOC`) and specific AE term (`AEDECOD`))
+- pagenization: there are many AE terms that can not be covered in one page. 
+Column headers and SOC information needs to be repeated in every page.
+
+In the code below, we count number of subjects in each AE term by SOC and treatment group.
+
+```{r}
+ana <- adae %>% 
+  mutate(AESOC = toTitleCase(tolower(AESOC)), 
+         AEDECOD =  toTitleCase(tolower(AEDECOD)))
+
+t1 <-ana %>% 
+  group_by(TRTAN, AESOC) %>%
+  summarise(n = fmt_num(n_distinct(USUBJID), digits = 0)) %>% 
+  mutate(AEDECOD = AESOC, order = 0)
+
+t1 %>% head(4)
+```
+In the code below, we count number of subjects in each AE term by SOC, AE term, and treatment group.
+
+```{r}
+t2 <- ana %>% 
+  group_by(TRTAN, AESOC, AEDECOD) %>%
+  summarise(n = fmt_num(n_distinct(USUBJID), digits = 0)) %>%
+  mutate(order = 1)
+
+t2 %>% head(4)
+```
+We prepare reporting data for AE information.
+
+```{r}
+t_ae <- bind_rows(t1, t2) %>% 
+        pivot_wider(id_cols = c(AESOC, order, AEDECOD), 
+                    names_from = TRTAN, 
+                    names_prefix = "n_",
+                    values_from = n, 
+                    values_fill = fmt_num(0, digits = 0)) %>%
+        arrange(AESOC, order, AEDECOD) %>% 
+        select(AESOC, AEDECOD, starts_with("n"))
+
+t_ae %>% head(4)
+```
+We prepare reporting data for analysis population.
+
+```{r}
+t_pop <- adsl %>%
+  filter(SAFFL == "Y") %>%
+  count_by("TRT01AN", "SAFFL",
+    var_label = "Participants in population"
+  ) %>%
+  mutate(
+    AESOC = "pop", 
+    AEDECOD = var_label) %>%
+  select(AESOC,AEDECOD, starts_with("n_"))
+
+t_pop
+```
+The final report data is saved in `tbl_ae_spec`.
+We also add a blank row between population and AE information in the reporting table.
+
+```{r}
+tbl_ae_spec <- bind_rows(t_pop,
+                         data.frame(AESOC = "pop"), 
+                         t_ae) %>% 
+  mutate(AEDECOD = ifelse(AEDECOD == AESOC , 
+                          AEDECOD, paste0("  ", AEDECOD)))
+
+tbl_ae_spec %>% head(4)
+```
+We start to define the format of the output.
+
+To obtain the nested layout, we use `page_by` argument in `rtf_body` function.
+By defining `page_by="AESOC"`, 
+`r2rtf` recognize the variable as a group indicator.
+
+After setting `pageby_row = "first_row"`, the first row is displayed as group header. If a group of information is breaked into multiple page, 
+the group header row is repeated in each page by default.
+
+We can also customize the text format by providing a matrix that have 
+same dimention of the input dataset (i.e. `tbl_ae_spec`). 
+In the code below, we illustrate how to display **bold** text for group headers
+to highlight the nested structure of the table layout.
+
+```{r}
+n_row <- nrow(tbl_ae_spec)
+n_col <- ncol(tbl_ae_spec)
+id <- tbl_ae_spec$AESOC == tbl_ae_spec$AEDECOD
+id <- ifelse(is.na(id), FALSE, id)
+
+text_format   <- ifelse(id, "b", "")
+```
+
+More discussion on `page_by`, `group_by` and `subline_by` 
+features can be found in the 
+[`r2rtf` package website](https://merck.github.io/r2rtf/articles/example-sublineby-pageby-groupby.html.
+
+```{r}
+tbl_ae_spec %>% 
+  rtf_title("Analysis of Participants With Specific Adverse Events", 
+            "(Safty Analysis Population)") %>%
+
+    rtf_colheader(" | Placebo | Xanomeline Low Dose| Xanomeline High Dose",
+    col_rel_width = c(3, rep(1, 3))
+  ) %>%
+  rtf_colheader(" | n |  n | n ",
+    border_top = "",
+    border_bottom = "single",
+    col_rel_width = c(3, rep(1, 3))
+  ) %>%
+
+  rtf_body(
+    col_rel_width = c(1, 3, rep(1, 3)),
+    text_justification = c("l", "l", rep("c", 3)),
+    text_format = matrix(text_format, nrow = n_row, ncol = n_col), 
+    page_by = "AESOC", 
+    pageby_row = "first_row") %>%
+
+  rtf_footnote("Every subject is counted a single time for each applicable row and column.") %>%
+
+  rtf_encode() %>%
+
+  write_rtf("tlf/tlf_spec_ae.rtf")
+
+```
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/tlf_spec_ae.pdf")
+```

### 2021-08-29 tlf-population.Rmd
update minor formating item
pass spell checking

@@
-# Table for Analysis Population {#population}
+# Analysis Population {#population}
 
@@
 
-The function can be enhanced in multiple ways, but here we only focus on simplificy and readability.
-More details about writting R functions can be found in [Stat454 course](https://stat545.com/functions-part1.html).
+The function can be enhanced in multiple ways, but here we only focus on simplicity and readability.
+More details about writing R functions can be found in [Stat454 course](https://stat545.com/functions-part1.html).
 

### 2021-08-29 index.Rmd
add efficacy ananylsis example
update manifest file

@@
 title: "R for Clinical Study Reports and Submission"
-author: ["Yilong Zhang"]
+author: ["Yilong Zhang", "Nan Xiao", "Keaven Anderson"]
 knit: "bookdown::render_book"
@@
 library(r2rtf) 
+library(knitr)
+library(kableExtra)
 ```
@@
 ```{r, eval = FALSE, include=FALSE}
-r2rtf:::rtf_convert_format(list.files("tlf", pattern = ".rtf", full.names = TRUE), output_dir = "tlf")
+r2rtf:::rtf_convert_format(list.files("tlf", pattern = ".rtf", full.names = TRUE), 
+                           output_dir = "tlf")
 ```
@@
 This is the work-in-progress draft of the book.
+
+## List of Authors and Contributors
+
+```{r, echo = FALSE}
+contributors <- c(person("Yilong", "Zhang", email = "yilong.zhang@merck.com", role = "aut"), 
+                  person("Nan", "Xiao", role = "aut"), 
+                  person("Keaven", "Anderson", role = "aut"),
+                  person("Yujie", "Zhao", role = "ctr"))
+```
+
+The document is maintained by a community. 
+While reading the document, you can be a contributor as well. 
+The quality of this document relies on you.
+
+- Author: contribute majority content of at least one chapter. 
+- Contributor: [contribute at least one commit to the source code](https://stash.merck.com/projects/BARDS/repos/r4csr/browse)
+
+```{r}
+data.frame(contributors = as.character(contributors)) %>%   
+  kable(format = "html") %>%
+  kable_styling()
+```

### 2021-08-29 index.Rmd
update index.Rmd

@@
 
+```{r, eval = FALSE, include=FALSE}
+bookdown::render_book("index.Rmd", "bookdown::pdf_book")
+```
+
 # Welcome! {-}
@@
 
-```{r}
+```{r, echo = FALSE, eval = TRUE}
 data.frame(contributors = as.character(contributors)) %>%   

### 2021-08-29 tlf-efficacy.Rmd
flx typo: tlf-efficacy

@@
 efficacy population flag `EFFFL` and 
-all records post baseline (`AVISITN > 1`) and on or before Week 24 (`AVISITN <= 12`).
+all records post baseline (`AVISITN > 1`) and on or before Week 24 (`AVISITN <= 24`).
 

### 2021-09-02 tlf-ae-spec.Rmd
pass spell checking
proofread of chapter 1-8

--- 2021-08-29 tlf-specific-ae.Rmd
@@
 we need to summarize which patients were included in each efficacy analysis 
-in the section 12.2, adverse events of a CSR.
+in Section 12.2, aAdverse Events (AEs).
 
@@
 
-In the code below, we count number of subjects in each AE term by SOC and treatment group.
+In the code below, we count the number of subjects in each AE term by SOC and treatment group, and we create a new variable `order` and set it as `0` for these counted numbers. 
+The variable `order` will help with the data manipulation later.
 
@@
 In the code below, we count number of subjects in each AE term by SOC, AE term, and treatment group.
+Here we also create a new variable `order` and set it as `1` for these counted numbers.
 
@@
 We can also customize the text format by providing a matrix that have 
-same dimention of the input dataset (i.e. `tbl_ae_spec`). 
+same dimension of the input dataset (i.e. `tbl_ae_spec`). 
 In the code below, we illustrate how to display **bold** text for group headers
@@
 ```
+
+In conclusion, the procedure to generate a specific AE summary table as shown in the above example is listed as follows:
+
+- Step 1: Read data into R, i.e., `adae` and `adsl`.
+- Step 2: Count the number of subjects by SOC and treatment group (rows with bold text) and store it in a data frame `t1`.
+- Step 3: Count the number of subjects in each AE term by SOC, AE term, and treatment group (rows without bold text) and store it in a data frame `t2`.
+- Step 4: Rowly bind `t1` and `t2` into `t_ae`.
+- Step 5: Count the number of subjects in each arm as `t_pop`.
+- Step 6: Rowly bind `t_pop` with `t_ae` as `tbl_ae_spec`.
+- Step 7: Format the output by `r2rtf`.

### 2021-09-02 tlf-ae-summary.Rmd
add specific AE example
proofread of chapter 1-8

@@
-# AE Summary
+# AE Summary {#aesummary}
 
@@
 we need to summarize which patients were included in each efficacy analysis 
-in the section 12.2, adverse events of a CSR.
+in Section 12.2, Adverse Events (AEs).
 
@@
 ```
+
+In conclusion, the procedure to generate a AE summary table as shown in the above example is listed as follows:
+
+- Step 1: Read data into R, i.e., `adae` and `adsl`.
+- Step 2: Summarize participants in population by treatment group, and we name the dataset as `t_pop`.
+- Step 3: Summarize participants in population by required AE criteria of interest, and we name the dataset as `t_ae`.
+- Step 4: Rowly bind `t_pop` and `t_ae` and format it by `r2rtf`.

### 2021-09-02 tlf-disposition.Rmd
add analysis population example
update minor formating item
proofread of chapter 1-8

@@
-# Table for Study Patients
+# Disposition {#disposition}
 
 Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to summarize accounting of all patients who entered the study, 
-in the section 10.1, disposition of patients of a CSR.
+we need to summarize accounting of all patients who entered the study, in Section 10.1, Disposition of Patients.
+The disposition of patients reports the numbers of patients who were randomized, and who entered and completed each phase of the study, as well as the reasons for all post-randomization discontinuations, grouped by treatment and by major reason (lost to follow-up, adverse event, poor compliance etc.).
 
@@
 ```
+
+In conclusion, the procedure to generate a AE summary table as shown in the above example is listed as follows:
+
+- Step 1: Read data into R, i.e. `adsl`.
+- Step 2: Count patients in the analysis population and name the dataset as `n_rand`.
+- Step 3: Count the number and percentage of patients who discontinued the study, and name the dataset as `n_disc`.
+- Step 4: Count the number and percentage of of patients who discontinued the study in different reasons, and name the dataset as `n_reason`.
+format it by `r2rtf`.
+- Step 5: Count the number and percentage of of patients who completed the study, and name the dataset as `n_complete`.
+- Step 6: Rowly bind `n_rand`, `n_disc`, `n_reason` and `n_complete` and format it by `r2rtf`.

### 2021-09-02 tlf-efficacy.Rmd
proofread of chapter 1-8

@@
 Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to summarize primary and secondary efficacy endpoints. 
-in the section 11.4, efficacy results and tabulations of individual patient data of a CSR.
+we need to summarize primary and secondary efficacy endpoints,  
+in Section 11.4, Efficacy Results and Tabulations of Individual Patient.
 
@@
 all records post baseline (`AVISITN > 1`) and on or before Week 24 (`AVISITN <= 24`).
+Here the variable `AVISITN` is the numerical analysis visit. 
+For example, if the analysis visit is recorded as "Baseline", i.e., `AVISIT = Baseline`, then `AVISITN = 0`; 
+if the analysis visit is recorded as "Week 24", i.e., `AVISIT = Week 24`, then `AVISITN = 24`;
+if the analysis visit is blank, then `AVISITN` is also blank.
+We will discuss these missing values in Section 6.4.
 
@@
   left_join(adsl %>% select(USUBJID, EFFFL), by = "USUBJID") %>%
-  filter(EFFFL == "Y" & PARAMCD == "GLUC") %>% 
+  filter(EFFFL == "Y" & PARAMCD == "GLUC") %>% # PARAMCD is parameter code and here we focus on Glucose (mg/dL) 
   arrange(TRTPN) %>% 
@@
 ana %>%
-  select(USUBJID, TRTPN, AVISIT, AVAL, BASE, CHG) %>%
+  select(USUBJID, TRTPN, AVISIT, AVAL, BASE, CHG) %>%  # AVAL: analysis value
+                                                       # BASE: baseline value
+                                                       # CHG: change from baseline
   head(4)
@@
 ```
+
+In conclusion, the procedure to generate the above efficacy results table is summarized as follows.
+
+- Step 1: Read the data into R, i.e., `adsl` and `adlb`.
+- Step 2: Define the analysis dataset.
+In this example, we define two analysis datasets.
+The first dataset is the efficacy population (`gluc`).
+The second dataset is the collection of all records post baseline and on or before week 24 (`ana`).
+- Step 3: Impute the missing values. In our example, we name the `ana` dataset after imputation as `ana_locf`.
+- Step 4: Summarize `gluc`, i.e., calculate the mean and standard derivation, and then format it into a rtf table.
+- Step 5: Summarize `ana_locf`, i.e., calculate the pairwise comparison by ANCOVA model, and then format it into a rtf table.
+- Step 6: Rowly bind the output in Step 4 and Step 5.

### 2021-09-02 tlf-ae-spec.Rmd
fix typo. reorg comments

@@
 we need to summarize which patients were included in each efficacy analysis 
-in Section 12.2, aAdverse Events (AEs).
+in Section 12.2, Adverse Events (AEs).
 

### 2021-09-02 tlf-ae-summary.Rmd
update ch7 `t_pop` code

@@
 ```{r}
-t_pop <- adsl %>%
-  filter(SAFFL == "Y") %>%
-  count_by("TRT01AN", "SAFFL",
-    var_label = "Participants in population"
-  ) %>%
-  rename(name = var_label) %>%
-  select(name, starts_with("n_"))
+t_pop <- pop %>% 
+  mutate(name = "Participants in population", 
+         tot = fmt_num(tot, digits = 0)) %>% 
+  pivot_wider(id_cols = name, 
+              names_from = TRTAN, 
+              names_prefix = "n_", 
+              values_from = tot)
 

### 2021-09-02 tlf-baseline.Rmd
pass spell checking

@@
 we need to summarize critical demographic and baseline characteristics of the patients
-in the section 11.2, demographic and other baseline characteristics of a CSR.
+in Section 11.2, Demographic and Other Baseline Characteristics.
 
@@
 
-There is many R packages that can effciently summarize baseline information.
+There is many R packages that can efficiently summarize baseline information.
 The [`table1`](https://github.com/benjaminrich/table1) R package is one of them.
@@
 More details of the `table` R package can be found in the package
-[vignettes](https://benjaminrich.github.io/table1/vignettes/table1-examples.html)
+[vignettes](https://benjaminrich.github.io/table1/vignettes/table1-examples.html).
 
@@
 
-We start to define the format of the output. We highlighted items that 
-is not discussed in previous discussion.
+We start to define the format of the output. We highlight items that 
+are not discussed in previous discussion.
 
@@
 ```
+
+In conclusion, the procedure to generate demographic and baseline characteristics table is summarized as follows.
+
+- Step 1: Read the data set.
+- Step 2: Use `table1::table1()` to get the baseline characteristics table.
+- Step 3: Transfer the output in Step 2 into a data frame that only contains ASCII character.
+- Step 4: Define the format of the rtf table by using R package `r2rtf`.

### 2021-09-02 tlf-efficacy.Rmd

@@
   left_join(adsl %>% select(USUBJID, EFFFL), by = "USUBJID") %>%
-  filter(EFFFL == "Y" & PARAMCD == "GLUC") %>% # PARAMCD is parameter code and here we focus on Glucose (mg/dL) 
+  # PARAMCD is parameter code and here we focus on Glucose (mg/dL) 
+  filter(EFFFL == "Y" & PARAMCD == "GLUC") %>% 
   arrange(TRTPN) %>% 
@@
 
+- AVAL: analysis value
+- BASE: baseline value
+- CHG: change from baseline
+
 ```{r}
 ana %>%
-  select(USUBJID, TRTPN, AVISIT, AVAL, BASE, CHG) %>%  # AVAL: analysis value
-                                                       # BASE: baseline value
-                                                       # CHG: change from baseline
+  select(USUBJID, TRTPN, AVISIT, AVAL, BASE, CHG) %>%  
   head(4)

### 2021-09-13 intro.Rmd
Changes on Intro

@@
 
-In clinical trial development, developers needs to organize source code to deliver Study Data Tabulation Model (SDTM) or Analysis Dataset Model (ADaM) datasets and table, listing and figures (TLFs).
+In clinical trial development, developers needs to organize source code to deliver Study Data Tabulation Model (SDTM) or Analysis Dataset Model (ADaM) datasets and tables, listings and figures (TLFs).
 For example, a Phase 3 clinical trial typically require hundreds of TLFs. 
@@
 
-- For an organization, it simplify process, tool, template, and training development, because a unified folder structure is used to develop and maintain standard tool and analysis project.
+- For an organization, it simplifies process, tool, template, and training development, because a unified folder structure is used to develop and maintain standard tool and analysis project.
 
@@
 
-- Part 1, "Delivering TLFs in CSR" provides general information with examples to create table, listing and figures. 
-In this part, we assume readers are an individual contributor of a clinical project with some experience in R programming. 
+- Part 1, "Delivering TLFs in CSR" provides general information with examples to create tables, listings and figures. 
+In this part, we assume readers are individual contributors of a clinical project with some experience in R programming. 
 We expect readers are familiar with data manipulation in R. Some good reference includes [Hands-On Programming with R](https://rstudio-education.github.io/hopr/), 
@@
 
-- Part 3, "eCTD Submission package" provides general information in preparing 
+- Part 3, "eCTD Submission package" (maybe provide the full spelling for "eCTD" here) provides general information in preparing 
 submission package related to clinical study report (CSR) in 

### 2021-09-15 tlf-overview.Rmd
init commit

@@
 Within eCTD, clinical study reports (CSR) are located at module 5. 
-[ICH E3 guidance](https://www.ich.org/page/efficacy-guidelines) provide 
+[ICH E3 guidance](https://www.ich.org/page/efficacy-guidelines) provides 
 compilation of structure and content of clinical study reports.
@@
 Here, we focus on the work and deliverables completed by statisticians and statistical programmers. 
-In an organization, A group of statisticians and statistical programmers commonly work together to 
+In an organization, a group of statisticians and statistical programmers commonly work together to 
 define, develop, validate and deliver tables, listings and figures (TLFs) required for a CSR.

### 2021-09-15 tlf-population.Rmd
init commit

@@
 we need to summarize which patients were included in each efficacy analysis 
-in the section 11.1, data sets analysed of a CSR.
+in Section 11.1, Data Sets Analysed.
 
@@
 "You should consider writing a function 
-whenever you’ve copied and pasted a block of code more than twice"
+whenever you’ve copied and pasted a block of code more than twice".
 
@@
 
-- format percentage using `formatC` 
-- calculate number and percentage by groups
+- format percentages using `formatC` 
+- calculate numbers and percentages by groups
 
@@
 
-To format number and percentage, we created a function called `fmt_num`.
-It is a very simple function that is a wrapper of `formatC`.
+To format numbers and percentages, we created a function called `fmt_num`.
+It is a very simple function wrapping `formatC`.
 
@@
 
-To calculate number and percentage of patients by groups, 
+To calculate numbers and percentages of patients by groups, 
 we provide a simple (but not robust) wrapper function, `count_by` using `dplyr` and `tidyr`.
@@
 
-Now we combined individual rows into the whole table for reporting purpose. 
+Now we combine individual rows into the whole table for reporting purpose. 
 `tbl_pop` is used as input for `r2rtf` to create final report.
@@
 ```
+
+In conclusion, the procedure to generate a population summary table as shown in the above example is listed as follows:
+
+- Step 1: Read data into R, i.e.,  `adsl`.
+- Step 2: Rowly bind the counts/percentages of the ITT population, the efficacy population and the safety population. These these types of population are generated by a help function `count_by`. 
+- Step 3: Format the output in Step 2 by `r2rtf`.

### 2021-09-15 index.Rmd
add bookdown

@@
 library(kableExtra)
+library(bookdown)
 ```

### 2021-09-15 project-folder.Rmd
pass spell checking
add placeholder

--- 2021-08-11 project-folder.Rmd
@@
 
-# Project Folder
+# Overview
 
-In a late-stage clinical trial, the number of A&R deliverables can easily be in the hundreds. 
-For an organization, it is possible to have multiple ongoing clinical trials in a clincial program.
-
-A clearly defined project folder structure can have many benefits to 
-clinical development teams in an organization. 
-Specifically a well defined project folder can achieve:
-
-- Consistency
-- Reproducibility
-- Automation
-- Compliance
-
-## Consistency
-
-For consistency, a well defined folder structure with potential templates
-ensures project teams organize A&R work consistently across multiple projects. 
-Consistent folder structure also reduce communication cost between study team members 
-and enhances the transparency of projects.
-
-For R user, you already benefit from a well defined and consistent folder structure. 
-That is the R package folder structure. 
-Every R package developer is required to follow the same convention to organize their R
-functions, before the function package can be disseminated through the Comprehensive R Archive Network
-(CRAN). As a user, you can easily install and user those packages after installing from CRAN.
-
-We recommend to also use R package folder structure to organize analysis scripts 
-for clincial trial development. 
-Using the R package folder structure to streamline data analysis work has been proposed 
-(@marwick2018packaging, @wuanalysis).
-
-In this book, we refer an R package as **project specific R package** 
-if the purpose of an R package is to organize analysis scripts for a clinical project.
-We refer an R package as **standard R package** if the purpose of an R package is to 
-share commonly used R functions to be hosted in an code repository such as CRAN.
-
-Below is a minimal sufficient folders and files based on the R package folder structure.
-
-- `*.Rproj`: RStudio project file used to open RStudio project.
-- `DESCRIPTION`: Metadata for a package including authors, license, dependency etc.
-- `vignettes/`: Analysis scripts using Rmarkdown.
-- `R/`: Project specific R functions.
-- `man/`: Manual of project specific R functions.
-
-We demonstrate the the idea using the [`esubdemo`](https://github.com/elong0527/esubdemo) project.
-
-A general discussion of the R package folder structure can be found in Chapter 4 of “R packages:
-organize, test, document, and share your code” book (@wickham2015r).
-
-## Reproducibility
-
-Reproducibility of analysis can be challenge. 
-By using R package folder structure and proper tools (e.g. `renv`, `packrat`), 
-we can easily achieve reproducibility for R and R package versions.
-
-> This is the same level of reproducibility in most SAS environment.
-> https://support.sas.com/techsup/pcn/altopsys.html
-
-In `esubdemo` project, a reproducible environment is created when you open the 
-`esubdemo.Rproj` from RStudio. 
-While RStudio open `esubdemo` project, RStudio will automatically execute R code in `.Rprofile`.
-So we are using `.Rprofile` to setup a reproducible environment. 
-More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html).
-
-> `.Rprofile` is only for project specific R package. 
-> A standard R package should not use `.Rprofile`.
-
-After the `esubdemo` project is open, the code in `.Rprofile` will automatically check 
-current R version is the same as we defined in `.Rprofile`
-
-```{r, eval = FALSE}
-R_version <- "4.0.1"	# set up project R version
-```
-
-If the R version mismatch, an error message is displayed.
-
-```
-Error: The current R version is not the same as the current project in R4.0.1
-```
-
-If we are running under the same R version, 
-you shall be able to see similar messages to inform how we control R package version.
-
-```
-* Project '~/esubdemo' loaded. [renv 0.14.0]
-Current oroject R package repository:
-    https://mran.microsoft.com/snapshot/2021-08-06
-```
-
-The R package version is controlled in two layers. 
-Firstly, we define a snapshot date in `.Rprofile`. 
-The snapshot date allows to freeze the repository. 
-In other words, all R packages installed in this RStudio project 
-are based on the frozen R version at the snapshot date. 
-Here it is `2021-08-06` by using Microsoft R Application Network (MRAN) server.
-
-```{r, eval = FALSE}
-snapshot  <- "2021-08-06" 									                           # set up snapshot date
-repos     <- paste0("https://mran.microsoft.com/snapshot/", snapshot)  # set up repository based on snapshot
-```
-
-> RStudio Package Manager (RSPM) provided a solution to host both 
-> public avaiable and internally developed R packages. 
-> However, public RSPM server (https://packagemanager.rstudio.com/client/#/repos/2/overview) 
-> did not provide daily snapshot as MRAN had.
-
-Secondly, we use `renv` to lock R package versions and saved in `renv.lock` file. 
-The exact same R package versions can be restored with `renv::restore()`.
-An introduction of `renv` can be found in its [package website](https://rstudio.github.io/renv/articles/renv.html).
-
-```
-* Project '~/github-repo/esubdemo' loaded. [renv 0.14.0]
-```
-
-> `renv.lock` file and `renv/` folder are only for project specific R package. 
-> A standard R package should not use `renv`.
-
-Once R packages have been properly installed, 
-the system will use the R packages located in the search path defined based on the order of `.libPaths()`.
-The startup message also provided the R package search path.
-
-```
-Below R package path are searching in order to find installed R pacakges in this R session:
-    /home/esubdemo/renv/library/R-4.0/x86_64-pc-linux-gnu
-    /rtmp/Rtmp8d5mcL/renv-system-library
-```
-
-For an organization, a cloud based R environment (e.g. RStudio Workbench) 
-can enhance the development and operation of a clinical development team by 
-centralize managing R and R package version. 
-More details can be found in [https://environments.rstudio.com/](https://environments.rstudio.com/).
-
-In summary, to achieve reproducibility for a project specific R package, 
-a clinical project team can work under a controlled R environment 
-in the same R version and R package versions defined by a repository snapshot date.
-
-## Automation
-
-- ToDo
-
-There are additional folders and files required in this demo for an analysis project
-
-> people may use different folder name as it is not a standard R package folder.
-> These folders and files needs to be added in `.Rbuildignore` to pass
-> R package check
-
-- `adam`: ADaM datasets in `.sas7bdat` format.
-  + one may also put it in `inst/exdata` following [R package convention](https://r-pkgs.org/data.html)
-  + in reality, we suggest to have real data saved outside of this project. (e.g. in a database)
-- `output`: TLFs output 
-- `renv.lock` and `renv`: R package management using `renv` package. ([Introduction](https://rstudio.github.io/renv/articles/renv.html)) 
-- `_pkgdown.yml`: [pkgdown](https://pkgdown.r-lib.org/articles/pkgdown.html) configuration file
-- `.Rprofile`: Project startup file to setup running environment including R version, repository, folder path etc. 
-  - We further use `inst/startup.R` and `R/zzz.R` to allow the startup file is executed while running. `devtools::load_all()` and RStudio build panel.
-
-## Compliance
-
-- ToDo
+ToDo

### 2021-09-15 _book/libs/bs4_book-1.0.0/bs4_book.js
ignroe _book folder

@@ -1,131 +1,0 @@
-$(function () {
-  var url = new URL(window.location.href);
-  var toMark = url.searchParams.get("q");
-  var mark = new Mark("main");
-  if (toMark) {
-    mark.mark(toMark, {
-      accuracy: {
-        value: "complementary",
-        limiters: [",", ".", ":", "/"],
-      }
-    });
-  }
-
-  // Activate popovers
-  $('[data-toggle="popover"]').popover({
-    container: 'body',
-    html: true,
-    trigger: 'focus',
-    placement: "top",
-    sanitize: false,
-  });
-  $('[data-toggle="tooltip"]').tooltip();
-})
-
-// Search ----------------------------------------------------------------------
-
-var fuse;
-
-$(function () {
-  // Initialise search index on focus
-  $("#search").focus(async function(e) {
-    if (fuse) {
-      return;
-    }
-
-    $(e.target).addClass("loading");
-
-    var response = await fetch('search.json');
-    var data = await response.json();
-
-    var options = {
-      keys: ["heading", "text", "code"],
-      ignoreLocation: true,
-      threshold: 0.1,
-      includeMatches: true,
-      includeScore: true,
-    };
-    fuse = new Fuse(data, options);
-
-    $(e.target).removeClass("loading");
-  });
-
-  // Use algolia autocomplete
-  var options = {
-    autoselect: true,
-    debug: true,
-    hint: false,
-    minLength: 2,
-  };
-
-  $("#search").autocomplete(options, [
-    {
-      name: "content",
-      source: searchFuse,
-      templates: {
-        suggestion: (s) => {
-          if (s.chapter == s.heading) {
-            return `${s.chapter}`;
-          } else {
-            return `${s.chapter} /<br> ${s.heading}`;
-          }
-        },
-      },
-    },
-  ]).on('autocomplete:selected', function(event, s) {
-    window.location.href = s.path + "?q=" + q + "#" + s.id;
-  });
-});
-
-var q;
-async function searchFuse(query, callback) {
-  await fuse;
-
-  var items;
-  if (!fuse) {
-    items = [];
-  } else {
-    q = query;
-    var results = fuse.search(query, { limit: 20 });
-    items = results
-      .filter((x) => x.score <= 0.75)
-      .map((x) => x.item);
-  }
-
-  callback(items);
-}
-
-// Copy to clipboard -----------------------------------------------------------
-
-function changeTooltipMessage(element, msg) {
[…]

### 2021-09-15 index.Rmd
update merck.com link
update index.Rmd
fix typo

@@
 Welcome to R for clinical study report and submission. 
-Clinical study reports (CSR) are the essential part of clinical trials development. 
+Clinical study reports (CSR) are essential parts of clinical trials development. 
 a CSR is an "integrated" full scientific report of an individual clinical trial.
 
-The [ICH E3: strcuture and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) 
+The [ICH E3: structure and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) 
 provide guidance to assist sponsors in the development of a CSR. 
-In this book, you will learn how to use R to prepare CSR and submit to regulatory agency.
+In this book, you will learn how to use R to prepare CSR and submit to regulatory agencies.
 
-This is the work-in-progress draft of the book.
+**This is the work-in-progress draft of the book.**
 
-## List of Authors and Contributors
+## List of Authors and Contributors {-}
 
@@
 - Author: contribute majority content of at least one chapter. 
-- Contributor: [contribute at least one commit to the source code](https://stash.merck.com/projects/BARDS/repos/r4csr/browse)
+- Contributor: [contribute at least one commit to the source code](https://github.com/elong0527/r4csr)
 

### 2021-09-15 intro.Rmd
fix typo

--- 2021-08-27 intro.Rmd
@@
 
-In clinical trial development, developers needs to organize source code to deliver Study Data Tabulation Model (SDTM) or Analysis Dataset Model (ADaM) datasets and table, listing and figures (TLFs).
-For example, a Phase 3 clinical trial typically require hundreds of TLFs. 
-A consistent and well-defined folder structure is crucial to manage a clinical trial analysis and reporting (A&R) project.
+In clinical trial development, developers need to organize source code to deliver 
+Study Data Tabulation Model (SDTM) or Analysis Dataset Model (ADaM) datasets 
+and tables, listings, and figures (TLFs).
+For example, a Phase 3 clinical trial typically requires hundreds of TLFs. 
+A consistent and well-defined folder structure is crucial 
+to managing a clinical trial analysis and reporting (A&R) project.
 
-We recommend to use R package folder structure to organize all the A&R related source code and documentation for a clinical trial A&R project. 
-By using the R package folder structure, you can follow the convention to organize your code as other contributors on CRAN.
-This consistent approach simplifies the communication for all developers within and across organizations.
+We recommend using R package folder structure to organize all the A&R related 
+source code and documentation for a clinical trial A&R project. 
+By using the R package folder structure, 
+you can follow the convention to organize your code as other contributors on CRAN.
+This consistent approach simplifies the communication for all developers 
+within and across organizations.
 
-- For a new R developer, it is an essential step to learn R package when you want to share your work with others. 
-You will learn one folder structure that is widely used in R community with outstanding tutorial and tools for free.
+- For a new R developer, 
+it is an essential step to develop R packages 
+when you want to share your work with others. 
+You will learn one folder structure that is widely used 
+in the R community with outstanding tutorials and tools for free.
 
-- For an experienced R developer, there is minimal learning curve.
+- For an experienced R developer, there is a minimal learning curve.
 
-- For an organization, it simplify process, tool, template, and training development, because a unified folder structure is used to develop and maintain standard tool and analysis project.
+- For an organization, it simplifies the process, tool, template, 
+and training development, because a unified folder structure is used 
+to develop and maintain standard tool and analysis projects.
 
-The workflow around an R package can also improve the tractability and reproducibility for an analysis project [@marwick2018packaging].
+The workflow around an R package can also improve 
+the tractability and reproducibility for an analysis project [@marwick2018packaging].
 
-In addition, R package folder structure are also recommended to develop Shiny app as discussed in [Chapter 20 of Master Shiny](https://mastering-shiny.org/scaling-packaging.html) book and [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/golem.html) book.
+In addition, R package folder structure is also recommended 
+to develop Shiny app as discussed 
+in [Chapter 20 of Master Shiny](https://mastering-shiny.org/scaling-packaging.html) book 
+and [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/golem.html) book.
 
@@
 
-This book is an intermediate level book by assuming reader has equipped some R programming and clinical development knowledge.
+This book is an intermediate-level book by 
+assuming readers have R programming and clinical development knowledge.
 The assumption of each part is as below:
 
-- Part 1, "Delivering TLFs in CSR" provides general information with examples to create table, listing and figures. 
-In this part, we assume readers are an individual contributor of a clinical project with some experience in R programming. 
-We expect readers are familiar with data manipulation in R. Some good reference includes [Hands-On Programming with R](https://rstudio-education.github.io/hopr/), 
-[R for Data Science](https://r4ds.had.co.nz/) and [Data Manipulation with R](https://www.amazon.com/Data-Manipulation-R-Use/dp/0387747303).
+- Part 1, "Delivering TLFs in CSR" provides general information with examples 
+to create tables, listings, and figures. 
+In this part, we assume readers are individual contributors to a clinical project
+with some experience in R programming. 
+We expect readers are familiar with data manipulation in R. 
+Some good reference includes [Hands-On Programming with R](https://rstudio-education.github.io/hopr/),
+[R for Data Science](https://r4ds.had.co.nz/) and 
+[Data Manipulation with R](https://www.amazon.com/Data-Manipulation-R-Use/dp/0387747303).
 
-- Part 2, "Clinical Trial Project" provides general information with examples to manage a clinical trial A&R 
-project. In this part, we assume a reader is a project lead with experience in R package development. 
-Some good reference includes [R packages](https://r-pkgs.org/) and [The tidyverse style guide](https://style.tidyverse.org/).
+- Part 2, "Clinical Trial Project" provides general information 
+with examples to manage a clinical trial A&R project. 
+In this part, we assume a reader is a project lead with experience in R package development. 
+Some good reference includes [R packages](https://r-pkgs.org/) and 
+[The tidyverse style guide](https://style.tidyverse.org/).
 
-- Part 3, "eCTD Submission package" provides general information in preparing 
+- Part 3, "eCTD Submission package"
+provides general information in preparing 
 submission package related to clinical study report (CSR) in 
@@
 
-We share the same philosophy described in [Section 1.1 of R Packages book](https://r-pkgs.org/intro.html?q=Marwick#intro-phil) and quote here.
+We share the same philosophy described in 
+[Section 1.1 of R Packages book](https://r-pkgs.org/intro.html?q=Marwick#intro-phil) 
+and quote here.
 
@@
 - "The goal is to spend your time thinking about what you want to do 
-rather than thinking about the minutiae of package structure."
+rather than thinking about the minutiae of the package structure."

### 2021-09-15 tlf-overview.Rmd
fix typo

@@
 
-In clinical trial development, it is a critical step to submit information on medicinal products to regulatory agencies.
+In clinical trial development, it is a critical step to 
+submit information on medical products to regulatory agencies. 
 [Electronic Common Technical Document (eCTD)](https://en.wikipedia.org/wiki/Electronic_common_technical_document) 
@@
 Clinical Data Interchange Standards Consortium (CDISC) provided a
-[pilot project
-following ICH E3 guidance](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse).
+[pilot project following ICH E3 guidance](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse).
 
@@
 including clinical and statistical description, presentations, and analyses. 
-A large number of tables and figures are incorporated into the main text and appendices 
-of a CSR.
-In CDISC pilot project, an [example CSR](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/53-clin-stud-rep/535-rep-effic-safety-stud/5351-stud-rep-contr/cdiscpilot01/cdiscpilot01.pdf)
+A large number of tables and figures are incorporated into the main text and appendices of a CSR. 
+In CDISC pilot project, an 
+[example CSR](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/53-clin-stud-rep/535-rep-effic-safety-stud/5351-stud-rep-contr/cdiscpilot01/cdiscpilot01.pdf)
 is also provided. If you are interested in more examples of clinical study reports, 
@@
 clinical study reports submitted by pharmaceutical companies 
-on [EMA clincal data website](https://clinicaldata.ema.europa.eu/web/cdp/home).
+on [EMA clinical data website](https://clinicaldata.ema.europa.eu/web/cdp/home).
 
-Building CSR is a team work with clinicians, medical writers, statisticians and statistical programmers.
+Building CSR is teamwork with clinicians, medical writers, statisticians, and statistical programmers. 
 Here, we focus on the work and deliverables completed by statisticians and statistical programmers. 
 In an organization, a group of statisticians and statistical programmers commonly work together to 
-define, develop, validate and deliver tables, listings and figures (TLFs) required for a CSR.
-Microsoft Word is widely used to prepare CSR in pharmaceutical industry. 
-Therefore, `rtf`, `doc`, `docx` are commonly used format to deliver TLFs.
+define, develop, validate and deliver tables, listings, and figures (TLFs) required for a CSR.
+Microsoft Word is widely used to prepare CSR in the pharmaceutical industry. 
+Therefore, `rtf`, `doc`, `docx` are commonly used formats to deliver TLFs.
 
-In this chapter, our focus is to illustrate how to create tables, listings and figures (TLFs) in RTF format 
+In this chapter, our focus is to illustrate how to create tables, listings, and figures (TLFs) in the RTF format 
 that is commonly used in a CSR. The examples are in compliance of 
@@
 
-> FDA's PDF specification is a general reference, each organization may define 
+> FDA's PDF specification is a general reference, each organization can define 
 > more specific TLF format requirement that can be different from the examples in this book.
 
-These TLFs are typically used to summarize efficacy and safety statistical analysis results of a drug (or treatment).
+These TLFs are typically used to summarize efficacy and safety 
+statistical analysis results of a drug (or treatment).
 
@@
 - Section 12: Safety evaluation 
-- Section 14: Tables, Figures and Graphs referred to but not included in the text
+- Section 14: ables, Figures, and Graphs referred to but not included in the text
 - Section 16: Appendices
@@
 
-We used public available CDISC pilot [study data located at CDISC Bitbucket repository](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).
+We used publicly available CDISC pilot 
+[study data located at CDISC Bitbucket repository](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).
 
@@
 
-The dataset structure is following [CDISC Analysis Data Model (ADaM)](https://www.cdisc.org/standards/foundational/adam).
+The dataset structure follows 
+[CDISC Analysis Data Model (ADaM)](https://www.cdisc.org/standards/foundational/adam).
 
@@
 
-In this part, we mainly use R packages below to illustrate 
+In this part, we mainly use the R packages below to illustrate 
 how to deliver TLFs in CSR.
@@
 visualize and analyze data in R. 
-Those R packages share [the tidy tools manifesto](https://cran.r-project.org/web/packages/tidyverse/vignettes/manifesto.html) 
-and easy to use for interactive data analysis.
+Those R packages share 
+[the tidy tools manifesto](https://cran.r-project.org/web/packages/tidyverse/vignettes/manifesto.html) 
+and are easy to use for interactive data analysis.
 
@@
 
-`r2rtf` is an R package to create production ready tables and figures in RTF format.
+`r2rtf` is an R package to create production-ready tables and figures in RTF format.
 The R package is designed to
@@
 - provide simple "verb" functions that correspond to each component of a table, 
-  to help you translate data frame to table in RTF file.
+  to help you translate data frame to a table in RTF file.
 - enable pipes (`%>%`). 
@@
 
-Other extended examples and features are covered in 
+Other extended examples and features are covered in the 
 [`r2rtf` package website](https://merck.github.io/r2rtf/articles/index.html).
@@
 Below is the meaning of relevant variables. 
-More information can be found in help page of the dataset (`?r2rtf_adae`)
+More information can be found in the help page of the dataset (`?r2rtf_adae`)
 
@@
 ```
+
 `dplyr` and `tidyr` packages within `tidyverse` are used 
@@
 
-All these verbs is designed to enables usage of pipes (`%>%`).
+All these verbs are designed to enable the usage of pipes (`%>%`).
 A full list of all functions can be found in [package reference](https://merck.github.io/r2rtf/reference/index.html)
@@
 
-In this example, the input of `col_rel_width` is a vector with same length for number of columns. 
+In this example, the input of `col_rel_width` is a vector 
+with the same length for the number of columns. 
 This argument defines the relative width of each column 
@@
 
-In previous example, we found issue of misaligned column header. 
-We can fix the issue by using `rtf_colheader` function.
+In the previous example, we found the issue of a misaligned column header. 
+We can fix the issue by using the `rtf_colheader` function.
 
-In `rtf_colheader`, `colheader` argument is used to provide content of column header. 
+In `rtf_colheader`, `colheader` argument is used to provide the content of the column header. 
 We use `"|"` to separate the columns.
@@
 
-Below is an example to make first column left aligned and center aligned for the rest.
+Below is an example to make the first column left-aligned and center-aligned for the rest.
 
@@
 
-If we want to remove the top border of `"Adverse Events"` in header, 
+If we want to remove the top border of `"Adverse Events"` in the header, 
 we can change default value `"single"` to `""` in `border_top` argument as showed below.

### 2021-09-15 tlf-overview.Rmd
update tlf-overview

@@
 - provide simple "verb" functions that correspond to each component of a table, 
-  to help you translate data frame to a table in RTF file.
+  to help you translate a data frame to a table in RTF file.
 - enable pipes (`%>%`). 
-- only focus on **table format**. 
+- only focus on the **table format**. 
   Data manipulation and analysis shall be handled by other R packages. (e.g. `tidyverse`)
@@
 we'll use the dataset `r2rtf_adae` saved in the `r2rtf` package. 
-This dataset contain adverse events (AE) information from a clinical trial.
+This dataset contains adverse events (AE) information from a clinical trial.
 
@@
 `r2rtf` aims to provide one function for each type of table layout.
-Commonly used verbs includes:
+Commonly used verbs include:
 
@@
 All these verbs are designed to enable the usage of pipes (`%>%`).
-A full list of all functions can be found in [package reference](https://merck.github.io/r2rtf/reference/index.html)
+A full list of all functions can be found in the 
+[r2rtf package reference](https://merck.github.io/r2rtf/reference/index.html)
 
@@
 provide more space to the first column, 
-this can be achieved by updating `col_rel_width` argument in `rtf_body`.
+this can be achieved by updating the `col_rel_width` argument 
+in the `rtf_body` function.
 
@@
 If we want to remove the top border of `"Adverse Events"` in the header, 
-we can change default value `"single"` to `""` in `border_top` argument as showed below.
+we can change the default value `"single"` to `""` in the `border_top` argument as showed below.
 
-`r2rtf` support 26 different border types. The details can be found in 
+`r2rtf` support 26 different border types. The details can be found on 
 the [r2rtf package website](https://merck.github.io/r2rtf/articles/rtf-row.html#border-type).
@@
 
-Those features will be introduced when we first use them in the rest of chapters.
+Those features will be introduced when we first use them in the rest of the chapters.
 
 There are other R packages that can create TLFs in `.rtf` or `.docx` format.
-Interested reader can refer "Microsoft/LibreOffice Formats" section in 
+Interested readers can refer "Microsoft/LibreOffice Formats" section in 
 [CRAN Task View: Reproducible Research](https://cran.r-project.org/web/views/ReproducibleResearch.html)

### 2021-09-15 tlf-disposition.Rmd
fix typo for disposition table

--- 2021-08-27 tlf-disposition.Rmd
@@
-# Table for Study Patients
+# Disposition {#disposition}
 
-Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to summarize accounting of all patients who entered the study, 
-in the section 10.1, disposition of patients of a CSR.
+Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), we need to summarize the accounting of all patients who entered the study, in Section 10.1, Disposition of Patients.
+The disposition of patients reports the numbers of patients who were randomized, and who entered and completed each phase of the study, as well as the reasons for all post-randomization discontinuations, grouped by treatment and by major reason (lost to follow-up, adverse event, poor compliance, etc.).
 
@@
 The first step is to read relevant datasets into R.
-For disposition table, all the required information is saved in the ADSL dataset.
-We can use `haven` package to read the dataset.
+For a disposition table, all the required information is saved in the ADSL dataset.
+We can use the `haven` package to read the dataset.
 
@@
 
-We illustrate how to prepare a report data for a simplified disposition 
-of patients table using variables below:
+We illustrate how to prepare a report data for a simplified disposition of patients table using variables below:
 
@@
 ```
+
 In the code below, we calculated patients in the analysis population.
@@
 ```
-In the code below, we calculate number and percentage of patients who discontinued the study.
 
-Here we use `formatC()` to customize the numeric value to be 1 decimal with fixed width of 5.
+In the code below, we calculate the number and percentage of patients who discontinued the study.
+
+Here we use the `formatC()` function to customize the numeric value to be 1 decimal with fixed width of 5.
 
@@
 ```
-In the code below, we calculate number and percentage of patients 
-who completed/discontinued the study in different reasons.
+
+In the code below, we calculate the number and percentage of patients who completed/discontinued the study for different reasons.
 
@@
 
-In the code below, we calculate number and percentage of patients 
-who complete the study. We split `n_reason` because we need to customize 
-the row order of the table.
+In the code below, we calculate the number and percentage of patients who complete the study.
+We split `n_reason` because we need to customize the row order of the table.
 
@@
 ```
-In the code below, we calculate number and percentage of patients 
-who discontinued the study in different reasons. 
+
+In the code below, we calculate the number and percentage of patients who discontinued the study for different reasons.
 Here we use `paste0("    ", row)` to add some leading space for individual reasons for study discontinuation.
@@
 
-Now we combined individual rows into the whole table for reporting purpose. 
-`tbl_disp` is used as input for `r2rtf` to create final report.
+Now we combined individual rows into the whole table for reporting purposes.
+`tbl_disp` is used as input for `r2rtf` to create the final report.
 
@@
 
-We start to define the format of the output. We highlighted items that 
-is not discussed in previous discussion.
+We start to define the format of the output.
+We highlighted items that is not discussed in the previous discussion.
 
-`rtf_title` define table title. We can provide a vector for the title argument. 
+The `rtf_title` defines table title.
+We can provide a vector for the title argument.
 Each value is a separate line.
@@
 ```
+
+The procedure to generate a disposition table can be summarized as follow:
+
+-   Step 1: Read data into R, i.e. `adsl`.
+-   Step 2: Count patients in the analysis population and name the dataset as `n_rand`.
+-   Step 3: Count the number and percentage of patients who discontinued the study, and name the dataset as `n_disc`.
+-   Step 4: Count the number and percentage of patients who discontinued the study in different reasons, and name the dataset as `n_reason`. Format it using `r2rtf`.
+-   Step 5: Count the number and percentage of patients who completed the study, and name the dataset as `n_complete`.
+-   Step 6: Bind `n_rand`, `n_disc`, `n_reason` and `n_complete` by row. Format it using `r2rtf`.

### 2021-09-15 tlf-population.Rmd
update tlf-population

@@
+---
+editor_options: 
+  markdown: 
+    wrap: 72
+---
+
 # Analysis Population {#population}
 
-Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to summarize which patients were included in each efficacy analysis 
-in Section 11.1, Data Sets Analysed.
+Following [ICH E3
+guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
+we need to summarize which patients were included in each efficacy
+analysis in Section 11.1, Data Sets Analysed.
 
@@
 
-The first step is to read relevant datasets into R. 
-For analysis population table, 
-all the required information is saved in the ADSL dataset.
-We can use `haven` package to read the dataset.
+The first step is to read relevant datasets into R. For the analysis
+population table, all the required information is saved in the ADSL
+dataset. We can use the `haven` package to read the dataset.
 
@@
 
-We illustrate how to prepare a report data for a simplified 
-analysis population table using variables below:
+We illustrate how to prepare a report data for a simplified analysis
+population table using variables below:
 
@@
 ```
+
 ## Helper Functions
 
-Before we write analysis code, let's discuss the possibility to
-reuse R code by writing some toy helper function.
+Before we write analysis code, let's discuss the possibility to reuse R
+code by writing some toy helper functions.
 
-As discussed in 
-[R for data science](https://r4ds.had.co.nz/functions.html#when-should-you-write-a-function),
-"You should consider writing a function 
-whenever you’ve copied and pasted a block of code more than twice".
+As discussed in [R for data
+science](https://r4ds.had.co.nz/functions.html#when-should-you-write-a-function),
+"You should consider writing a function whenever you've copied and
+pasted a block of code more than twice".
 
@@
 
-- format percentages using `formatC` 
-- calculate numbers and percentages by groups
+-   Format the percentages using the `formatC()` function.
+-   Calculate the numbers and percentages by groups.
 
-We create two ad-hoc functions and apply them to all rest of tables.
+We create two ad-hoc functions and apply them to all the rest of the
+tables.
 
-To format numbers and percentages, we created a function called `fmt_num`.
-It is a very simple function wrapping `formatC`.
+To format numbers and percentages, we created a function called
+`fmt_num()`. It is a very simple function wrapping `formatC()`.
 
@@
 
-The main reason to create `fmt_num` is to enhance readability of the analysis code.
+The main reason to create the `fmt_num()` function is to enhance the
+readability of the analysis code.
 
-For example, we can compare the two versions of code to format percentage 
-used in \@ref(disposition) and `fmt_num`.
+For example, we can compare the two versions of code to format the
+percentage used in \@ref(disposition) and `fmt_num`.
 
@@
 
-To calculate numbers and percentages of patients by groups, 
-we provide a simple (but not robust) wrapper function, `count_by` using `dplyr` and `tidyr`.
+To calculate the numbers and percentages of patients by groups, we
+provide a simple (but not robust) wrapper function, `count_by()` using
+the `dplyr` and `tidyr` package.
 
-The function can be enhanced in multiple ways, but here we only focus on simplicity and readability.
-More details about writing R functions can be found in [Stat454 course](https://stat545.com/functions-part1.html).
+The function can be enhanced in multiple ways, but here we only focus on
+simplicity and readability. More details about writing R functions can
+be found in the [Stat454
+course](https://stat545.com/functions-part1.html).
 
@@
 
-By using `count_by`, we can simplify the analysis code as below.
+By using the `count_by()` function, we can simplify the analysis code as
+below.
 
@@
 
-With the helper functions `count_by`, we can easily prepare report dataset as
+With the helper functions `count_by`, we can easily prepare a report
+dataset as
 
@@
 
-Now we combine individual rows into the whole table for reporting purpose. 
-`tbl_pop` is used as input for `r2rtf` to create final report.
+Now we combine individual rows into the whole table for reporting
+purposes. `tbl_pop` is used as input for `r2rtf` to create the final
+report.
 
@@
 
-In conclusion, the procedure to generate a population summary table as shown in the above example is listed as follows:
+The procedure to generate an analysis population table can be summarized
+as follow:
 
 -   Step 1: Read data into R, i.e., `adsl`.
-- Step 2: Rowly bind the counts/percentages of the ITT population, the efficacy population and the safety population. These these types of population are generated by a help function `count_by`. 
+-   Step 2: Bind the counts/percentages of the ITT population, the
+    efficacy population and the safety population by row using the
+    `count_by()` function.
 -   Step 3: Format the output in Step 2 by `r2rtf`.

### 2021-09-15 tlf-baseline.Rmd
update baseline table

@@
 
-Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to summarize critical demographic and baseline characteristics of the patients
-in Section 11.2, Demographic and Other Baseline Characteristics.
+Following [ICH E3
+guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
+we need to summarize critical demographic and baseline characteristics
+of the patients in Section 11.2, Demographic and Other Baseline
+Characteristics.
 
-In this chapter, we illustrate how to create a simplified 
-baseline characteristics table in a study.
+In this chapter, we illustrate how to create a simplified baseline
+characteristics table in a study.
 
@@
 
-There is many R packages that can efficiently summarize baseline information.
-The [`table1`](https://github.com/benjaminrich/table1) R package is one of them.
+There are many R packages that can efficiently summarize baseline
+information. The [`table1`](https://github.com/benjaminrich/table1) R
+package is one of them.
 
@@
 
-For simplicity, we only analyze `SEX`, `AGE` and `RACE` in this example
-using the `table1` R package. 
-More details of the `table` R package can be found in the package
+For simplicity, we only analyze `SEX`, `AGE` and, `RACE` in this example
+using the `table1` R package. More details of the `table` R package can
+be found in the package
 [vignettes](https://benjaminrich.github.io/table1/vignettes/table1-examples.html).
 
-The `table1` R package directly create an HTML report.
+The `table1` R package directly creates an HTML report.
 
@@
 
-The code below transfer the output into a dataframe 
-that only contain ASCII character 
-recommended by regulatory agencies. 
-`tbl_base` is used as input for `r2rtf` to create final report.
+The code below transfer the output into a data frame that only contain
+ASCII character recommended by regulatory agencies. `tbl_base` is used
+as input for `r2rtf` to create a final report.
 
@@
 
-We start to define the format of the output. We highlight items that 
-are not discussed in previous discussion.
+We start to define the format of the output. We highlight items that are
+not discussed in the previous discussion.
 
-`text_indent_first` and `test_indent_left` are used to control the indent
-space of text. They are helpful when you need to control the white space 
-of a long sentence. For example, "AMERICAN INDIAN OR ALASKA NATIVE" 
-in the table.
+The `text_indent_first` and `test_indent_left` arguments are used to
+control the indent space of the text. They are helpful when you need to
+control the white space of a long sentence. For example, "AMERICAN
+INDIAN OR ALASKA NATIVE" in the table.
 
@@
 
-In conclusion, the procedure to generate demographic and baseline characteristics table is summarized as follows.
+The procedure to generate a baseline characteristics table can be
+summarized as follow:
 
 -   Step 1: Read the data set.
-- Step 2: Use `table1::table1()` to get the baseline characteristics table.
-- Step 3: Transfer the output in Step 2 into a data frame that only contains ASCII character.
-- Step 4: Define the format of the rtf table by using R package `r2rtf`.
+-   Step 2: Use `table1::table1()` to get the baseline characteristics
+    table.
+-   Step 3: Transfer the output in Step 2 into a data frame that only
+    contains ASCII characters.
+-   Step 4: Define the format of the rtf table by using R package
+    `r2rtf`.

### 2021-09-16 tlf-ae-spec.Rmd
fix typo in ae tables

@@
 
-Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to summarize which patients were included in each efficacy analysis 
-in Section 12.2, Adverse Events (AEs).
+Following [ICH E3
+guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
+we need to summarize which patients were included in each efficacy
+analysis in Section 12.2, Adverse Events (AEs).
 
@@
 
-In this chapter, we illustrate how to summarize simplified specific AE information in a study.
+In this chapter, we illustrate how to summarize simplified specific AE
+information in a study.
 
@@
 
-The data used to summarize AE information is in `adsl` and `adae` datasets.
+The data used to summarize AE information is in `adsl` and `adae`
+datasets.
 
@@
 
-For illustration purpose, we only provide count in the simplified table. 
-The percentage of participants 
-for each AE criteria can be calculated as in Chapter \@ref(aesummary).
+For illustration purposes, we only provide count in the simplified
+table. The percentage of participants for each AE criteria can be
+calculated as in Chapter \@ref(aesummary).
 
-In this way, let's focus on the analysis script for advanced feature for table layout.
+In this way, let's focus on the analysis script for two advanced
+features for table layout.
 
-- group content: AE can be summarized in multiple nested layer. 
-(e.g. by system organ class (SOC, `AESOC`) and specific AE term (`AEDECOD`))
-- pagenization: there are many AE terms that can not be covered in one page. 
-Column headers and SOC information needs to be repeated in every page.
+-   group content: AE can be summarized in multiple nested layers. (e.g.
+    by system organ class (SOC, `AESOC`) and specific AE term
+    (`AEDECOD`))
+-   pagenization: there are many AE terms that can not be covered in one
+    page. Column headers and SOC information need to be repeated on
+    every page.
 
-In the code below, we count the number of subjects in each AE term by SOC and treatment group, and we create a new variable `order` and set it as `0` for these counted numbers. 
-The variable `order` will help with the data manipulation later.
+In the code below, we count the number of subjects in each AE term by
+SOC and treatment group, and we create a new variable `order` and set it
+as `0` for these counted numbers. The variable `order` will help with
+the data manipulation later.
 
@@
 ```
-In the code below, we count number of subjects in each AE term by SOC, AE term, and treatment group.
-Here we also create a new variable `order` and set it as `1` for these counted numbers.
+
+In the code below, we count the number of subjects in each AE term by
+SOC, AE term, and treatment group. Here we also create a new variable
+`order` and set it as `1` for these counted numbers.
 
@@
 ```
+
 We prepare reporting data for AE information.
@@
 ```
+
 We prepare reporting data for analysis population.
@@
 ```
-The final report data is saved in `tbl_ae_spec`.
-We also add a blank row between population and AE information in the reporting table.
+
+The final report data is saved in `tbl_ae_spec`. We also add a blank row
+between population and AE information in the reporting table.
 
@@
 ```
+
 We start to define the format of the output.
 
-To obtain the nested layout, we use `page_by` argument in `rtf_body` function.
-By defining `page_by="AESOC"`, 
-`r2rtf` recognize the variable as a group indicator.
+To obtain the nested layout, we use the `page_by` argument in the
+`rtf_body` function. By defining `page_by="AESOC"`, `r2rtf` recognize
+the variable as a group indicator.
 
-After setting `pageby_row = "first_row"`, the first row is displayed as group header. If a group of information is breaked into multiple page, 
-the group header row is repeated in each page by default.
+After setting `pageby_row = "first_row"`, the first row is displayed as
+group header. If a group of information is broken into multiple pages,
+the group header row is repeated on each page by default.
 
-We can also customize the text format by providing a matrix that have 
-same dimension of the input dataset (i.e. `tbl_ae_spec`). 
-In the code below, we illustrate how to display **bold** text for group headers
-to highlight the nested structure of the table layout.
+We can also customize the text format by providing a matrix that has the
+same dimension as the input dataset (i.e. `tbl_ae_spec`). In the code
+below, we illustrate how to display **bold** text for group headers to
+highlight the nested structure of the table layout.
 
@@
 
-More discussion on `page_by`, `group_by` and `subline_by` 
-features can be found in the 
-[`r2rtf` package website](https://merck.github.io/r2rtf/articles/example-sublineby-pageby-groupby.html.
+More discussion on `page_by`, `group_by`, and `subline_by` features can
+be found in the [`r2rtf` package
+website](https://merck.github.io/r2rtf/articles/example-sublineby-pageby-groupby.html).
 
@@
 
-In conclusion, the procedure to generate a specific AE summary table as shown in the above example is listed as follows:
+The procedure to generate a baseline characteristics table can be
+summarized as follow:
 
 -   Step 1: Read data into R, i.e., `adae` and `adsl`.
-- Step 2: Count the number of subjects by SOC and treatment group (rows with bold text) and store it in a data frame `t1`.
-- Step 3: Count the number of subjects in each AE term by SOC, AE term, and treatment group (rows without bold text) and store it in a data frame `t2`.
-- Step 4: Rowly bind `t1` and `t2` into `t_ae`.
+-   Step 2: Count the number of subjects by SOC and treatment group
+    (rows with bold text) and save into `t1`.
+-   Step 3: Count the number of subjects in each AE term by SOC, AE
+    term, and treatment group (rows without bold text) and save into
+    `t2`.
+-   Step 4: Bind `t1` and `t2` by row into `t_ae`.
 -   Step 5: Count the number of subjects in each arm as `t_pop`.
-- Step 6: Rowly bind `t_pop` with `t_ae` as `tbl_ae_spec`.
+-   Step 6: Bind `t_pop` and `t_ae` by row into `tbl_ae_spec`.
 -   Step 7: Format the output by `r2rtf`.

### 2021-09-16 tlf-ae-summary.Rmd
fix typo in ae tables

@@
 
-Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to summarize which patients were included in each efficacy analysis 
-in Section 12.2, Adverse Events (AEs).
+Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), we need to summarize which patients were included in each efficacy analysis in Section 12.2, Adverse Events (AEs).
 
@@
 -   With serious drug-related adverse events
-- Who died"
+-   Who "died"
 
@@
 ```
-We summarize the number and percentage of patients meet each AE criteria.
+
+We summarize the number and percentage of patients who meet each AE criteria.
 
@@
 ```
-We prepare reporting ready dataset for each AE criteria.
+
+We prepare reporting-ready dataset for each AE criteria.
 
@@
 
-We prepare reporting ready dataset for analysis population.
+We prepare reporting-ready dataset for the analysis population.
 
@@
 ```
-The final report data is saved in `tbl_ae_summary`
+
+The final report data is saved in `tbl_ae_summary`.
 
@@
 
-In conclusion, the procedure to generate a AE summary table as shown in the above example is listed as follows:
+The procedure to generate a AE summary table can be summarized as follow：
 

### 2021-09-16 index.Rmd
address Aiming's comments

@@
                   person("Keaven", "Anderson", role = "aut"),
-                  person("Yujie", "Zhao", role = "ctr"))
+                  person("Yujie", "Zhao", role = "ctr"), 
+                  person("Aiming", "Yang", role = "ctr"))
 ```

### 2021-09-16 index.Rmd
fix typo

@@
 library(kableExtra)
-library(bookdown)
 ```
@@
 
-Welcome to R for clinical study report and submission. 
+Welcome to R for clinical study reports and submission. 
 Clinical study reports (CSR) are essential parts of clinical trials development. 
@@
 
-- Author: contribute majority content of at least one chapter. 
+- Author: contribute the majority content of at least one chapter. 
 - Contributor: [contribute at least one commit to the source code](https://github.com/elong0527/r4csr)

### 2021-09-16 index.Rmd
General updates to improve flow and fixed typos.

@@
 Welcome to R for clinical study reports and submission. 
-Clinical study reports (CSR) are essential parts of clinical trials development. 
-a CSR is an "integrated" full scientific report of an individual clinical trial.
+Clinical study reports (CSR) are essential parts of clinical trial development. 
+A CSR is an "integrated" full scientific report of an individual clinical trial.
 
 The [ICH E3: structure and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) 
-provide guidance to assist sponsors in the development of a CSR. 
-In this book, you will learn how to use R to prepare CSR and submit to regulatory agencies.
+provides guidance to assist sponsors in the development of a CSR. 
+In this book, you will learn how to use R to prepare a CSR and how to submit it to regulatory agencies.
 
@@
                   person("Yujie", "Zhao", role = "ctr"), 
-                  person("Aiming", "Yang", role = "ctr"))
+                  person("Aiming", "Yang", role = "ctr",
+                  person("Steven", "Haesendonckx", role = "ctr")))
 ```

### 2021-09-16 intro.Rmd
address Aiming's comments
General updates to improve flow and fixed typos.

@@
 
-In clinical trial development, developers need to organize source code to deliver 
-Study Data Tabulation Model (SDTM) or Analysis Dataset Model (ADaM) datasets 
-and tables, listings, and figures (TLFs).
-For example, a Phase 3 clinical trial typically requires hundreds of TLFs. 
-A consistent and well-defined folder structure is crucial 
-to managing a clinical trial analysis and reporting (A&R) project.
+In clinical trial development, source code needs to be developed and maintained to generate and deliver 
+Study Data Tabulation Model (SDTM), Analysis Dataset Model (ADaM) datasets and tables, listings, and figures (TLFs).
+A typical example is a Phase 3 clinical trial where hundreds of TLFs are required for submission. 
+Considering the amount of programs required for such an effort, a consistent and well-defined folder structure is crucial in managing a clinical trial analysis and reporting (A&R) project.
 
-We recommend using R package folder structure to organize all the A&R related 
-source code and documentation for a clinical trial A&R project. 
-By using the R package folder structure, 
-you can follow the convention to organize your code as other contributors on CRAN.
-This consistent approach simplifies the communication for all developers 
-within and across organizations.
+During the course of this book we will provide a recommended folder structure that is aligned with CRAN submissions of R packages.
 
@@
 In addition, R package folder structure is also recommended 
-to develop Shiny app as discussed 
+to develop Shiny apps as discussed 
 in [Chapter 20 of Master Shiny](https://mastering-shiny.org/scaling-packaging.html) book 
@@
 We expect readers are familiar with data manipulation in R. 
-Some good reference includes [Hands-On Programming with R](https://rstudio-education.github.io/hopr/),
+Some good references include [Hands-On Programming with R](https://rstudio-education.github.io/hopr/),
 [R for Data Science](https://r4ds.had.co.nz/) and 
@@
 In this part, we assume a reader is a project lead of a clinical project 
-with experience in R package development as well.
+with experience in R package development and submission.
 

### 2021-09-17 intro.Rmd
Update intro.Rmd
Co-authored-by: yilong zhang <elong0527@gmail.com>

@@
 A typical example is a Phase 3 clinical trial where hundreds of TLFs are required for submission. 
-Considering the amount of programs required for such an effort, a consistent and well-defined folder structure is crucial in managing a clinical trial analysis and reporting (A&R) project.
+Considering the amount of programs required for such an effort, 
+a consistent and well-defined folder structure is crucial in 
+managing a clinical trial analysis and reporting (A&R) project.
 
-During the course of this book we will provide a recommended folder structure that is aligned with CRAN submissions of R packages.
+We recommend using 
+[R package folder structure](https://github.com/rstudio/cheatsheets/raw/master/package-development.pdf) 
+to organize all the A&R related source code and documentation for a clinical trial A&R project. 
+The R package folder structure is well defined and widely used in R community through repositories (e.g. CRAN).
+
+The consistent approach of using R package folder structure 
+simplifies the communication for all developers within and across organizations.
 

### 2021-09-17 tlf-overview.Rmd
Improve flow and fix typos

@@
 
-In clinical trial development, it is a critical step to 
-submit information on medical products to regulatory agencies. 
+In clinical trials, a critical step is
+submission to regulatory agencies. 
 [Electronic Common Technical Document (eCTD)](https://en.wikipedia.org/wiki/Electronic_common_technical_document) 
@@
 
-Within eCTD, clinical study reports (CSR) are located at module 5. 
+Within eCTD, clinical study reports (CSRs) are located at module 5. 
 [ICH E3 guidance](https://www.ich.org/page/efficacy-guidelines) provides 
@@
 
-A typical CSR contain full details of an individual clinical study 
-including clinical and statistical description, presentations, and analyses. 
-A large number of tables and figures are incorporated into the main text and appendices of a CSR. 
+A typical CSR contains full details on the methods and results of a individual clinical study.
+In support of the statistical anlaysis, a large number of tables, listings and figures are
+incorporated into the main text and appendices.
 In CDISC pilot project, an 
@@
 
-Building CSR is teamwork with clinicians, medical writers, statisticians, and statistical programmers. 
+Building CSRs is a teamwork between clinicians, medical writers, statisticians, statistical programmers
+and other relevant specialists such as experts on biomarkers. 
 Here, we focus on the work and deliverables completed by statisticians and statistical programmers. 
-In an organization, a group of statisticians and statistical programmers commonly work together to 
-define, develop, validate and deliver tables, listings, and figures (TLFs) required for a CSR.
+In an organization, they commonly work together to 
+define, develop, validate and deliver tables, listings, and figures (TLFs) required for a CSR to
+summarize the safety and/or safety of the pharmaceutical product.
 Microsoft Word is widely used to prepare CSR in the pharmaceutical industry. 
-Therefore, `rtf`, `doc`, `docx` are commonly used formats to deliver TLFs.
+Therefore, `rtf`, `doc`, `docx` are commonly used formats in their deliverables.
 
-In this chapter, our focus is to illustrate how to create tables, listings, and figures (TLFs) in the RTF format 
+In this chapter, our focus is to illustrate how to create tables, listings, and figures (TLFs) in RTF format
 that is commonly used in a CSR. The examples are in compliance of 
@@
 > more specific TLF format requirement that can be different from the examples in this book.
-
-These TLFs are typically used to summarize efficacy and safety 
-statistical analysis results of a drug (or treatment).
 
@@
 - Section 12: Safety evaluation 
-- Section 14: ables, Figures, and Graphs referred to but not included in the text
+- Section 14: Tables, Listings and Figures referrals but not included in the text
 - Section 16: Appendices
@@
 We provided a brief introduction of `r2rtf` and show how to transfer 
-data frames into Table, Listing, and Figure (TLFs).
+data frames into Table, Listing, and Figures (TLFs).
 

### 2021-09-17 intro.Rmd
Update intro.Rmd

@@
 The workflow around an R package can also improve 
-the tractability and reproducibility for an analysis project [@marwick2018packaging].
+the traceability and reproducibility for an analysis project [@marwick2018packaging].
+
+We will revisit folder structure when we discuss project management for a clinical trail project.
 

### 2021-09-18 project-folder.Rmd
clarify consistency

--- 2021-08-20 project-folder.Rmd
@@
 
-In this part, we would like to put you as an A&R leader for a clinical project.
+To deliver A&R results of a clincial tiral project, it is a team work that typically require
+collaborations from both statisticians and programmers. 
+In this part, let's consider how to organize a clinical trial project as an A&R leader.
 
@@
 
-A clearly defined project folder structure can have many benefits to 
+A clearly defined clinical project folder structure can have many benefits to 
 clinical development teams in an organization. 
@@
 
-- Consistency
-- Reproducibility
-- Automation
-- Compliance
+- Consistency: every team work on the same folder structure.
+- Reproducibility: analysis can be executed and reproduced by different team members months/years later.  
+- Automation: automatically check integration of a project.  
+- Compliance: reduce compliance issue
+
+We illustrate the R package project folder for A&R project using a demo project called 
+[`esubdemo`](https://github.com/elong0527/esubdemo.git). You can [clone](https://happygitwithr.com/rstudio-git-github.html#clone-the-new-github-repository-to-your-computer-via-rstudio) 
+the project using RStudio.
 
@@
 For R user, you already benefit from a well defined and consistent folder structure. 
-That is the R package folder structure. 
+That is the [R package folder structure](https://github.com/rstudio/cheatsheets/raw/master/package-development.pdf). 
 Every R package developer is required to follow the same convention to organize their R
-functions, before the function package can be disseminated through the Comprehensive R Archive Network
-(CRAN). As a user, you can easily install and user those packages after installing from CRAN.
+functions, before the R package can be disseminated through the Comprehensive R Archive Network
+(CRAN). As a user, you can easily install and use those R packages after installing from CRAN. 
+There are many good resources to guide developers on R package development.
+For example, [Hadley Wickmans' R packages book](https://r-pkgs.org/)
 
-We recommend to also use R package folder structure to organize analysis scripts 
-for clincial trial development. 
-Using the R package folder structure to streamline data analysis work has been proposed 
+We recommend to use R package folder structure to organize analysis scripts 
+for clinical trial development. 
+Using the R package folder structure to streamline data analysis work has also been proposed before 
 (@marwick2018packaging, @wuanalysis).
 
-In this book, we refer an R package as **project specific R package** 
-if the purpose of an R package is to organize analysis scripts for a clinical project.
+In this book, we refer an R package as **project specific R package**. 
+The purpose of project specific R package is to organize analysis scripts for a clinical project.
 We refer an R package as **standard R package** if the purpose of an R package is to 
@@
 - `DESCRIPTION`: Metadata for a package including authors, license, dependency etc.
+- `R/`: Project specific R functions.
 - `vignettes/`: Analysis scripts using Rmarkdown.
-- `R/`: Project specific R functions.
 - `man/`: Manual of project specific R functions.
 
-We demonstrate the the idea using the [`esubdemo`](https://github.com/elong0527/esubdemo) project.
+To illustrate the idea, you can open `esubdemo` project by click the `esubdemo.Rproj` project.
 
-A general discussion of the R package folder structure can be found in Chapter 4 of “R packages:
-organize, test, document, and share your code” book (@wickham2015r).
+For analysis code to create TLFs, we saved them in the `vignettes/` folder. 
+In `esubdemo`, we put analysis code we learned from previous chapters into the `vignettes` folder.
+
+> Here the `adam\` folder contain ADaM datasets. 
+> The `tlf` folder contain output TLFs in `rtf` and `pdf` format.
+> We put `adam\` and `tlf\` folder within `vignettes` folder only for illustration purpose. 
+> In a real A&R report, you may have different location to save your input and output.
+
+```
+vignettes
+├── adam
+├── tlf
+├── tlf-01-disposition.Rmd
+├── tlf-02-population.Rmd
+├── tlf-03-baseline.Rmd
+├── tlf-04-efficacy.Rmd
+├── tlf-05-ae-summary.Rmd
+└── tlf-06-ae-spec.Rmd
+```
+
+In previous chapters, we also create a few helper functions that can be used for different analysis. 
+(e.g. `fmt_num` and `count_by`). 
+We put those helper functions into the `R/` folder.
+
+> Here `zzz.R` is a template file that ensure startup file `inst/startup.R` is executed. 
+> We will revisit startup file while we discuss reproducibility.
+
+```
+R
+├── count_by.R
+├── fmt.R
+└── zzz.R
+```
+
+The `man` folder is used to save documentation automatically generated by `roxygen2`.
+A typical workflow is to add `roxygen2` documentations at the header of each function in the `R\` folder. 
+Then `devtools::documents()` is used to generate all the documentation files in the `man\` folder. 
+More details can be found in [Chapter 10 of the R packages book](https://r-pkgs.org/man.html).
 
@@
 
-- ToDo
+By using the R package folder structure, you can also benefit from many outstanding tools to 
+simplify and streamline your workflow.
 

### 2021-09-19 tlf-disposition.Rmd
- avoid 'we' in writing a text or book - Prefix R verbs (good programming practices) so the user sees which packages are used quickly without reading text + ensure correct package is used eg filter/lag - There are standard ADaM variables that capture discontinuation reason EOSSTT DCSREAS DCSREASP - participants in the study, not patients - represent treatments by actual names - calculation of count+percent is improved (show count and percent in same column, when count is 0 no percent should be shown)

--- 2021-08-27 tlf-disposition.Rmd
@@
-# Table for Study Patients
+# Disposition {#disposition}
 
-Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to summarize accounting of all patients who entered the study, 
-in the section 10.1, disposition of patients of a CSR.
+Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), a summary needs to be provided that accounts for all participants who entered the study within Section 10.1, Disposition of participants.
+This section summarizes the number of participants that were randomized, entered and completed each phase of the study.
+In addition, the reasons for all post-randomization discontinuations, grouped by treatment and by major reason of discontinuation are required (lost to follow-up, adverse event, ...).
 
@@
 
-In this chapter, we illustrate how to create a disposition table as below.
+In this chapter, the creation of a typical disposition table is shown.
 
@@
 
-The first step is to read relevant datasets into R. 
-For disposition table, all the required information is saved in the ADSL dataset.
-We can use `haven` package to read the dataset.
+The first step is to read in the relevant datasets into R.
+For a disposition table, all the required information is saved in a Subject-level Analysis Dataset (ADSL).
+The dataset is provided in SAS7bdat format, which is the current standard in current clinical trial reporting and analysis.The `haven` package is able to read in the dataset, while maintaining its attributes (eg.
+variable labels).
 
 ```{r}
-adsl <- read_sas("adam_data/adsl.sas7bdat")
+adsl <- haven::read_sas("adam_data/adsl.sas7bdat")
 ```
 
-We illustrate how to prepare a report data for a simplified disposition 
-of patients table using variables below:
+Following variables are used in the preparation of the disposition report:
 
@@
 -   TRT01P: Planed treatment
-- TRT01PN: Planned treatment encoding
+-   TRT01PN: Planned treatment numeric encoding
 -   DISCONFL: Discontinued from study flag
-- DCREASCD: Discontinued from study reason
+-   DCREASCD: Discontinued from Study reason coded
 
@@
 adsl %>%
-  select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>%
-  head(4)
+  dplyr::select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>%
+  utils::head(4)
 ```
-In the code below, we calculated patients in the analysis population.
+
+In the code below, the number of participants randomized is calculated.
 
@@
 n_rand <- adsl %>%
-  group_by(TRT01PN) %>%
-  summarize(n = n()) %>%
-  pivot_wider(
-    names_from = TRT01PN,
-    names_prefix = "n_",
+  dplyr::group_by(TRT01P) %>%
+  dplyr::summarize(n = n()) %>%
+  tidyr::pivot_wider(
+    names_from = TRT01P,
+    names_prefix = "",
     values_from = n
   ) %>%
-  mutate(row = "Patients in population")
+  dplyr::mutate(row = "participants Randomized")
+
 n_rand
 ```
-In the code below, we calculate number and percentage of patients who discontinued the study.
 
-Here we use `formatC()` to customize the numeric value to be 1 decimal with fixed width of 5.
+In the code below, the number of participants that discontinued the study is calculated.
+
+Here we use the `formatC()` function to customize the numeric value to be 1 decimal with fixed width of 5.
 
@@
 n_disc <- adsl %>%
-  group_by(TRT01PN) %>%
-  summarize(
-    n = sum(DISCONFL == "Y"),
-    pct = formatC(n / n() * 100,
+  dplyr::group_by(TRT01P) %>%
+  dplyr::summarize(
+    n = base::sum(DISCONFL == "Y"),
+    pct = base::formatC(n / n() * 100,
       digits = 1, format = "f", width = 5
@@
   ) %>%
-  pivot_wider(
-    names_from = TRT01PN,
-    values_from = c(n, pct)
+  dplyr:: mutate(npct = base::paste0(n, " (", pct, "%)")) %>%
+  dplyr::select(-c(n,pct)) %>%
+  tidyr::pivot_wider(
+    names_from = TRT01P,
+    values_from = npct,
+    values_fill = list(npct = "0")
   ) %>%
-  mutate(row = "Discontinued")
+  dplyr::mutate(row = "Discontinued")
 
@@
 ```
-In the code below, we calculate number and percentage of patients 
-who completed/discontinued the study in different reasons.
+
+To calculate the number and percentage of participants that completed/discontinued the study for each discontinuation reason, below code can be used.
 
@@
 n_reason <- adsl %>%
-  group_by(TRT01PN) %>%
-  mutate(n_total = n()) %>%
-  group_by(TRT01PN, DCREASCD) %>%
-  summarize(
-    n = n(),
-    pct = formatC(n / unique(n_total) * 100,
+  dplyr::group_by(TRT01P) %>%
+  dplyr::mutate(n_total = dplyr::n()) %>%
+  dplyr::group_by(TRT01P, DCREASCD) %>%
+  dplyr::summarize(
+    n = dplyr::n(),
+    pct = base::formatC(n / unique(n_total) * 100,
       digits = 1, format = "f", width = 5
@@
   ) %>%
-  pivot_wider(
+  dplyr:: mutate(npct = base::paste0(n, " (", pct, "%)")) %>%
+  tidyr::pivot_wider(
     id_cols = DCREASCD,
-    names_from = TRT01PN,
-    values_from = c(n, pct),
-    values_fill = list(n = 0, pct = "  0.0")
+    names_from = TRT01P,
+    values_from = npct,
+    values_fill = list(npct = "0")
   ) %>%
-  rename(row = DCREASCD)
+  dplyr::rename(row = DCREASCD)
 
@@
 
-In the code below, we calculate number and percentage of patients 
-who complete the study. We split `n_reason` because we need to customize 
-the row order of the table.
+From the step above, the number and percentage of participants that completed the study can be isolated so change the
+roworder in the final table.
 
 ```{r}
-n_complete <- n_reason %>% filter(row == "Completed")
+n_complete <- n_reason %>% 
+  dplyr::filter(row == "Completed")
 
@@
 ```
-In the code below, we calculate number and percentage of patients 
-who discontinued the study in different reasons. 
-Here we use `paste0("    ", row)` to add some leading space for individual reasons for study discontinuation.
+
+For display purposes leading spaces are added to study discontinuation. This will mimick identation in the final report.
 
@@
 n_reason <- n_reason %>%
-  filter(row != "Completed") %>%
-  mutate(row = paste0("    ", row))
+  dplyr::filter(row != "Completed") %>%
+  dplyr::mutate(row = paste0("    ", row))
 
@@
 
-Now we combined individual rows into the whole table for reporting purpose. 
-`tbl_disp` is used as input for `r2rtf` to create final report.
+In the final step, the individual rows are combined and used as input for `r2rtf.
 
 ```{r}
-tbl_disp <- bind_rows(n_rand, n_complete, n_disc, n_reason) %>%
-  select(row, ends_with(c("_0", "_54", "_81")))
+tbl_disp <- base::rbind(n_rand, n_complete, n_disc, n_reason)
 
@@
 
-We start to define the format of the output. We highlighted items that 
-is not discussed in previous discussion.
-
-`rtf_title` define table title. We can provide a vector for the title argument. 
-Each value is a separate line. 
-The format can also be controlled by providing a vector input in text format.
+In the below code, formatting of the final table is defined. Sections are highlighted that were not discussed before.
 
@@
   # Table title
-  rtf_title("Disposition of Patients") %>%
+  rtf_title("Disposition of participants") %>%
 
@@
 ```
+
+The procedure to generate a disposition table can be summarized as follow:
+
+-   Step 1: Read subject level data into R, i.e. `adsl`.
+-   Step 2: Count participants in the analysis population and name the dataset `n_rand`.
+-   Step 3: Count the number and percentage of participants that discontinued the study, and name the dataset `n_disc`.
+-   Step 4: Count the number and percentage of participants that discontinued the study for different reasons, and name the dataset `n_reason`.
+-   Step 5: Count the number and percentage of participants that completed the study, and name the dataset `n_complete`.
+-   Step 6: Bind `n_rand`, `n_disc`, `n_reason` and `n_complete` by row. 
+-   Step 7: Write the final table to RTF

### 2021-09-19 project-folder.Rmd
add project folder information

@@
 > Here the `adam\` folder contain ADaM datasets. 
-> The `tlf` folder contain output TLFs in `rtf` and `pdf` format.
+> The `tlf` folder contain output TLFs in RTF format.
 > We put `adam\` and `tlf\` folder within `vignettes` folder only for illustration purpose. 
@@
 vignettes
-├── adam
+├── adam_data
 ├── tlf
@@
 By using R package folder structure and proper tools (e.g. `renv`, `packrat`), 
-we can easily achieve reproducibility for R and R package versions.
+we illustrate how to achieve reproducibility for R and R package versions.
 
@@
 While RStudio open `esubdemo` project, RStudio will automatically execute R code in `.Rprofile`.
-So we are using `.Rprofile` to setup a reproducible environment. 
+So we can use `.Rprofile` to setup a reproducible environment. 
 More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html).
@@
 
-After the `esubdemo` project is open, the code in `.Rprofile` will automatically check 
-current R version is the same as we defined in `.Rprofile`
+In `esubdemo`, the `.Rprofile` contain two lines of R code.
 
+```{r}
+source("inst/startup.R")
+source("renv/activate.R")
+```
+
+The first line execute R code saved in `inst/startup.R`. 
+People can define a customized startup file to check running environment.
+
+For example, we can require R version to be `4.1.1` by using code below. 
 ```{r, eval = FALSE}
-R_version <- "4.0.1"	# set up project R version
+R_version <- "4.1.1"	# set up project R version
+
+# Check R Version
+if(paste(R.version$major, R.version$minor, sep = ".") != R_version){
+ warning("The current R version is not the same with the current project in ", R_version)
+}
 ```
@@
 ```
-Error: The current R version is not the same as the current project in R4.0.1
+Error: The current R version is not the same as the current project in R4.1.1
 ```
 
-If we are running under the same R version, 
-you shall be able to see similar messages to inform how we control R package version.
+We can also define the package repository to be a specific snapshot date.
+For example, we used Microsoft R Application Network (MRAN) to define the snapshot date to be `2021-08-06`. 
+The snapshot date freeze the R package repository. 
+In other words, all R packages installed in this RStudio project 
+are based on the frozen R version at the snapshot date. 
+Here it is `2021-08-06` by using MRAN server.
+
+```{r}
+# set up snapshot date
+snapshot  <- "2021-08-06"
+
+# set up repository based on snapshot
+repos     <- paste0("https://mran.microsoft.com/snapshot/", snapshot)
+
+# Define repo URL for project specific package installation
+options(repos = repos)
+```
+
+> RStudio Package Manager (RSPM) provided a solution to host both 
+> public avaiable and internally developed R packages. 
+> However, public RSPM server (https://packagemanager.rstudio.com/client/#/repos/2/overview) 
+> did not provide daily snapshot as MRAN had.
+
+The second line execute R code in `renv/activate.R`. 
+This is a default behavior by using the `renv` R package to manage snapshot 
+of the R packages. 
+The `renv` provide a robust and stable approach to manage R package version 
+for a project specific R packages.
+As a user, you can use `renv::init()`, `renv::snapshot()` and `renv::restore()` 
+to initialize, save and restore R packages used for the current analysis project.
+
+In the analysis project, the `renv` package will
+
+- create a `renv.lock` file to save the state of package versions.
+- create a `renv` folder to manage R packages for a project.
+
+An introduction of `renv` can be found in its [package website](https://rstudio.github.io/renv/articles/renv.html).
+
+> `renv.lock` file and `renv/` folder are only for project specific R package. 
+> A standard R package should not use `renv`.
+
+In summary, the R package version is controlled in two layers.
+
+- Define a snapshot date in `inst/startup.R`.
+- Using `renv` to lock R versions within a project.
+
+If the project is initiated properly  
+you shall be able to see similar messages to inform how we control R package versions.
 
@@
 
-The R package version is controlled in two layers. 
-Firstly, we define a snapshot date in `.Rprofile`. 
-The snapshot date allows to freeze the repository. 
-In other words, all R packages installed in this RStudio project 
-are based on the frozen R version at the snapshot date. 
-Here it is `2021-08-06` by using Microsoft R Application Network (MRAN) server.
-
-```{r, eval = FALSE}
-snapshot  <- "2021-08-06" 									                           # set up snapshot date
-repos     <- paste0("https://mran.microsoft.com/snapshot/", snapshot)  # set up repository based on snapshot
-```
-
-> RStudio Package Manager (RSPM) provided a solution to host both 
-> public avaiable and internally developed R packages. 
-> However, public RSPM server (https://packagemanager.rstudio.com/client/#/repos/2/overview) 
-> did not provide daily snapshot as MRAN had.
-
-Secondly, we use `renv` to lock R package versions and saved in `renv.lock` file. 
-The exact same R package versions can be restored with `renv::restore()`.
-An introduction of `renv` can be found in its [package website](https://rstudio.github.io/renv/articles/renv.html).
-
-```
-* Project '~/github-repo/esubdemo' loaded. [renv 0.14.0]
-```
-
-> `renv.lock` file and `renv/` folder are only for project specific R package. 
-> A standard R package should not use `renv`.
-
 Once R packages have been properly installed, 
-the system will use the R packages located in the search path defined based on the order of `.libPaths()`.
+the system will use the R packages located in the search 
+path defined based on the order of `.libPaths()`.
 The startup message also provided the R package search path.
@@
 
-For an organization, a cloud based R environment (e.g. RStudio Workbench) 
-can enhance the development and operation of a clinical development team by 
-centralize managing R and R package version. 
-More details can be found in [https://environments.rstudio.com/](https://environments.rstudio.com/).
+> A cloud based R environment (e.g. RStudio Workbench) 
+> can enhance the reproducibility within an organization 
+> by using the same operation system, R and R package versions for an A&R project. 
+> More details can be found in [https://environments.rstudio.com/](https://environments.rstudio.com/).
 
-In summary, to achieve reproducibility for a project specific R package, 
+> A container like [docker](https://github.com/rstudio/r-docker) can further enhance 
+> the reproducibility across an organization at operation system level, but beyond the scope of this book.
+
+In conclusion, to achieve reproducibility for a project specific R package, 
 a clinical project team can work under a controlled R environment 
@@
 
-By using the R package folder structure, you can also benefit from many outstanding tools to 
+By using the R package folder structure, you will benefit from many outstanding tools to 
 simplify and streamline your workflow.
 
-There are additional folders and files required in this demo for an analysis project
+Here is a short list of commonly used R packages
 
-> people may use different folder name as it is not a standard R package folder.
-> These folders and files needs to be added in `.Rbuildignore` to pass
-> R package check
-
-- `adam`: ADaM datasets in `.sas7bdat` format.
-  + one may also put it in `inst/exdata` following [R package convention](https://r-pkgs.org/data.html)
-  + in reality, we suggest to have real data saved outside of this project. (e.g. in a database)
-- `output`: TLFs output 
-- `renv.lock` and `renv`: R package management using `renv` package. ([Introduction](https://rstudio.github.io/renv/articles/renv.html)) 
-- `_pkgdown.yml`: [pkgdown](https://pkgdown.r-lib.org/articles/pkgdown.html) configuration file
-- `.Rprofile`: Project startup file to setup running environment including R version, repository, folder path etc. 
-  - We further use `inst/startup.R` and `R/zzz.R` to allow the startup file is executed while running. `devtools::load_all()` and RStudio build panel.
+- [`devtools`](https://devtools.r-lib.org/): make package development easier. 
+  + A good overview can be found in [Chapter 2 of the R packages](https://r-pkgs.org/whole-game.html) book  
+- [`usethis`](https://usethis.r-lib.org/): automates repetitive tasks that arise during project setup and development.
+- [`testthat`](https://testthat.r-lib.org/): streamline testing code.
+  + A discussion of using the `testthat` for an A&R project can be found in (@madhutest).  
+- [`pkgdown`](https://pkgdown.r-lib.org/): generate static html documentation for an R package
+  + It also allow you to run all analysis code in batch.
 

### 2021-09-19 tlf-ae-spec.Rmd
change patient to participants #6

@@
 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to summarize which patients were included in each efficacy
+we need to summarize which participants were included in each efficacy
 analysis in Section 12.2, Adverse Events (AEs).

### 2021-09-19 tlf-ae-summary.Rmd
change patient to participants #6

@@
 
-Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), we need to summarize which patients were included in each efficacy analysis in Section 12.2, Adverse Events (AEs).
+Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), 
+we need to summarize which participants were included in each efficacy analysis in Section 12.2, Adverse Events (AEs).
 
@@
 
-We summarize the number and percentage of patients who meet each AE criteria.
+We summarize the number and percentage of participants who meet each AE criteria.
 

### 2021-09-19 tlf-disposition.Rmd
change patient to participants #6

--- 2021-09-15 tlf-disposition.Rmd
@@
 
-Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), we need to summarize the accounting of all patients who entered the study, in Section 10.1, Disposition of Patients.
-The disposition of patients reports the numbers of patients who were randomized, and who entered and completed each phase of the study, as well as the reasons for all post-randomization discontinuations, grouped by treatment and by major reason (lost to follow-up, adverse event, poor compliance, etc.).
+Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), 
+we need to summarize the accounting of all participants who entered the study, 
+in Section 10.1, Disposition of Participants.
+The disposition of participant reports the numbers of participants who were randomized, 
+and who entered and completed each phase of the study, 
+as well as the reasons for all post-randomization discontinuations, 
+grouped by treatment and by major reason (lost to follow-up, adverse event, poor compliance, etc.).
 
@@
 
-We illustrate how to prepare a report data for a simplified disposition of patients table using variables below:
+We illustrate how to prepare a report data for a simplified disposition of participants table using variables below:
 
@@
 
-In the code below, we calculated patients in the analysis population.
+In the code below, we calculated participants in the analysis population.
 
@@
   ) %>%
-  mutate(row = "Patients in population")
+  mutate(row = "Participants in population")
 n_rand
@@
 
-In the code below, we calculate the number and percentage of patients who discontinued the study.
+In the code below, we calculate the number and percentage of participants who discontinued the study.
 
@@
 
-In the code below, we calculate the number and percentage of patients who completed/discontinued the study for different reasons.
+In the code below, we calculate the number and percentage of participants who completed/discontinued the study for different reasons.
 
@@
 
-In the code below, we calculate the number and percentage of patients who complete the study.
+In the code below, we calculate the number and percentage of participants who complete the study.
 We split `n_reason` because we need to customize the row order of the table.
@@
 
-In the code below, we calculate the number and percentage of patients who discontinued the study for different reasons.
+In the code below, we calculate the number and percentage of participants who discontinued the study for different reasons.
 Here we use `paste0("    ", row)` to add some leading space for individual reasons for study discontinuation.
@@
   # Table title
-  rtf_title("Disposition of Patients") %>%
+  rtf_title("Disposition of Participants") %>%
 
@@
 -   Step 1: Read data into R, i.e. `adsl`.
--   Step 2: Count patients in the analysis population and name the dataset as `n_rand`.
--   Step 3: Count the number and percentage of patients who discontinued the study, and name the dataset as `n_disc`.
--   Step 4: Count the number and percentage of patients who discontinued the study in different reasons, and name the dataset as `n_reason`. Format it using `r2rtf`.
--   Step 5: Count the number and percentage of patients who completed the study, and name the dataset as `n_complete`.
+-   Step 2: Count participants in the analysis population and name the dataset as `n_rand`.
+-   Step 3: Count the number and percentage of participants who discontinued the study, and name the dataset as `n_disc`.
+-   Step 4: Count the number and percentage of participants who discontinued the study in different reasons, and name the dataset as `n_reason`. Format it using `r2rtf`.
+-   Step 5: Count the number and percentage of participants who completed the study, and name the dataset as `n_complete`.
 -   Step 6: Bind `n_rand`, `n_disc`, `n_reason` and `n_complete` by row. Format it using `r2rtf`.

### 2021-09-19 tlf-efficacy.Rmd
change patient to participants #6

@@
 we need to summarize primary and secondary efficacy endpoints,  
-in Section 11.4, Efficacy Results and Tabulations of Individual Patient.
+in Section 11.4, Efficacy Results and Tabulations of Individual Participant.
 

### 2021-09-19 tlf-overview.Rmd
change patient to participants #6

@@
 
-A typical CSR contains full details on the methods and results of a individual clinical study.
+A typical CSR contains full details on the methods and results of an individual clinical study.
 In support of the statistical anlaysis, a large number of tables, listings and figures are
@@
 
-- Section 10: Study patients 
+- Section 10: Study participants 
 - Section 11: Efficacy evaluation
 - Section 12: Safety evaluation 
-- Section 14: Tables, Listings and Figures referrals but not included in the text
+- Section 14: Tables, listings and figures referrals but not included in the text
 - Section 16: Appendices

### 2021-09-19 tlf-population.Rmd
update baseline table
change patient to participants #6

@@
----
-editor_options: 
-  markdown: 
-    wrap: 72
----
-
 # Analysis Population {#population}
@@
 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to summarize which patients were included in each efficacy
+we need to summarize which participants were included in each efficacy
 analysis in Section 11.1, Data Sets Analysed.
@@
 
-To calculate the numbers and percentages of patients by groups, we
+To calculate the numbers and percentages of participants by groups, we
 provide a simple (but not robust) wrapper function, `count_by()` using

### 2021-09-19 tlf-disposition.Rmd
address conflicts

--- 2021-09-19 tlf-disposition.Rmd
@@
 
-Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), a summary needs to be provided that accounts for all participants who entered the study within Section 10.1, Disposition of participants.
-This section summarizes the number of participants that were randomized, entered and completed each phase of the study.
-In addition, the reasons for all post-randomization discontinuations, grouped by treatment and by major reason of discontinuation are required (lost to follow-up, adverse event, ...).
+Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), 
+we need to summarize the accounting of all participants who entered the study in Section 10.1, Disposition of Participants.
+
+The disposition of participant reports the numbers of participants who were randomized, 
+and who entered and completed each phase of the study. 
+In addition, the reasons for all post-randomization discontinuations, 
+grouped by treatment and by major reason (lost to follow-up, adverse event, poor compliance, etc.).
 
@@
 ```{r}
-adsl <- haven::read_sas("adam_data/adsl.sas7bdat")
+adsl <- read_sas("adam_data/adsl.sas7bdat")
 ```
 
-Following variables are used in the preparation of the disposition report:
+Following variables are used in the preparation of a simplified disposition of participants table:
 
@@
 adsl %>%
-  dplyr::select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>%
-  utils::head(4)
+  select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>%
+  head(4)
 ```
 
-In the code below, the number of participants randomized is calculated.
+In the code below, we calculated participants in the analysis population.
 
@@
 n_rand <- adsl %>%
-  dplyr::group_by(TRT01P) %>%
-  dplyr::summarize(n = n()) %>%
-  tidyr::pivot_wider(
+  group_by(TRT01P) %>%
+  summarize(n = n()) %>%
+  pivot_wider(
     names_from = TRT01P,
@@
   ) %>%
-  dplyr::mutate(row = "participants Randomized")
+  mutate(row = "Participants Randomized")
 
@@
 
-In the code below, the number of participants that discontinued the study is calculated.
+In the code below, we calculate the number and percentage of participants who discontinued the study.
 
@@
 n_disc <- adsl %>%
-  dplyr::group_by(TRT01P) %>%
-  dplyr::summarize(
-    n = base::sum(DISCONFL == "Y"),
-    pct = base::formatC(n / n() * 100,
+  group_by(TRT01P) %>%
+  summarize(
+    n = sum(DISCONFL == "Y"),
+    pct = formatC(n / n() * 100,
       digits = 1, format = "f", width = 5
@@
   ) %>%
-  dplyr:: mutate(npct = base::paste0(n, " (", pct, "%)")) %>%
-  dplyr::select(-c(n,pct)) %>%
-  tidyr::pivot_wider(
+  mutate(npct = paste0(n, " (", pct, "%)")) %>%
+  select(-c(n,pct)) %>%
+  pivot_wider(
     names_from = TRT01P,
@@
   ) %>%
-  dplyr::mutate(row = "Discontinued")
+  mutate(row = "Discontinued")
 
@@
 
-To calculate the number and percentage of participants that completed/discontinued the study for each discontinuation reason, below code can be used.
+In the code below, we calculate the number and percentage of participants who completed/discontinued the study for different reasons.
 
@@
 n_reason <- adsl %>%
-  dplyr::group_by(TRT01P) %>%
-  dplyr::mutate(n_total = dplyr::n()) %>%
-  dplyr::group_by(TRT01P, DCREASCD) %>%
-  dplyr::summarize(
-    n = dplyr::n(),
-    pct = base::formatC(n / unique(n_total) * 100,
+  group_by(TRT01P) %>%
+  mutate(n_total = n()) %>%
+  group_by(TRT01P, DCREASCD) %>%
+  summarize(
+    n = n(),
+    pct = formatC(n / unique(n_total) * 100,
       digits = 1, format = "f", width = 5
@@
   ) %>%
-  dplyr:: mutate(npct = base::paste0(n, " (", pct, "%)")) %>%
-  tidyr::pivot_wider(
+  mutate(npct = paste0(n, " (", pct, "%)")) %>%
+  pivot_wider(
     id_cols = DCREASCD,
@@
   ) %>%
-  dplyr::rename(row = DCREASCD)
+  rename(row = DCREASCD)
 
@@
 
-From the step above, the number and percentage of participants that completed the study can be isolated so change the
-roworder in the final table.
+In the code below, we calculate the number and percentage of participants who complete the study.
+We split `n_reason` because we need to customize the row order of the table.
 
@@
 n_complete <- n_reason %>% 
-  dplyr::filter(row == "Completed")
+  filter(row == "Completed")
 
@@
 
-For display purposes leading spaces are added to study discontinuation. This will mimick identation in the final report.
+In the code below, we calculate the number and percentage of participants who discontinued the study for different reasons.
+For display purposes, we use `paste0("    ", row)` to add leading spaces to mimick identation in the final report.
 
@@
 n_reason <- n_reason %>%
-  dplyr::filter(row != "Completed") %>%
-  dplyr::mutate(row = paste0("    ", row))
+  filter(row != "Completed") %>%
+  mutate(row = paste0("    ", row))
 
@@
 ```{r}
-tbl_disp <- base::rbind(n_rand, n_complete, n_disc, n_reason)
+tbl_disp <- rbind(n_rand, n_complete, n_disc, n_reason)
 
@@
   # Table title
-  rtf_title("Disposition of participants") %>%
+  rtf_title("Disposition of Participants") %>%
 

### 2021-09-19 tlf-disposition.Rmd
revert TRT01P -> TRT01PN #7

@@
 n_rand <- adsl %>%
-  group_by(TRT01P) %>%
+  group_by(TRT01PN) %>%
   summarize(n = n()) %>%
   pivot_wider(
-    names_from = TRT01P,
+    names_from = TRT01PN,
     names_prefix = "",
@@
 n_disc <- adsl %>%
-  group_by(TRT01P) %>%
+  group_by(TRT01PN) %>%
   summarize(
@@
   pivot_wider(
-    names_from = TRT01P,
+    names_from = TRT01PN,
     values_from = npct,
@@
 n_reason <- adsl %>%
-  group_by(TRT01P) %>%
+  group_by(TRT01PN) %>%
   mutate(n_total = n()) %>%
-  group_by(TRT01P, DCREASCD) %>%
+  group_by(TRT01PN, DCREASCD) %>%
   summarize(
@@
     id_cols = DCREASCD,
-    names_from = TRT01P,
+    names_from = TRT01PN,
     values_from = npct,

### 2021-09-19 tlf-disposition.Rmd
revert analysis code

--- 2021-09-19 tlf-disposition.Rmd
@@
 For a disposition table, all the required information is saved in a Subject-level Analysis Dataset (ADSL).
-The dataset is provided in SAS7bdat format, which is the current standard in current clinical trial reporting and analysis.The `haven` package is able to read in the dataset, while maintaining its attributes (eg.
-variable labels).
+The dataset is provided in `sas7bdat` format that is a SAS data format currently used in many clinical trial analysis and reporting.
+The `haven` package is able to read in the dataset, while maintaining its attributes (e.g. variable labels).
 
@@
 -   DISCONFL: Discontinued from study flag
--   DCREASCD: Discontinued from Study reason coded
+-   DCREASCD: Discontinued from study reason coded
 
@@
 n_rand <- adsl %>%
-  group_by(TRT01P) %>%
+  group_by(TRT01PN) %>%
   summarize(n = n()) %>%
   pivot_wider(
-    names_from = TRT01P,
-    names_prefix = "",
+    names_from = TRT01PN,
+    names_prefix = "n_",
     values_from = n
   ) %>%
-  mutate(row = "Participants Randomized")
+  mutate(row = "Participants in population")
 
@@
   ) %>%
-  mutate(npct = paste0(n, " (", pct, "%)")) %>%
-  select(-c(n,pct)) %>%
   pivot_wider(
-    names_from = TRT01P,
-    values_from = npct,
-    values_fill = list(npct = "0")
+    names_from = TRT01PN,
+    values_from = c(n, pct)
   ) %>%
@@
   ) %>%
-  mutate(npct = paste0(n, " (", pct, "%)")) %>%
   pivot_wider(
     id_cols = DCREASCD,
-    names_from = TRT01P,
-    values_from = npct,
-    values_fill = list(npct = "0")
+    names_from = TRT01PN,
+    values_from = c(n, pct),
+    values_fill = list(n = 0, pct = "  0.0")
   ) %>%
@@
 ```{r}
-tbl_disp <- rbind(n_rand, n_complete, n_disc, n_reason)
+tbl_disp <- bind_rows(n_rand, n_complete, n_disc, n_reason) %>%
+  select(row, ends_with(c("_0", "_54", "_81")))
 
@@
   # Table title
-  rtf_title("Disposition of Participants") %>%
+  rtf_title("Disposition of Patients") %>%
 

### 2021-09-19 tlf-disposition.Rmd
add back `rtf_title` discussion

@@
 
-In the below code, formatting of the final table is defined. Sections are highlighted that were not discussed before.
+In the below code, formatting of the final table is defined. 
+We highlighted items that is not discussed in the previous discussion.
+
+The `rtf_title` defines table title.
+We can provide a vector for the title argument.
+Each value is a separate line.
+The format can also be controlled by providing a vector input in text format.
 

### 2021-09-19 index.Rmd
update snapshot

@@
                   person("Yujie", "Zhao", role = "ctr"), 
-                  person("Aiming", "Yang", role = "ctr",
-                  person("Steven", "Haesendonckx", role = "ctr")))
+                  person("Aiming", "Yang", role = "ctr"),
+                  person("Steven", "Haesendonckx", role = "ctr"))
 ```

### 2021-09-21 tlf-disposition.Rmd
Update tlf-disposition.Rmd

@@
 In the code below, we calculate the number and percentage of participants who discontinued the study for different reasons.
-For display purposes, we use `paste0("    ", row)` to add leading spaces to mimick identation in the final report.
+For display purposes, `paste0("    ", row)` is used to add leading spaces to mimick identation in the final report.
 
@@
 In the below code, formatting of the final table is defined. 
-We highlighted items that is not discussed in the previous discussion.
+Items that were not discussed in the previous sections, are highlighted below.
 

### 2021-09-21 tlf-disposition.Rmd
Update tlf-disposition.Rmd

@@
 Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), 
-we need to summarize the accounting of all participants who entered the study in Section 10.1, Disposition of Participants.
+A summary needs to be provided to account for of all participants who entered the study in Section 10.1, Disposition of Participants.
 

### 2021-10-04 project-folder.Rmd
avoid evaluate startup file
update project folder

--- 2021-09-15 project-folder.Rmd
@@
 
-ToDo
+In a late-stage clinical trial, the number of A&R deliverables can easily be in the hundreds. 
+For an organization, it is also common to have multiple ongoing clinical trials in a clinical program.
+
+In this part, we would like to put you as an A&R leader for a clinical project.
+
+In Chapter \@ref(folder), we will discuss how to organize source code, documents, 
+deliverables in an A&R clinical project. 
+We recommend to use the R package folder structure.
+
+In Chapter \@ref(manage), we will discuss a process or system development lifecycle 
+to manage A&R of a clinical project. 
+We recommend to follow an agile management approach to define, develop, validate and deliver work.
+
+# Project Folder {#folder}
+
+A clearly defined project folder structure can have many benefits to 
+clinical development teams in an organization. 
+Specifically a well defined project folder can achieve:
+
+- Consistency
+- Reproducibility
+- Automation
+- Compliance
+
+For R users, you already benefit from a well defined and consistent folder structure. 
+That is the R package folder structure. 
+Every R package developer is required to follow the same convention to organize their R
+functions, before the package can be disseminated through the Comprehensive R Archive Network
+(CRAN). As a user, you can easily install and use those packages after installing from CRAN. 
+With a mature folder structure, you also get strong support from the community.
+
+We recommend to also use R package folder structure to organize analysis scripts 
+for clinical trial development. 
+Using the R package folder structure to streamline data analysis work has been proposed 
+(@marwick2018packaging, @wuanalysis).
+
+## Consistency
+
+For consistency, a well defined folder structure with potential templates
+ensures project teams organize A&R work consistently across multiple projects. 
+Consistent folder structure also reduce communication cost between study team members 
+and enhances the transparency of projects.
+
+In this book, we refer an R package as **project specific R package** 
+if the purpose of an R package is to organize analysis scripts for a clinical project.
+We refer an R package as **standard R package** if the purpose of an R package is to 
+share commonly used R functions to be hosted in an code repository such as CRAN.
+
+Below is a minimal sufficient folders and files for a project specific R package
+based on the R package folder structure.
+
+- `*.Rproj`: RStudio project file used to open RStudio project.
+- `DESCRIPTION`: Metadata for a package including authors, license, dependency etc.
+- `vignettes/`: Analysis scripts using Rmarkdown.
+- `R/`: Project specific R functions.
+- `man/`: Manual of project specific R functions.
+
+A general discussion of the R package folder structure can be found 
+in Chapter 4 of the R package book (@wickham2015r).
+
+We demonstrate the idea using the [`esubdemo`](https://github.com/elong0527/esubdemo) project.
+
+In `esubdemo` project, we saved all TLF generation scripts in previous chapters into the `vignettes/` folder.
+
+```
+vignettes/
+├── tlf-01-disposition.Rmd
+├── tlf-02-population.Rmd
+├── tlf-03-baseline.Rmd
+├── tlf-04-efficacy.Rmd
+├── tlf-05-ae-summary.Rmd
+└── tlf-06-ae-spec.Rmd
+```
+
+While creating those analysis scripts, we also defined a few helper functions. 
+Those functions are saved in the `R/` folder.
+
+```
+R/
+├── count_by.R
+└── fmt.R
+```
+
+With a formal project, it is also important to provide proper documentations for 
+those help functions. We used `Roxygen2` to document functions. 
+For example, the header below provide the definition of each variable in `fmt_est`. 
+More details can be found in [Chpater 10 of the R package book](https://r-pkgs.org/man.html)
+
+```{r, eval = FALSE}
+#' Format point estimator
+#'
+#' @param .mean mean of an estimator.
+#' @param .sd sd of an estimator.
+#' @param digits number of digits for `.mean` and `.sd`.
+#'
+#' @export
+fmt_est <- function(.mean,
+                    .sd,
+                    digits = c(1, 2)) {
+  .mean <- fmt_num(.mean, digits[1], width = digits[1] + 4)
+  .sd <- fmt_num(.sd, digits[2], width = digits[2] + 3)
+  paste0(.mean, " (", .sd, ")")
+}
+```
+
+Those `Roxygen2` documentation will be transferred to 
+standard R documentation `.Rd` files saved in the `man/` folder. 
+This step is automatically handled by `devtools::document()` function.
+
+```
+man
+├── count_by.Rd
+├── fmt_ci.Rd
+├── fmt_est.Rd
+├── fmt_num.Rd
+└── fmt_pval.Rd
+```
+
+## Reproducibility {#reproduce}
+
+Reproducibility of analysis is one of the most important aspects for regulatory deliverables. 
+By using R package folder structure and proper tools (e.g. `renv`, `packrat`), 
+we can easily achieve reproducibility for R and R package versions.
+
+> This is the same level of reproducibility in most SAS environment.
+> https://support.sas.com/techsup/pcn/altopsys.html
+
+In `esubdemo` project, a reproducible environment is created when you open the 
+`esubdemo.Rproj` from RStudio. 
+While RStudio open `esubdemo` project, RStudio will automatically execute R code in `.Rprofile`.
+So we are using `.Rprofile` to setup a reproducible environment. 
+More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html).
+
+> `.Rprofile` is only for project specific R package. 
+> A standard R package should not use `.Rprofile`.
+
+After we open the `esubdemo` project, the code in `.Rprofile` will automatically check 
+current R version is the same as we defined in `.Rprofile`
+
+```{r, eval = FALSE}
+R_version <- "4.1.1" # set up project R version
+```
+
+If the R version mismatch, an error message is displayed as below.
+
+```
+Error: The current R version is not the same as the current project in R4.1.1
+```
+
+If we are running under the same R version, 
+you shall be able to see similar messages to inform how we control R package version.
+
+```
+Current oroject R package repository:
+    https://mran.microsoft.com/snapshot/2021-08-06
+```
+
+The R package version is controlled in two layers. 
+Firstly, we define a snapshot date in `.Rprofile`. 
+The snapshot date allows us to freeze the source code repository. 
+In other words, all R packages installed in this RStudio project 
+are based on the frozen R version at the snapshot date. 
+Here it is `2021-08-06` by using Microsoft R Application Network (MRAN) server.
+
+```{r, eval = FALSE}
+snapshot <- "2021-08-06" # set up snapshot date
+repos <- paste0("https://mran.microsoft.com/snapshot/", snapshot) # set up repository based on snapshot
+```
+
+> RStudio Package Manager (RSPM) provided a solution to host both 
+> public avaiable and internally developed R packages. 
+> However, public RSPM server (https://packagemanager.rstudio.com/client/#/repos/2/overview) 
+> did not provide daily snapshot as MRAN had.
+
+Secondly, we use `renv` to lock R package versions and saved in `renv.lock` file. 
+The exact same R package versions can be restored with `renv::restore()`.
+An introduction of `renv` can be found in its 
+[package website](https://rstudio.github.io/renv/articles/renv.html).
+
+```
+* Project '~/github-repo/esubdemo' loaded. [renv 0.14.0]
+```
+
+> `renv.lock` file and `renv/` folder are only for project specific R package. 
+> A standard R package should not use `renv`.
+
+Once R packages have been properly installed, 
+the system will use the R packages located in the search path defined based on the order of `.libPaths()`.
+The startup message also provided the R package search path.
+
+```
+Below R package path are searching in order to find installed R pacakges in this R session:
+    "/home/zhanyilo/github-repo/esubdemo/renv/library/R-4.1/x86_64-pc-linux-gnu"
+    "/rtmp/RtmpT3ljoY/renv-system-library"   
+```
+
+For an organization, a cloud based R environment (e.g. RStudio Workbench) 
+can further enhance the reproducibility of clinical projects by 
+centralize managing operation system, R and R package version. 
+More details can be found in [https://environments.rstudio.com/](https://environments.rstudio.com/).
+
+In summary, to achieve reproducibility for a project specific R package, 
+a clinical project team can work under a controlled R environment 
+in the same R version and R package versions defined by a repository snapshot date.
+
+## Automation
+
+By using the R package folder structure for analysis and reporting clinical projects,
+there are many tools that can help project team simplify the workflow.
+
+We have learned a few functions in `devtools` to automatically generate content. 
+Here is a short list of tools that can enhance the workflow.
+
+- `devttols::load_all()`: load all functions in `R/` folder and running environment. 
+- `devtools::document()`: automatically create documentation using `Roxygen2`
+- `devttols::check()`: automatically perform compliance check as an R package. 
+- `devtools::build_site()`: automatically run analysis scripts in batch and create a `pkgdown` website. 
+- `usethis` package: automatically project and template setup.
+- `testthat` package: automatically execute validation code.
+
+You may further automatically execute routines by leveraging CI/CD workflow. 
+For example, the `esubdemo` project will rerun all required check and build pkgdown website
+by using [github action](https://usethis.r-lib.org/reference/github_actions.html).
+
+As the consistent folder is defined, it also become easier to create specific tools 
+that fit for the analysis and reporting purpose.
+
+Below are few potential tools that can be helpful:
+
+- Create project template using [RStudio Project Templates](https://rstudio.github.io/rstudio-extensions/rstudio_project_templates.html)
+- Add additional compliance check for analysis and reporting. 
+- Save log files for running in batch.
+
+## Compliance
+
+For a regulatory deliverable, it is important to maintain compliance. 
+With a consistent folder structure, we can define specific criteria for compliance. 
+Some compliance criteria can be implemented within the automatically checking steps.
+
+For an R package, there are already criteria to ensure R package integrity.  
+More details can be found in [Chapter 19 of the R package book](https://r-pkgs.org/r-cmd-check.html).

### 2021-10-04 project-folder.Rmd

--- 2021-09-19 project-folder.Rmd
@@
 We illustrate the R package project folder for A&R project using a demo project called 
-[`esubdemo`](https://github.com/elong0527/esubdemo.git). You can [clone](https://happygitwithr.com/rstudio-git-github.html#clone-the-new-github-repository-to-your-computer-via-rstudio) 
-the project using RStudio.
-
-## Consistency
-
-For consistency, a well defined folder structure with potential templates
-ensures project teams organize A&R work consistently across multiple projects. 
-Consistent folder structure also reduce communication cost between study team members 
-and enhances the transparency of projects.
+[`esubdemo`](https://github.com/elong0527/esubdemo.git). 
+You can [clone](https://happygitwithr.com/rstudio-git-github.html#clone-the-new-github-repository-to-your-computer-via-rstudio) the project using RStudio.
 
@@
 
-In this book, we refer an R package as **project specific R package**. 
-The purpose of project specific R package is to organize analysis scripts for a clinical project.
+## Consistency
+
+For consistency, a well defined folder structure with potential templates
+ensures project teams organize A&R work consistently across multiple projects. 
+Consistent folder structure also reduce communication cost between study team members 
+and enhances the transparency of projects.
+
+In this book, we refer an R package as **project specific R package** 
+if the purpose of an R package is to organize analysis scripts for a clinical project.
+
 We refer an R package as **standard R package** if the purpose of an R package is to 
@@
 
-Below is a minimal sufficient folders and files based on the R package folder structure.
+Below is a minimal sufficient folders and files for a project specific R package
+based on the R package folder structure.
 
@@
 
-To illustrate the idea, you can open `esubdemo` project by click the `esubdemo.Rproj` project.
+A general discussion of the R package folder structure can be found 
+in Chapter 4 of the R package book (@wickham2015r).
 
-For analysis code to create TLFs, we saved them in the `vignettes/` folder. 
-In `esubdemo`, we put analysis code we learned from previous chapters into the `vignettes` folder.
+We demonstrate the idea using the [`esubdemo`](https://github.com/elong0527/esubdemo) project.
+
+In `esubdemo` project, we saved all TLF generation scripts in previous chapters into the `vignettes/` folder.
 
@@
 
-In previous chapters, we also create a few helper functions that can be used for different analysis. 
-(e.g. `fmt_num` and `count_by`). 
-We put those helper functions into the `R/` folder.
-
-> Here `zzz.R` is a template file that ensure startup file `inst/startup.R` is executed. 
-> We will revisit startup file while we discuss reproducibility.
+While creating those analysis scripts, we also defined a few helper functions (e.g. `fmt_num` and `count_by`). 
+Those functions are saved in the `R/` folder.
 
 ```
-R
+R/
 ├── count_by.R
-├── fmt.R
-└── zzz.R
+└── fmt.R
+```
+
+With a formal project, it is also important to provide proper documentations for 
+those help functions. We used `Roxygen2` to document functions. 
+For example, the header below provide the definition of each variable in `fmt_est`. 
+More details can be found in [Chpater 10 of the R package book](https://r-pkgs.org/man.html)
+
+```{r, eval = FALSE}
+#' Format point estimator
+#'
+#' @param .mean mean of an estimator.
+#' @param .sd sd of an estimator.
+#' @param digits number of digits for `.mean` and `.sd`.
+#'
+#' @export
+fmt_est <- function(.mean,
+                    .sd,
+                    digits = c(1, 2)) {
+  .mean <- fmt_num(.mean, digits[1], width = digits[1] + 4)
+  .sd <- fmt_num(.sd, digits[2], width = digits[2] + 3)
+  paste0(.mean, " (", .sd, ")")
+}
+```
+
+Those `Roxygen2` documentation will be transferred to 
+standard R documentation `.Rd` files saved in the `man/` folder. 
+This step is automatically handled by `devtools::document()` function.
+
+```
+man
+├── count_by.Rd
+├── fmt_ci.Rd
+├── fmt_est.Rd
+├── fmt_num.Rd
+└── fmt_pval.Rd
 ```
@@
 
-Reproducibility of analysis can be challenge. 
+Reproducibility of analysis is one of the most important aspects for regulatory deliverables. 
 By using R package folder structure and proper tools (e.g. `renv`, `packrat`), 
@@
 
-In `esubdemo`, the `.Rprofile` contain two lines of R code.
+After we open the `esubdemo` project, the code in `.Rprofile` will automatically check 
+current R version is the same as we defined in `.Rprofile`
 
-```{r}
-source("inst/startup.R")
-source("renv/activate.R")
-```
-
-The first line execute R code saved in `inst/startup.R`. 
-People can define a customized startup file to check running environment.
-
-For example, we can require R version to be `4.1.1` by using code below. 
 ```{r, eval = FALSE}
 R_version <- "4.1.1" # set up project R version
-
-# Check R Version
-if(paste(R.version$major, R.version$minor, sep = ".") != R_version){
- warning("The current R version is not the same with the current project in ", R_version)
-}
 ```
@@
 
-We can also define the package repository to be a specific snapshot date.
-For example, we used Microsoft R Application Network (MRAN) to define the snapshot date to be `2021-08-06`. 
-The snapshot date freeze the R package repository. 
-In other words, all R packages installed in this RStudio project 
-are based on the frozen R version at the snapshot date. 
-Here it is `2021-08-06` by using MRAN server.
+The R package version is controlled in two layers. 
+Firstly, we define a snapshot date in `.Rprofile`. 
+The snapshot date allows us to freeze the source code repository.
 
@@
 
+We can also define the package repository to be a specific snapshot date.
+For example, we used Microsoft R Application Network (MRAN) to 
+define the snapshot date to be `2021-08-06`. 
+The snapshot date freeze the R package repository.
+
+In other words, all R packages installed in this RStudio project 
+are based on the frozen R version at the snapshot date. 
+Here it is `2021-08-06` by using MRAN server.
+
+Below information will be displayed after a new R session is opened.
+
+```
+Current oroject R package repository:
+    https://mran.microsoft.com/snapshot/2021-08-06
+```
+
 > RStudio Package Manager (RSPM) provided a solution to host both 
@@
 
-The second line execute R code in `renv/activate.R`. 
-This is a default behavior by using the `renv` R package to manage snapshot 
-of the R packages. 
+Secondly, we use `renv` to lock R package versions and saved in `renv.lock` file. 
 The `renv` provide a robust and stable approach to manage R package version 
 for a project specific R packages.
+An introduction of `renv` can be found in its 
+[package website](https://rstudio.github.io/renv/articles/renv.html).
+
+```{r}
+source("renv/activate.R")
+```
+
+The R code above in the `.Rprofile` initiate the `renv` running environment.  
 As a user, you can use `renv::init()`, `renv::snapshot()` and `renv::restore()` 
@@
 - create a `renv` folder to manage R packages for a project.
-
-An introduction of `renv` can be found in its [package website](https://rstudio.github.io/renv/articles/renv.html).
 
@@
 * Project '~/esubdemo' loaded. [renv 0.14.0]
-Current oroject R package repository:
-    https://mran.microsoft.com/snapshot/2021-08-06
 ```
@@
 Below R package path are searching in order to find installed R pacakges in this R session:
-    /home/esubdemo/renv/library/R-4.0/x86_64-pc-linux-gnu
-    /rtmp/Rtmp8d5mcL/renv-system-library
+    "/home/zhanyilo/github-repo/esubdemo/renv/library/R-4.1/x86_64-pc-linux-gnu"
+    "/rtmp/RtmpT3ljoY/renv-system-library"   
 ```
@@
 
-Here is a short list of commonly used R packages
+We have learned a few functions in `devtools` to automatically generate content. 
+Here is a short list of tools that can enhance the workflow.
 
@@
   + A good overview can be found in [Chapter 2 of the R packages](https://r-pkgs.org/whole-game.html) book  
+  + `devttols::load_all()`: load all functions in `R/` folder and running environment. 
+  + `devtools::document()`: automatically create documentation using `Roxygen2`
+  + `devttols::check()`: automatically perform compliance check as an R package. 
+  + `devtools::build_site()`: automatically run analysis scripts in batch and create a `pkgdown` website. 
 - [`usethis`](https://usethis.r-lib.org/): automates repetitive tasks that arise during project setup and development.
@@
 
+You may further automatically execute routines by leveraging CI/CD workflow. 
+For example, the `esubdemo` project will rerun all required check and build pkgdown website
+by using [github action](https://usethis.r-lib.org/reference/github_actions.html).
+
+As the consistent folder is defined, it also become easier to create specific tools 
+that fit for the analysis and reporting purpose.
+
+Below are few potential tools that can be helpful:
+
+- Create project template using [RStudio Project Templates](https://rstudio.github.io/rstudio-extensions/rstudio_project_templates.html)
+- Add additional compliance check for analysis and reporting. 
+- Save log files for running in batch.
+
 ## Compliance
 
-- ToDo
+For a regulatory deliverable, it is important to maintain compliance. 
+With a consistent folder structure, we can define specific criteria for compliance. 
+Some compliance criteria can be implemented within the automatically checking steps.
+
+For an R package, there are already criteria to ensure R package integrity.  
+More details can be found in [Chapter 19 of the R package book](https://r-pkgs.org/r-cmd-check.html).

### 2021-10-04 project-folder.Rmd
update running environment.
fix typo

@@
 clinical development teams in an organization. 
-Specifically a well defined project folder can achieve:
+Specifically, a well defined project folder can achieve:
 
-- Consistency: every team work on the same folder structure.
+- Consistency: everyone work on the same folder structure.
 - Reproducibility: analysis can be executed and reproduced by different team members months/years later.  
-- Automation: automatically check integration of a project.  
+- Automation: automatically check the integration of a project.  
 - Compliance: reduce compliance issue
 
-We illustrate the R package project folder for A&R project using a demo project called 
+We illustrate the R package project folder for the A&R project using a demo project called 
 [`esubdemo`](https://github.com/elong0527/esubdemo.git). 
@@
 
-For R user, you already benefit from a well defined and consistent folder structure. 
+For R users, you already benefit from a well-defined and consistent folder structure. 
 That is the [R package folder structure](https://github.com/rstudio/cheatsheets/raw/master/package-development.pdf). 
 Every R package developer is required to follow the same convention to organize their R
-functions, before the R package can be disseminated through the Comprehensive R Archive Network
+functions before the R package can be disseminated through the Comprehensive R Archive Network
 (CRAN). As a user, you can easily install and use those R packages after installing from CRAN. 
@@
 
-We recommend to use R package folder structure to organize analysis scripts 
+We recommend using the R package folder structure to organize analysis scripts 
 for clinical trial development. 
@@
 
-For consistency, a well defined folder structure with potential templates
+For consistency, a well-defined folder structure with potential templates
 ensures project teams organize A&R work consistently across multiple projects. 
-Consistent folder structure also reduce communication cost between study team members 
+Consistent folder structure also reduces communication cost between study team members 
 and enhances the transparency of projects.
 
-In this book, we refer an R package as **project specific R package** 
+In this book, we refer to an R package as **project specific R package** 
 if the purpose of an R package is to organize analysis scripts for a clinical project.
 
-We refer an R package as **standard R package** if the purpose of an R package is to 
-share commonly used R functions to be hosted in an code repository such as CRAN.
+We refer to an R package as a **standard R package** if the purpose of an R package is to 
+share commonly used R functions to be hosted in a code repository such as CRAN.
 
-Below is a minimal sufficient folders and files for a project specific R package
+Below is minimal sufficient folders and files for a project specific R package
 based on the R package folder structure.
@@
 
-> Here the `adam\` folder contain ADaM datasets. 
-> The `tlf` folder contain output TLFs in RTF format.
-> We put `adam\` and `tlf\` folder within `vignettes` folder only for illustration purpose. 
-> In a real A&R report, you may have different location to save your input and output.
+> Here the `adam\` folder contains ADaM datasets. 
+> The `tlf` folder contains output TLFs in RTF format.
+> We put `adam\` and `tlf\` folder within the `vignettes` folder only for illustration purposes. 
+> In a real A&R report, you may have a different location to save your input and output.
 
@@
 
-With a formal project, it is also important to provide proper documentations for 
-those help functions. We used `Roxygen2` to document functions. 
-For example, the header below provide the definition of each variable in `fmt_est`. 
-More details can be found in [Chpater 10 of the R package book](https://r-pkgs.org/man.html)
+With a formal project, it is also important to provide proper documentation for those help functions. 
+We used Roxygen2 to document functions. 
+For example, the header below defines each variable in fmt_est. 
+More details can be found in [Chapter 10 of the R package book](https://r-pkgs.org/man.html)
 
@@
 
-```{r}
+```{r, eval = FALSE}
 # set up snapshot date
@@
 
-```{r}
+```{r, eval = FALSE}
 source("renv/activate.R")

### 2021-10-04 project-folder.Rmd
fix typo

@@
 The `man` folder is used to save documentation automatically generated by `roxygen2`.
-A typical workflow is to add `roxygen2` documentations at the header of each function in the `R\` folder. 
+A typical workflow is to add `roxygen2` documentation at the header of each function in the `R\` folder. 
 Then `devtools::documents()` is used to generate all the documentation files in the `man\` folder. 
@@
 
-Reproducibility of analysis is one of the most important aspects for regulatory deliverables. 
-By using R package folder structure and proper tools (e.g. `renv`, `packrat`), 
+Reproducibility of analysis is one of the most important aspects of regulatory deliverables. 
+By using the R package folder structure and proper tools (e.g. `renv`, `packrat`), 
 we illustrate how to achieve reproducibility for R and R package versions.
 
-> This is the same level of reproducibility in most SAS environment.
+> This is the same level of reproducibility in most SAS environments.
 > https://support.sas.com/techsup/pcn/altopsys.html
@@
 define the snapshot date to be `2021-08-06`. 
-The snapshot date freeze the R package repository.
+The snapshot date freezes the R package repository.
 
@@
 > RStudio Package Manager (RSPM) provided a solution to host both 
-> public avaiable and internally developed R packages. 
-> However, public RSPM server (https://packagemanager.rstudio.com/client/#/repos/2/overview) 
-> did not provide daily snapshot as MRAN had.
+> publicly available and internally developed R packages. 
+> However, the public RSPM server (https://packagemanager.rstudio.com/client/#/repos/2/overview) 
+> did not provide a daily snapshot as MRAN had.
 
 Secondly, we use `renv` to lock R package versions and saved in `renv.lock` file. 
-The `renv` provide a robust and stable approach to manage R package version 
+The `renv` provides a robust and stable approach to manage R package versions 
 for a project specific R packages.
-An introduction of `renv` can be found in its 
+An introduction of `renv` can be found on its 
 [package website](https://rstudio.github.io/renv/articles/renv.html).
@@
 
-> A cloud based R environment (e.g. RStudio Workbench) 
+> A cloud-based R environment (e.g. RStudio Workbench) 
 > can enhance the reproducibility within an organization 
@@
 - [`pkgdown`](https://pkgdown.r-lib.org/): generate static html documentation for an R package
-  + It also allow you to run all analysis code in batch.
+  + It also allows you to run all analysis code in batch.
 
@@
 
-As the consistent folder is defined, it also become easier to create specific tools 
+As the consistent folder is defined, it also becomes easier to create specific tools 
 that fit for the analysis and reporting purpose.
@@
 - Create project template using [RStudio Project Templates](https://rstudio.github.io/rstudio-extensions/rstudio_project_templates.html)
-- Add additional compliance check for analysis and reporting. 
+- Add additional compliance checks for analysis and reporting. 
 - Save log files for running in batch.

### 2021-10-05 index.Rmd
create CNAME file

@@
 link-citations: yes
-
 ---
@@
 options(dplyr.summarise.inform = FALSE)
+file.copy("static/CNAME", "_book", overwrite = TRUE)
 ```

### 2021-10-05 tlf-baseline.Rmd

@@
 we need to summarize critical demographic and baseline characteristics
-of the patients in Section 11.2, Demographic and Other Baseline
+of participants in Section 11.2, Demographic and Other Baseline
 Characteristics.

### 2021-10-06 slides/workshop-slides.Rmd
add slides

---
title: "R for Clinical Study Reports and Submission"
subtitle: "R/Pharma Conference 2021"
author: "Yilong Zhang, Nan Xiao, Keaven Anderson"
date: "October, 2021"
output:
  mkdocs::mk_ioslides_presentation:
    mk_logo: "invent"
    mk_col_title: "white"
    mk_col_section: "teal"
    mk_class: "notclassified"
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, eval = FALSE, include = FALSE}
rmarkdown::render("slides/workshop-slides.Rmd")
```

# Welcome

## Outline

In this workshop, we have three parts:

- Delivering TLFs in CSR (Yilong Zhang)
  + Session 1 (45 min)
  + Break (10 min)
  + Session 2 (45 min)
  + Break (20 min) 
- Clinical trial project (Nan Xiao)
  + Session 3 (40 min)
  + Break (10 min)
- eCTD submission package (Nan Xiao)
  + Session 4 (30 min)
  + Q&A (10 min)

## Preparation

In this workshop, we assume you have 
some R programming experience and clinical development knowledge.

- Data manipulation: `tidyverse`, `dplyr`, `tidyr`, `ggplot2` etc. 
- ADaM data: `adsl`, `adae` etc.

```{r, echo = TRUE, eval = FALSE}
install.packages(c(
 "tidyverse",        # Data Manipulation 
 "r2rtf",            # TLF generation
 "pkglite"           # eCTD Submission 
))
```

## Resource

- Training Material: https://r4csr.org/
- Demo project: https://github.com/elong0527/esubdemo
- R Consortium FDA submission pilot: https://github.com/RConsortium/submissions-pilot1

## Philosophy

We share the same philosophy described in [Section 1.1 of R Packages book](https://r-pkgs.org/intro.html?q=Marwick#intro-phil) and quote here.

- "Anything that can be automated, should be automated." 
- "Do as little as possible by hand. Do as much as possible with functions." 
- "The goal is to spend your time thinking about what you want to do 
rather than thinking about the minutiae of package structure."

# Delivering TLFs in CSR

## ICH E3 guidance

The [ICH E3: strcuture and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) provide guidance to assist sponsors in the development of a CSR.

In a CSR, most of TLFs are located in

- Section 10: Study patients
- Section 11: Efficacy evaluation
- Section 12: Safety evaluation
- Section 14: Tables, Figures and Graphs referred to but not included in the text
- Section 16: Appendices

## Datasets

- Public available CDISC pilot [study data located at CDISC Bitbucket repository](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).

- The dataset structure is following [CDISC Analysis Data Model (ADaM)](https://www.cdisc.org/standards/foundational/adam).

- Source data: https://github.com/elong0527/r4csr/tree/master/adam_data

## Tools

- `tidyverse`: a collection of R packages to simplify the workflow to manipulate, 
visualize and analyze data in R.

- `r2rtf`: an R package to create production ready tables and figures in RTF format.
The R package is designed to 
  + `r2rtf` Introduction: https://r4csr.org/overview.html#r2rtf

## Disposition Table

https://r4csr.org/disposition.html

## Analysis Population

https://r4csr.org/population.html

# Break (10 min)

## Baseline Characteristics

https://r4csr.org/baseline-characteristics.html

## Efficacy Table

https://r4csr.org/efficacy.html

## AE Summary Table

https://r4csr.org/aesummary.html

## Specific AE Table

https://r4csr.org/specific-ae.html

# Clinical Trial Project

# Break (20 min)

## Overview

In a late-stage clinical trial, the number of A&R deliverables can easily be in the hundreds.

For an organization, it is also common to have multiple ongoing clinical trials in a clinical program.

In this part, let’s consider how to organize a clinical trial project as an A&R leader.

## Folder Structure

https://r4csr.org/folder.html

## Project Management

https://r4csr.org/manage.html

# Break (10 min)

# eCTD Submission

# Q&A

### 2021-10-06 project-management.Rmd
update project managment

@@
 
-SDLC for A&R deliverables can be defined in stages:
+SDLC for A&R deliverables can be defined in four stages:
 
@@
 - operation: an operation stage to deliver work to stakeholders.
-
-The define stage is an important stage in the SDLC lifecycle 
-as the requirements for all A&R deliverables are gathered and documented.
 
@@
 The goal is to embrace an iterative approach that continue improve target deliverables based on
-frequent customer feedback.
+frequent stakeholder feedback.
 
@@
 
+The planning stage is an important stage in the SDLC lifecycle 
+as the requirements for all A&R deliverables are gathered and documented.
+
 In the planning stage, a project leader shall identifies all the deliverables, 
@@
 
-- mock-up
+- mock-up tables
 - validation level (e.g. independent review or double programming)
 - etc.
-
-ToDo: add a simplified mock-up and A&R grid.
 
@@
 
-ToDo: add a simplified validation tracker
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("images/validation_tracker.png")
+```
 
@@
 To enable reproducibility, project leader shall also review the startup file 
-(i.e. `.Rprofile` discussed in Section \@ref(manage)) and define:
+(i.e. `.Rprofile` discussed in Section \@ref(reproduce)) and define:
 
@@
 
-> After project initiate, modifying `.Rprofile` will be a risk of reproducibility 
+> After project initiation, modifying `.Rprofile` will be a risk of reproducibility 
 > and shall be handled carefully if necessary.
@@
 in the `vignettes` folder.
+
 The use of R markdown allow developer to assemble narrative text, code and its
 comments in one place to simplify documentation. 
-It would be helpful to define a name convention for all TLFs deliverables. 
+It would be helpful to create a template and define a name convention for all TLFs deliverables. 
 For example, we can use `tlf_` prefix in the filename to indicate that the Rmarkdown is to deliver TLFs. 
-For example, `tlf_001_ds.Rmd` is an Rmarkdown file for disposition table. 
 Multiple TLFs with similar design can be included in one Rmarkdown file.
 
-ToDo: add a file path for `esubdemo`
+For example, in `esubdemo` project, we have 6 Rmarkdown files to create TLFs.
 
-To validate the results, 
-developer shall also save the last dataset before creating TLFs in RTF format to `.Rdata` files.
+If there is any project specific R function that needs to be developed, 
+the R functions can be placed in the `R\` folder as discussed in Section \@ref(consistency).
 
@@
 
+Validation is an important stage to ensure the deliverables are accurate and consistent. After the
+development stage is completed, the project team needs to validate the deliverables including Rmarkdown
+files for TLFs deliverables and project specific R functions. The level of validation is determined at the define
+stage.
+
+In an R package development, the validation or testing is completed under the `test/` folder. 
+The testthat R package can be used to streamline the validation process. 
+More details of `testthat` package for R package validation can be found in Chapter 12 of the R package book.
+
+It is recommended to have a name convention to indicate the type of validation. 
+For example, we can use `test-developer-test`, `test-independent-test`, `test-double-programming` 
+to classify the validation type.
+
+It is recommended to follow the same organization
+for files in testthat folder as `R/` folder and `vignettes/` folder. 
+Every single file in the `R/` folder and `vignettes/` folder should have a testing file saved in 
+the `tests/testthat/` folder to validate the content.
+
+For example, in `esubdemo` project, we can have a list of testing files below.
+
+```
+tests/testthat
+├── test-independent-test-tlf-01-disposition.Rmd
+├── test-independent-test-tlf-02-population.Rmd
+├── test-independent-test-tlf-03-baseline.Rmd
+├── test-independent-test-tlf-04-efficacy.Rmd
+├── test-independent-test-tlf-05-ae-summary.Rmd
+├── test-independent-test-tlf-06-ae-spec.Rmd
+└── test-independent-test-fmt.R
+```
+
+To validate the content of a Table, we can save the last datasets ready for table generation as a `.Rdata` file.
+A validator can reproduces the TLF and compares it with the original result saved in `.Rdata` file. 
+A test is passed when the results match. Customers can directly review the formatting of the TLFs by
+customers by comparing with the mock-up.
+
+To validate a Figure, we can use [snapshot testing](https://testthat.r-lib.org/articles/snapshotting.html) strategy
+
+After the validator completes the testing of project specific functions and Rmarkdown files, 
+the process to execute and report testing results is the same for a standard R package. 
+The `devtools::test()` function automatically executes all testing cases and summarizes the testing results in a report.
+
+After completing the validation, the validator updates the status in a validation tracker. 
+The project lead reviews the tracking sheet to make sure all required activities in the SDLC are completed and
+the tracking sheet has been filled correctly. The deliverables are ready for customer review after all the
+validation steps are completed. Any changes to the output requested by customers are documented.
+
 ## Operation
+
+After completion of development and required validation of all A&R deliverables, project lead runs compliance
+check for a project specific R package similar to other R packages. Using `devtools::check()` is a
+convenient way to run compliance check or R CMD checks. R CMD check is an automated check of the
+contents in the R package for frequently encountered issues, before submission to CRAN. Since project
+specific R package is not submitted to CRAN, some checks can be customized and skipped in
+`devtools::check()`. The project lead shall work with study team to ensure all reported errors, warnings and
+notes by `devtools::check()` are fixed.
+
+The project lead can also use `pkgdown` R package to build a complete website for a project specific R
+package. The `pkgdown` website is a convenient way to run all analysis in batch and 
+integrate outputs in a website.
+
+Project specific R functions, TLF generation programs, outputs and validation tracking information can all
+be explored from the website.
+
+For example, in `esubdemo` project, we created the `pkgdown` website in https://elong0527.github.io/esubdemo/.
+
+Many of the tasks in SDLC can be completed automatically. 
+An organization can leverage CI/CD workflow to automatically enable those tasks 
+such as running testing cases and create a `pkgdown` website.
+
+For example, in `esubdemo` project, we setup [github actions](https://github.com/elong0527/esubdemo/actions)
+for these work. It can be setup by using `usethis::use_github_action()` function.

### 2021-10-06 project-management.Rmd
fix typo

@@
 
-- [github project board](https://docs.github.com/en/issues/organizing-your-work-with-project-boards/managing-project-boards/about-project-boards)
+- [GitHub project board](https://docs.github.com/en/issues/organizing-your-work-with-project-boards/managing-project-boards/about-project-boards)
 - [Jira](https://www.atlassian.com/software/jira)

### 2021-10-07 project-management.Rmd
add section to project management

@@
 
-For A&R deliverables in a clinical project development, 
+## Setting up for success
+
+A clinical data analysis project is not unlike typical data analysis projects
+or software projects. Therefore, the conventional wisdom and tricks for
+managing a successful project are also applicable here.
+At the same time, clinical projects also have unique traits,
+such as high standards for planning, development, validation,
+and delivery under strict time constraints.
+
+Although many factors could decide if a project could execute efficiently,
+we believe a few aspects are critical for the long-term success,
+especially when managing clinical data analysis projects at scale.
+
+### Work as a team
+
+As a general principle, all the team members involved in a project
+should take basic training on project management and understand how to work as
+a software development team. @teamgeek provides some valuable tips on this topic.
+As always, setting a clear goal and following a system development lifecycle
+(SDLC) are essential.
+
+### Design clean code architecture
+
+Having a clean architecture design for your code improves the project's
+robustness and flexibility for future changes. For example,
+we should understand how to separate business logic from other layers;
+know what should be created as reusable components and what should be
+written as one-off analysis scripts; write low coupling, high cohesion code,
+and so on. @cleanarch offers some helpful insights on this topic.
+
+### Set capability boundaries
+
+Knowing what you can do is essential. Create a core capabilities list
+for your team.
+
+Sometimes, it is also critical to understand **what not to do**.
+For example, the hidden cost of integrating with external systems
+or involving other programming languages can be prohibitively high.
+Remember, a simple, robust solution is almost always preferable to
+a complex solution that requires high maintenance and constant attention.
+
+### Contribute to the community
+
+Every individual is limited in some way. The collective thinking
+from a community could benefit a project in the long term.
+When designing reusable components, make a plan to share with
+internal communities, or even better, with the open-source community.
+
+## The SDLC
+
+For A&R deliverables in clinical project development, 
 a clearly defined process or system development lifecycle (SDLC) 
-is crucial to ensure regulatory compliant.
+is crucial to ensure regulatory compliance.
 
@@
 Importantly, we should not consider SDLC as a linear process. 
-For example, if study team identify new requirement in a development or validation stage, 
-team shall return to planning stage to discuss and align the scope. 
+For example, if the study team identifies new requirement in a development or validation stage, 
+the team should return to the planning stage to discuss and align the scope. 
 An [agile project management](https://www.atlassian.com/agile/project-management) approach is suitable 
 and recommended to an A&R clinical development project.
-The goal is to embrace an iterative approach that continue improve target deliverables based on
+The goal is to embrace an iterative approach that continuously improves target deliverables based on
 frequent stakeholder feedback.
@@
 
-The planning stage is an important stage in the SDLC lifecycle 
+The planning stage is important in the SDLC lifecycle 
 as the requirements for all A&R deliverables are gathered and documented.
 
-In the planning stage, a project leader shall identifies all the deliverables, 
-e.g. a list of table, listing, and figures (TLFs). 
-For each TLFs, team shall prepare necessary specifications:
+In the planning stage, a project leader should identify all the deliverables, 
+e.g., a list of tables, listings, and figures (TLFs). 
+For each TLFs, the team should prepare the necessary specifications:
 
 - mock-up tables
-- validation level (e.g. independent review or double programming)
+- validation level (e.g., independent review or double programming)
 - etc.
 
-project leader shall also align work assignment with team members.
-The purpose is to answer the question on "who is doing what?".
+The project leader should also align work assignments with team members.
+The purpose is to answer the question of "who is doing what?"
 
@@
 
-Project lead shall also setup a project folder as discussed in Chapter \@ref(folder).
+The project lead should also set up a project folder as discussed in Chapter \@ref(folder).
 The project initiation can be simplified by creating an 
@@
 
-To enable reproducibility, project leader shall also review the startup file 
+To enable reproducibility, the project leader should also review the startup file 
 (i.e. `.Rprofile` discussed in Section \@ref(reproduce)) and define:
@@
 > After project initiation, modifying `.Rprofile` will be a risk of reproducibility 
-> and shall be handled carefully if necessary.
+> and should be handled carefully if necessary.
 
@@
 
-The use of R markdown allow developer to assemble narrative text, code and its
+The use of R markdown allows developers to assemble narrative text, code, and its
 comments in one place to simplify documentation. 
 It would be helpful to create a template and define a name convention for all TLFs deliverables. 
-For example, we can use `tlf_` prefix in the filename to indicate that the Rmarkdown is to deliver TLFs. 
-Multiple TLFs with similar design can be included in one Rmarkdown file.
+For example, we can use `tlf_` prefix in the filename to indicate that the R Markdown is to deliver TLFs. 
+Multiple TLFs with similar designs can be included in one R Markdown file.
 
-For example, in `esubdemo` project, we have 6 Rmarkdown files to create TLFs.
+For example, in `esubdemo` project, we have 6 R Markdown files to create TLFs.
 
-If there is any project specific R function that needs to be developed, 
-the R functions can be placed in the `R\` folder as discussed in Section \@ref(consistency).
+If there is any project-specific R functions that need to be developed, 
+the R functions can be placed in the `R/` folder as discussed in Section \@ref(consistency).
 
@@
 
-Validation is an important stage to ensure the deliverables are accurate and consistent. After the
-development stage is completed, the project team needs to validate the deliverables including Rmarkdown
-files for TLFs deliverables and project specific R functions. The level of validation is determined at the define
+Validation is an crucial stage to ensure the deliverables are accurate and consistent. After the
+development stage is completed, the project team needs to validate the deliverables, including R Markdown
+files for TLFs deliverables and project-specific R functions. The level of validation is determined at the define
 stage.
@@
 The testthat R package can be used to streamline the validation process. 
-More details of `testthat` package for R package validation can be found in Chapter 12 of the R package book.
+More details of the `testthat` package for R package validation can be found in Chapter 12 of the R package book.
 
@@
 To validate the content of a Table, we can save the last datasets ready for table generation as a `.Rdata` file.
-A validator can reproduces the TLF and compares it with the original result saved in `.Rdata` file. 
+A validator can reproduce the TLF and compares it with the original result saved in `.Rdata` file. 
 A test is passed when the results match. Customers can directly review the formatting of the TLFs by
-customers by comparing with the mock-up.
+customers by comparing them with the mock-up.
 
@@
 
-After the validator completes the testing of project specific functions and Rmarkdown files, 
+After the validator completes the testing of project-specific functions and R Markdown files, 
 the process to execute and report testing results is the same for a standard R package. 
@@
 After completion of development and required validation of all A&R deliverables, project lead runs compliance
-check for a project specific R package similar to other R packages. Using `devtools::check()` is a
-convenient way to run compliance check or R CMD checks. R CMD check is an automated check of the
-contents in the R package for frequently encountered issues, before submission to CRAN. Since project
-specific R package is not submitted to CRAN, some checks can be customized and skipped in
-`devtools::check()`. The project lead shall work with study team to ensure all reported errors, warnings and
+check for a project-specific R package similar to other R packages. Using `devtools::check()` is a
+convenient way to run compliance checks or `R CMD check`. `R CMD check` is an automated check of the
+contents in the R package for frequently encountered issues, before submission to CRAN. Since the project-specific
+R package is not submitted to CRAN, some checks can be customized and skipped in
+`devtools::check()`. The project lead should work with the study team to ensure all reported errors, warnings, and
 notes by `devtools::check()` are fixed.
 
-The project lead can also use `pkgdown` R package to build a complete website for a project specific R
-package. The `pkgdown` website is a convenient way to run all analysis in batch and 
+The project lead can also use the R package `pkgdown` to build a complete website for a project-specific R
+package. The `pkgdown` website is a convenient way to run all analyses in batch and 
 integrate outputs in a website.
 
-Project specific R functions, TLF generation programs, outputs and validation tracking information can all
+Project-specific R functions, TLF generation programs, outputs and validation tracking information can all
 be explored from the website.
 
-For example, in `esubdemo` project, we created the `pkgdown` website in https://elong0527.github.io/esubdemo/.
+For example, in the `esubdemo` project, we created the `pkgdown` website in https://elong0527.github.io/esubdemo/.
 
@@
 An organization can leverage CI/CD workflow to automatically enable those tasks 
-such as running testing cases and create a `pkgdown` website.
+such as running testing cases and creating a `pkgdown` website.
 

### 2021-10-07 slides/workshop-slides.Rmd
slides/workshop-slides.Rmd edited online with Bitbucket

@@
     mk_col_section: "teal"
-    mk_class: "notclassified"
+    mk_class: "public"
     widescreen: true

### 2021-10-07 slides/workshop-slides.Rmd

@@
   + Q&A (10 min)
+
+## Disclaimer
+
+- All opinions expressed are those of the presenter and not 
+  Merck Sharp & Dohme Corp., a subsidiary of Merck & Co., Inc., Kenilworth, NJ, USA.
+
+- Some slides need to be scrolled down to see the full content.
 
@@
 - `r2rtf`: an R package to create production ready tables and figures in RTF format.
-The R package is designed to 
-  + `r2rtf` Introduction: https://r4csr.org/overview.html#r2rtf
+
+## r2rtf Motivation
+
+In the pharmaceutical industry, RTF/Microsoft Word play a central role 
+in preparing clinical study reports
+
+Different organization can have different table standards
+
+- E.g., Table layout, Font size, Border type, Footnote, Data source
+
+`r2rtf` is designed to:
+
+- Generate highly customized tables
+- Limit package dependency
+- Target regulatory deliverables   
+- Support pipes (`%>%`)
+
+- `r2rtf` Introduction: https://r4csr.org/overview.html#r2rtf
 

### 2021-10-07 project-management.Rmd

@@
 should take basic training on project management and understand how to work as
-a software development team. @teamgeek provides some valuable tips on this topic.
+a development team. @teamgeek provides some valuable tips on this topic.
 As always, setting a clear goal and following a system development lifecycle

### 2021-10-09 project-folder.Rmd
update project managment
minor fixes for ch10 ch11

@@
 
-To deliver A&R results of a clincial tiral project, it is a team work that typically require
+To deliver A&R results of a clinical trial project, it is teamwork that typically requires
 collaborations from both statisticians and programmers. 
@@
 deliverables in an A&R clinical project. 
-We recommend to use the R package folder structure.
+We recommend using the R package folder structure.
 
 In Chapter \@ref(manage), we will discuss a process or system development lifecycle 
-to manage A&R of a clinical project. 
-We recommend to follow an agile management approach to define, develop, validate and deliver work.
+to manage the A&R of a clinical project. 
+We recommend following an agile management approach to define, develop, validate and deliver work.
 
@@
 clinical development teams in an organization. 
-Specifically, a well defined project folder can achieve:
+Specifically, a well-defined project structure can achieve:
 
-- Consistency: everyone work on the same folder structure.
+- Consistency: everyone works on the same folder structure.
 - Reproducibility: analysis can be executed and reproduced by different team members months/years later.  
@@
 There are many good resources to guide developers on R package development.
-For example, [Hadley Wickmans' R packages book](https://r-pkgs.org/)
+For example, Hadley Wickham's [R Packages book](https://r-pkgs.org/).
 
@@
 
-## Consistency
+## Consistency {#consistency}
 
@@
 ensures project teams organize A&R work consistently across multiple projects. 
-Consistent folder structure also reduces communication cost between study team members 
+Consistent folder structure also reduces communication costs between study team members 
 and enhances the transparency of projects.
 
-In this book, we refer to an R package as **project specific R package** 
+In this book, we refer to an R package as **project-specific R package** 
 if the purpose of an R package is to organize analysis scripts for a clinical project.
@@
 
-Below is minimal sufficient folders and files for a project specific R package
+Below is minimal sufficient folders and files for a project-specific R package
 based on the R package folder structure.
@@
 - `*.Rproj`: RStudio project file used to open RStudio project.
-- `DESCRIPTION`: Metadata for a package including authors, license, dependency etc.
-- `R/`: Project specific R functions.
+- `DESCRIPTION`: Metadata for a package including authors, license, dependencies, etc.
+- `R/`: Project-specific R functions.
 - `vignettes/`: Analysis scripts using Rmarkdown.
-- `man/`: Manual of project specific R functions.
+- `man/`: Manual of project-specific R functions.
 
@@
 
-In `esubdemo` project, we saved all TLF generation scripts in previous chapters into the `vignettes/` folder.
+In the `esubdemo` project, we saved all TLF generation scripts in previous chapters into the `vignettes/` folder.
 
-> Here the `adam\` folder contains ADaM datasets. 
+> Here, the `adam/` folder contains ADaM datasets. 
 > The `tlf` folder contains output TLFs in RTF format.
-> We put `adam\` and `tlf\` folder within the `vignettes` folder only for illustration purposes. 
-> In a real A&R report, you may have a different location to save your input and output.
+> We put `adam/` and `tlf/` folders within the `vignettes` folder only for illustration purposes. 
+> In an actual A&R report, you may have a different location to save your input and output.
 
@@
 The `man` folder is used to save documentation automatically generated by `roxygen2`.
-A typical workflow is to add `roxygen2` documentation at the header of each function in the `R\` folder. 
-Then `devtools::documents()` is used to generate all the documentation files in the `man\` folder. 
+A typical workflow is to add `roxygen2` documentation at the header of each function in the `R/` folder. 
+Then `devtools::documents()` is used to generate all the documentation files in the `man/` folder. 
 More details can be found in [Chapter 10 of the R packages book](https://r-pkgs.org/man.html).
@@
 
-In `esubdemo` project, a reproducible environment is created when you open the 
+In the `esubdemo` project, a reproducible environment is created when you open the 
 `esubdemo.Rproj` from RStudio. 
-While RStudio open `esubdemo` project, RStudio will automatically execute R code in `.Rprofile`.
-So we can use `.Rprofile` to setup a reproducible environment. 
+When we open the `esubdemo` project, RStudio will execute the R code in `.Rprofile` automatically.
+So we can use `.Rprofile` to set up a reproducible environment. 
 More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html).
 
-> `.Rprofile` is only for project specific R package. 
+> `.Rprofile` is only for project-specific R package. 
 > A standard R package should not use `.Rprofile`.
@@
 After we open the `esubdemo` project, the code in `.Rprofile` will automatically check 
-current R version is the same as we defined in `.Rprofile`
+the current R version is the same as we defined in `.Rprofile`
 
 ```{r, eval = FALSE}
-R_version <- "4.1.1" # set up project R version
+# set up project R version
+R_version <- "4.1.1"
 ```
@@
 
-# set up repository based on snapshot
+# set up repository based on the snapshot date
 repos <- paste0("https://mran.microsoft.com/snapshot/", snapshot)
 
-# Define repo URL for project specific package installation
+# define repo URL for project-specific package installation
 options(repos = repos)
@@
 are based on the frozen R version at the snapshot date. 
-Here it is `2021-08-06` by using MRAN server.
+Here it is `2021-08-06` by using the MRAN server.
 
@@
 ```
-Current oroject R package repository:
+Current project R package repository:
     https://mran.microsoft.com/snapshot/2021-08-06
@@
 
-Secondly, we use `renv` to lock R package versions and saved in `renv.lock` file. 
-The `renv` provides a robust and stable approach to manage R package versions 
-for a project specific R packages.
+Secondly, we use `renv` to lock R package versions and save them in the `renv.lock` file.
+`renv` provides a robust and stable approach to managing R package versions 
+for project-specific R packages.
 An introduction of `renv` can be found on its 
-[package website](https://rstudio.github.io/renv/articles/renv.html).
+[website](https://rstudio.github.io/renv/articles/renv.html).
 
@@
 
-The R code above in the `.Rprofile` initiate the `renv` running environment.  
+The R code above in the `.Rprofile` initiates the `renv` running environment.  
 As a user, you can use `renv::init()`, `renv::snapshot()` and `renv::restore()` 
@@
 
-> `renv.lock` file and `renv/` folder are only for project specific R package. 
+> `renv.lock` file and `renv/` folder are only for project-specific R package.
 > A standard R package should not use `renv`.
@@
 
-> A cloud-based R environment (e.g. RStudio Workbench) 
+> A cloud-based R environment (e.g., RStudio Workbench)
 > can enhance the reproducibility within an organization
-> by using the same operation system, R and R package versions for an A&R project. 
+> by using the same operating system, R version, and R package versions for an A&R project.
 > More details can be found in [https://environments.rstudio.com/](https://environments.rstudio.com/).
 
-> A container like [docker](https://github.com/rstudio/r-docker) can further enhance 
-> the reproducibility across an organization at operation system level, but beyond the scope of this book.
+> A container solution like [Docker](https://github.com/rstudio/r-docker) could further enhance 
+> the reproducibility across an organization at the operating system level, but beyond the scope of this book.
 
-In conclusion, to achieve reproducibility for a project specific R package, 
+In conclusion, to achieve reproducibility for a project-specific R package, 
 a clinical project team can work under a controlled R environment 
@@
 
-We have learned a few functions in `devtools` to automatically generate content. 
-Here is a short list of tools that can enhance the workflow.
+We have learned a few functions in `devtools` to generate content automatically.
+Here is a shortlist of tools that can enhance the workflow.
 
@@
   + `devttols::load_all()`: load all functions in `R/` folder and running environment. 
-  + `devtools::document()`: automatically create documentation using `Roxygen2`
+  + `devtools::document()`: automatically create documentation using `roxygen2`.
   + `devttols::check()`: automatically perform compliance check as an R package. 
@@
   + A discussion of using the `testthat` for an A&R project can be found in (@madhutest).  
-- [`pkgdown`](https://pkgdown.r-lib.org/): generate static html documentation for an R package
+- [`pkgdown`](https://pkgdown.r-lib.org/): generate static HTML documentation website for an R package
   + It also allows you to run all analysis code in batch.
@@
 You may further automatically execute routines by leveraging CI/CD workflow. 
-For example, the `esubdemo` project will rerun all required check and build pkgdown website
-by using [github action](https://usethis.r-lib.org/reference/github_actions.html).
+For example, the `esubdemo` project will rerun all required checks and build a pkgdown website
+by using [Github Actions](https://usethis.r-lib.org/reference/github_actions.html).
 

### 2021-10-10 project-management.Rmd

@@
 
-## Setting up for success
+## Setting up for Success
 
@@
 
-### Work as a team
+### Work as a Team
 
@@
 
-### Design clean code architecture
+### Design Clean Code Architecture
 
@@
 
-### Set capability boundaries
+### Set Capability Boundaries
 
@@
 
-### Contribute to the community
+### Contribute to the Community
 
@@
 
-For example, in the `esubdemo` project, we created the `pkgdown` website in https://elong0527.github.io/esubdemo/.
+For example, in the `esubdemo` project, we created the `pkgdown` website at <https://elong0527.github.io/esubdemo/>.
 
@@
 
-For example, in `esubdemo` project, we setup [github actions](https://github.com/elong0527/esubdemo/actions)
+For example, in `esubdemo` project, we setup [GitHub Actions](https://github.com/elong0527/esubdemo/actions)
 for these work. It can be setup by using `usethis::use_github_action()` function.

### 2021-10-12 submission-package.Rmd
finish submission chapters

# Submission Package {#submission-package}

In this chapter, we will first give a high-level overview of what
assets in the eCTD submission package we should focus on when
submitting R code.

Secondly, we will discuss how to prepare the proprietary R packages
(if any), and make them part of the submission package.

Lastly, we will provide reusable templates for updating ADRG and ARM,
so that the reviewers receive proper instructions to reproduce the
analysis results.

## Prerequisites

This chapter uses `pkglite` to convert R source packages
into text files and back.

```{r, eval=FALSE}
install.packages("pkglite")
```

The demo project (R package) we will prepare for submission is
available on GitHub at
[elong0527/esubdemo](https://github.com/elong0527/esubdemo).
You can download or clone it:

```bash
git clone https://github.com/elong0527/esubdemo.git
```

## The Whole Game

The files in two directories within the eCTD folder structure
are critical for documenting analysis using R: `datasets`
and `programs`. We assume the eCTD folder is `esub`.

```
/esub/analysis/adam/
├── datasets
│   ├── adrg.docx
│   ├── analysis-results-metadata.docx
│   ├── define-v1-updated-html.xsl
│   ├── define.pdf
│   └── define.xml
└── programs
    ├── tlf-01-disposition.txt
    ├── tlf-02-population.txt
    ├── tlf-03-baseline.txt
    ├── tlf-04-efficacy.txt
    ├── tlf-05-ae-summary.txt
    ├── tlf-06-ae-spec.txt
    └── r0pkgs.txt
```

The special considerations for each component are listed below.

### `datasets`

Folder path: `esub/analysis/adam/datasets/`.

- ADaM data in `.xpt` format: created by SAS
- `define.xml`: created by Pinnacle 21
- ADRG (ADaM Reviewer's Guide)
  - "Macro Programs" section: provide R and R package versions with a snapshot date
  - Appendix: provide step-by-step instructions for reviewers to
    reproduce the running environment and rerun analyses
- ARM (Analysis Results Metadata): provide the links between TLFs and
  analysis programs in tables

### `programs`

Folder path: `esub/analysis/adam/programs/`.

- `r0pkgs.txt`: contains all internally developed R packages
- Other `.txt` files: each contains R code for a specific analysis

### Notes

To verify if the submission package works, rerun all analyses following the
instructions defined in ADRG.

A few things to pay attention to for passing compliance checks:

- The filenames under `programs` should all use lower case
(with no underscores or other special characters).
- The `.txt` files should only contain ASCII characters.
This can be verified by `pkglite::verify_ascii()`
- The `.docx` files will be converted to PDF files for formal submission.

Now you have a general idea about the relevant components of the
submission package. We will focus on preparing the internal
R packages in the following sections.

## Practical Considerations for R Package Submissions

Before we start, there are a few aspects to figure out,
to accurately identify the R packages for submission.

### Source Location

There are a few common places to host R (source) packages:

1. CRAN
1. Public Git repository
1. Private Git repository (accessible externally)
1. Private Git repository (inaccessible externally)

For R packages hosted on CRAN or a public Git repository,
you probably do not need to submit them as part of the submission package,
as the reviewers can install them directly by following the instructions in ADRG.

For R packages hosted in private repositories, to avoid any complications
in infrastructure, authentication, and communication, it is often
recommended to submit them as part of the submission package.

### Dependency Locations

R package dependencies is another major factor to consider
before preparing your proprietary R package for submission.

For dependencies available from CRAN or public Git repositories,
you can declare them directly using the regular `Imports` and `Suggests`
syntax or the
[remotes dependency syntax](https://remotes.r-lib.org/articles/dependencies.html).

For dependencies hosted in private Git repositories, you should pack them
with the primary R package(s) you want to submit,
as `pkglite` supports packing multiple R packages into a single text file;
then restore and install them in the order they are packed.

### R Version

Always use a consistent version of R for developing the TLFs and for submission.
For example, you could enforce a rule to only use R x.y.z where z = 1,
such as R 4.0.1 or R 4.1.1.
This can be automatically checked using a
[startup script](https://github.com/elong0527/esubdemo/blob/master/inst/startup.R)
when the R project is opened.

### Package Repo Version

Always use the same snapshot package repo for developing the TLFs and for submission.
Again, this can be checked in the project startup script,
discussed in section \@ref(reproduce).

### System Environments

Introducing any extra external dependencies will likely increase the cost
of qualification, validation, and testing, and maintenance,
especially under Windows.
Therefore, it is recommended to keep the dependency chain simple,
especially when involving compiled code (e.g., C, C++, Fortran).

## Prepare R Packages for Submission

### Pack

Let's pack the `esubdemo` package into a text file.
Assume the source package path is `~/esubdemo/`,
you should be able to pack the package with a single pipe:

```{r, eval=FALSE}
library("pkglite")

"~/esubdemo/" %>%
  collate(file_ectd(), file_auto("inst")) %>%
  pack(output = "r0pkgs.txt")
```

```{r, fig.cap="Output of pkglite::pack()", out.width="100%", echo=FALSE}
knitr::include_graphics("images/pack.png")
```

Let's open the generated text file:

```{r, eval=FALSE}
file.edit("r0pkgs.txt")
```

```{r, fig.cap="Preview the generated text file", out.width="100%", echo=FALSE}
knitr::include_graphics("images/preview-txt.png")
```

What happened in the pipe? The function `pkglite::collate()`
evaluates a specified scope of folders and files defined by
a list of **file specifications**, and generate a **file collection** object.
This file collection contains the metadata required to properly convert
the files into text which is then used by `pkglite::pack()`.
With this flow, you can define the scope of the R source package
to be packed for submission in a flexible yet principled way.

To pack multiple R packages, simply feed multiple file collection as inputs:

```{r, eval=FALSE}
pack(
  "/path/to/pkg1/" %>% collate(file_ectd()),
  "/path/to/pkg2/" %>% collate(file_ectd()),
  output = "r0pkgs.txt"
)
```

The packages are always packed in the specified order and are always
unpacked and installed in the same order. Therefore, make sure to
pack the low-level dependencies first.

For more details on how to customize file specifications
and operate on file collections,
check out the vignette
[Generate File Specifications](https://merck.github.io/pkglite/articles/filespec.html)
and
[Curate File Collections](https://merck.github.io/pkglite/articles/filecollection.html).

### Verify

You should always verify if the file only contains ASCII characters:

```{r, eval=FALSE}
verify_ascii("r0pkgs.txt")
```

This should give `TRUE` if the file only contains ASCII characters,
or `FALSE` with the affected lines otherwise.

### Unpack

One can unpack and install the package from the text file, too.
For example:

```{r, eval = FALSE}
pkglite::unpack("r0pkgs.txt", output = "/tmp/", install = TRUE)
```

If successfully tested, this command will be used in the ADRG instructions
for restoring and installing the packed R package(s).

You can then proceed to save the file `r0pkgs.txt` to the folder
`esub/analysis/adam/programs` (or specify the output text file
path above directly).

## Prepare Analysis Programs for Submission

Besides the R packages, we need to convert the R Markdown (`.Rmd`) files
into `.txt` files and saved them in the `programs` folder.
You can do this with `knitr::purl()`:

```{r, eval=FALSE}
knitr::purl(
  "tlf-01-disposition.Rmd",
  output = "esub/analysis/adam/programs/tlf-01-disposition.txt"
)
```

Repeat this for all the R Markdown files.

Optionally, you can add a header to the individual `.txt` files to explain
the context and help the reviewers rerun the code. For example:

```r
# Note to Reviewer
# To rerun the code below, please refer to the ADRG appendix.
# After the required packages are installed,
# the path variable needs to be defined by using the example code below.
#
# path = list(adam = "/path/to/esub/analysis/adam/datasets") # Modify to use actual location
# path$outtable = path$outgraph = "." # Outputs saved to the current folder
```

## Update ADRG

After we converted the R packages and R Markdown files into the appropriate
formats and verified that they can be restored and executed correctly,
we should update the ADRG (ADaM Reviewer's Guide) to provide
guidelines on how to use them.

Specifically, we need to update two sections.
The first is the "Macro Programs" section, where
R and R package versions with a snapshot date are provided.

For example:

```markdown
7.x Macro Programs

Submitted R programs have [specific patterns] in filenames.
All internally developed R functions are saved in the r0pkgs.txt file.
The recommended steps to unpack these R functions for analysis output
programs are described in Appendix.

The table below contains the software version and instructions
for executing the R analysis output programs:
```

| Software or Package                            | Version |
|------------------------------------------------|---------|
| R                                              | 4.1.1   |
| dplyr                                          | 1.0.7   |
| r2rtf                                          | 0.3.0   |
| pkglite                                        | 0.2.0   |
| ...                                            | ...     |
| R package versions based on CRAN at 2021-08-06 |         |

The second section (Appendix) should include step-by-step instructions
to reproduce the running environment and rerun analyses. For example:

```markdown
Appendix: Instructions to Execute Analysis Program in R

1. Install R

Download and install R 4.1.1 for Windows from
https://cran.r-project.org/bin/windows/base/old/4.1.1/R-4.1.1-win.exe

2. Define Working Directory

Create a temporary working directory, for example, "C:\tempwork".
Copy all submitted R programs into the temporary folder.
All steps below should be executed in this working directory
represented as "." in the example R code below.

3. Specify R Package Repository

The R packages are based on CRAN at 2021-08-06. To install the exact
R package versions used in this project, run the code below to set
the snapshot repository.

options(repos = "https://mran.microsoft.com/snapshot/2021-08-06/")

4. Install Open-Source R Packages

In the same R session, install the required packages by running the code below.

install.packages(c("pkglite", "publicpkg1", "publicpkg2"))

5. Install Proprietary R Packages

All internal R packages are packed in the file r0pkgs.txt. In the same R session,
restore the package structures and install them by running the code below.
Adjust the output path as needed to use a writable local directory.

pkglite::unpack("r0pkgs.txt", output = ".", install = TRUE)

6. Update path to dataset and TLFs

INPUT path: to rerun the analysis programs, define the path variable

- Path for ADaM data: path$adam

OUTPUT path: to save the analysis results, define the path variable

- Path for output TLFs: path$output

All these paths need to be defined before executing the analysis program. For example:

path = list(adam = "/path/to/esub/analysis/adam/datasets/") # Modify to use actual location
path$outtable = path$outgraph = "." # Outputs saved to the current folder

7. Execute Analysis Program

To reproduce the analysis results, rerun the following programs:

- tlf-01-disposition.txt
- tlf-02-population.txt
- tlf-03-baseline.txt
- tlf-04-efficacy.txt
- tlf-05-ae-summary.txt
- tlf-06-ae-spec.txt
```

## Update ARM

The ARM (Analysis Results Metadata) should provide specific information
related to R in two sections:

- Section 2: indicate the **Programming Language**
- Section 3: document the details of the R programs listed in section 3.

For example, in ARM section 2, "Analysis Results Metadata Summary":

+--------------------------------------+-------------+----------------------+-------------------------+---------------------------------------+
| Table Reference                      | Table Title | Programming Language | Program Name (programs) | Input File Name / Analysis (datasets) |
+======================================+=============+======================+=========================+=======================================+
| [Ref. x.y.z: P001ZZZ9999: Table 1-1] | ...         | R                    | tlf-01-disposition.txt  | adsl.xpt                              |
+--------------------------------------+-------------+----------------------+-------------------------+---------------------------------------+
| ...                                  | ...         | ...                  | ...                     | ...                                   |
+--------------------------------------+-------------+----------------------+-------------------------+---------------------------------------+

In ARM section 3, "Analysis Results Metadata Details":

|                                                       |                                                               |
|-------------------------------------------------------|---------------------------------------------------------------|
| Table Reference: [Ref. x.y.z: P001ZZZ9999: Table 1-1] | ...                                                           |
| Analysis Result                                       | ...                                                           |
| Analysis Parameters (s)                               | ...                                                           |
| Analysis Reason                                       | ...                                                           |
| Analysis Purpose                                      | ...                                                           |
| ...                                                   | ...                                                           |
| Programming Statements                                | (R version 4.1.1), [P001ZZZ9999: programs-tlf-01-disposition] |

### 2021-10-12 index.Rmd

--- 2021-09-19 index.Rmd
@@
 library(kableExtra)
+library(pkglite)
 ```
@@
 
-Welcome to R for clinical study reports and submission. 
-Clinical study reports (CSR) are essential parts of clinical trial development. 
-A CSR is an "integrated" full scientific report of an individual clinical trial.
+Welcome to R for clinical study report and submission. 
+Clinical study reports (CSR) are the essential part of clinical trials development. 
+a CSR is an "integrated" full scientific report of an individual clinical trial.
 
-The [ICH E3: structure and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) 
-provides guidance to assist sponsors in the development of a CSR. 
-In this book, you will learn how to use R to prepare a CSR and how to submit it to regulatory agencies.
+The [ICH E3: strcuture and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) 
+provide guidance to assist sponsors in the development of a CSR. 
+In this book, you will learn how to use R to prepare CSR and submit to regulatory agency.
 
-**This is the work-in-progress draft of the book.**
+This is the work-in-progress draft of the book.
 
-## List of Authors and Contributors {-}
+## List of Authors and Contributors
 
@@
                   person("Keaven", "Anderson", role = "aut"),
-                  person("Yujie", "Zhao", role = "ctr"), 
-                  person("Aiming", "Yang", role = "ctr"),
-                  person("Steven", "Haesendonckx", role = "ctr"))
+                  person("Yujie", "Zhao", role = "ctr"))
 ```
@@
 
-- Author: contribute the majority content of at least one chapter. 
-- Contributor: [contribute at least one commit to the source code](https://github.com/elong0527/r4csr)
+- Author: contribute majority content of at least one chapter. 
+- Contributor: [contribute at least one commit to the source code](https://stash.merck.com/projects/BARDS/repos/r4csr/browse)
 

### 2021-10-12 submission-environment.Rmd

# Running Environment {#running-environment}

In the previous chapter, we have generated instructions to manually
create the running environments for reproducing the A&R deliverables.

In this chapter, we will focus on automating the creation of
the R environments with R code, to accelerate the dry run testing process,
simplify the ADRG instructions, and make it easy to recreate
different environment settings with reproducible analysis results.

## Prerequisites

`cleanslate` is an R package that offers a solution to
create portable R environments.

::: {.rmdnote}
As of Q4 2021, the cleanslate package used in this chapter
is still under active development and validation.
This chapter gives a preview of the planned APIs.
They may change in the future.
:::

Install `cleanslate` from CRAN (once available):

```{r, eval=FALSE}
install.packages("cleanslate")
```

Or from GitHub (once available):

```{r, eval=FALSE}
remotes::install_github("Merck/cleanslate")
```

## Practical Considerations

The `cleanslate` package supports:

- Creating a project folder with project-specific context
  (`.Rproj`, `.Rprofile`, `.Renviron`)
- Installing a specific version of R into the project folder
- Installing a specific version of Rtools into the project folder

An essential feature of `cleanslate` is that it does **not** require
administrator privileges to run R and Rtools installers.
This makes it easier to deploy under enterprise settings
and avoids security and portability concerns.

As many of the A&R deliverables are currently created, validated, and delivered
under Windows, the primary focus is Windows at the moment,
while the support for other platforms might be added in future versions.

## Create Canonical Environments

One can create a running environment with "canonical" settings with a
single function call to `use_cleanslate()`:

```{r, eval=FALSE}
cleanslate::use_cleanslate(
  "C:/temp/",
  r_version = "4.1.1",
  from = "https://cran.r-project.org/",
  repo = "https://mran.microsoft.com/snapshot/2021-08-06/"
)
```

This will

- Create a project folder under `C:/temp/` with a `.Rproj` file;
- Download R 4.1.1 installer from CRAN, install it into `C:/temp/R/`;
- Not install Rtools (by default, `rtools_version = NULL`);
- Create a `.Rprofile` file under the project folder, set `options(repos)`
  to use the specified `repo` (an MRAN snapshot in this example),
  and instructions to set the R binary path in RStudio;
- Create a `.Renviron` file under the project folder and set the
  library path to be the library of the project-specific R installation.

As a principle, one should always double-click the `.Rproj` file to
open the project.
This will ensure some sanity checks in the `.Rprofile`, such as whether the
R and library are located within the project folder.

## Create Tailored Environments

To create a more customized running environment, one can use the specific
functions to tailor each aspect, for example:

```{r, eval=FALSE}
library("cleanslate")

"C:/temp/" %>%
  use_project() %>%
  use_rprofile() %>%
  use_renviron() %>%
  use_r_version(version = "4.1.1") %>%
  use_rtools(version = "rtools40")
```

The project context functions
(`use_project()`, `use_rprofile()`, `use_renviron`) support custom templates
using [brew](https://cran.r-project.org/package=brew).

The `use_r_*()` functions have variations that serve as shortcuts
to use R versions defined by release lifecycles, for example,
`use_r_release()`, `use_r_oldrel()`, and `use_r_devel()`.
Note that to ensure better reproducibility, one should still use
`use_r_version()` as the release, oldrel, and devel versions will
shift as time goes by.

The helper functions `version_*()` and `snapshot_*()` could assist you
to determine specific versions of R and Rtools that are currently available,
besides generating and verifying the snapshot repo links.

## Update ADRG

If using cleanslate, remember to update the ADRG instructions for
executing the analysis programs in R. Mostly, this would simplify
the first three steps on creating a project, installing a specific
version of R, and configuring the package repo location.

## RStudio Addin

To make it convenient to use `cleanslate` in experiments,
one can also use its RStudio addin. After `cleanslate`
is installed, click `Addins` -> `cleanslate` -> `Create portable R environment`
in RStudio, or call `cleanslate:::create_env_addin()` to open it.

```{r, out.width="65%", echo=FALSE, fig.cap="cleanslate RStudio addin"}
knitr::include_graphics("images/cleanslate-addin.png")
```

The addin provides a wizard-like interface to help create the environment
with the most important options, although with less flexibility compared
to the functional API demonstrated above.

### 2021-10-12 project-folder.Rmd
proofread chapter 9 - 11

@@
 In a late-stage clinical trial, the number of A&R deliverables can easily be in the hundreds. 
-For an organization, it is also common to have multiple ongoing clinical trials in a clinical program.
+For an organization, it is common to have multiple ongoing clinical trials in a clinical program.
 
-To deliver A&R results of a clinical trial project, it is teamwork that typically requires
-collaborations from both statisticians and programmers. 
+To deliver the A&R results of a clinical trial project, it is teamwork that typically requires collaborations from both statisticians and programmers. 
 In this part, let's consider how to organize a clinical trial project as an A&R leader.
 
-In Chapter \@ref(folder), we will discuss how to organize source code, documents, 
-deliverables in an A&R clinical project. 
+In Chapter \@ref(folder), we will discuss how to organize source code, documents, deliverables in an A&R clinical project. 
 We recommend using the R package folder structure.
 
-In Chapter \@ref(manage), we will discuss a process or system development lifecycle 
-to manage the A&R of a clinical project. 
+In Chapter \@ref(manage), we will discuss a process or system development lifecycle to manage the A&R of a clinical project. 
 We recommend following an agile management approach to define, develop, validate and deliver work.
@@
 
-A clearly defined clinical project folder structure can have many benefits to 
-clinical development teams in an organization. 
+A clearly defined clinical project folder structure can have many benefits to clinical development teams in an organization. 
 Specifically, a well-defined project structure can achieve:
@@
 That is the [R package folder structure](https://github.com/rstudio/cheatsheets/raw/master/package-development.pdf). 
-Every R package developer is required to follow the same convention to organize their R
-functions before the R package can be disseminated through the Comprehensive R Archive Network
-(CRAN). As a user, you can easily install and use those R packages after installing from CRAN. 
+Every R package developer is required to follow the same convention to organize their R functions before the R package can be disseminated through the Comprehensive R Archive Network (CRAN). 
+As a user, you can easily install and use those R packages after installing from CRAN. 
 There are many good resources to guide developers on R package development.
@@
 
-We recommend using the R package folder structure to organize analysis scripts 
-for clinical trial development. 
+We recommend using the R package folder structure to organize analysis scripts for clinical trial development. 
 Using the R package folder structure to streamline data analysis work has also been proposed before 
-(@marwick2018packaging, @wuanalysis).
+(see @marwick2018packaging, @wuanalysis).
 
@@
 For consistency, a well-defined folder structure with potential templates
-ensures project teams organize A&R work consistently across multiple projects. 
-Consistent folder structure also reduces communication costs between study team members 
-and enhances the transparency of projects.
+ensures project teams organize the A&R work consistently across multiple projects. 
+Consistent folder structure also reduces communication costs between study team members and enhances the transparency of projects.
 
@@
 
-We refer to an R package as a **standard R package** if the purpose of an R package is to 
-share commonly used R functions to be hosted in a code repository such as CRAN.
+We refer to an R package as a **standard R package** if the purpose of an R package is to share commonly used R functions to be hosted in a code repository such as CRAN.
 
@@
 
-A general discussion of the R package folder structure can be found 
-in Chapter 4 of the R package book (@wickham2015r).
+A general discussion of the R package folder structure can be found in Chapter 4 of the R package book (@wickham2015r).
 
@@
 
-> Here, the `adam/` folder contains ADaM datasets. 
+> Undeer the `vignettes/` folder, there are `adam/` folder and `tlf` folder.
+> The `adam/` folder contains ADaM datasets. 
 > The `tlf` folder contains output TLFs in RTF format.
@@
 ```
-R/
+R
 ├── count_by.R
@@
 We used Roxygen2 to document functions. 
-For example, the header below defines each variable in fmt_est. 
+For example, the header below defines each variable in `fmt_est`. 
 More details can be found in [Chapter 10 of the R package book](https://r-pkgs.org/man.html)
@@
 Reproducibility of analysis is one of the most important aspects of regulatory deliverables. 
+To ensure a successful reproduce, we need a controlled R environment, including the control of R version and control of R package version.  
 By using the R package folder structure and proper tools (e.g. `renv`, `packrat`), 
@@
 
+First, we introduce the control of R version. 
 In the `esubdemo` project, a reproducible environment is created when you open the 
@@
 More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html). 
-
-> `.Rprofile` is only for project-specific R package. 
-> A standard R package should not use `.Rprofile`.
-
-After we open the `esubdemo` project, the code in `.Rprofile` will automatically check 
-the current R version is the same as we defined in `.Rprofile`
-
+After we open the `esubdemo` project, the code in `.Rprofile` will automatically check the current R version is the same as we defined in `.Rprofile`.
 ```{r, eval = FALSE}
@@
 ```
-
 If the R version mismatch, an error message is displayed as below.
@@
 
-The R package version is controlled in two layers. 
+> `.Rprofile` is only for project-specific R package. 
+> A standard R package should not use `.Rprofile`.
+
+Next, we introduce the control of R package version, which is controlled in two layers. 
 Firstly, we define a snapshot date in `.Rprofile`. 
@@
 
-If the project is initiated properly  
-you shall be able to see similar messages to inform how we control R package versions.
+If the project is initiated properly, you shall be able to see similar messages to inform how we control R package versions.
 
@@
 Once R packages have been properly installed, 
-the system will use the R packages located in the search 
-path defined based on the order of `.libPaths()`.
+the system will use the R packages located in the search path defined based on the order of `.libPaths()`.
 The startup message also provided the R package search path.
@@
 As the consistent folder is defined, it also becomes easier to create specific tools 
-that fit for the analysis and reporting purpose.
-
-Below are few potential tools that can be helpful:
+that fit for the analysis and reporting purpose. Below are few potential tools that can be helpful:
 

### 2021-10-12 project-management.Rmd
proofread chapter 9 - 11

@@
 
-A clinical data analysis project is not unlike typical data analysis projects
-or software projects. Therefore, the conventional wisdom and tricks for
-managing a successful project are also applicable here.
-At the same time, clinical projects also have unique traits,
-such as high standards for planning, development, validation,
-and delivery under strict time constraints.
+A clinical data analysis project is not unlike typical data analysis projects or software projects. 
+Therefore, the conventional wisdom and tricks for managing a successful project are also applicable here.
+At the same time, clinical projects also have unique traits, such as high standards for planning, development, validation, and delivery under strict time constraints.
 
@@
 should take basic training on project management and understand how to work as
-a development team. @teamgeek provides some valuable tips on this topic.
+a development team. 
+@teamgeek provides some valuable tips on this topic.
 As always, setting a clear goal and following a system development lifecycle
@@
 Having a clean architecture design for your code improves the project's
-robustness and flexibility for future changes. For example,
-we should understand how to separate business logic from other layers;
+robustness and flexibility for future changes. 
+For example, we should understand how to separate business logic from other layers;
 know what should be created as reusable components and what should be
 written as one-off analysis scripts; write low coupling, high cohesion code,
-and so on. @cleanarch offers some helpful insights on this topic.
+and so on. 
+@cleanarch offers some helpful insights on this topic.
 
@@
 
-Knowing what you can do is essential. Create a core capabilities list
-for your team.
+Knowing what you can do is essential. 
+Create a core capabilities list for your team.
 
@@
 
-Every individual is limited in some way. The collective thinking
-from a community could benefit a project in the long term.
+Every individual is limited in some way. 
+The collective thinking from a community could benefit a project in the long term.
 When designing reusable components, make a plan to share with
@@
 
-SDLC for A&R deliverables can be defined in four stages:
+SDLC for the A&R deliverables can be defined in four stages:
 
@@
 
-Validation is an crucial stage to ensure the deliverables are accurate and consistent. After the
-development stage is completed, the project team needs to validate the deliverables, including R Markdown
-files for TLFs deliverables and project-specific R functions. The level of validation is determined at the define
-stage.
+Validation is an crucial stage to ensure the deliverables are accurate and consistent. 
+After the development stage is completed, the project team needs to validate the deliverables, including R Markdown
+files for TLFs deliverables and project-specific R functions. 
+The level of validation is determined at the define stage.
 
@@
 The testthat R package can be used to streamline the validation process. 
-More details of the `testthat` package for R package validation can be found in Chapter 12 of the R package book.
+More details of the `testthat` package for R package validation can be found in [Chapter 12 of the R package book](https://r-pkgs.org/tests.html).
 
@@
 tests/testthat
-├── test-independent-test-tlf-01-disposition.Rmd
-├── test-independent-test-tlf-02-population.Rmd
-├── test-independent-test-tlf-03-baseline.Rmd
-├── test-independent-test-tlf-04-efficacy.Rmd
-├── test-independent-test-tlf-05-ae-summary.Rmd
-├── test-independent-test-tlf-06-ae-spec.Rmd
+├── test-independent-test-tlf-01-disposition.R
+├── test-independent-test-tlf-02-population.R
+├── test-independent-test-tlf-03-baseline.R
+├── test-independent-test-tlf-04-efficacy.R
+├── test-independent-test-tlf-05-ae-summary.R
+├── test-independent-test-tlf-06-ae-spec.R
 └── test-independent-test-fmt.R
@@
 
-To validate the content of a Table, we can save the last datasets ready for table generation as a `.Rdata` file.
+To validate the content of a table, we can save the last datasets ready for table generation as a `.Rdata` file.
 A validator can reproduce the TLF and compares it with the original result saved in `.Rdata` file. 
-A test is passed when the results match. Customers can directly review the formatting of the TLFs by
+A test is passed when the results match. 
+Customers can directly review the formatting of the TLFs by
 customers by comparing them with the mock-up.
 
-To validate a Figure, we can use [snapshot testing](https://testthat.r-lib.org/articles/snapshotting.html) strategy
+To validate a figure, we can use [snapshot testing](https://testthat.r-lib.org/articles/snapshotting.html) strategy.
 
@@
 The project lead reviews the tracking sheet to make sure all required activities in the SDLC are completed and
-the tracking sheet has been filled correctly. The deliverables are ready for customer review after all the
-validation steps are completed. Any changes to the output requested by customers are documented.
+the tracking sheet has been filled correctly. 
+The deliverables are ready for customer review after all the validation steps are completed. 
+Any changes to the output requested by customers are documented.
 
@@
 
-After completion of development and required validation of all A&R deliverables, project lead runs compliance
-check for a project-specific R package similar to other R packages. Using `devtools::check()` is a
-convenient way to run compliance checks or `R CMD check`. `R CMD check` is an automated check of the
-contents in the R package for frequently encountered issues, before submission to CRAN. Since the project-specific
-R package is not submitted to CRAN, some checks can be customized and skipped in
-`devtools::check()`. The project lead should work with the study team to ensure all reported errors, warnings, and
-notes by `devtools::check()` are fixed.
+After completion of development and required validation of all A&R deliverables, project lead runs compliance check for a project-specific R package similar to other R packages. 
+`devtools::check()` is a convenient way to run compliance checks or `R CMD check`. 
+`R CMD check` is an automated check of the contents in the R package for frequently encountered issues, before submission to CRAN. 
+Since the project-specific R package is not submitted to CRAN, some checks can be customized and skipped in `devtools::check()`. 
+The project lead should work with the study team to ensure all reported errors, warnings, and notes by `devtools::check()` are fixed.
 
-The project lead can also use the R package `pkgdown` to build a complete website for a project-specific R
-package. The `pkgdown` website is a convenient way to run all analyses in batch and 
-integrate outputs in a website.
-
-Project-specific R functions, TLF generation programs, outputs and validation tracking information can all
-be explored from the website.
-
+The project lead can also use the R package `pkgdown` to build a complete website for a project-specific R package. 
+The `pkgdown` website is a convenient way to run all analyses in batch and 
+integrate outputs in a website, which comprehensively covers project-specific R functions, TLF generation programs, outputs and validation tracking information, etc... 
 For example, in the `esubdemo` project, we created the `pkgdown` website at <https://elong0527.github.io/esubdemo/>.
@@
 such as running testing cases and creating a `pkgdown` website. 
-
 For example, in `esubdemo` project, we setup [GitHub Actions](https://github.com/elong0527/esubdemo/actions)
-for these work. It can be setup by using `usethis::use_github_action()` function.
+for these work.
+It can be setup by using `usethis::use_github_action()` function.

### 2021-10-12 project-folder.Rmd
project-folder.Rmd edited online with Bitbucket

@@
 
-To deliver the A&R results of a clinical trial project, it is teamwork that typically requires collaborations from both statisticians and programmers. 
+To deliver the A&R results of a clinical trial project, 
+it is a teamwork that typically requires collaborations from both statisticians and programmers. 
 In this part, let's consider how to organize a clinical trial project as an A&R leader.

### 2021-10-12 submission-package.Rmd
add review comments

@@
 
-The files in two directories within the eCTD folder structure
-are critical for documenting analysis using R: `datasets`
-and `programs`. We assume the eCTD folder is `esub`.
+In eCTD deliverable, the analysis dataset and source code are saved 
+under the module 5 folder `m5/datasets/<study>/analysis/adam`.
+
+For illustration and preparation purpose, we temporarily save
+the under the `vignettes/esub/` folder.
+
+The files in two directories within the `adam/` folder
+are critical for documenting analysis using R: `datasets/`
+and `programs/`.
 
 ```
-/esub/analysis/adam/
+vignettes/esub/analysis/adam/
 ├── datasets
@@
 
-Folder path: `esub/analysis/adam/datasets/`.
+Folder path: `vignettes/esub/analysis/adam/datasets/`.
 
-- ADaM data in `.xpt` format: created by SAS
-- `define.xml`: created by Pinnacle 21
+- ADaM data in `.xpt` format: created by SAS or R.
+- `define.xml`: created by Pinnacle 21.
 - ADRG (ADaM Reviewer's Guide)
-  - "Macro Programs" section: provide R and R package versions with a snapshot date
+  - "Macro Programs" section: provide R and R package versions with a snapshot date.
   - Appendix: provide step-by-step instructions for reviewers to
-    reproduce the running environment and rerun analyses
+    reproduce the running environment and rerun analyses.
 - ARM (Analysis Results Metadata): provide the links between TLFs and
-  analysis programs in tables
+  analysis programs in tables.
 
@@
 
-Folder path: `esub/analysis/adam/programs/`.
+Folder path: `vignettes/esub/analysis/adam/programs/`.
 
-- `r0pkgs.txt`: contains all internally developed R packages
-- Other `.txt` files: each contains R code for a specific analysis
+- `r0pkgs.txt`: contains all internally developed proprietary R packages.
+- Other `.txt` files: each contains R code for a specific analysis.
 
@@
 Always use a consistent version of R for developing the TLFs and for submission.
-For example, you could enforce a rule to only use R x.y.z where z = 1,
-such as R 4.0.1 or R 4.1.1.
+For example, you could enforce a rule to only use R `x.y.z` where `z = 1`,
+such as R `4.0.1` or R `4.1.1`.
 This can be automatically checked using a

### 2021-10-12 submission-overview.Rmd
proofread chapter 12-13

# (PART) eCTD Submission {-}

# Overview

The electronic Common Technical Document (eCTD) is a standard format for
the electronic submission of applications, amendments, supplements,
and reports from the applicant to the regulator.
The eCTD offers a solution to submit documents stored in a
standard directory structure, with file integrity validation
mechanisms in place.

To submit TLFs created by R to regulatory agencies, we should follow
the spirit of the existing eCTD submission guidelines to prepare
the deliverables, and provide the essential details in the relevant
documents for review.

The goal of the following two chapters is to provide guidance to follow
Section 4.1.2.10 of the
[FDA Study Data Technical Conformance Guide](https://www.fda.gov/media/88173/download):

> "Sponsors should provide the software programs used to create all ADaM
datasets and **generate tables and figures associated with primary and
secondary efficacy analyses**. Furthermore, sponsors should submit software
programs used to generate additional information included in Section 14
CLINICAL STUDIES of the Prescribing Information (PI)26 if applicable.
**The specific software utilized should be specified in the ADaM Reviewer's Guide (ADRG)**.
The main purpose of requesting the submission of these programs
**is to understand the process by which the variables for the respective
analyses were created and to confirm the analysis algorithms**.
Sponsors should submit software programs in **ASCII text format**;
however, executable file extensions should not be used."

In Chapter \@ref(submission-package), we will focus on the approach
to prepare proprietary R packages and analysis code into proper
formats for submission.

In Chapter \@ref(running-environment), we will discuss
the recommendations to make the R code running environment
reproducible for dry run tests and reviews.

### 2021-10-12 submission-package.Rmd
proofread chapter 12-13

--- 2021-10-12 submission-package.Rmd
@@
 submitting R code.
-
 Secondly, we will discuss how to prepare the proprietary R packages
 (if any), and make them part of the submission package.
-
-Lastly, we will provide reusable templates for updating ADRG and ARM,
+Lastly, we will provide reusable templates for updating ADaM Reviewer's Guide (ADRG) and Analysis Results Metadata (ARM),
 so that the reviewers receive proper instructions to reproduce the
@@
 
-The demo project (R package) we will prepare for submission is
-available on GitHub at
-[elong0527/esubdemo](https://github.com/elong0527/esubdemo).
+The demo project (R package) we will prepare for submission is called [`esubdemo`](https://github.com/elong0527/esubdemo.git) available on GitHub.
 You can download or clone it:
@@
 
-- ADaM data in `.xpt` format: created by SAS
-- `define.xml`: created by Pinnacle 21
-- ADRG (ADaM Reviewer's Guide)
-  - "Macro Programs" section: provide R and R package versions with a snapshot date
-  - Appendix: provide step-by-step instructions for reviewers to
-    reproduce the running environment and rerun analyses
-- ARM (Analysis Results Metadata): provide the links between TLFs and
-  analysis programs in tables
+- ADaM data in `.xpt` format: created by SAS.
+- `define.xml`: created by Pinnacle 21.
+- ADRG: 
+  - "Macro Programs" section: provide R and R package versions with a snapshot date.
+  - Appendix: provide step-by-step instructions for reviewers to reproduce the running environment and rerun analyses.
+- ARM: provide the links between TLFs and analysis programs in tables.
 
@@
 
-- `r0pkgs.txt`: contains all internally developed R packages
-- Other `.txt` files: each contains R code for a specific analysis
+- `r0pkgs.txt`: contains all internally developed R packages.
+- Other `.txt` files: each contains R code for a specific analysis.
 
@@
 Now you have a general idea about the relevant components of the
-submission package. We will focus on preparing the internal
-R packages in the following sections.
+submission package. 
+We will focus on preparing the internal R packages in the following sections.
 
@@
 syntax or the
-[remotes dependency syntax](https://remotes.r-lib.org/articles/dependencies.html).
+[remotes dependency syntax](https://remotes.r-lib.org/articles/dependencies.html) in the `DESCRIPTION` file.
 
@@
 This can be automatically checked using a
-[startup script](https://github.com/elong0527/esubdemo/blob/master/inst/startup.R)
-when the R project is opened.
+[startup script](https://github.com/elong0527/esubdemo/blob/master/inst/startup.R) when the R project is opened.
 
@@
 ## Prepare R Packages for Submission
+
+To prepare R packages for submission, one needs to pack the packages into a text files, and then verify if the file only contains ASSCII.
+With packed packages, one can unpack and install them from the text file, too.
 
@@
 
-What happened in the pipe? The function `pkglite::collate()`
-evaluates a specified scope of folders and files defined by
-a list of **file specifications**, and generate a **file collection** object.
+What happened in the pipe? 
+The function `pkglite::collate()` evaluates a specified scope of folders and files defined by a list of **file specifications**, and generate a **file collection** object.
 This file collection contains the metadata required to properly convert
@@
 formats and verified that they can be restored and executed correctly,
-we should update the ADRG (ADaM Reviewer's Guide) to provide
-guidelines on how to use them.
+we should update the ADRG to provide guidelines on how to use them.
 

### 2021-10-13 submission-package.Rmd

@@
 
-The demo project (R package) we will prepare for submission is called [`esubdemo`](https://github.com/elong0527/esubdemo.git) available on GitHub.
+The demo project (R package) we will prepare for submission is called [`esubdemo`](https://github.com/elong0527/esubdemo) available on GitHub.
 You can download or clone it:
@@
 
-To prepare R packages for submission, one needs to pack the packages into a text files, and then verify if the file only contains ASSCII.
+To prepare R packages for submission, one needs to pack the packages into text files, and then verify if the file only contains ASSCII characters.
 With packed packages, one can unpack and install them from the text file, too.

### 2021-10-14 submission-package.Rmd

--- 2021-10-12 submission-package.Rmd
@@
 
-The demo project (R package) we will prepare for submission is
-available on GitHub at
-[elong0527/esubdemo](https://github.com/elong0527/esubdemo).
+The demo project (R package) we will prepare for submission is called [`esubdemo`](https://github.com/elong0527/esubdemo) available on GitHub.
 You can download or clone it:
@@
 ## Prepare R Packages for Submission
+
+To prepare R packages for submission, one needs to pack the packages into text files, and then verify if the file only contains ASSCII characters.
+With packed packages, one can unpack and install them from the text file, too.
 

### 2021-10-14 slides/workshop-slides.Rmd

@@
 - Clinical trial project (Nan Xiao)
-  + Session 3 (40 min)
+  + Session 3 (30 min)
   + Break (10 min)
 - eCTD submission package (Nan Xiao)
-  + Session 4 (30 min)
+  + Session 4 (40 min)
   + Q&A (10 min)

### 2021-10-14 submission-environment.Rmd

@@
 ::: {.rmdnote}
-As of Q4 2021, the cleanslate package used in this chapter
+As of Q4 2021, the `cleanslate` package used in this chapter
 is still under active development and validation.
@@
 - Create a project folder under `C:/temp/` with a `.Rproj` file;
-- Download R 4.1.1 installer from CRAN, install it into `C:/temp/R/`;
+- Download R `4.1.1` installer from CRAN, install it into `C:/temp/R/`;
 - Not install Rtools (by default, `rtools_version = NULL`);
@@
 
-If using cleanslate, remember to update the ADRG instructions for
+If using `cleanslate`, remember to update the ADRG instructions for
 executing the analysis programs in R. Mostly, this would simplify

### 2021-10-14 tlf-ae-spec.Rmd

@@
 
-Following [ICH E3
-guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
+Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
 we need to summarize which participants were included in each efficacy
@@
 
-The data used to summarize AE information is in `adsl` and `adae`
-datasets.
+The data used to summarize AE information is in `adsl` and `adae` datasets.
 
@@
 
-The final report data is saved in `tbl_ae_spec`. We also add a blank row
-between population and AE information in the reporting table.
+The final report data is saved in `tbl_ae_spec`.
+We also add a blank row between population and AE information in the reporting table.
 
@@
 
-More discussion on `page_by`, `group_by`, and `subline_by` features can
-be found in the [`r2rtf` package
-website](https://merck.github.io/r2rtf/articles/example-sublineby-pageby-groupby.html).
+More discussion on `page_by`, `group_by` and `subline_by` 
+features can be found in the 
+[`r2rtf` package website](https://merck.github.io/r2rtf/articles/example-sublineby-pageby-groupby.html.
 

### 2021-10-14 tlf-ae-summary.Rmd

@@
 - With serious drug-related adverse events
--   Who "died"
+- Who died"
 
@@
 -   Step 2: Summarize participants in population by treatment group, and we name the dataset as `t_pop`.
--   Step 3: Summarize participants in population by required AE criteria of interest, and we name the dataset as `t_ae`.
+-   Step 3: Summarize participants in population by required AE criteria of interest, 
+   and we name the dataset as `t_ae`.
 -   Step 4: Rowly bind `t_pop` and `t_ae` and format it by `r2rtf`.

### 2021-10-14 tlf-disposition.Rmd

@@
 
-In the code below, we calculate the number and percentage of participants who discontinued the study.
-
-Here we use the `formatC()` function to customize the numeric value to be 1 decimal with fixed width of 5.
-
 ```{r}
@@
 
-In the code below, we calculate the number and percentage of participants who completed/discontinued the study for different reasons.
+In the code below, we calculate the number and percentage of 
+participants who completed/discontinued the study for different reasons.
 
@@
 
-In the final step, the individual rows are combined and used as input for `r2rtf.
+Now we combined individual rows into the whole table for reporting purpose. 
+`tbl_disp` is used as input for `r2rtf` to create final report.
 
@@
 -   Step 3: Count the number and percentage of participants that discontinued the study, and name the dataset `n_disc`.
--   Step 4: Count the number and percentage of participants that discontinued the study for different reasons, and name the dataset `n_reason`.
+-   Step 4: Count the number and percentage of participants that discontinued the study for different reasons, 
+    and name the dataset `n_reason`.
 -   Step 5: Count the number and percentage of participants that completed the study, and name the dataset `n_complete`.

### 2021-10-14 index.Rmd
use sentence case instead of title case for headings; run styler on Rmds

@@
 description: >
-  This book will teach you how to prepare tables, listings and figures for clinical study report and submit to regulatory agency, 
+  This book will teach you how to prepare tables, listings, and figures for
+  clinical study report and submit to regulatory agencies, 
   the essential part of clinical trial development.
@@
 ```{r, eval = FALSE, include=FALSE}
-r2rtf:::rtf_convert_format(list.files("tlf", pattern = ".rtf", full.names = TRUE), 
-                           output_dir = "tlf")
+r2rtf:::rtf_convert_format(
+  list.files("tlf", pattern = ".rtf", full.names = TRUE),
+  output_dir = "tlf"
+)
 ```
@@
 provides guidance to assist sponsors in the development of a CSR. 
-In this book, you will learn how to use R to prepare a CSR and how to submit it to regulatory agencies.
+In this book, you will learn how to use R to prepare a CSR and
+how to submit it to regulatory agencies.
 
@@
 
-## List of Authors and Contributors {-}
+## Authors and contributors {-}
 
 ```{r, echo = FALSE}
-contributors <- c(person("Yilong", "Zhang", email = "yilong.zhang@merck.com", role = "aut"), 
+contributors <- c(
+  person("Yilong", "Zhang", email = "yilong.zhang@merck.com", role = "aut"),
   person("Nan", "Xiao", role = "aut"),
@@
   person("Aiming", "Yang", role = "ctr"),
-                  person("Steven", "Haesendonckx", role = "ctr"))
+  person("Steven", "Haesendonckx", role = "ctr")
+)
 ```

### 2021-10-14 intro.Rmd
use sentence case instead of title case for headings; run styler on Rmds

@@
 
-- Part 3, "eCTD Submission package"
+- Part 3, "eCTD submission package"
 provides general information in preparing 
@@
 We share the same philosophy described in 
-[Section 1.1 of R Packages book](https://r-pkgs.org/intro.html?q=Marwick#intro-phil) 
+[Section 1.1 of R Packages book](https://r-pkgs.org/intro.html#intro-phil) 
 and quote here.

### 2021-10-14 tlf-ae-spec.Rmd
use sentence case instead of title case for headings; run styler on Rmds

@@
 ana <- adae %>%
-  mutate(AESOC = toTitleCase(tolower(AESOC)), 
-         AEDECOD =  toTitleCase(tolower(AEDECOD)))
+  mutate(
+    AESOC = toTitleCase(tolower(AESOC)),
+    AEDECOD = toTitleCase(tolower(AEDECOD))
+  )
 
-t1 <-ana %>% 
+t1 <- ana %>%
   group_by(TRTAN, AESOC) %>%
@@
 t_ae <- bind_rows(t1, t2) %>%
-        pivot_wider(id_cols = c(AESOC, order, AEDECOD), 
+  pivot_wider(
+    id_cols = c(AESOC, order, AEDECOD),
     names_from = TRTAN,
@@
     values_from = n,
-                    values_fill = fmt_num(0, digits = 0)) %>%
+    values_fill = fmt_num(0, digits = 0)
+  ) %>%
   arrange(AESOC, order, AEDECOD) %>%
@@
     AESOC = "pop",
-    AEDECOD = var_label) %>%
-  select(AESOC,AEDECOD, starts_with("n_"))
+    AEDECOD = var_label
+  ) %>%
+  select(AESOC, AEDECOD, starts_with("n_"))
 
@@
 ```{r}
-tbl_ae_spec <- bind_rows(t_pop,
+tbl_ae_spec <- bind_rows(
+  t_pop,
   data.frame(AESOC = "pop"),
-                         t_ae) %>% 
-  mutate(AEDECOD = ifelse(AEDECOD == AESOC , 
-                          AEDECOD, paste0("  ", AEDECOD)))
+  t_ae
+) %>%
+  mutate(AEDECOD = ifelse(AEDECOD == AESOC,
+    AEDECOD, paste0("  ", AEDECOD)
+  ))
 
@@
 tbl_ae_spec %>%
-  rtf_title("Analysis of Participants With Specific Adverse Events", 
-            "(Safty Analysis Population)") %>%
-
+  rtf_title(
+    "Analysis of Participants With Specific Adverse Events",
+    "(Safty Analysis Population)"
+  ) %>%
   rtf_colheader(" | Placebo | Xanomeline Low Dose| Xanomeline High Dose",
@@
   ) %>%
-
   rtf_body(
@@
     page_by = "AESOC",
-    pageby_row = "first_row") %>%
-
+    pageby_row = "first_row"
+  ) %>%
   rtf_footnote("Every subject is counted a single time for each applicable row and column.") %>%
-
   rtf_encode() %>%
-
   write_rtf("tlf/tlf_spec_ae.rtf")
-
 ```
@@
 - Step 3: Count the number of subjects in each AE term by SOC, AE
-    term, and treatment group (rows without bold text) and save into
-    `t2`.
+  term, and treatment group (rows without bold text) and save into `t2`.
 - Step 4: Bind `t1` and `t2` by row into `t_ae`.

### 2021-10-14 tlf-ae-summary.Rmd
use sentence case instead of title case for headings; run styler on Rmds

@@
-# AE Summary {#aesummary}
+# AE summary {#aesummary}
 
@@
 t_pop <- pop %>%
-  mutate(name = "Participants in population", 
-         tot = fmt_num(tot, digits = 0)) %>% 
-  pivot_wider(id_cols = name, 
+  mutate(
+    name = "Participants in population",
+    tot = fmt_num(tot, digits = 0)
+  ) %>%
+  pivot_wider(
+    id_cols = name,
     names_from = TRTAN,
     names_prefix = "n_",
-              values_from = tot)
+    values_from = tot
+  )
 

### 2021-10-14 tlf-baseline.Rmd
use sentence case instead of title case for headings; run styler on Rmds

@@
-# Baseline Characteristics
+# Baseline characteristics
 
@@
 ana <- adsl %>%
-  mutate(SEX = factor(SEX, c("F", "M"), c("Female", "Male")), 
-         RACE = toTitleCase(tolower(RACE)))
+  mutate(
+    SEX = factor(SEX, c("F", "M"), c("Female", "Male")),
+    RACE = toTitleCase(tolower(RACE))
+  )
 
@@
   as_tibble() %>%
-              mutate( across(everything(), 
-                      ~ str_replace_all(.x, intToUtf8(160), " ") ) 
-                    )
+  mutate(across(
+    everything(),
+    ~ str_replace_all(.x, intToUtf8(160), " ")
+  ))
 
@@
 tbl_base[-1, ] %>%
-      rtf_title("Participant Baseline Characteristics", 
-                "(All Participants Randomized)") %>%
+  rtf_title(
+    "Participant Baseline Characteristics",
+    "(All Participants Randomized)"
+  ) %>%
   rtf_colheader(colheader1,
-                    col_rel_width = rel_width) %>% 
+    col_rel_width = rel_width
+  ) %>%
   rtf_colheader(colheader2,
     border_top = "",
-                    col_rel_width = rel_width) %>%
-      rtf_body(col_rel_width = rel_width,
+    col_rel_width = rel_width
+  ) %>%
+  rtf_body(
+    col_rel_width = rel_width,
     text_justification = c("l", rep("c", 4)),
     text_indent_first = -240,
-               text_indent_left = 180) %>% 
+    text_indent_left = 180
+  ) %>%
   rtf_encode() %>%

### 2021-10-14 tlf-disposition.Rmd
use sentence case instead of title case for headings; run styler on Rmds

@@
 tbl_disp %>%
-
   # Table title
   rtf_title("Disposition of Participants") %>%
-
   # First row of column header
@@
   ) %>%
-
   # Table body

### 2021-10-14 tlf-efficacy.Rmd
use sentence case instead of title case for headings; run styler on Rmds

@@
 
-## Analysis Dataset
+## Analysis dataset
 
@@
 ```
-## Helper Functions
+## Helper functions
 
@@
 
-## Summary of Observed Data
+## Summary of observed data
 
@@
 
-## Missing Data Imputation
+## Missing data imputation
 
@@
   ) %>%
-  rtf_body(text_justification = c("l", rep("c", 7)), 
-           col_rel_width = c(2.5, rep(c(0.5, 1.5), 3), 2)) %>%
+  rtf_body(
+    text_justification = c("l", rep("c", 7)),
+    col_rel_width = c(2.5, rep(c(0.5, 1.5), 3), 2)
+  ) %>%
   rtf_footnote(c(
@@
   ) %>%
-  rtf_body(text_justification = c("l", "c", "c"), 
-           col_rel_width = c(4.5, 4, 2))
+  rtf_body(
+    text_justification = c("l", "c", "c"),
+    col_rel_width = c(4.5, 4, 2)
+  )
 

### 2021-10-14 tlf-overview.Rmd
use sentence case instead of title case for headings; run styler on Rmds

@@
 
-## Structure and Content
+## Structure and content
 
@@
 - Figure out table layout. 
-
 - Split the layout into small tasks in the form of a computer program.
-
 - Execute the program.
@@
   rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
-
   # define relative width
@@
 head(tbl) %>%
-
   rtf_colheader(
@@
   ) %>%
-
   rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
-
   rtf_encode() %>%
-
   write_rtf("tlf/intro-ae3.rtf")
@@
 head(tbl) %>%
-
   rtf_body(text_justification = c("l", "c", "c", "c")) %>%
-
   rtf_encode() %>%
-
   write_rtf("tlf/intro-ae5.rtf")
@@
   ) %>%
-
   rtf_colheader(
@@
   ) %>%
-
   rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
-
   rtf_encode() %>%
-
   write_rtf("tlf/intro-ae7.rtf")

### 2021-10-14 tlf-population.Rmd
use sentence case instead of title case for headings; run styler on Rmds

@@
-# Analysis Population {#population}
+# Analysis population {#population}
 
@@
 
-## Helper Functions
+## Helper functions
 
@@
 count_by(adsl, "TRT01PN", "EFFFL") %>%
-  select(- ends_with(c("_54", "_81")))
+  select(-ends_with(c("_54", "_81")))
 ```
 
-## Analysis Code
+## Analysis code
 
@@
 pop <- count_by(adsl, "TRT01PN", "RANDFL",
-                var_label = "Participants in Population") %>%
+  var_label = "Participants in Population"
+) %>%
   select(var_label, starts_with("n_"))

### 2021-10-27 slides/workshop-slides.Rmd
update slides

@@
 ```{r, eval = FALSE, include = FALSE}
-rmarkdown::render("slides/workshop-slides.Rmd")
+rmarkdown::render("workshop-slides.Rmd")
 ```
@@
 
+## Training Objective
+
+- Learning how to create tables for clinical study reports.
+- Learning how to organize a clinical development project.
+- Learning how to prepare eCTD submission package to FDA.
+
+> The tool, process and format may be different in different organization/company.
+> We only provide a recommended way to address them.
+
+## Acknowledgement
+
+- Team members from Merck & Co. 
+  + [We are hiring!](https://jobs.merck.com/bards) 
+  + Contact us if interested.
+
+- [Contributors](https://r4csr.org/) of the training materials.
+  + Please consider to submit issues or PR in the repo.
+
+- [R consortium](https://www.r-consortium.org/projects/isc-working-groups)
+  + R validation hub
+  + Submission working group
+  + R Tables for Regulatory Submission (RTRS) working group
+  + Please consider to join and contribute to a working group.
+
+- R / Pharma organizers
+  + It is a fun and productive annual gathering. 
+  + Please consider to share story and use cases to enlarge the community.
+
 ## Preparation
@@
 - Training Material: https://r4csr.org/
-- Demo project: https://github.com/elong0527/esubdemo
-- R Consortium FDA submission pilot: https://github.com/RConsortium/submissions-pilot1
+- Demo project: 
+  + Analysis project: https://github.com/elong0527/esubdemo
+  + eCTD submission package: https://github.com/elong0527/ectddemo
+- R Consortium FDA submission pilot: 
+  + https://github.com/RConsortium/submissions-pilot1
+
+- During the workshop, we will use `esubdemo` project.
+  + `esubdemo` is shared in RStudio Cloud (https://rstudio.cloud/) 
+  + Project link will be shared in chat
+  + Post questions in group chat.
 
@@
 
-We share the same philosophy described in [Section 1.1 of R Packages book](https://r-pkgs.org/intro.html?q=Marwick#intro-phil) and quote here.
+We share the same philosophy described in 
+[Section 1.1 of R Packages book](https://r-pkgs.org/intro.html?q=Marwick#intro-phil) and quote here.
 

### 2021-10-27 slides/workshop-slides.Rmd
reorg r2rtf examples

@@
 knitr::opts_chunk$set(echo = FALSE)
+```
+
+```{r}
+library(dplyr)
+library(tidyr)
+library(r2rtf)
 ```
@@
 
-## r2rtf Motivation
+# r2rtf Introduction
+
+## Motivation
 
@@
 in preparing clinical study reports
+
+- [ICH E3 Structure and Content of Clinical Study Reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf)
 
@@
 
+- `r2rtf` is an R package to create production-ready tables and figures in RTF format.
+The R package is designed to
+
 `r2rtf` is designed to:
 
-- Generate highly customized tables
-- Limit package dependency
-- Target regulatory deliverables   
-- Support pipes (`%>%`)
+- provide simple "verb" functions that correspond to each component of a table, 
+  to help you translate a data frame to a table in RTF file.
+- enable pipes (`%>%`). 
+- only focus on the **table format**. 
+  Data manipulation and analysis shall be handled by other R packages. (e.g. `tidyverse`)
 
-- `r2rtf` Introduction: https://r4csr.org/overview.html#r2rtf
+## Workflow
+
+Before creating an RTF table we need to:
+
+- Figure out table layout.
+
+- Split the layout into small tasks in the form of a computer program.
+
+- Execute the program.
+
+## Minimal Example
+
+`r2rtf` is designed to enable pipes (`%>%`)
+
+```{r, echo = TRUE, eval = FALSE}
+head(iris) %>%
+  rtf_body() %>%                     # Step 1 Add table attributes 
+  rtf_encode() %>%                   # Step 2 Convert attributes to RTF encode 
+  write_rtf("minimal.rtf")           # Step 3 Write to a .rtf file
+```
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("minimal.pdf")
+```
+
+## Package Overview
+
+`r2rtf` package provides the flexibility to customize table appearance for
+
+- **Table component**: title, column header, footnote, etc.
+- **Table cell style**: size, border type, color, font size, text color, alignment, etc.
+- **Flexible control**: the specification of the cell style can be row or column vectorized.
+- **Complicated format**: pagination, section grouping, multiple table concatenations, etc.
+
+`r2rtf` package also provides the flexibility to convert figures in RTF format.
+
+## Simple Example - Adverse Events
+
+`r2rtf` only focuses on **table format**. 
+Data manipulation and analysis shall be handled by other R packages. (e.g. `tidyverse`)
+
+- Let's refer examples in the [Section 2.4.2](https://r4csr.org/overview.html#r2rtf)
+
+## Function Summary
+
+`r2rtf` provides simple "verb" functions that correspond to each component of a table, 
+  to help you translate data frame to table in RTF file.
+
+```{r, echo = FALSE}
+r2rtf_funs <- read.csv("r2rtf-functions.csv")
+names(r2rtf_funs)[3] <- "Optional/required"
+knitr::kable(r2rtf_funs)
+```
+
+## Function Illustration
+
+```{r, out.width="80%", echo = FALSE}
+knitr::include_graphics("function_summary.PNG")
+```
+
+# Break (10 min)
+
+# CSR Examples
 
@@
 https://r4csr.org/population.html
-
-# Break (10 min)
 

### 2021-10-27 slides/workshop-slides.Rmd
change break time

@@
   + Session 1 (45 min)
+  + Break and/or Exercise (20 min)
+  + Session 2 (45 min)
   + Break (10 min) 
-  + Session 2 (45 min)
-  + Break (20 min) 
 - Clinical trial project (Nan Xiao)
@@
 
-# Break (10 min)
+# Break and/or Exercise (20 min)
 
@@
 
-# Break (20 min)
+# Break (10 min)
 

### 2021-10-27 tlf-disposition.Rmd

@@
 Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), 
-A summary needs to be provided to account for of all participants who entered the study in Section 10.1, Disposition of Participants.
+a summary needs to be provided to account for all participants who entered the study in Section 10.1, Disposition of Participants.
 
@@
 
-The procedure to generate a disposition table can be summarized as follow:
+The procedure to generate a disposition table can be summarized as follows:
 

### 2021-10-27 tlf-ae-spec.Rmd
obvious fixes

@@
     "Analysis of Participants With Specific Adverse Events",
-    "(Safty Analysis Population)"
+    "(Safety Analysis Population)"
   ) %>%

### 2021-10-27 tlf-efficacy.Rmd
obvious fixes

@@
     "LOCF",
-    "Efficacly Analysis Population"
+    "Efficacy Analysis Population"
   )) %>%

### 2021-10-27 submission-package.Rmd
use sentence case instead of title case for headings; run styler on Rmds

@@
-# Submission Package {#submission-package}
+# Submission package {#submission-package}
 
@@
 
-## The Whole Game
+## The whole game
 
@@
 
-## Practical Considerations for R Package Submissions
+## Practical considerations for R package submissions
 
@@
 
-### Source Location
+### Source location
 
@@
 
-### Dependency Locations
+### Dependency locations
 
@@
 
-### R Version
+### R version
 
@@
 
-### Package Repo Version
+### Package repo version
 
@@
 
-### System Environments
+### System environments
 
@@
 
-## Prepare R Packages for Submission
+## Prepare R Packages for submission
 
@@
 
-## Prepare Analysis Programs for Submission
+## Prepare analysis programs for submission
 
@@
 ```{r, eval=FALSE}
-knitr::purl(
-  "tlf-01-disposition.Rmd",
-  output = "esub/analysis/adam/programs/tlf-01-disposition.txt"
-)
+input_path <- "~/esubdemo/vignettes/"
+output_path <- "~/ectddemo/m5/datasets/ectddemo/analysis/adam/programs/"
+knitr::purl(file.path(input_path, "tlf-01-disposition.Rmd"), output = file.path(output_path, "tlf-01-disposition.txt"))
+knitr::purl(file.path(input_path, "tlf-02-population.Rmd"), output = file.path(output_path, "tlf-02-population.txt"))
+knitr::purl(file.path(input_path, "tlf-03-baseline.Rmd"), output = file.path(output_path, "tlf-03-baseline.txt"))
+knitr::purl(file.path(input_path, "tlf-04-efficacy.Rmd"), output = file.path(output_path, "tlf-04-efficacy.txt"))
+knitr::purl(file.path(input_path, "tlf-05-ae-summary.Rmd"), output = file.path(output_path, "tlf-05-ae-summary.txt"))
+knitr::purl(file.path(input_path, "tlf-06-ae-spec.Rmd"), output = file.path(output_path, "tlf-06-ae-spec.txt"))
 ```
 
-Repeat this for all the R Markdown files.
+Run this on all the R Markdown files under `vignettes/`.
 
@@
 # path$outtable = path$outgraph = "." # Outputs saved to the current folder
+```
+
+To automate this process:
+
+```{r, eval=FALSE}
+header <- readLines(textConnection("# Note to Reviewer
+# To rerun the code below, please refer to the ADRG appendix.
+# After the required packages are installed,
+# the path variable needs to be defined by using the example code below.
+#
+# path = list(adam = \"/path/to/esub/analysis/adam/datasets\") # Modify to use actual location
+# path$outtable = path$outgraph = \".\" # Outputs saved to the current folder"))
+
+append_header <- function(file, header) {
+  x <- readLines(file)
+  y <- c(header, "\n", x)
+  writeLines(y, con = file)
+  invisible(file)
+}
+
+file.path(output_path, "tlf-01-disposition.txt") %>% append_header(header)
+file.path(output_path, "tlf-02-population.txt") %>% append_header(header)
+file.path(output_path, "tlf-03-baseline.txt") %>% append_header(header)
+file.path(output_path, "tlf-04-efficacy.txt") %>% append_header(header)
+file.path(output_path, "tlf-05-ae-summary.txt") %>% append_header(header)
+file.path(output_path, "tlf-06-ae-spec.txt") %>% append_header(header)
 ```
@@
 
-2. Define Working Directory
+2. Define working directory
 
@@
 
-3. Specify R Package Repository
+3. Specify R package repository
 
@@
 
-4. Install Open-Source R Packages
+4. Install open-source R packages
 
@@
 
-5. Install Proprietary R Packages
+5. Install proprietary R packages
 
@@
 
-7. Execute Analysis Program
+7. Execute analysis program
 

### 2021-10-27 index.Rmd
Added myself as a contributor

@@
   person("Aiming", "Yang", role = "ctr"),
-  person("Steven", "Haesendonckx", role = "ctr")
+  person("Steven", "Haesendonckx", role = "ctr"),
+  person("Howard", "Baek", role = "ctr")
 )

### 2021-10-27 slides/workshop-slides.Rmd
add slides for ch 9 - 14

@@
 author: "Yilong Zhang, Nan Xiao, Keaven Anderson"
-date: "October, 2021"
+date: "October 2021"
 output:
@@
 
-> The tool, process and format may be different in different organization/company.
+> The toolchain, process, and formats may be different in different organizations.
 > We only provide a recommended way to address them.
@@
 - [Contributors](https://r4csr.org/) of the training materials.
-  + Please consider to submit issues or PR in the repo.
+  + Please consider submitting issues or PR in the repo.
 
-- [R consortium](https://www.r-consortium.org/projects/isc-working-groups)
+- [R Consortium](https://www.r-consortium.org/projects/isc-working-groups)
   + R validation hub
@@
   + It is a fun and productive annual gathering.
-  + Please consider to share story and use cases to enlarge the community.
+  + Please consider sharing stories and use cases to expand the community.
 
@@
 install.packages(c(
- "tidyverse",        # Data Manipulation 
+  "tidyverse",      # Data manipulation
   "r2rtf",          # TLF generation
- "pkglite"           # eCTD Submission 
+  "pkglite"         # eCTD submission
 ))
@@
 
-- Training Material: https://r4csr.org/
+- Training Material: <https://r4csr.org/>
 - Demo project: 
-  + Analysis project: https://github.com/elong0527/esubdemo
-  + eCTD submission package: https://github.com/elong0527/ectddemo
+  + Analysis project: <https://github.com/elong0527/esubdemo>
+  + eCTD submission package: <https://github.com/elong0527/ectddemo>
 - R Consortium FDA submission pilot: 
-  + https://github.com/RConsortium/submissions-pilot1
+  + <https://github.com/RConsortium/submissions-pilot1>
 
 - During the workshop, we will use `esubdemo` project.
-  + `esubdemo` is shared in RStudio Cloud (https://rstudio.cloud/) 
+  + `esubdemo` is shared in RStudio Cloud (<https://rstudio.cloud/>)
   + Project link will be shared in chat
@@
 
-## ICH E3 guidance
+## ICH E3 Guidance
 
-The [ICH E3: strcuture and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) provide guidance to assist sponsors in the development of a CSR.
+The [ICH E3: structure and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) provide guidance to assist sponsors in the development of a CSR.
 
@@
 
-- The dataset structure is following [CDISC Analysis Data Model (ADaM)](https://www.cdisc.org/standards/foundational/adam).
+- The dataset structure follows the [CDISC Analysis Data Model (ADaM)](https://www.cdisc.org/standards/foundational/adam).
 
-- Source data: https://github.com/elong0527/r4csr/tree/master/adam_data
+- Source data: <https://github.com/elong0527/r4csr/tree/master/adam_data>
 
@@
 
-- `r2rtf`: an R package to create production ready tables and figures in RTF format.
+- `r2rtf`: an R package to create production-ready tables and figures in RTF format.
 
@@
 
-Different organization can have different table standards
+Different organizations can have different table standards
 
@@
 
-Before creating an RTF table we need to:
+Before creating an RTF table, we need to:
 
@@
 
-- Let's refer examples in the [Section 2.4.2](https://r4csr.org/overview.html#r2rtf)
+- Let's refer to examples in the [Section 2.4.2](https://r4csr.org/overview.html#r2rtf)
 
@@
 `r2rtf` provides simple "verb" functions that correspond to each component of a table, 
-  to help you translate data frame to table in RTF file.
+  to help you translate data frame to tables in RTF file.
 
@@
 
-https://r4csr.org/disposition.html
+<https://r4csr.org/disposition.html>
 
@@
 
-https://r4csr.org/population.html
+<https://r4csr.org/population.html>
 
@@
 
-https://r4csr.org/baseline-characteristics.html
+<https://r4csr.org/baseline-characteristics.html>
 
@@
 
-https://r4csr.org/efficacy.html
+<https://r4csr.org/efficacy.html>
 
@@
 
-https://r4csr.org/aesummary.html
+<https://r4csr.org/aesummary.html>
 
@@
 
-https://r4csr.org/specific-ae.html
+<https://r4csr.org/specific-ae.html>
+
+# Break (10 min)
 
 # Clinical Trial Project
-
-# Break (10 min)
 
@@
 
-In this part, let’s consider how to organize a clinical trial project as an A&R leader.
+In this part, let's consider how to organize a clinical trial project as an A&R lead.
 
@@
 
-https://r4csr.org/folder.html
+Our primary focus is creating a standard R package structure to organize the
+project, with 4 goals in mind:
+
+- Consistency
+- Reproducibility
+- Automation
+- Compliance
+
+<https://r4csr.org/folder.html>
 
@@
 
-https://r4csr.org/manage.html
+- Setting up for success
+  - Teamwork
+  - Architecture
+  - Boundaries
+  - Community
+
+- The SDLC
+  - Planning
+  - Development
+  - Validation
+  - Operation
+
+<https://r4csr.org/manage.html>
 
@@
 
+## Overview
+
+<https://r4csr.org/overview-2.html>
+
+## Submission Package
+
+We will discuss strategies to prepare proprietary R packages and analysis code
+into proper formats for submission:
+
+- The whole game
+- Practical considerations
+- Prepare R packages using `pkglite`
+- Prepare analysis programs
+- Update ADRG and ARM
+
+<https://r4csr.org/submission-package.html>
+
+## Running Environment
+
+We will also give recommendations to make the R code
+running environment reproducible for dry run tests and reviews
+using `cleanslate`.
+
+<https://r4csr.org/running-environment.html>
+
+# pkglite
+
+## pkglite: Compact Package Representations
+
+- To provide a **tool** for packing and restoring R packages as plaintext
+assets that are easy to store, transfer, and review.
+- To provide a **grammar** for specifying the file packing scope that is
+functional, precise, and extendable.
+- To provide a **standard** for exchanging the packed asset that is
+unambiguous, human-friendly, and machine-readable.
+
+## Pipe-Friendly Workflow
+
+```{r, eval=FALSE, echo=TRUE}
+library("pkglite")
+
+"/path/to/pkg/" %>%
+  collate(file_ectd()) %>%
+  pack()
+
+pack(
+  "/path/to/pkg1/" %>% collate(file_ectd()),
+  "/path/to/pkg2/" %>% collate(file_ectd()),
+  output = "/path/to/pkglite.txt"
+)
+
+"/path/to/pkglite.txt" %>%
+  unpack(output = "/path/to/output/")
+```
+
+## File Specifications + File Collections
+
+File specifications offer flexibility and brevity in specifying the files to include.
+
+| File specification type | Functions                                                                           |
+|-------------------------|-------------------------------------------------------------------------------------|
+| Manual discovery        | `file_spec()`                                                                       |
+| Automatic discovery     | `file_auto()`                                                                       |
+| Common patterns         | `file_root_core()`, `file_r()`, `file_man()`, `file_src()`, `file_vignettes()`, ... |
+| Default sets            | `file_default()`, `file_ectd()`                                                     |
+
+<br>
+
+File collections contain the evaluation results of file specifications for packing.
+
+```{r, eval=FALSE, echo=TRUE}
+library("pkglite")
+
+"/path/to/pkg" %>%
+  collate(file_root_core(), file_r(), file_auto("inst/"))
+```
+
+## pkglite.txt
+
+pkglite.txt follows the standard Debian Control File (DCF) format used by
+Debian, R, and RStudio IDE, to be both machine-readable and human-readable.
+
+```
+# Generated by pkglite: do not edit by hand
+# Use pkglite::unpack() to restore the packages
+
+Package: pkg1
+File: DESCRIPTION
+Format: text
+Content:
+  Package: pkg1
+  Type: Package
+  Title: Example Package One
+  Version: 0.1.0
+```
+
 # Q&A

### 2021-10-27 submission-package.Rmd
use ectddemo + update paths

--- 2021-10-14 submission-package.Rmd
@@
 
-## The Whole Game
+The demo submission package (not to be confused with the R package above)
+is [`ectddemo`](https://github.com/elong0527/ectddemo), also available on GitHub.
+You can download or clone it:
+
+```bash
+git clone https://github.com/elong0527/ectddemo.git
+```
+
+We will assume the paths to the two folders are `esubdemo/` and `ectddemo/` below.
+
+## The whole game
 
 In eCTD deliverable, the analysis dataset and source code are saved 
-under the module 5 folder `m5/datasets/<study>/analysis/adam`.
+under the eCTD module 5 (clinical study reports) folder
 
-For illustration and preparation purpose, we temporarily save
-the under the `vignettes/esub/` folder.
+```
+ectddemo/m5/datasets/<study>/analysis/adam/
+```
 
 The files in two directories within the `adam/` folder
-are critical for documenting analysis using R: `datasets/`
-and `programs/`.
+are critical for documenting analysis using R: `datasets/` and `programs/`.
 
 ```
-vignettes/esub/analysis/adam/
+ectddemo/m5/datasets/ectddemo/analysis/adam/
 ├── datasets
-│   ├── adrg.docx
-│   ├── analysis-results-metadata.docx
-│   ├── define-v1-updated-html.xsl
-│   ├── define.pdf
-│   └── define.xml
+│   ├── adae.xpt
+│   ├── ...
+│   ├── adrg.pdf
+│   ├── analysis-results-metadata.pdf
+│   ├── define.xml
+│   └── define2-0-0.xsl
 └── programs
+    ├── r0pkgs.txt
     ├── tlf-01-disposition.txt
@@
     ├── tlf-05-ae-summary.txt
-    ├── tlf-06-ae-spec.txt
-    └── r0pkgs.txt
+    └── tlf-06-ae-spec.txt
 ```
@@
 
-Folder path: `vignettes/esub/analysis/adam/datasets/`.
+Folder path: `ectddemo/m5/datasets/ectddemo/analysis/adam/datasets/`.
 
@@
 
-Folder path: `vignettes/esub/analysis/adam/programs/`.
+Folder path: `ectddemo/m5/datasets/ectddemo/analysis/adam/programs/`.
 
@@
 
-- The filenames under `programs` should all use lower case
+- The filenames under `programs/` should all use lower case
 (with no underscores or other special characters).
@@
 This can be verified by `pkglite::verify_ascii()`
-- The `.docx` files will be converted to PDF files for formal submission.
+- Any `.docx` files should be converted to PDF files for formal submission.
 
@@
 submission package. 
-We will focus on preparing the internal R packages in the following sections.
+We will start preparing the proprietary R packages in the following sections.
 
-## Practical Considerations for R Package Submissions
+## Practical considerations for R package submissions
 
@@
 Let's pack the `esubdemo` package into a text file.
-Assume the source package path is `~/esubdemo/`,
+Assume the source package path is `esubdemo/`,
 you should be able to pack the package with a single pipe:
@@
 
-"~/esubdemo/" %>%
+"esubdemo/" %>%
   collate(file_ectd(), file_auto("inst")) %>%
@@
 What happened in the pipe?
-The function `pkglite::collate()` evaluates a specified scope of folders and files defined by a list of **file specifications**, and generate a **file collection** object.
+The function `pkglite::collate()` evaluates a specified scope of
+folders and files defined by a list of **file specifications**,
+and generate a **file collection** object.
 This file collection contains the metadata required to properly convert
@@
 
-To pack multiple R packages, simply feed multiple file collection as inputs:
+To pack multiple R packages, simply feed multiple file collections as inputs:
 
@@
 
-The packages are always packed in the specified order and are always
+The R packages are always packed in the specified order and are always
 unpacked and installed in the same order. Therefore, make sure to
@@
 check out the vignette
-[Generate File Specifications](https://merck.github.io/pkglite/articles/filespec.html)
+[generate file specifications](https://merck.github.io/pkglite/articles/filespec.html)
 and
-[Curate File Collections](https://merck.github.io/pkglite/articles/filecollection.html).
+[curate file collections](https://merck.github.io/pkglite/articles/filecollection.html).
 
@@
 
-You should always verify if the file only contains ASCII characters:
+You should always verify if the text file only contains ASCII characters:
 
@@
 ```{r, eval = FALSE}
-pkglite::unpack("r0pkgs.txt", output = "/tmp/", install = TRUE)
+unpack("r0pkgs.txt", output = "/tmp/", install = TRUE)
 ```
@@
 
-You can then proceed to save the file `r0pkgs.txt` to the folder
-`esub/analysis/adam/programs` (or specify the output text file
-path above directly).
+You can then proceed to move the file `r0pkgs.txt` to the folder
+`ectddemo/m5/datasets/ectddemo/analysis/adam/programs`,
+or specify the output text file path above directly.
 
-## Prepare Analysis Programs for Submission
+## Prepare analysis programs for submission
 
@@
 ```{r, eval=FALSE}
+input_path <- "esubdemo/vignettes/"
+output_path <- "ectddemo/m5/datasets/ectddemo/analysis/adam/programs/"
+
+convert_rmd <- function(filename, input_dir, output_dir) {
   knitr::purl(
-  "tlf-01-disposition.Rmd",
-  output = "esub/analysis/adam/programs/tlf-01-disposition.txt"
+    file.path(input_dir, paste0(filename, ".Rmd")),
+    output = file.path(output_dir, paste0(filename, ".txt"))
   )
+}
+
+"tlf-01-disposition" %>% convert_rmd(input_path, output_path)
+"tlf-02-population" %>% convert_rmd(input_path, output_path)
+"tlf-03-baseline" %>% convert_rmd(input_path, output_path)
+"tlf-04-efficacy" %>% convert_rmd(input_path, output_path)
+"tlf-05-ae-summary" %>% convert_rmd(input_path, output_path)
+"tlf-06-ae-spec" %>% convert_rmd(input_path, output_path)
 ```
-
-Repeat this for all the R Markdown files.
 
@@
 
+To automate this process:
+
+```{r, eval=FALSE}
+header <- readLines(textConnection("# Note to Reviewer
+# To rerun the code below, please refer to the ADRG appendix.
+# After the required packages are installed,
+# the path variable needs to be defined by using the example code below.
+#
+# path = list(adam = \"/path/to/esub/analysis/adam/datasets\") # Modify to use actual location
+# path$outtable = path$outgraph = \".\" # Outputs saved to the current folder"))
+
+append_header <- function(filename, output_dir, header) {
+  file <- file.path(output_dir, paste0(filename, ".txt"))
+  x <- readLines(file)
+  y <- c(header, "", x)
+  writeLines(y, con = file)
+  invisible(file)
+}
+
+"tlf-01-disposition" %>% append_header(output_path, header)
+"tlf-02-population" %>% append_header(output_path, header)
+"tlf-03-baseline" %>% append_header(output_path, header)
+"tlf-04-efficacy" %>% append_header(output_path, header)
+"tlf-05-ae-summary" %>% append_header(output_path, header)
+"tlf-06-ae-spec" %>% append_header(output_path, header)
+```
+
 ## Update ADRG
@@
 
-Specifically, we need to update two sections.
-The first is the "Macro Programs" section, where
-R and R package versions with a snapshot date are provided.
+Specifically, we need to update two sections in ADRG.
 
-For example:
+The first section is "Macro Programs", where R and R package versions
+with a snapshot date are provided. For example:
 
@@
 The recommended steps to unpack these R functions for analysis output
-programs are described in Appendix.
+programs are described in the Appendix.
 
@@
 
+An example ADRG following this template can be found in
+[ectddemo](https://github.com/elong0527/ectddemo/blob/master/m5/datasets/ectddemo/analysis/adam/datasets/adrg.pdf).
+
 ## Update ARM
@@
 
-- Section 2: indicate the **Programming Language**
+- Section 2: indicate the **Programming Language**;
 - Section 3: document the details of the R programs listed in section 3.

### 2021-10-27 submission-environment.Rmd
use sentence case instead of title case for headings; run styler on Rmds
add adrg template for cleanslate

@@
-# Running Environment {#running-environment}
+# Running environment {#running-environment}
 
@@
 
-## Practical Considerations
+## Practical considerations
 
@@
 
-## Create Canonical Environments
+## Create canonical environments
 
@@
 
-## Create Tailored Environments
+## Create tailored environments
 
@@
 version of R, and configuring the package repo location.
+For example:
 
-## RStudio Addin
+```markdown
+Appendix: Instructions to Execute Analysis Program in R
+
+1. Setup R environment
+
+Open the existing R, install the required packages by running the code below.
+
+install.packages("cleanslate")
+
+Create a temporary working directory, for example, "C:\tempwork".
+Copy all submitted R programs into the temporary folder.
+In the same R session, run the code below to create a project
+with a portable R environment.
+
+cleanslate::use_cleanslate(
+  "C:/temp/",
+  r_version = "4.1.1",
+  from = "https://cran.r-project.org/",
+  repo = "https://mran.microsoft.com/snapshot/2021-08-06/"
+)
+
+2. Open the project
+
+Go to the working directory created above, double click the .Rproj file
+to open the project in RStudio. Follow the instructions to select the
+project-specific R version then restart RStudio. If successful,
+the R version and package repo should be printed as defined above.
+
+3. Install open-source R packages
+
+In the new R session, install the required packages by running the code below.
+
+install.packages(c("pkglite", "publicpkg1", "publicpkg2"))
+
+4. Install proprietary R packages
+
+All internal R packages are packed in the file r0pkgs.txt. In the same R session,
+restore the package structures and install them by running the code below.
+Adjust the output path as needed to use a writable local directory.
+
+pkglite::unpack("r0pkgs.txt", output = ".", install = TRUE)
+
+5. Update path to dataset and TLFs
+
+INPUT path: to rerun the analysis programs, define the path variable
+
+- Path for ADaM data: path$adam
+
+OUTPUT path: to save the analysis results, define the path variable
+
+- Path for output TLFs: path$output
+
+All these paths need to be defined before executing the analysis program. For example:
+
+path = list(adam = "/path/to/esub/analysis/adam/datasets/") # Modify to use actual location
+path$outtable = path$outgraph = "." # Outputs saved to the current folder
+
+6. Execute analysis program
+
+To reproduce the analysis results, rerun the following programs:
+
+- tlf-01-disposition.txt
+- tlf-02-population.txt
+- tlf-03-baseline.txt
+- tlf-04-efficacy.txt
+- tlf-05-ae-summary.txt
+- tlf-06-ae-spec.txt
+```
+
+## RStudio addin
 

### 2021-10-27 slides/render.R
render slides into html

@@ -1,0 +1,1 @@
+rmarkdown::render("slides/workshop-slides.Rmd")

### 2021-10-28 index.Rmd
copy slides

@@
 file.copy("static/CNAME", "_book", overwrite = TRUE)
+fs::dir_copy("slides/", "_book/slides", overwrite = TRUE)
 ```

### 2021-10-28 slides/workshop-slides.Rmd
Deploying to gh-pages from @ elong0527/r4csr@f80ebc7f5c0d9a903568bd8f037d75336c7c182c 🚀

@@
 
-```{r}
+```{r, message=FALSE}
 library(dplyr)
@@
 library(r2rtf)
-```
-
-```{r, eval = FALSE, include = FALSE}
-rmarkdown::render("workshop-slides.Rmd")
 ```
@@
 
-```{r, out.width="80%", echo = FALSE}
-knitr::include_graphics("function_summary.PNG")
+```{r, out.width="80%", echo=FALSE, fig.align="center"}
+knitr::include_graphics("function-summary.png")
 ```

### 2021-11-01 project-folder.Rmd
project-folder.Rmd edited online with Bitbucket
use sentence case instead of title case for headings; run styler on Rmds
fix grammar, trailing spaces, file organization

--- 2021-10-12 project-folder.Rmd
@@
-# (PART) Clinical Trial Project {-}
-
-# Overview
-
-In a late-stage clinical trial, the number of A&R deliverables can easily be in the hundreds. 
-For an organization, it is common to have multiple ongoing clinical trials in a clinical program.
-
-To deliver the A&R results of a clinical trial project, it is teamwork that typically requires collaborations from both statisticians and programmers. 
-In this part, let's consider how to organize a clinical trial project as an A&R leader.
-
-In Chapter \@ref(folder), we will discuss how to organize source code, documents, deliverables in an A&R clinical project. 
-We recommend using the R package folder structure.
-
-In Chapter \@ref(manage), we will discuss a process or system development lifecycle to manage the A&R of a clinical project. 
-We recommend following an agile management approach to define, develop, validate and deliver work.
-
-# Project Folder {#folder}
+# Project folder {#folder}
 
@@
 - Automation: automatically check the integration of a project.
-- Compliance: reduce compliance issue
+- Compliance: reduce compliance issues.
 
@@
 [`esubdemo`](https://github.com/elong0527/esubdemo.git).
-You can [clone](https://happygitwithr.com/rstudio-git-github.html#clone-the-new-github-repository-to-your-computer-via-rstudio) the project using RStudio.
+You can
+[clone](https://happygitwithr.com/rstudio-git-github.html#clone-the-new-github-repository-to-your-computer-via-rstudio)
+the project using RStudio.
 
@@
 That is the [R package folder structure](https://github.com/rstudio/cheatsheets/raw/master/package-development.pdf).
-Every R package developer is required to follow the same convention to organize their R functions before the R package can be disseminated through the Comprehensive R Archive Network (CRAN). 
+Every R package developer is required to follow the same convention to organize
+their R functions before the R package can be disseminated through the Comprehensive R Archive Network (CRAN).
 As a user, you can easily install and use those R packages after installing from CRAN.
-There are many good resources to guide developers on R package development.
-For example, Hadley Wickham's [R Packages book](https://r-pkgs.org/).
+There are many good resources to guide developers on R package development,
+for example, the [R Packages book](https://r-pkgs.org/) by Hadley Wickham.
 
@@
 ensures project teams organize the A&R work consistently across multiple projects.
-Consistent folder structure also reduces communication costs between study team members and enhances the transparency of projects.
+Consistent folder structure also reduces communication costs
+between study team members and enhances the transparency of projects.
 
-In this book, we refer to an R package as **project-specific R package** 
+In this book, we refer to an R package as a **project-specific R package**
 if the purpose of an R package is to organize analysis scripts for a clinical project.
 
-We refer to an R package as a **standard R package** if the purpose of an R package is to share commonly used R functions to be hosted in a code repository such as CRAN.
+We refer to an R package as a **standard R package** if the purpose of an R package is
+to share commonly used R functions to be hosted in a code repository such as CRAN.
 
@@
 - `R/`: Project-specific R functions.
-- `vignettes/`: Analysis scripts using Rmarkdown.
+- `vignettes/`: Analysis scripts using R Markdown.
 - `man/`: Manual of project-specific R functions.
 
-A general discussion of the R package folder structure can be found in Chapter 4 of the R package book (@wickham2015r).
+A general discussion of the R package folder structure can be found
+in Chapter 4 of the R Packages book (@wickham2015r).
 
@@
 
-> Undeer the `vignettes/` folder, there are `adam/` folder and `tlf` folder.
+> Under the `vignettes/` folder, there are `adam/` folder and `tlf` folder.
 > The `adam/` folder contains ADaM datasets.
@@
 
-While creating those analysis scripts, we also defined a few helper functions (e.g. `fmt_num` and `count_by`). 
+While creating those analysis scripts, we also defined a few helper functions (e.g., `fmt_num` and `count_by`).
 Those functions are saved in the `R/` folder.
@@
 ```
-R
+R/
 ├── count_by.R
@@
 With a formal project, it is also important to provide proper documentation for those help functions.
-We used Roxygen2 to document functions. 
+We used `roxygen2` to document functions.
 For example, the header below defines each variable in `fmt_est`.
-More details can be found in [Chapter 10 of the R package book](https://r-pkgs.org/man.html)
+More details can be found in [Chapter 10 of the R Packages book](https://r-pkgs.org/man.html).
 
@@
 
-Those `Roxygen2` documentation will be transferred to 
-standard R documentation `.Rd` files saved in the `man/` folder. 
-This step is automatically handled by `devtools::document()` function.
+The `roxygen2` documentation will be converted into
+standard R documentation format, saved as `.Rd` files in the `man/` folder.
+This step is automatically handled by `devtools::document()`.
 
@@
 
-The `man` folder is used to save documentation automatically generated by `roxygen2`.
-A typical workflow is to add `roxygen2` documentation at the header of each function in the `R/` folder. 
+The `man/` folder is used to save documentation automatically generated by `roxygen2`.
+A typical workflow is to add `roxygen2` documentation before each function in the `R/` folder.
 Then `devtools::documents()` is used to generate all the documentation files in the `man/` folder.
-More details can be found in [Chapter 10 of the R packages book](https://r-pkgs.org/man.html).
+More details can be found in [Chapter 10 of the R Packages book](https://r-pkgs.org/man.html).
 
@@
 Reproducibility of analysis is one of the most important aspects of regulatory deliverables.
-To ensure a successful reproduce, we need a controlled R environment, including the control of R version and control of R package version.  
-By using the R package folder structure and proper tools (e.g. `renv`, `packrat`), 
+To ensure a successful reproduce, we need a controlled R environment,
+including the control of the R version and the R package versions.
+By using the R package folder structure and proper tools (e.g., `renv`, `packrat`),
 we illustrate how to achieve reproducibility for R and R package versions.
@@
 
-First, we introduce the control of R version. 
-In the `esubdemo` project, a reproducible environment is created when you open the 
-`esubdemo.Rproj` from RStudio. 
+First, we introduce the control of the R version.
+In the `esubdemo` project, a reproducible environment is created when you open
+the `esubdemo.Rproj` from RStudio.
 When we open the `esubdemo` project, RStudio will execute the R code in `.Rprofile` automatically.
 So we can use `.Rprofile` to set up a reproducible environment.
-More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html). 
-After we open the `esubdemo` project, the code in `.Rprofile` will automatically check the current R version is the same as we defined in `.Rprofile`.
+More details can be found in <https://rstats.wtf/r-startup.html>.
+After we open the `esubdemo` project, the code in `.Rprofile` will
+automatically check the current R version is the same as we defined in `.Rprofile`.
+
 ```{r, eval = FALSE}
-# set up project R version
+# Set project R version
 R_version <- "4.1.1"
 ```
+
 If the R version mismatch, an error message is displayed as below.
@@
 
-> `.Rprofile` is only for project-specific R package. 
+> `.Rprofile` is only for project-specific R packages.
 > A standard R package should not use `.Rprofile`.
 
-Next, we introduce the control of R package version, which is controlled in two layers. 
+Next, we introduce the control of the R package version, which is controlled in two layers.
 Firstly, we define a snapshot date in `.Rprofile`.
@@
 The R code above in the `.Rprofile` initiates the `renv` running environment.
-As a user, you can use `renv::init()`, `renv::snapshot()` and `renv::restore()` 
+As a user, you can use `renv::init()`, `renv::snapshot()`, and `renv::restore()`
 to initialize, save and restore R packages used for the current analysis project.
@@
 
-> `renv.lock` file and `renv/` folder are only for project-specific R package.
+> The `renv.lock` file and `renv/` folder are only for project-specific R package.
 > A standard R package should not use `renv`.
@@
 
-If the project is initiated properly, you shall be able to see similar messages to inform how we control R package versions.
+If the project is initiated properly,
+you shall be able to see similar messages to inform how we control R package versions.
 
@@
 Once R packages have been properly installed,
-the system will use the R packages located in the search path defined based on the order of `.libPaths()`.
+the system will use the R packages located in the search path
+defined based on the order of `.libPaths()`.
 The startup message also provided the R package search path.
@@
 ```
-Below R package path are searching in order to find installed R pacakges in this R session:
+Below R package path are searching in order to find installed R packages in this R session:
     "/home/zhanyilo/github-repo/esubdemo/renv/library/R-4.1/x86_64-pc-linux-gnu"
@@
 > by using the same operating system, R version, and R package versions for an A&R project.
-> More details can be found in [https://environments.rstudio.com/](https://environments.rstudio.com/).
+> More details can be found at <https://environments.rstudio.com/>.
 
 > A container solution like [Docker](https://github.com/rstudio/r-docker) could further enhance
-> the reproducibility across an organization at the operating system level, but beyond the scope of this book.
+> the reproducibility across an organization at the operating system level but beyond the scope of this book.
 
@@
 - [`devtools`](https://devtools.r-lib.org/): make package development easier.
-  + A good overview can be found in [Chapter 2 of the R packages](https://r-pkgs.org/whole-game.html) book  
+  + A good overview can be found in [Chapter 2 of the R Packages book](https://r-pkgs.org/whole-game.html).
   + `devttols::load_all()`: load all functions in `R/` folder and running environment.
@@
 As the consistent folder is defined, it also becomes easier to create specific tools
-that fit for the analysis and reporting purpose. Below are few potential tools that can be helpful:
+that fit the analysis and reporting purpose. Below are a few potential tools that can be helpful:
 
@@
 For an R package, there are already criteria to ensure R package integrity.
-More details can be found in [Chapter 19 of the R package book](https://r-pkgs.org/r-cmd-check.html).
+More details can be found in [Chapter 19 of the R Packages book](https://r-pkgs.org/r-cmd-check.html).

### 2021-11-01 project-management.Rmd
project-management.Rmd edited online with Bitbucket
use sentence case instead of title case for headings; run styler on Rmds
fix grammar, trailing spaces, file organization

@@
-# Project Management {#manage}
+# Project management {#manage}
 
-## Setting up for Success
+## Setting up for success
 
@@
 Therefore, the conventional wisdom and tricks for managing a successful project are also applicable here.
-At the same time, clinical projects also have unique traits, such as high standards for planning, development, validation, and delivery under strict time constraints.
+At the same time, clinical projects also have unique traits,
+such as high standards for planning, development, validation, and delivery
+under strict time constraints.
 
 Although many factors could decide if a project could execute efficiently,
-we believe a few aspects are critical for the long-term success,
+we believe a few aspects are critical for long-term success,
 especially when managing clinical data analysis projects at scale.
 
-### Work as a Team
+### Work as a team
 
@@
 As always, setting a clear goal and following a system development lifecycle
-(SDLC) are essential.
+(SDLC) is essential.
 
-### Design Clean Code Architecture
+### Design clean code architecture
 
@@
 
-### Set Capability Boundaries
+### Set capability boundaries
 
@@
 
-### Contribute to the Community
+### Contribute to the community
 
@@
 
-SDLC for the A&R deliverables can be defined in four stages:
+SDLC for the A&R deliverables can be defined in four stages.
 
-- planning: a planning stage to define the scope of a project.  
-- development: a development stage to implement target deliverables. 
-- validation: a validation stage to verify target deliverables. 
-- operation: an operation stage to deliver work to stakeholders.
+- Planning: a planning stage to define the scope of a project.
+- Development: a development stage to implement target deliverables.
+- Validation: a validation stage to verify target deliverables.
+- Operation: an operation stage to deliver work to stakeholders.
 
 Importantly, we should not consider SDLC as a linear process.
-For example, if the study team identifies new requirement in a development or validation stage, 
+For example, if the study team identifies a new requirement in a development or validation stage,
 the team should return to the planning stage to discuss and align the scope.
 An [agile project management](https://www.atlassian.com/agile/project-management) approach is suitable
-and recommended to an A&R clinical development project.
+and recommended for an A&R clinical development project.
 The goal is to embrace an iterative approach that continuously improves target deliverables based on
@@
 
-The project lead should also set up a project folder as discussed in Chapter \@ref(folder).
+The project lead should also set up a project folder, as discussed in Chapter \@ref(folder).
 The project initiation can be simplified by creating an
@@
 
-After a project is initiated, the study team start to develop TLFs based on 
+After a project is initiated, the study team starts to develop TLFs based on
 pre-defined mock-up tables assigned to each team member.
 
-The analysis code and relevant description can be saved in R markdown files 
+The analysis code and relevant description can be saved in R Markdown files
 in the `vignettes` folder.
 
-The use of R markdown allows developers to assemble narrative text, code, and its
+The use of R Markdown allows developers to assemble narrative text, code, and its
 comments in one place to simplify documentation.
 It would be helpful to create a template and define a name convention for all TLFs deliverables.
-For example, we can use `tlf_` prefix in the filename to indicate that the R Markdown is to deliver TLFs. 
+For example, we can use the `tlf_` prefix in the filename to indicate that the R Markdown file is for delivering TLFs.
 Multiple TLFs with similar designs can be included in one R Markdown file.
 
-For example, in `esubdemo` project, we have 6 R Markdown files to create TLFs.
+For example, in the `esubdemo` project, we have 6 R Markdown files to create TLFs.
 
-If there is any project-specific R functions that need to be developed, 
+If there are any project-specific R functions that need to be developed,
 the R functions can be placed in the `R/` folder as discussed in Section \@ref(consistency).
@@
 
-Validation is an crucial stage to ensure the deliverables are accurate and consistent. 
+Validation is a crucial stage to ensure the deliverables are accurate and consistent.
 After the development stage is completed, the project team needs to validate the deliverables, including R Markdown
@@
 To validate the content of a table, we can save the last datasets ready for table generation as a `.Rdata` file.
-A validator can reproduce the TLF and compares it with the original result saved in `.Rdata` file. 
+A validator can reproduce the TLF and compare it with the original result saved in the `.Rdata` file.
 A test is passed when the results match.
-Customers can directly review the formatting of the TLFs by
-customers by comparing them with the mock-up.
+Customers can directly review the formatting of the TLFs by comparing them with the mock-up.
 
-To validate a figure, we can use [snapshot testing](https://testthat.r-lib.org/articles/snapshotting.html) strategy.
+To validate a figure, we can use the [snapshot testing](https://testthat.r-lib.org/articles/snapshotting.html) strategy.
 
@@
 After completing the validation, the validator updates the status in a validation tracker.
-The project lead reviews the tracking sheet to make sure all required activities in the SDLC are completed and
-the tracking sheet has been filled correctly. 
+The project lead reviews the tracking sheet to make sure all required activities in the SDLC are completed,
+and the tracking sheet has been filled correctly.
 The deliverables are ready for customer review after all the validation steps are completed.
@@
 
-After completion of development and required validation of all A&R deliverables, project lead runs compliance check for a project-specific R package similar to other R packages. 
+After completion of development and required validation of all A&R deliverables,
+the project lead runs compliance checks for a project-specific R package similar to other R packages.
 `devtools::check()` is a convenient way to run compliance checks or `R CMD check`.
-`R CMD check` is an automated check of the contents in the R package for frequently encountered issues, before submission to CRAN. 
+`R CMD check` is an automated check of the contents in the R package
+for frequently encountered issues before submission to CRAN.
 Since the project-specific R package is not submitted to CRAN, some checks can be customized and skipped in `devtools::check()`.
-The project lead should work with the study team to ensure all reported errors, warnings, and notes by `devtools::check()` are fixed.
+The project lead should work with the study team to ensure
+all reported errors, warnings, and notes by `devtools::check()` are fixed.
 
@@
 The `pkgdown` website is a convenient way to run all analyses in batch and
-integrate outputs in a website, which comprehensively covers project-specific R functions, TLF generation programs, outputs and validation tracking information, etc... 
+integrate outputs in a website, which comprehensively covers
+project-specific R functions, TLF generation programs, outputs and validation tracking information, etc.
 For example, in the `esubdemo` project, we created the `pkgdown` website at <https://elong0527.github.io/esubdemo/>.
@@
 Many of the tasks in SDLC can be completed automatically.
-An organization can leverage CI/CD workflow to automatically enable those tasks 
+An organization can leverage CI/CD workflow to automatically enable those tasks,
 such as running testing cases and creating a `pkgdown` website.
-For example, in `esubdemo` project, we setup [GitHub Actions](https://github.com/elong0527/esubdemo/actions)
-for these work.
-It can be setup by using `usethis::use_github_action()` function.
+For example, in the `esubdemo` project, we set up
+[GitHub Actions](https://github.com/elong0527/esubdemo/actions)
+for it. This can be done by using `usethis::use_github_action()`.

### 2021-11-01 submission-package.Rmd
fix grammar, trailing spaces, file organization

@@
 (if any), and make them part of the submission package.
-Lastly, we will provide reusable templates for updating ADaM Reviewer's Guide (ADRG) and Analysis Results Metadata (ARM),
+Lastly, we will provide reusable templates for updating
+ADaM Reviewer's Guide (ADRG) and Analysis Results Metadata (ARM)
 so that the reviewers receive proper instructions to reproduce the
@@
 
-The demo project (R package) we will prepare for submission is called [`esubdemo`](https://github.com/elong0527/esubdemo) available on GitHub.
+The demo project (R package) we will prepare for submission is called
+[`esubdemo`](https://github.com/elong0527/esubdemo), available on GitHub.
 You can download or clone it:
@@
 
-R package dependencies is another major factor to consider
+R package dependency is another major factor to consider
 before preparing your proprietary R package for submission.
@@
 
-To prepare R packages for submission, one needs to pack the packages into text files, and then verify if the file only contains ASSCII characters.
+To prepare R packages for submission, one needs to pack the packages into text files, and then verify if the file only contains ASCII characters.
 With packed packages, one can unpack and install them from the text file, too.
@@
 Let's pack the `esubdemo` package into a text file.
-Assume the source package path is `esubdemo/`,
-you should be able to pack the package with a single pipe:
+Assume the source package path is `esubdemo/`.
+You should be able to pack the package with a single pipe:
 
@@
 folders and files defined by a list of **file specifications**,
-and generate a **file collection** object.
+and generates a **file collection** object.
 This file collection contains the metadata required to properly convert

### 2021-11-01 tlf-baseline.Rmd
fix grammar, trailing spaces, file organization

@@
 
-There is many R packages that can efficiently summarize baseline information.
+There are many R packages that can efficiently summarize baseline information.
 The [`table1`](https://github.com/benjaminrich/table1) R package is one of them.
@@
 
-As in previous chapters, we first read `adsl` dataset that contain all 
-required information for baseline characteristics table.
+As in previous chapters, we first read the `adsl` dataset that contains all
+the required information for the baseline characteristics table.
 
@@
 - Step 2: Use `table1::table1()` to get the baseline characteristics table.
-- Step 3: Transfer the output in Step 2 into a data frame that only contains ASCII character.
-- Step 4: Define the format of the rtf table by using R package `r2rtf`.
+- Step 3: Transfer the output in Step 2 into a data frame that only contains ASCII characters.
+- Step 4: Define the format of the RTF table by using the R package `r2rtf`.

### 2021-11-01 tlf-disposition.Rmd
fix grammar, trailing spaces, file organization

@@
 The dataset is provided in `sas7bdat` format that is a SAS data format currently used in many clinical trial analysis and reporting.
-The `haven` package is able to read in the dataset, while maintaining its attributes (e.g. variable labels).
+The `haven` package is able to read in the dataset, while maintaining its attributes (e.g., variable labels).
 
@@
 In the code below, we calculate the number and percentage of participants who discontinued the study for different reasons.
-For display purposes, `paste0("    ", row)` is used to add leading spaces to mimick identation in the final report.
+For display purposes, `paste0("    ", row)` is used to add leading spaces to mimick indentation in the final report.
 
@@
 
-- Step 1: Read subject level data into R, i.e. `adsl`.
+- Step 1: Read subject level data into R, i.e., `adsl`.
 - Step 2: Count participants in the analysis population and name the dataset `n_rand`.
@@
 - Step 5: Count the number and percentage of participants that completed the study, and name the dataset `n_complete`.
-- Step 6: Bind `n_rand`, `n_disc`, `n_reason` and `n_complete` by row. 
+- Step 6: Bind `n_rand`, `n_disc`, `n_reason`, and `n_complete` by row.
 - Step 7: Write the final table to RTF

### 2021-11-01 tlf-efficacy.Rmd
fix grammar, trailing spaces, file organization

@@
 
-For efficacy analysis, we only analyze change from baseline glucose data at Week 24.
+For efficacy analysis, we only analyze the change from baseline glucose data at week 24.
 
@@
 
-To prepare analysis dataset, 
+To prepare the analysis dataset,
 we need both `adsl` and `adlbc` datasets for this analysis.
@@
 ```
+
 ## Helper functions
 
-To prepare report dataframe, we created a few helper functions 
-by using the `fmt_num` function defined in Chapter \@ref(population).
+To prepare the report data frame, we created a few helper functions
+by using the `fmt_num()` function defined in Chapter \@ref(population).
 
@@
 
-We start to analyze the imputed data using ANCOVA model with treatment and baseline glucose as covariates.
+We start to analyze the imputed data using the ANCOVA model with treatment and baseline glucose as covariates.
 
@@
 Then we use `r2rtf` to prepare the table format for `t1`.
-We also highlight how to handle special character in this example.
+We also highlight how to handle special characters in this example.
 
@@
 For example, `{^a}` provide a superscript `a` for footnote notation.
-`r2rtf` also support most LaTeX characters. 
-Examples can be found in
-[`r2rtf` Get Started Page](https://merck.github.io/r2rtf/articles/r2rtf.html#special-character). 
-The `text_convert` argument in `r2rtf_xxx` functions control whether convert special characters.
+`r2rtf` also supports most LaTeX characters.
+Examples can be found on the
+[`r2rtf` get started page](https://merck.github.io/r2rtf/articles/r2rtf.html#special-character).
+The `text_convert` argument in `r2rtf_*()` functions controls whether convert special characters.
 
@@
 - Step 3: Impute the missing values. In our example, we name the `ana` dataset after imputation as `ana_locf`.
-- Step 4: Summarize `gluc`, i.e., calculate the mean and standard derivation, and then format it into a rtf table.
-- Step 5: Summarize `ana_locf`, i.e., calculate the pairwise comparison by ANCOVA model, and then format it into a rtf table.
+- Step 4: Summarize `gluc`, i.e., calculate the mean and standard derivation, and then format it into an RTF table.
+- Step 5: Summarize `ana_locf`, i.e., calculate the pairwise comparison by ANCOVA model, and then format it into an RTF table.
 - Step 6: Rowly bind the output in Step 4 and Step 5.

### 2021-11-01 tlf-overview.Rmd
fix grammar, trailing spaces, file organization

@@
 
-# Overview
+# Overview {#tlf-overview}
 
@@
 
-In clinical trials, a critical step is
-submission to regulatory agencies. 
+In clinical trials, a critical step is submission to regulatory agencies.
 [Electronic Common Technical Document (eCTD)](https://en.wikipedia.org/wiki/Electronic_common_technical_document)
 has become a worldwide regulatory submission standard format.
-For example, [the United States Food and Drug Administration (US FDA) required new drug applications, biologics license applications must be submitted using eCTD format](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd).
-Clinical Data Interchange Standards Consortium (CDISC) provided a
+For example, the United States Food and Drug Administration (US FDA) required
+new drug applications and biologics license applications
+[must be submitted using the eCTD format](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd).
+The Clinical Data Interchange Standards Consortium (CDISC) provided a
 [pilot project following ICH E3 guidance](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse).
@@
 [ICH E3 guidance](https://www.ich.org/page/efficacy-guidelines) provides
-compilation of structure and content of clinical study reports.
+a compilation of the structure and content of clinical study reports.
 
 A typical CSR contains full details on the methods and results of an individual clinical study.
-In support of the statistical anlaysis, a large number of tables, listings and figures are
+In support of the statistical analysis, a large number of tables, listings, and figures are
 incorporated into the main text and appendices.
-In CDISC pilot project, an 
+In the CDISC pilot project, an
 [example CSR](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/53-clin-stud-rep/535-rep-effic-safety-stud/5351-stud-rep-contr/cdiscpilot01/cdiscpilot01.pdf)
 is also provided. If you are interested in more examples of clinical study reports,
-the European Medicines Agency (EMA) publishes clinical data including 
+the European Medicines Agency (EMA) publishes clinical data, including
 clinical study reports submitted by pharmaceutical companies
-on [EMA clinical data website](https://clinicaldata.ema.europa.eu/web/cdp/home).
+on the [EMA clinical data website](https://clinicaldata.ema.europa.eu/web/cdp/home).
 
-Building CSRs is a teamwork between clinicians, medical writers, statisticians, statistical programmers
+Building CSRs is teamwork between clinicians, medical writers, statisticians, statistical programmers,
 and other relevant specialists such as experts on biomarkers.
@@
 In this chapter, our focus is to illustrate how to create tables, listings, and figures (TLFs) in RTF format
-that is commonly used in a CSR. The examples are in compliance of 
-[FDA's Portable Document Format (PDF) Specifications](https://www.fda.gov/media/76797/download)
+that is commonly used in a CSR. The examples are in compliance with the
+[FDA's Portable Document Format (PDF) Specifications](https://www.fda.gov/media/76797/download).
 
-> FDA's PDF specification is a general reference, each organization can define 
-> more specific TLF format requirement that can be different from the examples in this book.
+> FDA's PDF specification is a general reference. Each organization can define
+> more specific TLF format requirements that can be different from the examples in this book.
 
@@
 [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf)
-on structure and content of clinical study reports.
+on the structure and content of clinical study reports.
 
@@
 - Section 12: Safety evaluation
-- Section 14: Tables, listings and figures referrals but not included in the text
+- Section 14: Tables, listings, and figures referrals but not included in the text
 - Section 16: Appendices
@@
 We used publicly available CDISC pilot
-[study data located at CDISC Bitbucket repository](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).
+[study data located in the CDISC Bitbucket repository](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).
 
-For simplicity, we have downloaded all these datasets into `adam_data` folder of 
-this project and transfer them from `.xpt` format to `sas7bdat` format.
+For simplicity, we have downloaded all these datasets into the `adam_data` folder of
+this project and converted them from the `.xpt` format to the `.sas7bdat` format.
 
@@
 
-- [tidyverse](https://www.tidyverse.org/): prepare reporting ready datasets. 
+- [tidyverse](https://www.tidyverse.org/): prepare datasets ready for reporting.
 - [r2rtf](https://merck.github.io/r2rtf/): create RTF outputs
@@
 
-RStudio provided outstanding [cheatsheets](https://www.rstudio.com/resources/cheatsheets/), 
+RStudio provided outstanding [cheatsheets](https://www.rstudio.com/resources/cheatsheets/)
 and [tutorials](https://github.com/rstudio-education/remaster-the-tidyverse) for `tidyverse`.
@@
 There are also books to introduce tidyverse.
-We assume reader to have experience in using `tidyverse` in this book.
+We assume the reader to have experience in using `tidyverse` in this book.
 
@@
 - provide simple "verb" functions that correspond to each component of a table,
-  to help you translate a data frame to a table in RTF file.
-- enable pipes (`%>%`). 
-- only focus on the **table format**. 
-  Data manipulation and analysis shall be handled by other R packages. (e.g. `tidyverse`)
+  to help you translate a data frame to a table in an RTF file;
+- enable pipes (`%>%`);
+- focus on the **table format** only.
+  Data manipulation and analysis shall be handled by other R packages (e.g., `tidyverse`).
 
-Before creating an RTF table we need to:
+Before creating an RTF table, we need to
 
-- Figure out table layout. 
-- Split the layout into small tasks in the form of a computer program.
-- Execute the program.
+- figure out the table layout;
+- split the layout into small tasks in the form of a computer program;
+- execute the program.
 
-We provided a brief introduction of `r2rtf` and show how to transfer 
+We provided a brief introduction of `r2rtf` and showed how to transfer
 data frames into Table, Listing, and Figures (TLFs).
 
-Other extended examples and features are covered in the 
+Other extended examples and features are covered on the
 [`r2rtf` package website](https://merck.github.io/r2rtf/articles/index.html).
@@
 To explore the basic RTF generation verbs in `r2rtf`,
-we'll use the dataset `r2rtf_adae` saved in the `r2rtf` package. 
+we will use the dataset `r2rtf_adae` saved in the `r2rtf` package.
 This dataset contains adverse events (AE) information from a clinical trial.
 
+We will begin by loading the packages:
+
+```{r}
+library(dplyr) # Manipulate data
+library(tidyr) # Manipulate data
+library(r2rtf) # Reporting in RTF format
+```
+
 Below is the meaning of relevant variables.
-More information can be found in the help page of the dataset (`?r2rtf_adae`)
+More information can be found on the help page of the dataset (`?r2rtf_adae`)
 
-In this example we consider three variables:
+In this example, we consider three variables:
 
@@
 
-- `rtf_page`: RTF page information
-- `rtf_title`: RTF title information
-- `rtf_colheader`: RTF column header information
-- `rtf_body`: RTF table body information
-- `rtf_footnote`: RTF footnote information
-- `rtf_source`: RTF data source information
+- `rtf_page()`: RTF page information
+- `rtf_title()`: RTF title information
+- `rtf_colheader()`: RTF column header information
+- `rtf_body()`: RTF table body information
+- `rtf_footnote()`: RTF footnote information
+- `rtf_source()`: RTF data source information
 
@@
 A full list of all functions can be found in the
-[r2rtf package reference](https://merck.github.io/r2rtf/reference/index.html)
+[r2rtf package function reference manual](https://merck.github.io/r2rtf/reference/index.html).
 
@@
 
-- `rtf_body` is used to define table body layout. 
-- `rtf_encode` transfers table layout information into RTF syntax.
-- `write_rtf` save RTF encoding into a file with file extension `.rtf`
+- `rtf_body()` is used to define table body layout.
+- `rtf_encode()` transfers table layout information into RTF syntax.
+- `write_rtf()` save RTF encoding into a file with file extension `.rtf`
 
@@
 this can be achieved by updating the `col_rel_width` argument
-in the `rtf_body` function.
+in the `rtf_body()` function.
 
@@
 Only the ratio of `col_rel_width` is used.
-Therefore it is equivalent to use `col_rel_width = c(6,4,4,4)` 
-or `col_rel_width = c(1.5,1,1,1)`.
+Therefore it is equivalent to use `col_rel_width = c(6, 4, 4, 4)`
+or `col_rel_width = c(1.5, 1, 1, 1)`.
 
@@
 In the previous example, we found the issue of a misaligned column header.
-We can fix the issue by using the `rtf_colheader` function.
+We can fix the issue by using the `rtf_colheader()` function.
 
-In `rtf_colheader`, `colheader` argument is used to provide the content of the column header. 
+In `rtf_colheader()`, the `colheader` argument is used to provide the content of the column header.
 We use `"|"` to separate the columns.
@@
 
-In `rtf_*` functions such as `rtf_body`, `rtf_footnote`, 
-`text_justification` argument is used to align text. 
+In `rtf_*()` functions such as `rtf_body()`, `rtf_footnote()`,
+the `text_justification` argument is used to align text.
 Default is `"c"` for center justification.
 To vary text justification by column, use character vector with length of vector equal to
-number of columns displayed (e.g., `c("c","l","r")`).
+number of columns displayed (e.g., `c("c", "l", "r")`).
 
@@
 
-In `rtf_*` functions such as `rtf_body`, `rtf_footnote`, etc.,
+In `rtf_*()` functions such as `rtf_body()`, `rtf_footnote()`, etc.,
 `border_left`, `border_right`, `border_top`, and `border_bottom` control cell borders.
@@
 If we want to remove the top border of `"Adverse Events"` in the header,
-we can change the default value `"single"` to `""` in the `border_top` argument as showed below.
+we can change the default value `"single"` to `""` in the `border_top` argument, as shown below.
 
@@
 
-In this example, we also demonstrate the possibility to add multiple column headers.
+In this example, we also demonstrate the possibility of adding multiple column headers.
 
@@
 
-In the `r2rtf` R package [get started](https://merck.github.io/r2rtf/articles/r2rtf.html) page
+In the `r2rtf` R package [get started](https://merck.github.io/r2rtf/articles/r2rtf.html) page,
 there are more examples to illustrate how to customize

### 2021-11-01 tlf-population.Rmd
fix grammar, trailing spaces, file organization

@@
 In this chapter, we illustrate how to create a summary table for
-analysis population in a study.
+the analysis population in a study.
 
@@
 
-Before we write analysis code, let's discuss the possibility to reuse R
+Before we write the analysis code, let's discuss the possibility of reusing R
 code by writing some toy helper functions.
@@
 
-In Chapter \@ref(disposition), there are few repeating steps to:
+In Chapter \@ref(disposition), there are a few repeating steps to:
 
@@
 To calculate the numbers and percentages of participants by groups, we
-provide a simple (but not robust) wrapper function, `count_by()` using
+provide a simple (but not robust) wrapper function, `count_by()`, using
 the `dplyr` and `tidyr` package.
@@
 - Step 2: Bind the counts/percentages of the ITT population, the
-  efficacy population and the safety population by row using the
+  efficacy population, and the safety population by row using the
   `count_by()` function.

### 2021-11-01 submission-environment.Rmd

@@
 
-In the previous chapter, we have generated instructions to manually
+In the previous chapter, we generated instructions to manually
 create the running environments for reproducing the A&R deliverables.
@@
 In this chapter, we will focus on automating the creation of
-the R environments with R code, to accelerate the dry run testing process,
-simplify the ADRG instructions, and make it easy to recreate
+the R environments with R code to accelerate the dry run testing process,
+simplifying the ADRG instructions, and making it easy to recreate
 different environment settings with reproducible analysis results.
@@
 The helper functions `version_*()` and `snapshot_*()` could assist you
-to determine specific versions of R and Rtools that are currently available,
+in determining specific versions of R and Rtools that are currently available,
 besides generating and verifying the snapshot repo links.
@@
 to open the project in RStudio. Follow the instructions to select the
-project-specific R version then restart RStudio. If successful,
+project-specific R version, then restart RStudio. If successful,
 the R version and package repo should be printed as defined above.

### 2021-11-01 tlf-ae-spec.Rmd

@@
 In this way, let's focus on the analysis script for two advanced
-features for table layout.
+features for a table layout.
 
--   group content: AE can be summarized in multiple nested layers. (e.g.
-    by system organ class (SOC, `AESOC`) and specific AE term
+- group content: AE can be summarized in multiple nested layers.
+  (e.g., by system organ class (SOC, `AESOC`) and specific AE term
   (`AEDECOD`))
@@
 We can also customize the text format by providing a matrix that has the
-same dimension as the input dataset (i.e. `tbl_ae_spec`). In the code
+same dimension as the input dataset (i.e., `tbl_ae_spec`). In the code
 below, we illustrate how to display **bold** text for group headers to

### 2021-11-01 submission-package.Rmd
correct adrg full name

@@
 Lastly, we will provide reusable templates for updating
-ADaM Reviewer's Guide (ADRG) and Analysis Results Metadata (ARM)
+Analysis Data Reviewer's Guide (ADRG) and Analysis Results Metadata (ARM)
 so that the reviewers receive proper instructions to reproduce the
@@
 - `define.xml`: created by Pinnacle 21.
-- ADRG (ADaM Reviewer's Guide)
+- ADRG (Analysis Data Reviewer's Guide)
   - "Macro Programs" section: provide R and R package versions with a snapshot date.

### 2021-11-03 tlf-baseline.Rmd
Minor typo fix

@@
 For simplicity, we only analyze `SEX`, `AGE` and, `RACE` in this example
-using the `table1` R package. More details of the `table` R package can
+using the `table1` R package. More details of the `table1` R package can
 be found in the package
@@
 The code below transfer the output into a dataframe
-that only contain ASCII character
+that only contains ASCII characters
 recommended by regulatory agencies.
-`tbl_base` is used as input for `r2rtf` to create final report.
+`tbl_base` is used as input for `r2rtf` to create the final report.
 

### 2021-11-03 tlf-ae-spec.Rmd
Minor syntax fix

@@
 features can be found in the
-[`r2rtf` package website](https://merck.github.io/r2rtf/articles/example-sublineby-pageby-groupby.html.
+[`r2rtf` package website](https://merck.github.io/r2rtf/articles/example-sublineby-pageby-groupby.html).
 

### 2021-11-05 tlf-efficacy.Rmd
general updates to improve readability

@@
 Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to summarize primary and secondary efficacy endpoints,
+primary and secondary efficacy endpoints need to summarized
 in Section 11.4, Efficacy Results and Tabulations of Individual Participant.
@@
 
-For efficacy analysis, we only analyze the change from baseline glucose data at week 24.
+For efficacy analysis, only the change from baseline glucose data at week 24 is analyzed.
 
@@
 
-To prepare the analysis dataset,
-we need both `adsl` and `adlbc` datasets for this analysis.
+To prepare the analysis, both `adsl` and `adlbc` datasets are required.
 
@@
 
-We first define the analysis dataset using
-efficacy population flag `EFFFL` and
-all records post baseline (`AVISITN > 1`) and on or before Week 24 (`AVISITN <= 24`).
+First, both the population and the data in scope are selected.
+The analysis is done on the efficacy population, identified by `EFFFL == "Y"`, and
+all records post baseline (`AVISITN >= 1`) and on or before Week 24 (`AVISITN <= 24`).
 Here the variable `AVISITN` is the numerical analysis visit.
@@
 
-We first summarize observed data at Baseline and Week 24
+First the observed data at Baseline and Week 24 is summarized:
 
@@
 
-We also summarize observed change from baseline glucose at Week 24.
+Also the observed change from baseline glucose at Week 24 is summarized:
 
@@
 
-We start to analyze the imputed data using the ANCOVA model with treatment and baseline glucose as covariates.
+The imputed data is analyzed using the ANCOVA model with treatment and baseline glucose as covariates.
 
@@
 
-Then we use the `emmeans` R package to obtain
+The `emmeans` R package is used to obtain
 within and between group least square mean (LS mean)
@@
 
-We combine `t11`, `t12` and `t13` to get the first part of the report data
+`t11`, `t12` and `t13` are combined to get the first part of the report data
 
@@
 
-Then we use `r2rtf` to prepare the table format for `t1`.
+Then `r2rtf` is used to prepare the table format for `t1`.
 We also highlight how to handle special characters in this example.
@@
 - Step 5: Summarize `ana_locf`, i.e., calculate the pairwise comparison by ANCOVA model, and then format it into an RTF table.
-- Step 6: Rowly bind the output in Step 4 and Step 5.
+- Step 6: Set the results from Step 4 and Step 5 and output.

### 2021-11-05 tlf-efficacy.Rmd
Apply suggestions from code review
Co-authored-by: yilong zhang <elong0527@gmail.com>

@@
 Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-primary and secondary efficacy endpoints need to summarized
+primary and secondary efficacy endpoints need to be summarized
 in Section 11.4, Efficacy Results and Tabulations of Individual Participant.

### 2021-11-06 project-folder.Rmd
Minor syntax / spelling fix

@@
 
-> Under the `vignettes/` folder, there are `adam/` folder and `tlf` folder.
-> The `adam/` folder contains ADaM datasets.
+> Under the `vignettes` folder, there are two folders: `adam` folder and `tlf` folder.
+> The `adam` folder contains ADaM datasets.
 > The `tlf` folder contains output TLFs in RTF format.
-> We put `adam/` and `tlf/` folders within the `vignettes` folder only for illustration purposes.
+> We put `adam` and `tlf` folders within the `vignettes` folder only for illustration purposes.
 > In an actual A&R report, you may have a different location to save your input and output.
@@
 A typical workflow is to add `roxygen2` documentation before each function in the `R/` folder.
-Then `devtools::documents()` is used to generate all the documentation files in the `man/` folder.
+Then `devtools::document()` is used to generate all the documentation files in the `man/` folder.
 More details can be found in [Chapter 10 of the R Packages book](https://r-pkgs.org/man.html).
@@
 Reproducibility of analysis is one of the most important aspects of regulatory deliverables.
-To ensure a successful reproduce, we need a controlled R environment,
+To ensure a successful reproduction, we need a controlled R environment,
 including the control of the R version and the R package versions.
@@
 
-If the R version mismatch, an error message is displayed as below.
+If there is a R version mismatch, an error message is displayed as below.
 
@@
   + A good overview can be found in [Chapter 2 of the R Packages book](https://r-pkgs.org/whole-game.html).
-  + `devttols::load_all()`: load all functions in `R/` folder and running environment.
+  + `devtools::load_all()`: load all functions in `R/` folder and running environment.
   + `devtools::document()`: automatically create documentation using `roxygen2`.
-  + `devttols::check()`: automatically perform compliance check as an R package.
+  + `devtools::check()`: automatically perform compliance check as an R package.
   + `devtools::build_site()`: automatically run analysis scripts in batch and create a `pkgdown` website.

### 2021-11-06 submission-package.Rmd
rename `adam_data/` as `data-adam/`

@@
 You can then proceed to move the file `r0pkgs.txt` to the folder
-`ectddemo/m5/datasets/ectddemo/analysis/adam/programs`,
+`ectddemo/m5/datasets/ectddemo/analysis/adam/programs/`,
 or specify the output text file path above directly.
@@
 Besides the R packages, we need to convert the R Markdown (`.Rmd`) files
-into `.txt` files and saved them in the `programs` folder.
+into `.txt` files and saved them in the `programs/` folder.
 You can do this with `knitr::purl()`:

### 2021-11-06 tlf-ae-summary.Rmd
fix grammar, trailing spaces, file organization
rename `adam_data/` as `data-adam/`

@@
 ```{r}
-adsl <- read_sas("adam_data/adsl.sas7bdat")
-adae <- read_sas("adam_data/adae.sas7bdat")
+adsl <- read_sas("data-adam/adsl.sas7bdat")
+adae <- read_sas("data-adam/adae.sas7bdat")
 ```
@@
 
-The procedure to generate a AE summary table can be summarized as follow：
+The procedure to generate an AE summary table can be summarized as follow:
 

### 2021-11-06 tlf-efficacy.Rmd
Update tlf-efficacy.Rmd
rename `adam_data/` as `data-adam/`

@@
 ```{r}
-adsl <- read_sas("adam_data/adsl.sas7bdat")
-adlb <- read_sas("adam_data/adlbc.sas7bdat")
+adsl <- read_sas("data-adam/adsl.sas7bdat")
+adlb <- read_sas("data-adam/adlbc.sas7bdat")
 ```
@@
 - Step 5: Summarize `ana_locf`, i.e., calculate the pairwise comparison by ANCOVA model, and then format it into an RTF table.
-- Step 6: Set the results from Step 4 and Step 5 and output.
+- Step 6: Combine the outputs in step 4 and step 5 by rows.

### 2021-11-07 project-folder.Rmd

--- 2021-11-01 project-folder.Rmd
@@
 vignettes
-├── adam_data
+├── data-adam
 ├── tlf
@@
 - create a `renv.lock` file to save the state of package versions.
-- create a `renv` folder to manage R packages for a project.
+- create a `renv/` folder to manage R packages for a project.
 

### 2021-11-08 index.Rmd

--- 2021-10-27 index.Rmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, message = FALSE)
+knitr::opts_chunk$set(comment = "#>", echo = TRUE, message = FALSE)
 options(dplyr.summarise.inform = FALSE)
 file.copy("static/CNAME", "_book", overwrite = TRUE)
+fs::dir_copy("slides/", "_book/slides", overwrite = TRUE)
 ```

### 2021-11-08 intro.Rmd
Minor typo fix

--- 2021-09-17 intro.Rmd
@@
 A typical example is a Phase 3 clinical trial where hundreds of TLFs are required for submission.
-Considering the amount of programs required for such an effort, 
+Considering the number of programs required for such an effort,
 a consistent and well-defined folder structure is crucial in
@@
 
-We recommend using 
+We recommend using the
 [R package folder structure](https://github.com/rstudio/cheatsheets/raw/master/package-development.pdf)
-to organize all the A&R related source code and documentation for a clinical trial A&R project. 
-The R package folder structure is well defined and widely used in R community through repositories (e.g. CRAN).
+to organize all the A&R-related source code and documentation for a clinical trial A&R project.
+The R package folder structure is well defined and widely used in the R community through repositories (e.g., CRAN).
 
-The consistent approach of using R package folder structure 
+The consistent approach of using the R package folder structure
 simplifies the communication for all developers within and across organizations.
 
-- For a new R developer, 
-it is an essential step to develop R packages 
+- For a new R developer, it is an essential step to develop R packages
   when you want to share your work with others.
-You will learn one folder structure that is widely used 
-in the R community with outstanding tutorials and tools for free.
-
+  You will learn one folder structure widely used in the R community
+  with outstanding tutorials and tools for free.
 - For an experienced R developer, there is a minimal learning curve.
-
 - For an organization, it simplifies the process, tool, template,
@@
 The workflow around an R package can also improve
-the tractability and reproducibility for an analysis project [@marwick2018packaging].
+the traceability and reproducibility of an analysis project [@marwick2018packaging].
+
+We will revisit folder structure when we discuss project management for a clinical trial project.
 
@@
 to develop Shiny apps as discussed
-in [Chapter 20 of Master Shiny](https://mastering-shiny.org/scaling-packaging.html) book 
-and [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/golem.html) book.
+in [Chapter 20 of the Master Shiny book](https://mastering-shiny.org/scaling-packaging.html)
+and the [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/golem.html) book.
 
@@
 
-- Part 2, "Clinical Trial Project" provides general information 
+- Part 2, "Clinical trial project" provides general information
   with examples to manage a clinical trial A&R project.
   In this part, we assume a reader is a project lead with experience in R package development.
-Some good reference includes [R packages](https://r-pkgs.org/) and 
+  Some good references include [R Packages](https://r-pkgs.org/) and
   [The tidyverse style guide](https://style.tidyverse.org/).
 
-- Part 3, "eCTD Submission package"
-provides general information in preparing 
-submission package related to clinical study report (CSR) in 
+- Part 3, "eCTD submission package" provides general information in preparing
+  submission packages related to clinical study report (CSR) in
   electronic Common Technical Document (eCTD).
@@
 We share the same philosophy described in
-[Section 1.1 of R Packages book](https://r-pkgs.org/intro.html?q=Marwick#intro-phil) 
+[Section 1.1 of the R Packages book](https://r-pkgs.org/intro.html#intro-phil)
 and quote here.

### 2021-11-08 project-overview.Rmd

# (PART) Clinical trial project {-}

# Overview {#project-overview}

In a late-stage clinical trial, the number of A&R deliverables can easily be in the hundreds.
For an organization, it is common to have multiple ongoing clinical trials in a clinical program.

To deliver the A&R results of a clinical trial project,
it is teamwork that typically requires collaborations from both statisticians and programmers.
In this part, let's consider how to organize a clinical trial project as an A&R lead.

Chapter \@ref(folder) will discuss how to organize source code, documents, and
deliverables in an A&R clinical project.
We recommend using the R package folder structure.

Chapter \@ref(manage) will discuss a process or system development lifecycle
to manage the A&R of a clinical project.
We recommend following an agile management approach to define, develop, validate, and deliver work.

### 2021-11-08 slides/workshop-slides.Rmd
fix typo
rename `adam_data/` as `data-adam/`
set comment = "#>"

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(comment = "#>", echo = FALSE)
 ```
@@
 
-- Source data: <https://github.com/elong0527/r4csr/tree/master/adam_data>
+- Source data: <https://github.com/elong0527/r4csr/tree/master/data-adam>
 
@@
 
-`r2rtf` only focuses on **table format**. 
+`r2rtf` only focus on **table format**.
 Data manipulation and analysis shall be handled by other R packages. (e.g. `tidyverse`)

### 2021-11-08 submission-overview.Rmd
address Nan's comments
use sentence case instead of title case for headings; run styler on Rmds

@@
-# (PART) eCTD Submission {-}
+# (PART) eCTD submission {-}
 
-# Overview
+# Overview {#submission-overview}
 
@@
 CLINICAL STUDIES of the Prescribing Information (PI)26 if applicable.
-**The specific software utilized should be specified in the ADaM Reviewer's Guide (ADRG)**.
+**The specific software utilized should be specified in the ADRG**.
 The main purpose of requesting the submission of these programs
@@
 
-In Chapter \@ref(submission-package), we will focus on the approach
-to prepare proprietary R packages and analysis code into proper
+Chapter \@ref(submission-package) will focus on preparing
+proprietary R packages and analysis code into proper
 formats for submission.
 
-In Chapter \@ref(running-environment), we will discuss
+Chapter \@ref(running-environment) will discuss
 the recommendations to make the R code running environment

### 2021-11-08 tlf-overview.Rmd

@@
 
-For simplicity, we have downloaded all these datasets into the `adam_data` folder of
-this project and converted them from the `.xpt` format to the `.sas7bdat` format.
+For simplicity, we have downloaded all these datasets into the `data-adam/`
+folder of this project and converted them from the `.xpt` format to
+the `.sas7bdat` format.
 

### 2021-11-09 project-management.Rmd
rename `adam_data/` as `data-adam/`
update project folder

@@
 
-> After project initiation, modifying `.Rprofile` will be a risk of reproducibility
-> and should be handled carefully if necessary.
+::: {.rmdnote}
+After project initiation, modifying `.Rprofile` will be a risk of reproducibility
+and should be handled carefully if necessary.
+:::
 
@@
 The analysis code and relevant description can be saved in R Markdown files
-in the `vignettes` folder.
+in the `vignettes/` folder.
 

### 2021-11-09 slides/workshop-slides.Rmd
update project folder

@@
 We share the same philosophy described in
-[Section 1.1 of R Packages book](https://r-pkgs.org/intro.html?q=Marwick#intro-phil) and quote here.
+[Section 1.1 of R Packages book](https://r-pkgs.org/intro.html#intro-phil) and quote here.
 
@@
 
-- Let's refer to examples in the [Section 2.4.2](https://r4csr.org/overview.html#r2rtf)
+- Let's refer to examples in the [Section 2.4.2](https://r4csr.org/tlf-overview.html#r2rtf)
 
@@
 
-<https://r4csr.org/overview-2.html>
+<https://r4csr.org/submission-overview.html>
 

### 2021-11-09 project-folder.Rmd

--- 2021-11-06 project-folder.Rmd
@@
 
-A clearly defined clinical project folder structure can have many benefits to clinical development teams in an organization.
+A clearly defined clinical project folder structure can have many benefits
+to clinical development teams in an organization.
 Specifically, a well-defined project structure can achieve:
@@
 - Consistency: everyone works on the same folder structure.
-- Reproducibility: analysis can be executed and reproduced by different team members months/years later.
+- Reproducibility: analysis can be executed and reproduced by different
+  team members months/years later.
 - Automation: automatically check the integration of a project.
@@
 
-We illustrate the R package project folder for the A&R project using a demo project called
-[`esubdemo`](https://github.com/elong0527/esubdemo.git).
-You can
+We will use the [`esubdemo`](https://github.com/elong0527/esubdemo) R package to
+illustrate the project folder structure for the A&R project. You can
 [clone](https://happygitwithr.com/rstudio-git-github.html#clone-the-new-github-repository-to-your-computer-via-rstudio)
-the project using RStudio.
+the project using RStudio with
+
+```bash
+git clone https://github.com/elong0527/esubdemo.git
+```
 
@@
 
-> Under the `vignettes` folder, there are two folders: `adam` folder and `tlf` folder.
-> The `adam` folder contains ADaM datasets.
-> The `tlf` folder contains output TLFs in RTF format.
-> We put `adam` and `tlf` folders within the `vignettes` folder only for illustration purposes.
-> In an actual A&R report, you may have a different location to save your input and output.
+::: {.rmdnote}
+Under the `vignettes/` folder, there are two folders: `adam/` and `tlf/`.
+The `adam/` folder contains ADaM datasets.
+The `tlf/` folder contains output TLFs in RTF format.
+We put `adam/` and `tlf/` folders within the `vignettes/` folder only for illustration purposes.
+In an actual A&R report, you may have a different location to save your input and output.
+:::
 
@@
 #' @export
-fmt_est <- function(.mean,
-                    .sd,
-                    digits = c(1, 2)) {
+fmt_est <- function(.mean, .sd, digits = c(1, 2)) {
   .mean <- fmt_num(.mean, digits[1], width = digits[1] + 4)
@@
 
-> This is the same level of reproducibility in most SAS environments.
-> https://support.sas.com/techsup/pcn/altopsys.html
+::: {.rmdnote}
+This is the same level of reproducibility in most SAS environments:
+<https://support.sas.com/techsup/pcn/altopsys.html>
+:::
+
+### R version
 
@@
 
-If there is a R version mismatch, an error message is displayed as below.
+If there is an R version mismatch, an error message is displayed as below.
 
@@
 
-> `.Rprofile` is only for project-specific R packages.
-> A standard R package should not use `.Rprofile`.
+::: {.rmdnote}
+`.Rprofile` is only for project-specific R packages.
+A standard R package should not use `.Rprofile`.
+:::
+
+### R package version
 
@@
 
-> RStudio Package Manager (RSPM) provided a solution to host both
-> publicly available and internally developed R packages.
-> However, the public RSPM server (https://packagemanager.rstudio.com/client/#/repos/2/overview)
-> did not provide a daily snapshot as MRAN had.
+::: {.rmdnote}
+RStudio Package Manager (RSPM) provides a solution to host both
+publicly available and internally developed R packages.
+However, the [public RSPM server](https://packagemanager.rstudio.com/client/#/repos/2/overview)
+does not provide a daily snapshot as MRAN does.
+:::
 
@@
 - create a `renv.lock` file to save the state of package versions.
-- create a `renv` folder to manage R packages for a project.
+- create a `renv/` folder to manage R packages for a project.
 
-> The `renv.lock` file and `renv/` folder are only for project-specific R package.
-> A standard R package should not use `renv`.
+::: {.rmdnote}
+The `renv.lock` file and `renv/` folder are only for project-specific R package.
+A standard R package should not use `renv`.
+:::
 
@@
 If the project is initiated properly,
-you shall be able to see similar messages to inform how we control R package versions.
+you should be able to see similar messages to inform how we control R package versions.
 
@@
 
-> A cloud-based R environment (e.g., RStudio Workbench)
-> can enhance the reproducibility within an organization
-> by using the same operating system, R version, and R package versions for an A&R project.
-> More details can be found at <https://environments.rstudio.com/>.
+::: {.rmdnote}
+A cloud-based R environment (e.g., RStudio Workbench)
+can enhance the reproducibility within an organization
+by using the same operating system, R version, and R package versions for an A&R project.
+More details can be found at <https://environments.rstudio.com/>.
+:::
 
-> A container solution like [Docker](https://github.com/rstudio/r-docker) could further enhance
-> the reproducibility across an organization at the operating system level but beyond the scope of this book.
+::: {.rmdnote}
+A container solution like Docker [@RJ-2020-007] could further enhance
+the reproducibility across an organization at the operating system level
+but beyond the scope of this book.
+:::
 
@@
 
-- Create project template using [RStudio Project Templates](https://rstudio.github.io/rstudio-extensions/rstudio_project_templates.html)
-- Add additional compliance checks for analysis and reporting.
+- Create project template using
+  [RStudio project templates](https://rstudio.github.io/rstudio-extensions/rstudio_project_templates.html);
+- Add additional compliance checks for analysis and reporting;
 - Save log files for running in batch.

### 2021-11-10 preface.Rmd
clarify license; optimize contributor display

--- 2021-11-08 intro.Rmd
@@
-# Introduction {#intro}
+# Preface {-}
 
@@
 to develop Shiny apps as discussed
-in [Chapter 20 of the Master Shiny book](https://mastering-shiny.org/scaling-packaging.html)
+in Chapter 20 of the [Mastering Shiny](https://mastering-shiny.org/scaling-packaging.html) book
 and the [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/golem.html) book.
@@
   In this part, we assume a reader is a project lead with experience in R package development.
-  Some good references include [R Packages](https://r-pkgs.org/) and
-  [The tidyverse style guide](https://style.tidyverse.org/).
+  Some good references include [R Packages](https://r-pkgs.org/) and the
+  [tidyverse style guide](https://style.tidyverse.org/).
 
@@
 We share the same philosophy described in
-[Section 1.1 of the R Packages book](https://r-pkgs.org/intro.html#intro-phil)
+Section 1.1 of the [R Packages](https://r-pkgs.org/intro.html#intro-phil) book
 and quote here.
@@
 rather than thinking about the minutiae of the package structure."
+
+## Authors and contributors
+
+The document is maintained by a community.
+While reading the document, you can be a contributor as well.
+The quality of this document relies on you.
+
+- Authors: contributed the majority of content to at least one chapter.
+
+  Yilong Zhang <<yilong.zhang@merck.com>>, Nan Xiao, Keaven Anderson
+
+- Contributors: contributed at least one commit to the [source code](https://github.com/elong0527/r4csr).
+
+```{r, results = "asis", echo = FALSE, message = FALSE}
+contributors <- read.csv("contributors.csv", stringsAsFactors = FALSE, na.strings = "")
+contributors$desc <- with(contributors, ifelse(is.na(username), name, paste0(name, " (\\@", username, ")")))
+cat("  We are grateful for all the improvements brought by these contributors (in chronological order): ", sep = "")
+cat(paste0(contributors$desc, collapse = ", "))
+cat(".\n")
+```

### 2021-11-10 index.Rmd

@@
 library(knitr)
-library(kableExtra)
 library(pkglite)
@@
 
-**This is the work-in-progress draft of the book.**
+**This is a work-in-progress draft.**
 
-## Authors and contributors {-}
+## License {-}
 
-```{r, echo = FALSE}
-contributors <- c(
-  person("Yilong", "Zhang", email = "yilong.zhang@merck.com", role = "aut"),
-  person("Nan", "Xiao", role = "aut"),
-  person("Keaven", "Anderson", role = "aut"),
-  person("Yujie", "Zhao", role = "ctr"),
-  person("Aiming", "Yang", role = "ctr"),
-  person("Steven", "Haesendonckx", role = "ctr"),
-  person("Howard", "Baek", role = "ctr")
-)
-```
-
-The document is maintained by a community.
-While reading the document, you can be a contributor as well.
-The quality of this document relies on you.
-
-- Author: contribute the majority content of at least one chapter.
-- Contributor: [contribute at least one commit to the source code](https://github.com/elong0527/r4csr)
-
-```{r, echo = FALSE, eval = TRUE}
-data.frame(contributors = as.character(contributors)) %>%
-  kable(format = "html") %>%
-  kable_styling()
-```
+This book is licensed to you under [Creative Commons Attribution-NonCommercial-NoDerivatives 4.0 International License](https://creativecommons.org/licenses/by-nc-nd/4.0/).

### 2021-11-11 tlf-disposition.Rmd

@@
 ```{r}
-adsl <- read_sas("adam_data/adsl.sas7bdat")
+adsl <- read_sas("data-adam/adsl.sas7bdat")
 ```
@@
 - USUBJID: Unique subject identifier
-- TRT01P: Planed treatment
+- TRT01P: Planned treatment
 - TRT01PN: Planned treatment numeric encoding

### 2021-11-11 tlf-overview.Rmd

@@
 
-In clinical trials, a critical step is submission to regulatory agencies.
+In clinical trials, a critical step is to submit trial results to regulatory agencies.
 [Electronic Common Technical Document (eCTD)](https://en.wikipedia.org/wiki/Electronic_common_technical_document)
@@
 define, develop, validate and deliver tables, listings, and figures (TLFs) required for a CSR to
-summarize the safety and/or safety of the pharmaceutical product.
+summarize the efficacy and/or safety of the pharmaceutical product.
 Microsoft Word is widely used to prepare CSR in the pharmaceutical industry.
-Therefore, `rtf`, `doc`, `docx` are commonly used formats in their deliverables.
+Therefore, `.rtf`, `.doc`, `.docx` are commonly used formats in their deliverables.
 
@@
 
-We provided a brief introduction of `r2rtf` and showed how to transfer
-data frames into Table, Listing, and Figures (TLFs).
+We provide a brief introduction of `r2rtf` and showed how to transfer
+data frames into table, listing, and figures (TLFs).
 

### 2021-11-15 tlf-disposition.Rmd
Xiaoxia reviewed TLFs part on Nov14

@@
 Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-a summary needs to be provided to account for all participants who entered the study in Section 10.1, Disposition of Participants.
+a summary table needs to be provided to include all participants who entered the study in Section 10.1, Disposition of Participants.
 
-The disposition of participant reports the numbers of participants who were randomized,
+The disposition of participants table reports the numbers of participants who were randomized,
 and who entered and completed each phase of the study.
 In addition, the reasons for all post-randomization discontinuations,
-grouped by treatment and by major reason (lost to follow-up, adverse event, poor compliance, etc.).
+grouped by treatment and by major reason (lost to follow-up, adverse event, poor compliance, etc.) are reported.
 
@@
 
-In this chapter, the creation of a typical disposition table is shown.
+In this chapter, we show how to the create a typical disposition table.
 
@@
 For a disposition table, all the required information is saved in a Subject-level Analysis Dataset (ADSL).
-The dataset is provided in `sas7bdat` format that is a SAS data format currently used in many clinical trial analysis and reporting.
-The `haven` package is able to read in the dataset, while maintaining its attributes (e.g., variable labels).
+This dataset is provided in `sas7bdat` format, which is a SAS data format currently used in many clinical trial analysis and report.
+The `haven` package is able to read the dataset, while maintaining its attributes (e.g., variable labels).
 
@@
 
-Following variables are used in the preparation of a simplified disposition of participants table:
+The following variables are used in the preparation of a simplified disposition of participants table:
 
-- USUBJID: Unique subject identifier
-- TRT01P: Planned treatment
-- TRT01PN: Planned treatment numeric encoding
-- DISCONFL: Discontinued from study flag
-- DCREASCD: Discontinued from study reason coded
+- USUBJID: unique subject identifier
+- TRT01P: planned treatment
+- TRT01PN: planned treatment numeric encoding
+- DISCONFL: discontinued from study flag
+- DCREASCD: discontinued from study reason coded
 
@@
 
-In the code below, we calculated participants in the analysis population.
+In the code below, we calculate the number of participants in the analysis population by treatment arms.
 
@@
 In the code below, we calculate the number and percentage of
-participants who completed/discontinued the study for different reasons.
+participants who completed/discontinued the study for different reasons by treatment arms.
 
@@
 
-In the code below, we calculate the number and percentage of participants who complete the study.
-We split `n_reason` because we need to customize the row order of the table.
+In the code below, we calculate the number and percentage of participants who complete the study by treatment arms.
+We split `n_reason` because we want to customize the row order of the table.
 
@@
 
-In the code below, we calculate the number and percentage of participants who discontinued the study for different reasons.
-For display purposes, `paste0("    ", row)` is used to add leading spaces to mimick indentation in the final report.
+In the code below, we calculate the numbers and percentages of participants who discontinued the study for different reasons by treatment arms.
+For display purpose, `paste0("    ", row)` is used to add leading spaces to produce indentation in the final report.
 
@@
 
-Now we combined individual rows into the whole table for reporting purpose.
+Now we combine individual rows into one table for reporting purpose.
 `tbl_disp` is used as input for `r2rtf` to create final report.
@@
 
-- Step 1: Read subject level data into R, i.e., `adsl`.
+- Step 1: Read subject level data (i.e., `adsl`) into R.
 - Step 2: Count participants in the analysis population and name the dataset `n_rand`.
-- Step 3: Count the number and percentage of participants that discontinued the study, and name the dataset `n_disc`.
-- Step 4: Count the number and percentage of participants that discontinued the study for different reasons,
+- Step 3: Calculate the number and percentage of participants who discontinued the study by treatment arm, and name the dataset `n_disc`.
+- Step 4: Calculate the numbers and percentages of participants who discontinued the study for different reasons by treatment arm,
   and name the dataset `n_reason`.
-- Step 5: Count the number and percentage of participants that completed the study, and name the dataset `n_complete`.
+- Step 5: Calculate the number and percentage of participants who completed the study by treatment arm, and name the dataset `n_complete`.
 - Step 6: Bind `n_rand`, `n_disc`, `n_reason`, and `n_complete` by row.

### 2021-11-15 tlf-baseline.Rmd

@@
 Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to summarize critical demographic and baseline characteristics of the patients
+we need to summarize critical demographic and baseline characteristics of the participants
 in Section 11.2, Demographic and Other Baseline Characteristics.
@@
 In this chapter, we illustrate how to create a simplified
-baseline characteristics table in a study.
+baseline characteristics table for a study.
 
@@
 
-We start to define the format of the output. We highlight items that
+We define the format of the output. We highlight items that
 are not discussed in previous discussion.
@@
 space of text. They are helpful when you need to control the white space
-of a long sentence. For example, "AMERICAN INDIAN OR ALASKA NATIVE"
-in the table.
+of a long phrase "AMERICAN INDIAN OR ALASKA NATIVE"
+in the table provide an example.
 
@@
   rtf_title(
-    "Participant Baseline Characteristics",
+    "Baseline Characteristics of Participants",
     "(All Participants Randomized)"
@@
 
-In conclusion, the procedure to generate demographic and baseline characteristics table is summarized as follows.
+In conclusion, the procedure to generate demographic and baseline characteristics table is summarized as follows:
 
@@
 - Step 2: Use `table1::table1()` to get the baseline characteristics table.
-- Step 3: Transfer the output in Step 2 into a data frame that only contains ASCII characters.
+- Step 3: Transfer the output from Step 2 into a data frame that only contains ASCII characters.
 - Step 4: Define the format of the RTF table by using the R package `r2rtf`.

### 2021-11-15 tlf-efficacy.Rmd

@@
 
+In this chapter, we illustrate how to generate an efficacy table for a study.
 For efficacy analysis, only the change from baseline glucose data at week 24 is analyzed.
@@
 Here the variable `AVISITN` is the numerical analysis visit.
-For example, if the analysis visit is recorded as "Baseline", i.e., `AVISIT = Baseline`, then `AVISITN = 0`;
-if the analysis visit is recorded as "Week 24", i.e., `AVISIT = Week 24`, then `AVISITN = 24`;
-if the analysis visit is blank, then `AVISITN` is also blank.
+For example, if the analysis visit is recorded as "Baseline" (i.e., `AVISIT = Baseline`), 
+`AVISITN = 0`;
+if the analysis visit is recorded as "Week 24" (i.e., `AVISIT = Week 24`), `AVISITN = 24`;
+if the analysis visit is blank, `AVISITN` is also blank.
 We will discuss these missing values in Section 6.4.
@@
 
-To prepare the report data frame, we created a few helper functions
+To prepare the report, we create a few helper functions
 by using the `fmt_num()` function defined in Chapter \@ref(population).
 
-- Format Estimators
+- Format estimators
 
@@
 
-- Format Confidence Interval
+- Format confidence interval
 
@@
 
-- Format P-Value
+- Format p-value
 
@@
 
-First the observed data at Baseline and Week 24 is summarized:
+First the observed data at Baseline and Week 24 are summarized using code below:
 
@@
 
-Also the observed change from baseline glucose at Week 24 is summarized:
+Also the observed change from baseline glucose at Week 24 is summarized using code below:
 
@@
 In clinical trials, missing data is inevitable.
-In this study, there are also missing values in glucose data.
+In this study, there are missing values in glucose data.
 
@@
 LOCF approach is a single imputation approach that is **not recommended**
-in real studies.
+in real application.
 Interested readers can find more discussion on missing data approaches in the book:
@@
 The `emmeans` R package is used to obtain
-within and between group least square mean (LS mean)
+within and between group least square (LS) mean
 
@@
 
-`t11`, `t12` and `t13` are combined to get the first part of the report data
+`t11`, `t12` and `t13` are combined to get the first part of the report table
 
@@
 Special characters `^` and `_` are used to define superscript and subscript of text. And `{}` is to define the part that will be impacted.
-For example, `{^a}` provide a superscript `a` for footnote notation.
+For example, `{^a}` provides a superscript `a` for footnote notation.
 `r2rtf` also supports most LaTeX characters.
@@
 [`r2rtf` get started page](https://merck.github.io/r2rtf/articles/r2rtf.html#special-character).
-The `text_convert` argument in `r2rtf_*()` functions controls whether convert special characters.
+The `text_convert` argument in `r2rtf_*()` functions controls whether to convert special characters.
 
@@
 
-we also use `r2rtf` to prepare the table format for `t2`
+We also use `r2rtf` to prepare the table format for `t2`
 
@@
 
-Finally we combine the two parts to get the final table using `r2rtf`.
+Finally, we combine the two parts to get the final table using `r2rtf`.
 This is achieved by providing a list of `t1_rtf` and `t2_rtf` as input for
@@
 
-- Step 1: Read the data into R, i.e., `adsl` and `adlb`.
+- Step 1: Read the data (i.e., `adsl` and `adlb`) into R.
 - Step 2: Define the analysis dataset.
@@
 The second dataset is the collection of all records post baseline and on or before week 24 (`ana`).
-- Step 3: Impute the missing values. In our example, we name the `ana` dataset after imputation as `ana_locf`.
-- Step 4: Summarize `gluc`, i.e., calculate the mean and standard derivation, and then format it into an RTF table.
-- Step 5: Summarize `ana_locf`, i.e., calculate the pairwise comparison by ANCOVA model, and then format it into an RTF table.
-- Step 6: Combine the outputs in step 4 and step 5 by rows.
+- Step 3: Impute the missing values. In this example, we name the `ana` dataset after imputation as `ana_locf`.
+- Step 4: Calculate the mean and standard derivation of efficacy endpoint (i.e., `gluc`), and then format it into an RTF table.
+- Step 5: Calculate the pairwise comparison by ANCOVA model, and then format it into an RTF table.
+- Step 6: Combine the outputs from steps 4 and 5 by rows.

### 2021-11-15 tlf-population.Rmd
rename `adam_data/` as `data-adam/`

@@
 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to summarize which participants were included in each efficacy
+we need to summarize the number of participants included in each efficacy
 analysis in Section 11.1, Data Sets Analysed.
@@
 In this chapter, we illustrate how to create a summary table for
-the analysis population in a study.
+the analysis population for a study.
 
@@
 
--   USUBJID: Unique subject identifier
--   ITTFL: Intent-to-treat population flag
--   EFFFL: Efficacy population flag
--   SAFFL: Safty population flag
+-   USUBJID: unique subject identifier
+-   ITTFL: intent-to-treat population flag
+-   EFFFL: efficacy population flag
+-   SAFFL: safty population flag
 
@@
 Before we write the analysis code, let's discuss the possibility of reusing R
-code by writing some toy helper functions.
+code by writing helper functions.
 
@@
 -   Format the percentages using the `formatC()` function.
--   Calculate the numbers and percentages by groups.
+-   Calculate the numbers and percentages by treatment arm.
 
-We create two ad-hoc functions and apply them to all the rest of the
-tables.
+We create two ad-hoc functions and use them to create the
+tables in the rest of this book.
 
-To format numbers and percentages, we created a function called
+To format numbers and percentages, we create a function called
 `fmt_num()`. It is a very simple function wrapping `formatC()`.
@@
 For example, we can compare the two versions of code to format the
-percentage used in \@ref(disposition) and `fmt_num`.
+percentage used in Chapter \@ref(disposition) and `fmt_num`.
 
@@
 
-With the helper functions `count_by`, we can easily prepare a report
+With the helper function `count_by`, we can easily prepare a report
 dataset as
@@
 
-Now we combine individual rows into the whole table for reporting
-purposes. `tbl_pop` is used as input for `r2rtf` to create the final
+Now we combine individual rows into one table for reporting
+purpose. `tbl_pop` is used as input for `r2rtf` to create the final
 report.
@@
 
-We start to define the format of the output.
+We define the format of the output using code below.
 
@@
   rtf_title(
-    "Participants Accounting in Analysis Population",
+    "Summary of Analysis Sets", ## "Participants Accounting in Analysis Population" is a weird title for me. Merck uses this table title for CSR? 
     "(All Participants Randomized)"
@@
 The procedure to generate an analysis population table can be summarized
-as follow:
+as follows:
 
-- Step 1: Read data into R, i.e., `adsl`.
+- Step 1: Read data (i.e., `adsl`) into R.
 - Step 2: Bind the counts/percentages of the ITT population, the
@@
   `count_by()` function.
-- Step 3: Format the output in Step 2 by `r2rtf`.
+- Step 3: Format the output from Step 2 using `r2rtf`.

### 2021-11-16 project-management.Rmd
Xiaoxia review project and submission Nov 15

@@
 
-Although many factors could decide if a project could execute efficiently,
+Although many factors determine if a project can execute efficiently,
 we believe a few aspects are critical for long-term success,
@@
 
-For example, in the `esubdemo` project, we have 6 R Markdown files to create TLFs.
+For example, in the `esubdemo` project, we have six R Markdown files to create TLFs.
 

### 2021-11-16 submission-package.Rmd
Xiaoxia review project and submission Nov 15

@@
 submitting R code.
-Secondly, we will discuss how to prepare the proprietary R packages
-(if any), and make them part of the submission package.
-Lastly, we will provide reusable templates for updating
+Then, we will discuss how to prepare the proprietary R packages
+(if any), and make them be part of the submission package.
+In the end, we will provide reusable templates for updating
 Analysis Data Reviewer's Guide (ADRG) and Analysis Results Metadata (ARM)
@@
 The demo project (R package) we will prepare for submission is called
-[`esubdemo`](https://github.com/elong0527/esubdemo), available on GitHub.
+[`esubdemo`](https://github.com/elong0527/esubdemo), which is available on GitHub.
 You can download or clone it:
@@
 The demo submission package (not to be confused with the R package above)
-is [`ectddemo`](https://github.com/elong0527/ectddemo), also available on GitHub.
+is [`ectddemo`](https://github.com/elong0527/ectddemo), which is also available on GitHub.
 You can download or clone it:
@@
 
-We will assume the paths to the two folders are `esubdemo/` and `ectddemo/` below.
+We assume the paths to the two folders are `esubdemo/` and `ectddemo/` below.
 
@@
 
-A few things to pay attention to for passing compliance checks:
+A few things need to be paid attention to in order to pass compliance checks:
 
-- The filenames under `programs/` should all use lower case
+- The file names under `programs/` should be in lower case letters
 (with no underscores or other special characters).
@@
 This can be verified by `pkglite::verify_ascii()`
-- Any `.docx` files should be converted to PDF files for formal submission.
+- All `.docx` files should be converted to PDF files for formal submission.
 
@@
 submission package.
-We will start preparing the proprietary R packages in the following sections.
+We will prepare the proprietary R packages in the following sections.
 
@@
 
-Before we start, there are a few aspects to figure out,
+Before we start, there are a few aspects to figure out in order
 to accurately identify the R packages for submission.
@@
 Again, this can be checked in the project startup script,
-discussed in section \@ref(reproduce).
+as discussed in section \@ref(reproduce).
 
@@
 Introducing any extra external dependencies will likely increase the cost
-of qualification, validation, and testing, and maintenance,
+of qualification, validation, testing, and maintenance,
 especially under Windows.
@@
 
-## Prepare R Packages for submission
+## Prepare R packages for submission
 
-To prepare R packages for submission, one needs to pack the packages into text files, and then verify if the file only contains ASCII characters.
-With packed packages, one can unpack and install them from the text file, too.
+To prepare R packages for submission, one needs to pack the packages into text files, and then verify if the files only contain ASCII characters.
+With packed packages, one can unpack and install them from the text files, too.
 
@@
 
-If successfully tested, this command will be used in the ADRG instructions
+If the test is successful, this command can be used in the ADRG instructions
 for restoring and installing the packed R package(s).
@@
 formats and verified that they can be restored and executed correctly,
-we should update the ADRG to provide guidelines on how to use them.
+we need update the ADRG to provide guidelines on how to use them.
 

### 2021-11-16 project-folder.Rmd

@@
 their R functions before the R package can be disseminated through the Comprehensive R Archive Network (CRAN).
-As a user, you can easily install and use those R packages after installing from CRAN.
+As a user, you can easily install and use those R packages after downloading from CRAN.
 There are many good resources to guide developers on R package development,
-for example, the [R Packages book](https://r-pkgs.org/) by Hadley Wickham.
+such as, the [R Packages book](https://r-pkgs.org/) by Hadley Wickham.
 
@@
 
-With a formal project, it is also important to provide proper documentation for those help functions.
-We used `roxygen2` to document functions.
+For a clinical trial project, it is also important to provide proper documentation for those help functions.
+We use `roxygen2` to document functions.
 For example, the header below defines each variable in `fmt_est`.
@@
 The `roxygen2` documentation will be converted into
-standard R documentation format, saved as `.Rd` files in the `man/` folder.
+standard R documentation format, and saved as `.Rd` files in the `man/` folder.
 This step is automatically handled by `devtools::document()`.
@@
 We have learned a few functions in `devtools` to generate content automatically.
-Here is a shortlist of tools that can enhance the workflow.
+Here is a list of tools that can enhance the workflow.
 

### 2021-11-16 submission-environment.Rmd

@@
 
-In this chapter, we will focus on automating the creation of
+In this chapter, we focus on automating the creation of
 the R environments with R code to accelerate the dry run testing process,
@@
 - Create a project folder under `C:/temp/` with a `.Rproj` file;
-- Download R `4.1.1` installer from CRAN, install it into `C:/temp/R/`;
+- Download R `4.1.1` installer from CRAN, and install it into `C:/temp/R/`;
 - Not install Rtools (by default, `rtools_version = NULL`);
@@
   to use the specified `repo` (an MRAN snapshot in this example),
-  and instructions to set the R binary path in RStudio;
+  and give instruction to set the R binary path in RStudio;
 - Create a `.Renviron` file under the project folder and set the
@@
 
-The helper functions `version_*()` and `snapshot_*()` could assist you
+The helper functions `version_*()` and `snapshot_*()` can assist you
 in determining specific versions of R and Rtools that are currently available,
@@
 
-If using `cleanslate`, remember to update the ADRG instructions for
-executing the analysis programs in R. Mostly, this would simplify
+If you use `cleanslate`, remember to update the ADRG instructions for
+executing the analysis programs in R. Mostly, this can simplify
 the first three steps on creating a project, installing a specific
@@
 The addin provides a wizard-like interface to help create the environment
-with the most important options, although with less flexibility compared
+with the most important options, yet with less flexibility compared
 to the functional API demonstrated above.

### 2021-11-16 tlf-efficacy-km.Rmd
Create efficacy KM plot chapter

# Efficacy Figure

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
primary and secondary efficacy endpoints need to be summarized
in Section 11.4, Efficacy Results and Tabulations of Individual Participant.

```{r}
library(haven) # Read SAS data
library(dplyr) # Manipulate data
library(tidyr) # Manipulate data
library(r2rtf) # Reporting in RTF format
library(survival) # fit survival model
library(survminer) # plot nice KM plot with risk table
```

In this chapter, we illustrate how to create a Kaplan-Meier plot in a study.
For the survival analysis in efficacy, time to dermatologic event (TTDE) will be analyzed.

```{r, out.width = "100%", out.height = "450px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/fig_km.png")
```

## Analysis dataset

To prepare the analysis, `adtte` dataset is required.

```{r}
adtte <- read_sas("data-adam/adtte.sas7bdat")
```

First of all, to perpare the analysis ready data, filter all records for the efficacy endpoint of TTDE using `PARAMCD` (or `PARAM`, `PRAMN`), then select the survival analysis related variables:

  + Treatment Arm: `TRTP` (using corresponding numeric code `TRTAN` to re-order the levels, "Placebo" will be the reference level)
  + Time-to-event Analysis Value: `AVAL`
  + Event (Censoring) Status: `CNSR`

```{r}
adtte_ttde <- adtte %>% filter(PARAMCD=="TTDE") %>% select(TRTP, TRTAN, AVAL, CNSR) %>%
  mutate(TRTP = forcats::fct_reorder(TRTP, TRTAN)) # recorder levels
```

## Create Kaplan-Meier Curve

The `survival` package is used to create the curve data points, then `survminer` package can be used to plot the KM curve with an elegant ggplot2 style. There are tons of customized options for users to create a KM plot with risk table and testing result (nominal p-value). More details can be found in  https://rpkgs.datanovia.com/survminer/reference/ggsurvplot.html

User can directly use `Surv` with formula, or use a fitted Cox model with `coxph` function if also want to obtain other information, i.e. hazard ratio, survival rates, etc. [TBDev: "Analysis of Survival Endpoint Table" Chapter]

```{r}
fit <- survfit(Surv(AVAL, 1-CNSR) ~ TRTP, data=adtte_ttde)
# sum_fit <- summary(coxph(Surv(AVAL, 1-CNSR) ~ ARM  + strata(STRATAP), data = adtte_efs))
fig_km <- ggsurvplot(fit,  
           xlab = "Time in Months",
           ylab = "Survival Probability",
           break.time.by = 30.4367,
           xscale = "d_m",
           xlim = c(0, 200),
           linetype = c(1, 2, 3),
           risk.table = TRUE,
           risk.table.height = 0.3,# Useful when you have multiple groups
           legend = c(0.8, 0.9), legend.title = "", 
 legend.labs = sort(unique(adtte_ttde$TRTP)), censor.size=4, size = 0.8, title = "", censor.shape="|",

 pval = TRUE, pval.method = TRUE, pval.coord = c(0, 0.1), pval.method.coord = c(0, 0.2), pval.size = 5,

 ggtheme = theme_classic2(base_family = "Times New Roman", base_size = 18)

)
```

Then we can save the figure as a `.png` file, and use `r2rtf` package to create a formatted RTF figure. More details can be found in https://merck.github.io/r2rtf/articles/example-figure.html

```{r}
# Save as PNG file
ggsave("tlf/fig_km.png", plot = print(fig_km, newpage = FALSE), width = 10, height = 6)

# Create RTF figure
  rtf_read_png("tlf/fig_km.png") %>%              # read PNG files from the file path
  rtf_title("Time to Dermatologic Event by Treatment Group", "All Participants") %>%     # add title or subtitle
  rtf_footnote("footnote") %>%           # add footnote
  rtf_source("[datasource: adam-adtte]") %>%  # add data source
  rtf_figure(fig_width = 6, fig_height = 3.6) %>%      # set proportional figure size to the original PNG figure size
  rtf_encode(doc_type = "figure") %>%    # encode rtf as figure
  write_rtf(file = "tlf/fig_km.rtf") 
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/fig_km.pdf")
```

In conclusion, the procedure to generate the above efficacy KM plot is summarized as follows.

- Step 1: Read the data into R, i.e., `adtte`.
- Step 2: Define the analysis ready dataset. In this example, we define the analysis dataset for TTDE endpoint `adtte_ttde`.
- Step 3: Create the curve's data points, using `Surv` formula directly or fitted survival models beforehand. Then plot and customize the KM curve with risk table using `survminer` package. 
- Step 4: Save the output as a PNG file, then format it into an RTF table.

### 2021-11-16 tlf-efficacy-km.Rmd
loaded package update

@@
 library(dplyr) # Manipulate data
-library(tidyr) # Manipulate data
 library(r2rtf) # Reporting in RTF format
@@
 library(survminer) # plot nice KM plot with risk table
+library(forcats) # Reorder ordinal variable level based on preference 
 ```

### 2021-11-17 tlf-ae-spec.Rmd
streamline paragraphs

--- 2021-11-01 tlf-ae-spec.Rmd
@@
 Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to summarize which participants were included in each efficacy
-analysis in Section 12.2, Adverse Events (AEs).
+we need to summarize number of participants for each specific AE in Section 12.2, Adverse Events (AEs).
 
@@
 In this chapter, we illustrate how to summarize simplified specific AE
-information in a study.
+information for a study.
 
@@
 ```{r}
-adsl <- read_sas("adam_data/adsl.sas7bdat")
-adae <- read_sas("adam_data/adae.sas7bdat")
+adsl <- read_sas("data-adam/adsl.sas7bdat")
+adae <- read_sas("data-adam/adae.sas7bdat")
 ```
 
-For illustration purposes, we only provide count in the simplified
-table. The percentage of participants for each AE criteria can be
-calculated as in Chapter \@ref(aesummary).
+For illustration purpose, we only provide counts in the simplified
+table. The percentage of participants for each AE can be
+calculated as shown in Chapter \@ref(aesummary).
 
-In this way, let's focus on the analysis script for two advanced
+Here, we focus on the analysis script for two advanced
 features for a table layout.
@@
 
-In the code below, we count the number of subjects in each AE term by
-SOC and treatment group, and we create a new variable `order` and set it
-as `0` for these counted numbers. The variable `order` will help with
+In the code below, we count the number of participants in each AE term by
+SOC and treatment arm, and we create a new variable `order` and set it
+as `0`. The variable `order` can help with
 the data manipulation later.
@@
 In the code below, we count the number of subjects in each AE term by
-SOC, AE term, and treatment group. Here we also create a new variable
-`order` and set it as `1` for these counted numbers.
+SOC, AE term, and treatment arm. Here we also create a new variable
+`order` and set it as `1`.
 
@@
 
-We prepare reporting data for AE information.
+We prepare reporting data for AE information using code below:
 
@@
 
-We prepare reporting data for analysis population.
+We prepare reporting data for analysis population using code below:
 
@@
 
-We start to define the format of the output.
+We define the format of the output as below:
 
 To obtain the nested layout, we use the `page_by` argument in the
-`rtf_body` function. By defining `page_by="AESOC"`, `r2rtf` recognize
+`rtf_body` function. By defining `page_by="AESOC"`, `r2rtf` recognizes
 the variable as a group indicator.
@@
 ```
-
-More discussion on `page_by`, `group_by` and `subline_by`
-features can be found in the
-[`r2rtf` package website](https://merck.github.io/r2rtf/articles/example-sublineby-pageby-groupby.html.
 
@@
 
+More discussion on `page_by`, `group_by` and `subline_by`
+features can be found on the
+[`r2rtf` package website](https://merck.github.io/r2rtf/articles/example-sublineby-pageby-groupby.html).
+
 The procedure to generate a baseline characteristics table can be
-summarized as follow:
+summarized as follows:
 
-- Step 1: Read data into R, i.e., `adae` and `adsl`.
-- Step 2: Count the number of subjects by SOC and treatment group
+- Step 1: Read data (i.e., `adae` and `adsl`) into R.
+- Step 2: Count the number of participants by SOC and treatment arm
   (rows with bold text) and save into `t1`.
-- Step 3: Count the number of subjects in each AE term by SOC, AE
-  term, and treatment group (rows without bold text) and save into `t2`.
+- Step 3: Count the number of participants in each AE term by SOC, AE
+  term, and treatment arm (rows without bold text) and save into `t2`.
 - Step 4: Bind `t1` and `t2` by row into `t_ae`.
-- Step 5: Count the number of subjects in each arm as `t_pop`.
-- Step 6: Bind `t_pop` and `t_ae` by row into `tbl_ae_spec`.
-- Step 7: Format the output by `r2rtf`.
+- Step 5: Count the number of participants in each arm as `t_pop`.
+- Step 6: Row-wise combine `t_pop` and `t_ae` into `tbl_ae_spec`.
+- Step 7: Format the output by using `r2rtf`.

### 2021-11-17 tlf-ae-summary.Rmd

@@
 Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to summarize which participants were included in each efficacy analysis in Section 12.2, Adverse Events (AEs).
+we need to the number of participants were included in each safety analysis in Section 12.2, Adverse Events (AEs).
 
@@
 
-In this chapter, we illustrate how to summarize AE information in a study.
+In this chapter, we illustrate how to summarize AEs information for a study.
 
@@
 
-We first summarize participants in population by treatment group.
+We first summarize participants in population by treatment arm.
 
@@
 - With serious drug-related adverse events
-- Who died"
+- Who died
 
@@
 
-We prepare reporting-ready dataset for each AE criteria.
+We prepare reporting-ready dataset for each AE group.
 
@@
 
-We start to define the format of the output.
+We define the format of the output using code below:
 
@@
 
-The procedure to generate an AE summary table can be summarized as follow:
+The procedure to generate an AE summary table can be summarized as follows:
 
-- Step 1: Read data into R, i.e., `adae` and `adsl`.
-- Step 2: Summarize participants in population by treatment group, and we name the dataset as `t_pop`.
+- Step 1: Read data (i.e., `adae` and `adsl`) into R.
+- Step 2: Summarize participants in population by treatment arm, and name the dataset as `t_pop`.
 - Step 3: Summarize participants in population by required AE criteria of interest,
-  and we name the dataset as `t_ae`.
-- Step 4: Rowly bind `t_pop` and `t_ae` and format it by `r2rtf`.
+  and name the dataset as `t_ae`.
+- Step 4: Row-wise combine `t_pop` and `t_ae` and format it by using `r2rtf`.

### 2021-11-17 tlf-overview.Rmd

@@
 has become a worldwide regulatory submission standard format.
-For example, the United States Food and Drug Administration (US FDA) required
+For example, the United States Food and Drug Administration (US FDA) requires
 new drug applications and biologics license applications
 [must be submitted using the eCTD format](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd).
-The Clinical Data Interchange Standards Consortium (CDISC) provided a
+The Clinical Data Interchange Standards Consortium (CDISC) provides a
 [pilot project following ICH E3 guidance](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse).
@@
 is also provided. If you are interested in more examples of clinical study reports,
-the European Medicines Agency (EMA) publishes clinical data, including
-clinical study reports submitted by pharmaceutical companies
-on the [EMA clinical data website](https://clinicaldata.ema.europa.eu/web/cdp/home).
+you can go to the [European Medicines Agency (EMA) clinical data website](https://clinicaldata.ema.europa.eu/web/cdp/home).
 
@@
 In this part, we mainly use the R packages below to illustrate
-how to deliver TLFs in CSR.
+how to deliver TLFs in a CSR.
 
@@
 `r2rtf` is an R package to create production-ready tables and figures in RTF format.
-The R package is designed to
+This R package is designed to
 
@@
 
-We provide a brief introduction of `r2rtf` and showed how to transfer
+We provide a brief introduction of `r2rtf` and showe how to transfer
 data frames into table, listing, and figures (TLFs).
@@
 we will use the dataset `r2rtf_adae` saved in the `r2rtf` package.
-This dataset contains adverse events (AE) information from a clinical trial.
+This dataset contains adverse events (AEs) information from a clinical trial.
 

### 2021-11-17 tlf-efficacy-km.Rmd
apply style fixes

--- 2021-11-16 tlf-efficacy-km.Rmd
@@
-# Efficacy Figure
+# Efficacy figure
 
-Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
+Following the [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
 primary and secondary efficacy endpoints need to be summarized
@@
 library(dplyr) # Manipulate data
-library(tidyr) # Manipulate data
 library(r2rtf) # Reporting in RTF format
-library(survival) # fit survival model
-library(survminer) # plot nice KM plot with risk table
+library(survival) # Fit survival model
+library(survminer) # K-M plot with risk table
+library(forcats) # Reorder ordinal variable level based on preference
 ```
@@
 
-To prepare the analysis, `adtte` dataset is required.
+To prepare the analysis, the `adtte` dataset is required.
 
@@
 
-First of all, to perpare the analysis ready data, filter all records for the efficacy endpoint of TTDE using `PARAMCD` (or `PARAM`, `PRAMN`), then select the survival analysis related variables:
+First, to prepare the analysis ready data, filter all records for the efficacy endpoint of TTDE using `PARAMCD` (or `PARAM`, `PRAMN`), then select the survival analysis related variables:
 
-  + Treatment Arm: `TRTP` (using corresponding numeric code `TRTAN` to re-order the levels, "Placebo" will be the reference level)
-  + Time-to-event Analysis Value: `AVAL`
-  + Event (Censoring) Status: `CNSR`
+- `TRTP`: treatment arm: (using corresponding numeric code `TRTAN` to re-order the levels, "Placebo" will be the reference level)
+- `AVAL`: time-to-event analysis value
+- `CNSR`: event (censoring) status
 
 ```{r}
-adtte_ttde <- adtte %>% filter(PARAMCD=="TTDE") %>% select(TRTP, TRTAN, AVAL, CNSR) %>%
-  mutate(TRTP = forcats::fct_reorder(TRTP, TRTAN)) # recorder levels
+adtte_ttde <- adtte %>%
+  filter(PARAMCD == "TTDE") %>%
+  select(TRTP, TRTAN, AVAL, CNSR) %>%
+  mutate(TRTP = forcats::fct_reorder(TRTP, TRTAN)) # Recorder levels
 ```
 
-## Create Kaplan-Meier Curve
+## Create Kaplan-Meier curve
 
-The `survival` package is used to create the curve data points, then `survminer` package can be used to plot the KM curve with an elegant ggplot2 style.
-There are tons of customized options for users to create a KM plot with risk table and testing result (nominal p-value).
-More details can be found in  https://rpkgs.datanovia.com/survminer/reference/ggsurvplot.html
+The `survival` package is used to create the curve data points, then `survminer` package can be used to plot the K-M curve with an elegant ggplot2 style.
+There are tons of customized options for users to create a K-M plot with risk table and testing result (nominal p-value).
+More details can be found in `?survminer::ggsurvplot`.
 
-User can directly use `Surv` with formula, or use a fitted Cox model with `coxph` function if also want to obtain other information, i.e.
+Users can use `Surv()` with formula directly, or use a fitted Cox model with the `coxph()` function if also want to obtain other information, i.e.
 hazard ratio, survival rates, etc.
-[TBDev: "Analysis of Survival Endpoint Table" Chapter]
+<!-- TODO: "Analysis of Survival Endpoint Table" Chapter -->
 
 ```{r}
-fit <- survfit(Surv(AVAL, 1-CNSR) ~ TRTP, data=adtte_ttde)
+fit <- survfit(Surv(AVAL, 1 - CNSR) ~ TRTP, data = adtte_ttde)
 # sum_fit <- summary(coxph(Surv(AVAL, 1-CNSR) ~ ARM  + strata(STRATAP), data = adtte_efs))
-fig_km <- ggsurvplot(fit,  
+fig_km <- ggsurvplot(
+  fit,
   xlab = "Time in Months",
@@
   risk.table = TRUE,
-           risk.table.height = 0.3,# Useful when you have multiple groups
+  risk.table.height = 0.3, # Useful when you have multiple groups
   legend = c(0.8, 0.9), legend.title = "",
- legend.labs = sort(unique(adtte_ttde$TRTP)), censor.size=4, size = 0.8, title = "", censor.shape="|",
-
- pval = TRUE, pval.method = TRUE, pval.coord = c(0, 0.1), pval.method.coord = c(0, 0.2), pval.size = 5,
-
+  legend.labs = sort(unique(adtte_ttde$TRTP)),
+  censor.size = 4,
+  size = 0.8,
+  title = "",
+  censor.shape = "|",
+  pval = TRUE,
+  pval.method = TRUE,
+  pval.coord = c(0, 0.1),
+  pval.method.coord = c(0, 0.2),
+  pval.size = 5,
   ggtheme = theme_classic2(base_family = "Times New Roman", base_size = 18)
-
 )
@@
 
-Then we can save the figure as a `.png` file, and use `r2rtf` package to create a formatted RTF figure. More details can be found in https://merck.github.io/r2rtf/articles/example-figure.html
+Then we can save the figure as a `.png` file, and use the `r2rtf` package to create a formatted RTF figure. More details can be found in <https://merck.github.io/r2rtf/articles/example-figure.html>.
 
@@
 # Create RTF figure
-  rtf_read_png("tlf/fig_km.png") %>%              # read PNG files from the file path
-  rtf_title("Time to Dermatologic Event by Treatment Group", "All Participants") %>%     # add title or subtitle
-  rtf_footnote("footnote") %>%           # add footnote
-  rtf_source("[datasource: adam-adtte]") %>%  # add data source
-  rtf_figure(fig_width = 6, fig_height = 3.6) %>%      # set proportional figure size to the original PNG figure size
-  rtf_encode(doc_type = "figure") %>%    # encode rtf as figure
+rtf_read_png("tlf/fig_km.png") %>% # Read PNG files from the file path
+  rtf_title("Time to Dermatologic Event by Treatment Group", "All Participants") %>% # Add title or subtitle
+  rtf_footnote("footnote") %>% # Add footnote
+  rtf_source("[datasource: adam-adtte]") %>% # Add data source
+  rtf_figure(fig_width = 6, fig_height = 3.6) %>% # Set proportional figure size to the original PNG figure size
+  rtf_encode(doc_type = "figure") %>% # Encode rtf as figure
   write_rtf(file = "tlf/fig_km.rtf")
@@
 
-In conclusion, the procedure to generate the above efficacy KM plot is summarized as follows.
+In conclusion, the procedure to generate the above efficacy K-M plot is summarized as follows.
 
-- Step 1: Read the data into R, i.e., `adtte`.
-- Step 2: Define the analysis ready dataset. In this example, we define the analysis dataset for TTDE endpoint `adtte_ttde`.
-- Step 3: Create the curve's data points, using `Surv` formula directly or fitted survival models beforehand. Then plot and customize the KM curve with risk table using `survminer` package. 
+- Step 1: Read the data `adtte` into R.
+- Step 2: Define the analysis-ready dataset. In this example, we define the analysis dataset for the TTDE endpoint `adtte_ttde`.
+- Step 3: Create the curve's data points using the `Surv()` formula directly or fitted survival models beforehand. Then plot and customize the K-M curve with risk table using the `survminer` package.
 - Step 4: Save the output as a PNG file, then format it into an RTF table.

### 2021-11-17 tlf-efficacy-km.Rmd
update renv.lock

@@
 
-First, to prepare the analysis ready data, filter all records for the efficacy endpoint of TTDE using `PARAMCD` (or `PARAM`, `PRAMN`), then select the survival analysis related variables:
+First, to prepare the analysis ready data, 
+filter all records for the efficacy endpoint of time to event of interest (`TTDE`) 
+using `PARAMCD` (or `PARAM`, `PRAMN`), then select the survival analysis related variables:
 
@@
 
-The `survival` package is used to create the curve data points, then `survminer` package can be used to plot the K-M curve with an elegant ggplot2 style.
-There are tons of customized options for users to create a K-M plot with risk table and testing result (nominal p-value).
+The `survival` package is used to create the curve data points, 
+then `survminer` package can be used to plot the K-M curve with an elegant `ggplot2` style. 
+There are tons of customized options for users to create a K-M plot
+with risk table and testing result (nominal p-value). 
 More details can be found in `?survminer::ggsurvplot`.
 
-Users can use `Surv()` with formula directly, or use a fitted Cox model with the `coxph()` function if also want to obtain other information, i.e.
-hazard ratio, survival rates, etc.
+Users can use `Surv()` with formula directly, 
+or use a fitted Cox model with the `coxph()` function 
+if we also want to obtain other information, i.e. hazard ratio, survival rates, etc. 
 <!-- TODO: "Analysis of Survival Endpoint Table" Chapter -->
@@
 fit <- survfit(Surv(AVAL, 1 - CNSR) ~ TRTP, data = adtte_ttde)
-# sum_fit <- summary(coxph(Surv(AVAL, 1-CNSR) ~ ARM  + strata(STRATAP), data = adtte_efs))
+```
+
+```{r}
+sum_fit <- summary(coxph(Surv(AVAL, 1-CNSR) ~ ARM  + strata(STRATAP), data = adtte_efs))
+```
+
+```{r}
 fig_km <- ggsurvplot(

### 2021-11-17 submission-overview.Rmd
update format

@@
 
-> "Sponsors should provide the software programs used to create all ADaM
+::: {.rmdnote}
+"Sponsors should provide the software programs used to create all ADaM
 datasets and **generate tables and figures associated with primary and
@@
 however, executable file extensions should not be used."
+:::
 

### 2021-11-17 tlf-ae-summary.Rmd
fix minor typos

@@
 Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need to the number of participants were included in each safety analysis in Section 12.2, Adverse Events (AEs).
+we need the number of participants were included in each safety analysis in Section 12.2, Adverse Events (AEs).
 

### 2021-11-17 tlf-disposition.Rmd
fix minor typos

@@
 
-In this chapter, we show how to the create a typical disposition table.
+In this chapter, we show how to create a typical disposition table.
 
@@
 For a disposition table, all the required information is saved in a Subject-level Analysis Dataset (ADSL).
-This dataset is provided in `sas7bdat` format, which is a SAS data format currently used in many clinical trial analysis and report.
+This dataset is provided in `sas7bdat` format, which is a SAS data format currently used in many clinical trial analysis and reporting.
 The `haven` package is able to read the dataset, while maintaining its attributes (e.g., variable labels).

### 2021-11-17 tlf-overview.Rmd
fix minor typos

@@
 There are also books to introduce tidyverse.
-We assume the reader to have experience in using `tidyverse` in this book.
+We assume the reader have experience in using `tidyverse` in this book.
 
@@
 
-We provide a brief introduction of `r2rtf` and showe how to transfer
+We provide a brief introduction of `r2rtf` and show how to transfer
 data frames into table, listing, and figures (TLFs).
@@
 
-`r2rtf` support 26 different border types. The details can be found on
+`r2rtf` supports 26 different border types. The details can be found on
 the [r2rtf package website](https://merck.github.io/r2rtf/articles/rtf-row.html#border-type).

### 2021-11-17 tlf-ae-summary.Rmd
Update a few wording
Co-authored-by: yilong zhang <elong0527@gmail.com>

@@
 Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
-we need the number of participants were included in each safety analysis in Section 12.2, Adverse Events (AEs).
+we summarize number of participants that were included in each safety analysis in Section 12.2, Adverse Events (AEs).
 

### 2021-11-17 tlf-overview.Rmd

--- 2021-11-17 tlf-overview.Rmd
@@
 
-> FDA's PDF specification is a general reference. Each organization can define
-> more specific TLF format requirements that can be different from the examples in this book.
+::: {.rmdnote}
+FDA's PDF specification is a general reference. Each organization can define
+more specific TLF format requirements that can be different from the examples in this book.
+:::
 
@@
 - [r2rtf](https://merck.github.io/r2rtf/): create RTF outputs
+
+::: {.rmdnote}
+There are other R packages to create TLFs in ASCIII, RTF and Word format.
+For example, `rtables`, `huxtable`, `pharmaRTF`, `gt`, `officer`, `flextable` etc.
+Here we focus on `r2rtf` to illustrate the concept.  
+Reader is encourage to explore other R packages to find the proper tools to fit your purpose. 
+:::
 
@@
 Those features will be introduced when we first use them in the rest of the chapters.
-
-There are other R packages that can create TLFs in `.rtf` or `.docx` format.
-Interested readers can refer "Microsoft/LibreOffice Formats" section in
-[CRAN Task View: Reproducible Research](https://cran.r-project.org/web/views/ReproducibleResearch.html)
-for an overview.

### 2021-11-19 index.Rmd
automate rtf to pdf conversion

@@
 options(dplyr.summarise.inform = FALSE)
-file.copy("static/CNAME", "_book", overwrite = TRUE)
-fs::dir_copy("slides/", "_book/slides", overwrite = TRUE)
 ```
@@
 )
+```
+
+```{r, include=FALSE}
+source("common.R")
 ```

### 2021-11-19 common.R

@@ -1,0 +1,7 @@
+rtf2pdf <- function(input) {
+  input <- normalizePath(input)
+  x <- "export LD_LIBRARY_PATH=:/usr/lib/libreoffice/program:/usr/lib/x86_64-linux-gnu/"
+  y <- paste0("libreoffice --invisible --headless --nologo --convert-to pdf --outdir tlf/ ", input)
+  z <- paste(x, y, sep = " && ")
+  if (Sys.getenv("GITHUB_ACTIONS") != "") system(z) else invisible(NULL)
+}

### 2021-11-19 post-processing.R

@@ -1,0 +1,3 @@
+fs::dir_copy("tlf/", "_book/tlf", overwrite = TRUE)
+fs::dir_copy("slides/", "_book/slides", overwrite = TRUE)
+file.copy("static/CNAME", "_book", overwrite = TRUE)

### 2021-11-19 tlf-efficacy-km.Rmd
update tlf-efficacy-km.Rmd
- update file format
automate rtf to pdf conversion
improve output figure details

@@
 in Section 11.4, Efficacy Results and Tabulations of Individual Participant.
+
+```{r, include=FALSE}
+source("common.R")
+```
 
@@
 
-```{r, out.width = "100%", out.height = "450px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/fig_km.png")
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/fig_km.pdf")
 ```
@@
 ```{r}
-sum_fit <- summary(coxph(Surv(AVAL, 1-CNSR) ~ ARM  + strata(STRATAP), data = adtte_efs))
-```
-
-```{r}
 fig_km <- ggsurvplot(
@@
   pval.size = 5,
+  font.family = "Times New Roman",
   ggtheme = theme_classic2(base_family = "Times New Roman", base_size = 18)
@@
 
-Then we can save the figure as a `.png` file, and use the `r2rtf` package to create a formatted RTF figure. More details can be found in <https://merck.github.io/r2rtf/articles/example-figure.html>.
+Then we can save the figure as a `.png` file, 
+and use the `r2rtf` package to create a formatted RTF figure. 
+More details can be found in <https://merck.github.io/r2rtf/articles/example-figure.html>.
 
@@
 # Save as PNG file
-ggsave("tlf/fig_km.png", plot = print(fig_km, newpage = FALSE), width = 10, height = 6)
+grid.draw.ggsurvplot <- function(x) survminer:::print.ggsurvplot(x, newpage = FALSE)
+ggsave("tlf/fig_km.png", plot = fig_km, width = 10, height = 6)
 
@@
 
+```{r, include=FALSE}
+rtf2pdf("tlf/fig_km.rtf")
+```
+
 ```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
@@
 - Step 2: Define the analysis-ready dataset. In this example, we define the analysis dataset for the TTDE endpoint `adtte_ttde`.
-- Step 3: Create the curve's data points using the `Surv()` formula directly or fitted survival models beforehand. Then plot and customize the K-M curve with risk table using the `survminer` package.
+- Step 3: Create the curve's data points using the `Surv()` formula directly 
+or fitted survival models beforehand. Then plot and customize the K-M curve with risk table using the `survminer` package.
 - Step 4: Save the output as a PNG file, then format it into an RTF table.

### 2021-11-19 slides/render.R
add china-r slides

@@ -1,1 +1,2 @@
 rmarkdown::render("slides/workshop-slides.Rmd")
+rmarkdown::render("slides/china-r.Rmd")

### 2021-11-19 slides/china-r.Rmd
Deploying to gh-pages from @ elong0527/r4csr@aebf17558a8a9e42e87d6f592ca578a896064c0d 🚀

---
title: "R for Clinical Study Reports and Submission"
subtitle: "The China-R Conference"
author: "Yilong Zhang, Nan Xiao, Yujie Zhao, Keaven Anderson"
date: "November 2021"
output:
  mkdocs::mk_ioslides_presentation:
    mk_logo: "msd"
    mk_col_title: "white"
    mk_col_section: "teal"
    mk_class: "public"
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(r2rtf)
library(ggplot2)
data("r2rtf_adae")
```

## Disclaimer

- All opinions expressed are those of the presenter and not
  Merck Sharp & Dohme Corp., a subsidiary of Merck & Co., Inc., Kenilworth, NJ, USA.

- Some slides need to be scrolled down to see the full content.

## Background

- Clinical study report is a key deliverable for clinical trials to regulatory agencies. (e.g., FDA, CFDA)
  + [ICH E3 Structure and Content of Clinical Study Reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf)

- We try to fill in gaps to streamline workflow using R:
  + To develop, validate, deliver analysis results.
  + To submit analysis results to regulatory agencies in [eCTD format](https://en.wikipedia.org/wiki/Electronic_common_technical_document).

- Focus on table, listing, figure (TLFs) delivered in RTF/Microsoft Word format
  + In the pharmaceutical industry, RTF/Microsoft Word play a central role in preparing clinical study reports.
  + Different organization can have different table standards

## Clarification from FDA

- [FDA: Statistical Software Clarifying Statement](https://www.fda.gov/files/about%20fda/published/Statistical-Software-Clarifying-Statement-PDF.pdf)

"FDA does not require use of any specific software for statistical analyses, and statistical software
is not explicitly discussed in Title 21 of the Code of Federal Regulations [e.g., in 21CFR part
11]. However, the software package(s) used for statistical analyses should be fully documented
in the submission, including version and build identification."

## Requirement from FDA

- [FDA Study Data Technical Conformance Guide](https://www.fda.gov/media/88173/download):

> "Sponsors should provide the software programs used to create all ADaM
datasets and **generate tables and figures associated with primary and
secondary efficacy analyses**. Furthermore, sponsors should submit software
programs used to generate additional information included in Section 14
CLINICAL STUDIES of the Prescribing Information (PI)26 if applicable.
**The specific software utilized should be specified in the ADRG**.
The main purpose of requesting the submission of these programs
**is to understand the process by which the variables for the respective
analyses were created and to confirm the analysis algorithms**.
Sponsors should submit software programs in **ASCII text format**;
however, executable file extensions should not be used."

## Philosophy

We share the same philosophy described in
[Section 1.1 of R Packages book](https://r-pkgs.org/intro.html#intro-phil) and quote here.

- "Anything that can be automated, should be automated."
- "Do as little as possible by hand. Do as much as possible with functions."
- "The goal is to spend your time thinking about what you want to do
rather than thinking about the minutiae of package structure."

## Tools for study design

> Details will be skipped in today's talk.

Tools:

- [`gsDesign`](https://github.com/keaven/gsDesign): an R package for group sequential design under proportional hazards.
- [`simtrial`](https://github.com/Merck/simtrial): an experimental R package for simulation-based group sequential design under non-proportional hazards
- [`gsDesign2`](https://github.com/Merck/gsDesign2) and
[`gsdmvn`](https://github.com/Merck/gsdmvn):
an experimental R package for analytical-based group sequential design under proportional hazards.

Bookdown: <https://keaven.github.io/gsd-deming/>

Training: [December 6th, Deming Conference 2021](https://demingconference.org/programs/2021-program/)

## Tools for reporting and submission

Tools:

- [`r2rtf`](https://merck.github.io/r2rtf/): create production-ready tables and figures in RTF format.
- [`pkglite`](https://merck.github.io/pkglite/): represent and exchange R package source code as text files.
- `cleanslate` (under internal validation): create portable R environments.

Bookdown: <https://r4csr.org/>

Training: [December 2nd, Short course for ASA Princeton-Trenton Chapter](https://rutgers.webex.com/rutgers/j.php?RGID=rc0a71b2f239233a7850a121cf972045f)

## r2rtf: design

`r2rtf` is designed to:

- Generate highly customized tables
- Limit package dependency
- Target regulatory deliverable
- Support pipes (`%>%`)

## r2rtf: minimal example

`r2rtf` is designed to be pipe-friendly (`%>%`)

```{r, echo = TRUE, eval = FALSE}
head(iris) %>%
  rtf_body() %>%           # Step 1 Add table attributes
  rtf_encode() %>%         # Step 2 Convert attributes to RTF encode
  write_rtf("minimal.rtf") # Step 3 Write to a .rtf file
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("minimal.pdf")
```

## r2rtf: function illustration

```{r, out.width="80%", echo=FALSE, fig.align="center"}
knitr::include_graphics("function-summary.png")
```

## pkglite: compact package representations

`pkglite` reimagines the way to represent R packages.

- A **tool** for packing and restoring R packages as plaintext
  assets that are easy to store, transfer, and review
- A **grammar** for specifying the file packing scope that is
  functional, precise, and extendable
- A **standard** for exchanging the packed asset that is
  unambiguous, human-friendly, and machine-readable

```{r, eval=FALSE}
library("pkglite")

"/path/to/pkg/" %>%
  collate(file_ectd(), file_auto("inst/")) %>%
  pack()

pack(
  "/path/to/pkg1/" %>% collate(file_ectd()),
  "/path/to/pkg2/" %>% collate(file_ectd()),
  output = "/path/to/pkglite.txt"
)

"/path/to/pkglite.txt" %>% unpack(output = "/path/to/output/", install = TRUE)
```

Website: <https://merck.github.io/pkglite/>

## cleanslate: portable R environments

- Create a project folder with specific context (`.Rproj`, `.Rprofile`, `.Renviron`)
- Install a specific version of R into the project folder
- Install a specific version of Rtools into the project folder
- (without administrator privileges)

```{r, eval=FALSE}
library("cleanslate")

"portable-project/" %>%
  use_project(repo = "https://url/snapshot/2021-11-20/") %>%
  use_rprofile() %>%
  use_renviron() %>%
  use_r_version(version = "4.1.1") %>%
  use_rtools(version = "rtools40")
```

## Summary of workflow as user stories

Within a regulatory R environment:

- As a statistician, I use `tidyverse`, `r2rtf`, and internal tools to define the mock-up table, listing and figure (TLFs) for statistical analysis of a clinical trial.
  + More than 100 TLFs for efficacy and safety of a drug or vaccine.

- As a programmer, I use `tidyverse`, `r2rtf`, and internal tools to develop and/or validate analysis results based on mock-up TLFs.

- As a statistican/programmer, I use `pkglite` and internal tools to prepare proprietary R packages and analysis R scripts for eCTD submission package.

- As an internal/external reviewer, I use `cleanslate` to re-construct a portable environment (if required) to reproduce analysis results.

More details: <https://r4csr.org/>

## Folder structure

We recommended to use **R package structure** to organize standard tools, analysis projects, and Shiny apps.

- Consistency: everyone works on the same folder structure.
- Reproducibility: analysis can be executed and reproduced by different team members months/years later.
- Automation: automatically check the integration of a project.
- Compliance: reduce compliance issues.

More details: <https://r4csr.org/folder.html>

Example: <https://github.com/elong0527/esubdemo>

## Project management

- Setting up for success
  - Work as a team
  - Design clean code architecture
  - Set capability boundaries
  - Contribute to the community

- The Software Development Life Cycle
  - Planning: define the scope of a project
  - Development: mplement target deliverables
  - Validation: verify target deliverables
  - Operation: deliver work to stakeholders

More details: <https://r4csr.org/manage.html>

## Cross-industry collaborations

- R Validation Hub: <https://www.pharmar.org/>
  + Focus on designing a framework that assesses the quality of an R package
  + Presentations: <https://www.pharmar.org/present/>

- R-based submission pilots to FDA:
  + <https://rconsortium.github.io/r-submission-site/>
  + Focus on improving practices of R-based clinical trial regulatory submission.

- R/Pharma: <https://rinpharma.com/>
  + Annual conference focus on the use of R in the development of pharmaceuticals.

- R Consortium R Adoption Seminar Series
  + <https://www.r-consortium.org/webinars>

## Future directions

- Enhance compliance, reproducibility, traceability, and automation
  + Automation of analysis, documentation, review, and testing
  + Linkage among data, TLFs, and final reports
  + R package qualification
- Enable advanced study design and statistical methods
- Introduce interactive visualization and reporting (with/without backend server)

## Join us

- Contact us for R developer or Methodology Research Positions:
  + Yilong Zhang: <https://elong0527.github.io/>
  + Nan Xiao: <https://nanx.me/>
  + Keaven Anderson: <https://www.linkedin.com/in/keaven-anderson-09aa7538/>

- Multiple open positions (statisticians/programmers) in China, EU, and US
  + US: <https://jobs.merck.com/bards>
  + China/EU: <https://jobs.msd.com/cn/zh>

# Thank you {.thank-you}

### 2021-11-28 submission-package.Rmd
update adrg and arm tables with real-world formatting

@@
 
-The table below contains the software version and instructions
+The tables below contain the software version and instructions
 for executing the R analysis output programs:
@@
 
-| Software or Package                            | Version |
-|------------------------------------------------|---------|
-| R                                              | 4.1.1   |
-| dplyr                                          | 1.0.7   |
-| r2rtf                                          | 0.3.0   |
-| pkglite                                        | 0.2.0   |
-| ...                                            | ...     |
-| R package versions based on CRAN at 2021-08-06 |         |
+```{r, echo=FALSE}
+library("kableExtra")
+
+df1 <- data.frame(
+  program = c(
+    "tlf-01-disposition.txt",
+    "tlf-02-population.txt",
+    "tlf-03-baseline.txt",
+    "tlf-04-efficacy.txt",
+    "tlf-05-ae-summary.txt",
+    "tlf-06-ae-spec.txt"
+  ),
+  table = c(
+    "Table x.y.z"
+  ),
+  desc = c(
+    "Disposition of Patients",
+    "Participants Accounting in Analysis Population (All Participants Randomized)",
+    "Participant Baseline Characteristics (All Participants Randomized)",
+    "ANCOVA of Change from Baseline Glucose (mmol/L) at Week 24 <br> LOCF <br> Efficacy Analysis Population",
+    "Analysis of Adverse Event Summary (Safety Analysis Population)",
+    "Analysis of Participants With Specific Adverse Events (Safety Analysis Population)"
+  ),
+  stringsAsFactors = FALSE
+)
+
+names(df1) <- c("Program Name", "Output Table", "Title")
+
+df1 %>%
+  kbl(escape = FALSE) %>%
+  kable_classic(full_width = FALSE, html_font = "'Times New Roman', Times, serif", font_size = 18) %>%
+  column_spec(1, extra_css = "border: 1px solid #000; text-align: center;", width = "18rem") %>%
+  column_spec(2, extra_css = "border: 1px solid #000; text-align: center;", width = "16rem") %>%
+  column_spec(3, extra_css = "border: 1px solid #000;", width = "28rem") %>%
+  row_spec(0, background = "#DFDFDF", bold = TRUE, extra_css = "border: 1px solid #000; text-align: center;")
+
+df2 <- data.frame(
+  package = c("pkglite", "haven", "dplyr", "tidyr", "emmeans", "r2rtf"),
+  version = NA,
+  description = c(
+    "Prepare submission package",
+    "Read SAS datasets",
+    "Manipulate datasets",
+    "Manipulate datasets",
+    "Least-squares means estimation",
+    "Create RTF tables"
+  ),
+  stringsAsFactors = FALSE
+)
+
+x <- available.packages(contriburl = contrib.url("https://mran.microsoft.com/snapshot/2021-08-06/"))
+df2$version <- x[match(df2$package, x[, "Package"]), "Version"]
+
+names(df2) <- c(
+  "Open-Source R Analysis Package",
+  "Package Version",
+  "Analysis Package Description"
+)
+
+df2 %>%
+  kbl() %>%
+  kable_classic(full_width = FALSE, html_font = "'Times New Roman', Times, serif", font_size = 18) %>%
+  column_spec(1, extra_css = "border: 1px solid #000; text-align: center;", width = "18rem") %>%
+  column_spec(2, extra_css = "border: 1px solid #000; text-align: center;", width = "16rem") %>%
+  column_spec(3, extra_css = "border: 1px solid #000;", width = "28rem") %>%
+  row_spec(0, background = "#DFDFDF", bold = TRUE, extra_css = "border: 1px solid #000; text-align: center;") %>%
+  footnote(general = "R package versions based on CRAN on 2021-08-06", general_title = "")
+
+df3 <- data.frame(
+  package = "esubdemo",
+  version = "0.1.0",
+  description = "A demo package for analysis and reporting of clinical trials",
+  stringsAsFactors = FALSE
+)
+
+names(df3) <- c(
+  "Proprietary R Analysis Package",
+  "Package Version",
+  "Analysis Package Description"
+)
+
+df3 %>%
+  kbl() %>%
+  kable_classic(full_width = FALSE, html_font = "'Times New Roman', Times, serif", font_size = 18) %>%
+  column_spec(1, extra_css = "border: 1px solid #000; text-align: center;", width = "18rem") %>%
+  column_spec(2, extra_css = "border: 1px solid #000; text-align: center;", width = "16rem") %>%
+  column_spec(3, extra_css = "border: 1px solid #000;", width = "28rem") %>%
+  row_spec(0, background = "#DFDFDF", bold = TRUE, extra_css = "border: 1px solid #000; text-align: center;")
+```
 
@@
 
-+--------------------------------------+-------------+----------------------+-------------------------+---------------------------------------+
-| Table Reference                      | Table Title | Programming Language | Program Name (programs) | Input File Name / Analysis (datasets) |
-+======================================+=============+======================+=========================+=======================================+
-| [Ref. x.y.z: P001ZZZ9999: Table 1-1] | ...         | R                    | tlf-01-disposition.txt  | adsl.xpt                              |
-+--------------------------------------+-------------+----------------------+-------------------------+---------------------------------------+
-| ...                                  | ...         | ...                  | ...                     | ...                                   |
-+--------------------------------------+-------------+----------------------+-------------------------+---------------------------------------+
+```{r, echo=FALSE}
+df4 <- data.frame(
+  "ref" = c("[Ref. x.y.z: P001ZZZ9999: Table 1-1]", "..."),
+  "title" = c("Disposition of Patients", "..."),
+  "pl" = c("R", "..."),
+  "pn" = c("tlf-01-disposition.txt", "..."),
+  "input" = c("adsl.xpt", "..."),
+  stringsAsFactors = FALSE
+)
+
+names(df4) <- c(
+  "Table Reference",
+  "Table Title",
+  "Programming Language",
+  "Program Name (programs)",
+  "Input File Name / Analysis (datasets)"
+)
+
+df4 %>%
+  kbl() %>%
+  kable_classic(full_width = FALSE, html_font = "'Times New Roman', Times, serif", font_size = 18) %>%
+  column_spec(1, extra_css = "border: 1px solid #000; text-align: center;") %>%
+  column_spec(2, extra_css = "border: 1px solid #000; text-align: center;") %>%
+  column_spec(3, extra_css = "border: 1px solid #000; text-align: center;") %>%
+  column_spec(4, extra_css = "border: 1px solid #000; text-align: center;") %>%
+  column_spec(5, extra_css = "border: 1px solid #000; text-align: center;") %>%
+  row_spec(0, background = "#DFDFDF", bold = TRUE, extra_css = "border: 1px solid #000; text-align: center;")
+```
 
@@
 
-|                                                       |                                                               |
-|-------------------------------------------------------|---------------------------------------------------------------|
-| Table Reference: [Ref. x.y.z: P001ZZZ9999: Table 1-1] | ...                                                           |
-| Analysis Result                                       | ...                                                           |
-| Analysis Parameters (s)                               | ...                                                           |
-| Analysis Reason                                       | ...                                                           |
-| Analysis Purpose                                      | ...                                                           |
-| ...                                                   | ...                                                           |
-| Programming Statements                                | (R version 4.1.1), [P001ZZZ9999: programs-tlf-01-disposition] |
+```{r, echo=FALSE}
+df5 <- data.frame(
+  "x1" = c(
+    "Table Reference: [Ref. x.y.z: P001ZZZ9999: Table 1-1]",
+    "Analysis Result",
+    "Analysis Parameters (s)",
+    "Analysis Reason",
+    "Analysis Purpose",
+    "...",
+    "Programming Statements"
+  ),
+  "x2" = c(
+    "...",
+    "...",
+    "...",
+    "...",
+    "...",
+    "...",
+    "(R version 4.1.1), [P001ZZZ9999: programs-tlf-01-disposition]"
+  ),
+  stringsAsFactors = FALSE
+)
+
+df5 %>%
+  kbl(format = "html") %>%
+  kable_classic(full_width = FALSE, html_font = "'Times New Roman', Times, serif", font_size = 18) %>%
+  column_spec(1, extra_css = "border: 1px solid #000;", bold = TRUE) %>%
+  column_spec(2, extra_css = "border: 1px solid #000;") %>%
+  row_spec(0, background = "#DFDFDF", bold = TRUE, extra_css = "border: 1px solid #000;") %>%
+  gsub("<thead>.*</thead>", "", .)
+```

### 2021-11-30 slides/workshop-slides.Rmd
update slides info for ASA training

@@
 title: "R for Clinical Study Reports and Submission"
-subtitle: "R/Pharma Conference 2021"
-author: "Yilong Zhang, Nan Xiao, Keaven Anderson"
+subtitle: "ASA Princeton-Trenton Chapter, 2021"
+author: "Yilong Zhang, Nan Xiao"
 date: "October 2021"

### 2021-11-30 slides/workshop-slides.Rmd
minor fixes

@@
 title: "R for Clinical Study Reports and Submission"
-subtitle: "ASA Princeton-Trenton Chapter, 2021"
+subtitle: "ASA Princeton-Trenton Chapter"
 author: "Yilong Zhang, Nan Xiao"
-date: "October 2021"
+date: "December 2021"
 output:

### 2021-12-02 slides/workshop-slides.Rmd
update slides

--- 2021-11-30 slides/workshop-slides.Rmd
@@
 
+- [ASA Princeton-Trenton Chapter](https://community.amstat.org/princetontrenton/home)
+  + Please consider to be a member of our chapter.
+
+<!--
+
 - R / Pharma organizers
@@
   + Please consider sharing stories and use cases to expand the community.
+
+-->
 
@@
 - `r2rtf` is an R package to create production-ready tables and figures in RTF format.
-The R package is designed to
 
@@
 - provide simple "verb" functions that correspond to each component of a table,
-  to help you translate a data frame to a table in RTF file.
+  to help you translate a data frame to a table in RTF format.
 - enable pipes (`%>%`).
 - only focus on the **table format**.
-  Data manipulation and analysis shall be handled by other R packages. (e.g. `tidyverse`)
+  + Data manipulation and analysis shall be handled by other R packages. (e.g. `tidyverse`)
 
@@
 
-- Let's refer to examples in the [Section 2.4.2](https://r4csr.org/tlf-overview.html#r2rtf)
+- Let's refer to examples in the [Section 1.4.2](https://r4csr.org/tlf-overview.html#r2rtf)
 

### 2021-12-10 tlf-efficacy-km.Rmd
revise the efficacy figure chapter,  starting with KM plot using `survival` package, then `survminer` and `visR`.

@@
 library(survival) # Fit survival model
-library(survminer) # K-M plot with risk table
+library(survminer) # K-M plot with risk table; optional for user's preference
+library(visR) # K-M plot with risk table; optional for user's preference
 library(forcats) # Reorder ordinal variable level based on preference
@@
 For the survival analysis in efficacy, time to dermatologic event (TTDE) will be analyzed.
-
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/fig_km.pdf")
-```
 
@@
   select(TRTP, TRTAN, AVAL, CNSR) %>%
-  mutate(TRTP = forcats::fct_reorder(TRTP, TRTAN)) # Recorder levels
+  mutate(TRTP = forcats::fct_reorder(TRTP, TRTAN), # Recorder levels
+         AVAL_m = AVAL/30.4367) # convert Day to Month
 ```
@@
 
-The `survival` package is used to create the curve data points, 
-then `survminer` package can be used to plot the K-M curve with an elegant `ggplot2` style. 
-There are tons of customized options for users to create a K-M plot
-with risk table and testing result (nominal p-value). 
-More details can be found in `?survminer::ggsurvplot`.
-
-Users can use `Surv()` with formula directly, 
+The `survival` package is used to create the curve data points, users can use `Surv()` with formula directly, 
 or use a fitted Cox model with the `coxph()` function 
@@
 ```{r}
-fit <- survfit(Surv(AVAL, 1 - CNSR) ~ TRTP, data = adtte_ttde)
+# Fit survival model, convert the time value from Days to Month
+fit <- survfit(Surv(AVAL_m, 1 - CNSR) ~ TRTP, data = adtte_ttde)
 ```
 
+There are various ways to create a KM plot.
+In the following, we will illustrate how to use `survival`, `survminer` or `visR` to generate a KM plot with a risk table, respectively.
+There are pro and con for using these package.
+Compared with the basic `survival` package, `survminer` or `visR` package can be easily used to customize a nice-look KM plot.
+However, the `survminer` or `visR` package has relative heavy dependencies (Ref: https://github.com/elong0527/r4csr/issues/50#issuecomment-985109643), which may increase risk of reproducibility (Section 4.2 of the [White Paper](https://www.pharmar.org/white-paper/)).
+
+### Plot with `survival`
+
+Firstly, create a function `risk_tab()` to count cumulative number at risk, then display them as the risk table.
+
 ```{r}
-fig_km <- ggsurvplot(
+risk_tab <- function(tlim, time) {
+  n_risk <- NULL
+  for (i in tlim){
+    n_risk <- c(n_risk, sum(time >= i))
+  }
+  return(n_risk)
+}
+```
+
+Then create the KM plot and save as a `.png` file.
+
+```{r}
+# Save as a PNG file
+png(file = "tlf/fig_km1.png", width = 3000, height = 2000, res = 400)
+
+par(mfrow = c(1, 1), mar = c(9, 11, 1, 1), family = "Times") # Set plot margins
+
+plot(fit, xlab = "Time in Months", ylab = "Survival probability", lwd = 2, mark.time = T, col = c(2, 3, 4), lty = c(1, 2, 3))
+legend("topright", lwd = 2, cex = 0.9, col = c(2, 3, 4), lty = c(1, 2, 3), legend = levels(adtte_ttde$TRTP), bty = "n")
+mtext("Number at risk", side = 1, line = 4, at = 0)
+grid <- 0:6
+mtext(c("Placebo", risk_tab(grid,adtte_ttde$AVAL[adtte_ttde$TRTP=="Placebo"]/30.4367)), side = 1, adj = c(1, rep(NA, length(grid))), line = 5, at = c(-0.8, grid))
+mtext(c("Xanomeline Low Dose", risk_tab(grid,adtte_ttde$AVAL[adtte_ttde$TRTP=="Xanomeline Low Dose"]/30.4367)), side = 1, adj = c(1, rep(NA, length(grid))), line = 6, at = c(-0.8, grid))
+mtext(c("Xanomeline High Dose", risk_tab(grid,adtte_ttde$AVAL[adtte_ttde$TRTP=="Xanomeline High Dose"]/30.4367)), side = 1, adj = c(1, rep(NA, length(grid))), line = 7, at = c(-0.8, grid))
+
+dev.off()
+```
+
+Then we can use the `r2rtf` package to create a formatted RTF figure. 
+More details can be found in <https://merck.github.io/r2rtf/articles/example-figure.html>.
+
+```{r}
+# Create RTF figure
+rtf_read_png("tlf/fig_km1.png") %>% # Read the PNG file from the file path
+  rtf_title("Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group", "All Participants") %>% # Add title or subtitle
+  rtf_footnote("footnote") %>% # Add footnote
+  rtf_source("[datasource: adam-adtte]") %>% # Add data source
+  rtf_figure(fig_width = 6, fig_height = 4) %>% # Set proportional figure size to the original PNG figure size
+  rtf_encode(doc_type = "figure") %>% # Encode rtf as figure
+  write_rtf(file = "tlf/fig_km1.rtf")
+```
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/fig_km1.pdf")
+```
+
+### Plot with `survminer`
+
+Alternatively, the `survminer` package can be also used to plot the KM curve with an elegant `ggplot2` style. 
+There are tons of customized options for users to create a K-M plot
+with risk table and testing result (nominal p-value). 
+More details can be found in `?survminer::ggsurvplot`.
+
+```{r}
+fig_km2 <- ggsurvplot(
   fit,
@@
   ylab = "Survival Probability",
-  break.time.by = 30.4367,
-  xscale = "d_m",
-  xlim = c(0, 200),
+  break.time.by = 1,
+#  xscale = "d_m",
+#  xlim = c(0, 200),
   linetype = c(1, 2, 3),
@@
 grid.draw.ggsurvplot <- function(x) survminer:::print.ggsurvplot(x, newpage = FALSE)
-ggsave("tlf/fig_km.png", plot = fig_km, width = 10, height = 6)
+ggsave("tlf/fig_km2.png", plot = fig_km2, width = 10, height = 6)
 
@@
 rtf_read_png("tlf/fig_km.png") %>% # Read PNG files from the file path
-  rtf_title("Time to Dermatologic Event by Treatment Group", "All Participants") %>% # Add title or subtitle
+  rtf_title("Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group", "All Participants") %>% # Add title or subtitle
   rtf_footnote("footnote") %>% # Add footnote
@@
   rtf_encode(doc_type = "figure") %>% # Encode rtf as figure
-  write_rtf(file = "tlf/fig_km.rtf")
-```
-
-```{r, include=FALSE}
-rtf2pdf("tlf/fig_km.rtf")
+  write_rtf(file = "tlf/fig_km2.rtf")
 ```
@@
 ```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/fig_km.pdf")
+knitr::include_graphics("tlf/fig_km2.pdf")
+```
+
+### Plot with `visR`
+
+Equivalently, the `visR` package can be also used to plot the KM curve fitting CDISC ADaM standard.
+In the following we will illustrate how to create KM plot in RTF format using `visR`.
+Firstly, similar to the previous sections, create a png file.
+
+```{r}
+# Estimate KM
+surv_mod <- visR::estimate_KM(data = adtte_ttde, strata = "TRTP", AVAL = "AVAL_m")
+# save plot
+ggplot2::theme_set(theme_bw() + theme(text = element_text(family = 'serif')))
+
+png("tlf/fig_km3.png", width = 3000, height = 2000, res = 400)
+KM <- visR::visr(surv_mod,
+           y_label = "Survival Probability\n",
+           x_label = "Time in Months",
+           x_ticks = 0:6,
+           y_ticks = seq(0,1,0.1),
+           legend_position = "bottom") %>%
+    add_CNSR()
+
+KM %>%
+    visR::add_risktable(group = "statlist")  # seemingly current visR::add_risktable() does not support different fonts, will open an issue in the `visR` Github
+
+dev.off()
+```
+
+Then create RTF file
+
+```{r}
+# Create RTF figure
+rtf_read_png("tlf/fig_km3.png") %>% # Read the PNG file from the file path
+  rtf_title("Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group", "All Participants") %>% # Add title or subtitle
+  rtf_footnote("footnote") %>% # Add footnote
+  rtf_source("[datasource: adam-adtte]") %>% # Add data source
+  rtf_figure(fig_width = 6, fig_height = 4) %>% # Set proportional figure size to the original PNG figure size
+  rtf_encode(doc_type = "figure") %>% # Encode rtf as figure
+  write_rtf(file = "tlf/fig_km3.rtf")
+```
+
+Alternatively, we can also directly generate PDF file for KM plot with title and footnote (Ref: https://github.com/RConsortium/submissions-pilot1/blob/main/vignettes/tlf-kmplot.Rmd).
+
+```{r}
+# save plot as PDF file
+pdf.options(reset = TRUE, onefile = FALSE)
+pdf("tlf/fig_km3.pdf", width = 12, height = 6)
+KM <- visR::visr(surv_mod,
+           y_label = "Survival Probability\n",
+           x_label = "Time in Months",
+           x_ticks = 0:6,
+           y_ticks = seq(0,1,0.1)) %>%
+    add_CNSR()
+
+KM1 <- KM %>%
+    visR::add_risktable(group = "statlist")  # seemingly current visR::add_risktable() does not support different fonts, will open an issue in the `visR` Github
+
+title <- cowplot::ggdraw() + 
+  cowplot::draw_label(
+    "Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group\nAll Participants",
+    fontfamily = "serif",
+    size=12
+  ) 
+caption <- cowplot::ggdraw() + 
+  cowplot::draw_label(
+    paste0("\nProgram: tlf-efficacy-km.Rmd [", Sys.time(), "]\n[Datasource: adam-adtte]"),
+    fontfamily = "serif",
+    size=12
+  )
+
+KM2 <- cowplot::plot_grid(
+  title, KM1, caption,
+  ncol = 1,
+  rel_heights = c(0.1,0.8,0.1)
+) #%>%
+ # visR::apply_theme(theme_times)
+print(KM2)
+dev.off()
+```
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("pdf/fig_km3.pdf")
 ```
@@
 - Step 3: Create the curve's data points using the `Surv()` formula directly 
-or fitted survival models beforehand. Then plot and customize the K-M curve with risk table using the `survminer` package.
-- Step 4: Save the output as a PNG file, then format it into an RTF table.
+or fitted survival models beforehand. Then plot and customize the K-M curve with risk table using `survival`, `survminer` or `visR` package.
+- Step 4: Save the output as a PNG file, then format it into an RTF table. (The output format is based on user's preference)

### 2021-12-10 tlf-efficacy-km.Rmd
fix minor typo

@@
 # Create RTF figure
-rtf_read_png("tlf/fig_km.png") %>% # Read PNG files from the file path
+rtf_read_png("tlf/fig_km2.png") %>% # Read PNG files from the file path
   rtf_title("Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group", "All Participants") %>% # Add title or subtitle
@@
 ```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("pdf/fig_km3.pdf")
+knitr::include_graphics("tlf/fig_km3.pdf")
 ```

### 2021-12-10 tlf-efficacy-km.Rmd
reformat the code

@@
   mutate(TRTP = forcats::fct_reorder(TRTP, TRTAN), # Recorder levels
-         AVAL_m = AVAL/30.4367) # convert Day to Month
+         AVAL_m = AVAL / 30.4367) # convert Day to Month
 ```
@@
   n_risk <- NULL
-  for (i in tlim){
+  for (i in tlim) {
     n_risk <- c(n_risk, sum(time >= i))
@@
 # Save as a PNG file
-png(file = "tlf/fig_km1.png", width = 3000, height = 2000, res = 400)
+png(
+  file = "tlf/fig_km1.png",
+  width = 3000,
+  height = 2000,
+  res = 400
+)
 
-par(mfrow = c(1, 1), mar = c(9, 11, 1, 1), family = "Times") # Set plot margins
+par(mfrow = c(1, 1),
+    mar = c(9, 11, 1, 1),
+    family = "Times") # Set plot margins and font style
 
-plot(fit, xlab = "Time in Months", ylab = "Survival probability", lwd = 2, mark.time = T, col = c(2, 3, 4), lty = c(1, 2, 3))
-legend("topright", lwd = 2, cex = 0.9, col = c(2, 3, 4), lty = c(1, 2, 3), legend = levels(adtte_ttde$TRTP), bty = "n")
-mtext("Number at risk", side = 1, line = 4, at = 0)
+plot(
+  fit,
+  xlab = "Time in Months",
+  ylab = "Survival probability",
+  lwd = 2,
+  mark.time = T,
+  col = c(2, 3, 4),
+  lty = c(1, 2, 3)
+)
+legend(
+  "topright",
+  lwd = 2,
+  cex = 0.9,
+  col = c(2, 3, 4),
+  lty = c(1, 2, 3),
+  legend = levels(adtte_ttde$TRTP),
+  bty = "n"
+)
+mtext("Number at risk",
+      side = 1,
+      line = 4,
+      at = 0)
 grid <- 0:6
-mtext(c("Placebo", risk_tab(grid,adtte_ttde$AVAL[adtte_ttde$TRTP=="Placebo"]/30.4367)), side = 1, adj = c(1, rep(NA, length(grid))), line = 5, at = c(-0.8, grid))
-mtext(c("Xanomeline Low Dose", risk_tab(grid,adtte_ttde$AVAL[adtte_ttde$TRTP=="Xanomeline Low Dose"]/30.4367)), side = 1, adj = c(1, rep(NA, length(grid))), line = 6, at = c(-0.8, grid))
-mtext(c("Xanomeline High Dose", risk_tab(grid,adtte_ttde$AVAL[adtte_ttde$TRTP=="Xanomeline High Dose"]/30.4367)), side = 1, adj = c(1, rep(NA, length(grid))), line = 7, at = c(-0.8, grid))
+mtext(
+  c("Placebo", risk_tab(grid, adtte_ttde$AVAL_m[adtte_ttde$TRTP == "Placebo"])),
+  side = 1,
+  adj = c(1, rep(NA, length(grid))),
+  line = 5,
+  at = c(-0.8, grid)
+)
+mtext(
+  c("Xanomeline Low Dose",
+    risk_tab(grid, adtte_ttde$AVAL_m[adtte_ttde$TRTP == "Xanomeline Low Dose"])),
+  side = 1,
+  adj = c(1, rep(NA, length(grid))),
+  line = 6,
+  at = c(-0.8, grid)
+)
+mtext(
+  c("Xanomeline High Dose",
+    risk_tab(grid, adtte_ttde$AVAL_m[adtte_ttde$TRTP == "Xanomeline High Dose"])),
+  side = 1,
+  adj = c(1, rep(NA, length(grid))),
+  line = 7,
+  at = c(-0.8, grid)
+)
 
@@
 rtf_read_png("tlf/fig_km1.png") %>% # Read the PNG file from the file path
-  rtf_title("Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group", "All Participants") %>% # Add title or subtitle
+  rtf_title(
+    "Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group",
+    "All Participants"
+  ) %>% # Add title or subtitle
   rtf_footnote("footnote") %>% # Add footnote
@@
   risk.table = TRUE,
-  risk.table.height = 0.3, # Useful when you have multiple groups
-  legend = c(0.8, 0.9), legend.title = "",
+  risk.table.height = 0.3,
+  # Useful when you have multiple groups
+  legend = c(0.8, 0.9),
+  legend.title = "",
   legend.labs = sort(unique(adtte_ttde$TRTP)),
@@
 # Save as PNG file
-grid.draw.ggsurvplot <- function(x) survminer:::print.ggsurvplot(x, newpage = FALSE)
-ggsave("tlf/fig_km2.png", plot = fig_km2, width = 10, height = 6)
+# grid.draw.ggsurvplot <- function(x) survminer:::print.ggsurvplot(x, newpage = FALSE)
+ggsave("tlf/fig_km2.png",
+       plot = fig_km2,
+       width = 10,
+       height = 6)
 
@@
 rtf_read_png("tlf/fig_km2.png") %>% # Read PNG files from the file path
-  rtf_title("Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group", "All Participants") %>% # Add title or subtitle
+  rtf_title(
+    "Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group",
+    "All Participants"
+  ) %>% # Add title or subtitle
   rtf_footnote("footnote") %>% # Add footnote
@@
 ```{r}
-# Estimate KM
-surv_mod <- visR::estimate_KM(data = adtte_ttde, strata = "TRTP", AVAL = "AVAL_m")
-# save plot
+# Estimate KM data point
+surv_mod <-
+  visR::estimate_KM(data = adtte_ttde,
+                    strata = "TRTP",
+                    AVAL = "AVAL_m")
+# save plot with white background and Times New Roman font
 ggplot2::theme_set(theme_bw() + theme(text = element_text(family = 'serif')))
 
-png("tlf/fig_km3.png", width = 3000, height = 2000, res = 400)
-KM <- visR::visr(surv_mod,
+png("tlf/fig_km3.png",
+    width = 3000,
+    height = 2000,
+    res = 400)
+KM <- visR::visr(
+  surv_mod,
   y_label = "Survival Probability\n",
@@
   x_ticks = 0:6,
-           y_ticks = seq(0,1,0.1),
-           legend_position = "bottom") %>%
+  y_ticks = seq(0, 1, 0.1),
+  legend_position = "bottom"
+) %>%
   add_CNSR()
@@
 rtf_read_png("tlf/fig_km3.png") %>% # Read the PNG file from the file path
-  rtf_title("Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group", "All Participants") %>% # Add title or subtitle
+  rtf_title(
+    "Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group",
+    "All Participants"
+  ) %>% # Add title or subtitle
   rtf_footnote("footnote") %>% # Add footnote
@@
 pdf("tlf/fig_km3.pdf", width = 12, height = 6)
-KM <- visR::visr(surv_mod,
+
+KM <- visR::visr(
+  surv_mod,
   y_label = "Survival Probability\n",
@@
   x_ticks = 0:6,
-           y_ticks = seq(0,1,0.1)) %>%
+  y_ticks = seq(0, 1, 0.1)
+) %>%
   add_CNSR()
@@
     fontfamily = "serif",
-    size=12
+    size = 12
   )
@@
   cowplot::draw_label(
-    paste0("\nProgram: tlf-efficacy-km.Rmd [", Sys.time(), "]\n[Datasource: adam-adtte]"),
+    paste0(
+      "\nProgram: tlf-efficacy-km.Rmd [",
+      Sys.time(),
+      "]\n[Datasource: adam-adtte]"
+    ),
     fontfamily = "serif",
-    size=12
+    size = 12
   )
 
-KM2 <- cowplot::plot_grid(
-  title, KM1, caption,
+KM2 <- cowplot::plot_grid(title,
+                          KM1,
+                          caption,
                           ncol = 1,
-  rel_heights = c(0.1,0.8,0.1)
-) #%>%
+                          rel_heights = c(0.1, 0.8, 0.1)) #%>%
 # visR::apply_theme(theme_times)
 print(KM2)
+
 dev.off()

### 2022-01-06 tlf-efficacy-km.Rmd
simplify kmplot chapter

@@
 library(survival) # Fit survival model
-library(survminer) # K-M plot with risk table; optional for user's preference
-library(visR) # K-M plot with risk table; optional for user's preference
-library(forcats) # Reorder ordinal variable level based on preference
 ```
 
-In this chapter, we illustrate how to create a Kaplan-Meier plot in a study.
+In this chapter, we illustrate how to create a simplified Kaplan-Meier plot in a study.
 For the survival analysis in efficacy, time to dermatologic event (TTDE) will be analyzed.
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/fig_km.pdf")
+```
+
+::: {.rmdnote}
+There are R packages that can create more informative Kaplan-Meier plot such as `visR` and `survminer`.
+Interested reader can find examples in their package website. 
+:::
 
@@
 
-The `survival` package is used to create the curve data points, users can use `Surv()` with formula directly, 
-or use a fitted Cox model with the `coxph()` function 
-if we also want to obtain other information, i.e. hazard ratio, survival rates, etc. 
-<!-- TODO: "Analysis of Survival Endpoint Table" Chapter -->
+The `survival` package is used to obtain the K-M estimator.
 
@@
 
-There are various ways to create a KM plot.
-In the following, we will illustrate how to use `survival`, `survminer` or `visR` to generate a KM plot with a risk table, respectively.
-There are pro and con for using these package.
-Compared with the basic `survival` package, `survminer` or `visR` package can be easily used to customize a nice-look KM plot.
-However, the `survminer` or `visR` package has relative heavy dependencies (Ref: https://github.com/elong0527/r4csr/issues/50#issuecomment-985109643), which may increase risk of reproducibility (Section 4.2 of the [White Paper](https://www.pharmar.org/white-paper/)).
-
-### Plot with `survival`
-
-Firstly, create a function `risk_tab()` to count cumulative number at risk, then display them as the risk table.
-
-```{r}
-risk_tab <- function(tlim, time) {
-  n_risk <- NULL
-  for (i in tlim) {
-    n_risk <- c(n_risk, sum(time >= i))
-  }
-  return(n_risk)
-}
-```
-
-Then create the KM plot and save as a `.png` file.
+Using code below, we save the simplified KM plot into a `.png` file.
 
@@
 png(
-  file = "tlf/fig_km1.png",
+  file = "tlf/fig_km.png",
   width = 3000,
   height = 2000,
-  res = 400
+  res = 300
 )
-
-par(mfrow = c(1, 1),
-    mar = c(9, 11, 1, 1),
-    family = "Times") # Set plot margins and font style
 
@@
 )
-legend(
-  "topright",
-  lwd = 2,
-  cex = 0.9,
-  col = c(2, 3, 4),
-  lty = c(1, 2, 3),
-  legend = levels(adtte_ttde$TRTP),
-  bty = "n"
-)
-mtext("Number at risk",
-      side = 1,
-      line = 4,
-      at = 0)
-grid <- 0:6
-mtext(
-  c("Placebo", risk_tab(grid, adtte_ttde$AVAL_m[adtte_ttde$TRTP == "Placebo"])),
-  side = 1,
-  adj = c(1, rep(NA, length(grid))),
-  line = 5,
-  at = c(-0.8, grid)
-)
-mtext(
-  c("Xanomeline Low Dose",
-    risk_tab(grid, adtte_ttde$AVAL_m[adtte_ttde$TRTP == "Xanomeline Low Dose"])),
-  side = 1,
-  adj = c(1, rep(NA, length(grid))),
-  line = 6,
-  at = c(-0.8, grid)
-)
-mtext(
-  c("Xanomeline High Dose",
-    risk_tab(grid, adtte_ttde$AVAL_m[adtte_ttde$TRTP == "Xanomeline High Dose"])),
-  side = 1,
-  adj = c(1, rep(NA, length(grid))),
-  line = 7,
-  at = c(-0.8, grid)
-)
 
@@
 
-Then we can use the `r2rtf` package to create a formatted RTF figure. 
-More details can be found in <https://merck.github.io/r2rtf/articles/example-figure.html>.
+Now, we can use the `r2rtf` package to create a formatted RTF figure. 
+More details can be found in [r2rtf website](https://merck.github.io/r2rtf/articles/example-figure.html).
 
@@
 # Create RTF figure
-rtf_read_png("tlf/fig_km1.png") %>% # Read the PNG file from the file path
+rtf_read_figure("tlf/fig_km.png") %>% # Read the PNG file from the file path
   rtf_title(
@@
   rtf_encode(doc_type = "figure") %>% # Encode rtf as figure
-  write_rtf(file = "tlf/fig_km1.rtf")
+  write_rtf(file = "tlf/fig_km.rtf")
+```
+
+```{r, include=FALSE}
+rtf2pdf("tlf/tlf_eff.rtf")
 ```
@@
 ```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/fig_km1.pdf")
+knitr::include_graphics("tlf/fig_km.pdf")
 ```
 
-### Plot with `survminer`
-
-Alternatively, the `survminer` package can be also used to plot the KM curve with an elegant `ggplot2` style. 
-There are tons of customized options for users to create a K-M plot
-with risk table and testing result (nominal p-value). 
-More details can be found in `?survminer::ggsurvplot`.
-
-```{r}
-fig_km2 <- ggsurvplot(
-  fit,
-  xlab = "Time in Months",
-  ylab = "Survival Probability",
-  break.time.by = 1,
-  #  xscale = "d_m",
-  #  xlim = c(0, 200),
-  linetype = c(1, 2, 3),
-  risk.table = TRUE,
-  risk.table.height = 0.3,
-  # Useful when you have multiple groups
-  legend = c(0.8, 0.9),
-  legend.title = "",
-  legend.labs = sort(unique(adtte_ttde$TRTP)),
-  censor.size = 4,
-  size = 0.8,
-  title = "",
-  censor.shape = "|",
-  pval = TRUE,
-  pval.method = TRUE,
-  pval.coord = c(0, 0.1),
-  pval.method.coord = c(0, 0.2),
-  pval.size = 5,
-  font.family = "Times New Roman",
-  ggtheme = theme_classic2(base_family = "Times New Roman", base_size = 18)
-)
-```
-
-Then we can save the figure as a `.png` file, 
-and use the `r2rtf` package to create a formatted RTF figure. 
-More details can be found in <https://merck.github.io/r2rtf/articles/example-figure.html>.
-
-```{r}
-# Save as PNG file
-# grid.draw.ggsurvplot <- function(x) survminer:::print.ggsurvplot(x, newpage = FALSE)
-ggsave("tlf/fig_km2.png",
-       plot = fig_km2,
-       width = 10,
-       height = 6)
-
-# Create RTF figure
-rtf_read_png("tlf/fig_km2.png") %>% # Read PNG files from the file path
-  rtf_title(
-    "Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group",
-    "All Participants"
-  ) %>% # Add title or subtitle
-  rtf_footnote("footnote") %>% # Add footnote
-  rtf_source("[datasource: adam-adtte]") %>% # Add data source
-  rtf_figure(fig_width = 6, fig_height = 3.6) %>% # Set proportional figure size to the original PNG figure size
-  rtf_encode(doc_type = "figure") %>% # Encode rtf as figure
-  write_rtf(file = "tlf/fig_km2.rtf")
-```
-
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/fig_km2.pdf")
-```
-
-### Plot with `visR`
-
-Equivalently, the `visR` package can be also used to plot the KM curve fitting CDISC ADaM standard.
-In the following we will illustrate how to create KM plot in RTF format using `visR`.
-Firstly, similar to the previous sections, create a png file.
-
-```{r}
-# Estimate KM data point
-surv_mod <-
-  visR::estimate_KM(data = adtte_ttde,
-                    strata = "TRTP",
-                    AVAL = "AVAL_m")
-# save plot with white background and Times New Roman font
-ggplot2::theme_set(theme_bw() + theme(text = element_text(family = 'serif')))
-
-png("tlf/fig_km3.png",
-    width = 3000,
-    height = 2000,
-    res = 400)
-KM <- visR::visr(
-  surv_mod,
-  y_label = "Survival Probability\n",
-  x_label = "Time in Months",
-  x_ticks = 0:6,
-  y_ticks = seq(0, 1, 0.1),
-  legend_position = "bottom"
-) %>%
-  add_CNSR()
-
-KM %>%
-  visR::add_risktable(group = "statlist")  # seemingly current visR::add_risktable() does not support different fonts, will open an issue in the `visR` Github
-
-dev.off()
-```
-
-Then create RTF file
-
-```{r}
-# Create RTF figure
-rtf_read_png("tlf/fig_km3.png") %>% # Read the PNG file from the file path
-  rtf_title(
-    "Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group",
-    "All Participants"
-  ) %>% # Add title or subtitle
-  rtf_footnote("footnote") %>% # Add footnote
-  rtf_source("[datasource: adam-adtte]") %>% # Add data source
-  rtf_figure(fig_width = 6, fig_height = 4) %>% # Set proportional figure size to the original PNG figure size
-  rtf_encode(doc_type = "figure") %>% # Encode rtf as figure
-  write_rtf(file = "tlf/fig_km3.rtf")
-```
-
-Alternatively, we can also directly generate PDF file for KM plot with title and footnote (Ref: https://github.com/RConsortium/submissions-pilot1/blob/main/vignettes/tlf-kmplot.Rmd).
-
-```{r}
-# save plot as PDF file
-pdf.options(reset = TRUE, onefile = FALSE)
-pdf("tlf/fig_km3.pdf", width = 12, height = 6)
-
-KM <- visR::visr(
-  surv_mod,
-  y_label = "Survival Probability\n",
-  x_label = "Time in Months",
-  x_ticks = 0:6,
-  y_ticks = seq(0, 1, 0.1)
-) %>%
-  add_CNSR()
-
-KM1 <- KM %>%
-  visR::add_risktable(group = "statlist")  # seemingly current visR::add_risktable() does not support different fonts, will open an issue in the `visR` Github
-
-title <- cowplot::ggdraw() +
-  cowplot::draw_label(
-    "Kaplan-Meier Plot for Time to First Dermatologic Event by Treatment Group\nAll Participants",
-    fontfamily = "serif",
-    size = 12
-  )
-caption <- cowplot::ggdraw() +
-  cowplot::draw_label(
-    paste0(
-      "\nProgram: tlf-efficacy-km.Rmd [",
-      Sys.time(),
-      "]\n[Datasource: adam-adtte]"
-    ),
-    fontfamily = "serif",
-    size = 12
-  )
-
-KM2 <- cowplot::plot_grid(title,
-                          KM1,
-                          caption,
-                          ncol = 1,
-                          rel_heights = c(0.1, 0.8, 0.1)) #%>%
-# visR::apply_theme(theme_times)
-print(KM2)
-
-dev.off()
-```
-
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/fig_km3.pdf")
-```
-
-In conclusion, the procedure to generate the above efficacy K-M plot is summarized as follows.
+In conclusion, the steps to create a K-M plot are as follows.
 
@@
 - Step 2: Define the analysis-ready dataset. In this example, we define the analysis dataset for the TTDE endpoint `adtte_ttde`.
-- Step 3: Create the curve's data points using the `Surv()` formula directly 
-or fitted survival models beforehand. Then plot and customize the K-M curve with risk table using `survival`, `survminer` or `visR` package.
-- Step 4: Save the output as a PNG file, then format it into an RTF table. (The output format is based on user's preference)
+- Step 3: Save figures into `png` files based on required analysis specification.
+- Step 4: Create RTF table using `r2rtf` package.

### 2022-01-06 tlf-efficacy-km.Rmd
update file name

@@
 ```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/fig_km.pdf")
+# knitr::include_graphics("tlf/tlf_km.pdf")
 ```
@@
   rtf_encode(doc_type = "figure") %>% # Encode rtf as figure
-  write_rtf(file = "tlf/fig_km.rtf")
+  write_rtf(file = "tlf/tlf_km.rtf")
 ```
@@
 ```{r, include=FALSE}
-rtf2pdf("tlf/tlf_eff.rtf")
+rtf2pdf("tlf/tlf_km.rtf")
 ```
@@
 ```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/fig_km.pdf")
+knitr::include_graphics("tlf/tlf_km.pdf")
 ```

### 2022-01-06 tlf-efficacy-km.Rmd
minor fixes

@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-# knitr::include_graphics("tlf/tlf_km.pdf")
-```
-
 ::: {.rmdnote}
-There are R packages that can create more informative Kaplan-Meier plot such as `visR` and `survminer`.
-Interested reader can find examples in their package website. 
+R packages such as
+[visR](https://cran.r-project.org/package=visR) and
+[survminer](https://cran.r-project.org/package=survminer)
+can create more informative Kaplan-Meier plots.
+Interested readers can find examples on their websites.
 :::
@@
 
-- `TRTP`: treatment arm: (using corresponding numeric code `TRTAN` to re-order the levels, "Placebo" will be the reference level)
+- `TRTP`: treatment arm (using corresponding numeric code `TRTAN` to re-order the levels, "Placebo" will be the reference level)
 - `AVAL`: time-to-event analysis value
@@
   select(TRTP, TRTAN, AVAL, CNSR) %>%
-  mutate(TRTP = forcats::fct_reorder(TRTP, TRTAN), # Recorder levels
-         AVAL_m = AVAL / 30.4367) # convert Day to Month
+  mutate(
+    TRTP = forcats::fct_reorder(TRTP, TRTAN), # Recorder levels
+    AVAL_m = AVAL / 30.4367 # Convert Day to Month
+  )
 ```
@@
 
-The `survival` package is used to obtain the K-M estimator.
+The `survival` package is used to obtain the K-M estimate.
 
@@
 
-Using code below, we save the simplified KM plot into a `.png` file.
+We save the simplified K-M plot into a `.png` file using code below.
 
-```{r}
+```{r, results = FALSE}
 # Save as a PNG file
@@
   ylab = "Survival probability",
+  mark.time = TRUE,
   lwd = 2,
-  mark.time = T,
   col = c(2, 3, 4),
@@
 Now, we can use the `r2rtf` package to create a formatted RTF figure.
-More details can be found in [r2rtf website](https://merck.github.io/r2rtf/articles/example-figure.html).
+More details can be found on the [r2rtf website](https://merck.github.io/r2rtf/articles/example-figure.html).
 
@@
   rtf_figure(fig_width = 6, fig_height = 4) %>% # Set proportional figure size to the original PNG figure size
-  rtf_encode(doc_type = "figure") %>% # Encode rtf as figure
+  rtf_encode(doc_type = "figure") %>% # Encode figure as rtf
   write_rtf(file = "tlf/tlf_km.rtf")
@@
 - Step 3: Save figures into `png` files based on required analysis specification.
-- Step 4: Create RTF table using `r2rtf` package.
+- Step 4: Create RTF output using the `r2rtf` package.

### 2022-01-20 slides/render.R
update slides

@@ -1,2 +1,3 @@
 rmarkdown::render("slides/workshop-slides.Rmd")
 rmarkdown::render("slides/china-r.Rmd")
+rmarkdown::render("slides/r4csr-gwu.Rmd")

### 2022-01-20 slides/r4csr-gwu.Rmd

--- 2021-11-19 slides/china-r.Rmd
@@
 title: "R for Clinical Study Reports and Submission"
-subtitle: "The China-R Conference"
-author: "Yilong Zhang, Nan Xiao, Yujie Zhao, Keaven Anderson"
-date: "November 2021"
+subtitle: "The George Washington University Biostatistics Center"
+author: "Yilong Zhang"
+date: "Jan 2022"
 output:
   mkdocs::mk_ioslides_presentation:
-    mk_logo: "msd"
+    mk_logo: "merck"
     mk_col_title: "white"
@@
     widescreen: true
+bibliography: reference.bib  
 ---
@@
 
+## Clarification from FDA
+
+- [FDA: Statistical Software Clarifying Statement](https://www.fda.gov/files/about%20fda/published/Statistical-Software-Clarifying-Statement-PDF.pdf)
+
+> "FDA does not require use of any specific software for statistical analyses, and statistical software
+is not explicitly discussed in Title 21 of the Code of Federal Regulations [e.g., in 21CFR part
+11]. However, the software package(s) used for statistical analyses should be fully documented
+in the submission, including version and build identification."
+
+## Motivation
+
+As an organization, we need to ensure compliance and reduce the risk of using R and R packages in regulatory deliverables.
+
+- R is widely used in clinical trial study design.
+  + Following @lachin1986evaluation
+  + [`gsDesign`](https://github.com/keaven/gsDesign): an R package for group sequential design under proportional hazards.
+  + [`simtrial`](https://github.com/Merck/simtrial), 
+  [`gsDesign2`](https://github.com/Merck/gsDesign2) and 
+  [`gsdmvn`](https://github.com/Merck/gsdmvn): experimental R packages for group sequential design under non-proportional hazards
+  + Bookdown: <https://keaven.github.io/gsd-deming/>
+
+- R is flexible for handling novel missing data approaches following estimand framework in ICH E9 (R1)
+  + Recurrent event data: @gao2017control @diao2020efficient 
+  + Robustness: @liu2021multiply
+
+- R is used in Bayesian statistics.
+  + [stan](https://mc-stan.org/users/interfaces/rstan) and network meta-analysis for drug reimbursement analysis
+
+- R is widely used in visualization 
+  + [SafetyGraphics](https://github.com/SafetyGraphics/safetyGraphics) 
+  + [forestly](https://elong0527.github.io/forestly/articles/example.html#construct-interactive-forest-plot-1): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_194/)
+
 ## Background
@@
 
-- We try to fill in gaps to streamline workflow using R:
-  + To develop, validate, deliver analysis results.
+- We try to fill in gaps to streamline workflow using R for clinical trial development:
+  + To develop, validate, and deliver analysis results.
   + To submit analysis results to regulatory agencies in [eCTD format](https://en.wikipedia.org/wiki/Electronic_common_technical_document).
@@
 
-## Clarification from FDA
+# R Consortium Pilot Submission
 
-- [FDA: Statistical Software Clarifying Statement](https://www.fda.gov/files/about%20fda/published/Statistical-Software-Clarifying-Statement-PDF.pdf)
+## Challenge and Assumption
 
-"FDA does not require use of any specific software for statistical analyses, and statistical software
-is not explicitly discussed in Title 21 of the Code of Federal Regulations [e.g., in 21CFR part
-11]. However, the software package(s) used for statistical analyses should be fully documented
-in the submission, including version and build identification."
+### Challenge
+
+- How to submit internal developed (proprietary) R packages? 
+- How to follow [ICH and/or FDA guidances](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd) 
+in preparing eCTD package?
+- How to enhance reproducibility from FDA reviewer's perspective?
+
+### Assumption
+
+- Focus on analysis and reporting given available ADaM datasets.
+
+## Deliverables
+
+- Open source pilot eCTD submission package to FDA 
+  + Proprietary R package
+  + R scripts for analysis
+  + Analysis data reviewer guide (ADRG)
+  + etc. 
+- Development: <https://github.com/RConsortium/submissions-pilot1> 
+- eCTD package: <https://github.com/RConsortium/submissions-pilot1-to-fda> 
+- [R consortium announcement](https://www.r-consortium.org/blog/2021/12/08/successful-r-based-test-package-submitted-to-fda)
+
+## FDA response
+
+- "Using R version 4.1.1, FDA was able to run the submitted code and confirm the applicant's tables and the submitted figure in report-tlf pdf file."
+- "Using FDA developed code, a statistical analyst was able to independently generate tables using the submitted data."
+- [original FDA response](https://github.com/RConsortium/submissions-wg/blob/main/Documents/Summary_R_Pilot_Submission.pdf)
+
+## Future work
+
+- Address minor issues reported by FDA reviewers. 
+- Pilot R Shiny app submission.
+- Pilot submission with advanced analysis methods (e.g. study design, missing data, Bayesian, etc).
+
+# Reproducibility
+
+## Reproducibility spectrum
+
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("reproducibility.png")
+```
 
@@
 
+## Recommendations
+
+Although FDA did not expect submitted R code is executable, 
+sponsor shall enhance reproducibility.
+
+- Fixed R version: e.g. R4.1.0 
+- Fixed R package snapshot date: e.g. 2021-08-31 
+  + [RStudio package manager](https://packagemanager.rstudio.com/client/#/repos/1/overview)
+  + [MRAN](https://mran.microsoft.com/timemachine)
+- Flexibility of input and output path: 
+  + Define path as parameter. 
+- eCTD deliverables dry-run in Windows environment
+- Provide steps to follow in [ADRG appendix](https://github.com/RConsortium/submissions-pilot1-to-fda/blob/main/m5/datasets/rconsortiumpilot1/analysis/adam/datasets/adrg.pdf)
+
+# Implementation
+
 ## Philosophy
@@
 
-## Tools for study design
-
-> Details will be skipped in today's talk.
-
-Tools:
-
-- [`gsDesign`](https://github.com/keaven/gsDesign): an R package for group sequential design under proportional hazards.
-- [`simtrial`](https://github.com/Merck/simtrial): an experimental R package for simulation-based group sequential design under non-proportional hazards
-- [`gsDesign2`](https://github.com/Merck/gsDesign2) and
-[`gsdmvn`](https://github.com/Merck/gsdmvn):
-an experimental R package for analytical-based group sequential design under proportional hazards.
-
-Bookdown: <https://keaven.github.io/gsd-deming/>
-
-Training: [December 6th, Deming Conference 2021](https://demingconference.org/programs/2021-program/)
-
 ## Tools for reporting and submission
@@
 Bookdown: <https://r4csr.org/>
-
-Training: [December 2nd, Short course for ASA Princeton-Trenton Chapter](https://rutgers.webex.com/rutgers/j.php?RGID=rc0a71b2f239233a7850a121cf972045f)
 
@@
 
-- As a statistician, I use `tidyverse`, `r2rtf`, and internal tools to define the mock-up table, listing and figure (TLFs) for statistical analysis of a clinical trial.
+- As a statistician, I use `tidyverse`, `r2rtf`, 
+and internal tools to define the mock-up table, listing and figure (TLFs) for statistical analysis of a clinical trial.
   + More than 100 TLFs for efficacy and safety of a drug or vaccine.
@@
   - Planning: define the scope of a project
-  - Development: mplement target deliverables
+  - Development: implement target deliverables
   - Validation: verify target deliverables
@@
 
-## Join us
+# Thank you {.thank-you}
 
-- Contact us for R developer or Methodology Research Positions:
-  + Yilong Zhang: <https://elong0527.github.io/>
-  + Nan Xiao: <https://nanx.me/>
-  + Keaven Anderson: <https://www.linkedin.com/in/keaven-anderson-09aa7538/>
-
-- Multiple open positions (statisticians/programmers) in China, EU, and US
-  + US: <https://jobs.merck.com/bards>
-  + China/EU: <https://jobs.msd.com/cn/zh>
-
-# Thank you {.thank-you}
+## Reference

### 2022-01-29 project-management.Rmd
Replace validation tracker image with kable

@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("images/validation_tracker.png")
+```{r, echo=FALSE}
+library("kableExtra")
+
+df <- read.table("validation-tracker.tsv", stringsAsFactors = FALSE, header = FALSE, sep = "\t")
+
+df %>%
+  kbl(col.names = NULL) %>%
+  kable_classic(full_width = FALSE, html_font = "'Times New Roman', Times, serif", font_size = 16) %>%
+  add_header_above(
+    c(
+      "Program Name" = 1, "Program Validation Category" = 1,
+      "Who" = 1, "Status" = 1, "Who" = 1, "Status" = 1, "Who" = 1, "Status" = 1
+    ),
+    line = FALSE,
+    extra_css = "border: 1px solid #000;"
+  ) %>%
+  add_header_above(
+    c(
+      "<span></span>" = 2, "Requirement/Specification" = 2,
+      "Developer Testing" = 2, "Independent Testing" = 2
+    ),
+    escape = FALSE,
+    line = FALSE,
+    extra_css = "border: 1px solid #000;"
+  ) %>%
+  column_spec(4, background = ifelse(df$V4 == "C", "#00ff00", "#ffff00")) %>%
+  column_spec(6, background = ifelse(df$V6 == "C", "#00ff00", "#ffff00")) %>%
+  column_spec(8, background = ifelse(df$V8 == "C", "#00ff00", "#ffff00")) %>%
+  column_spec(1, extra_css = "border: 1px solid #000; text-align: left;") %>%
+  column_spec(2:8, extra_css = "border: 1px solid #000; text-align: center;")
 ```
+
+<p></p>
 

### 2022-02-12 tlf-baseline.Rmd
fix some minor issues

@@
 baseline characteristics table for a study.
+
+```{r, include=FALSE}
+source("common.R")
+```
 
@@
 
-`text_indent_first` and `test_indent_left` are used to control the indent
+`text_indent_first` and `text_indent_left` are used to control the indent
 space of text. They are helpful when you need to control the white space
 of a long phrase "AMERICAN INDIAN OR ALASKA NATIVE"
-in the table provide an example.
+in the table provides an example.
 
@@
 
+```{r, include=FALSE}
+rtf2pdf("tlf/tlf_base.rtf")
+```
+
 ```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}

### 2022-02-12 tlf-overview.Rmd
automate rtf to pdf conversion
source common.R in each Rmd to allow compile individually
fix some minor issues

@@
 # Overview {#tlf-overview}
+
+```{r, include=FALSE}
+source("common.R")
+```
 
@@
 ::: {.rmdnote}
-There are other R packages to create TLFs in ASCIII, RTF and Word format.
+There are other R packages to create TLFs in ASCII, RTF and Word format.
 For example, `rtables`, `huxtable`, `pharmaRTF`, `gt`, `officer`, `flextable` etc.
 Here we focus on `r2rtf` to illustrate the concept.  
-Reader is encourage to explore other R packages to find the proper tools to fit your purpose. 
+Reader is encouraged to explore other R packages to find the proper tools to fit your purpose. 
 :::
@@
 
-There are also books to introduce tidyverse.
+There are also books to introduce `tidyverse`.
 We assume the reader have experience in using `tidyverse` in this book.
@@
 
+```{r, include=FALSE}
+rtf2pdf("tlf/intro-ae1.rtf")
+```
+
 ```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
@@
 
+```{r, include=FALSE}
+rtf2pdf("tlf/intro-ae2.rtf")
+```
+
 ```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
@@
 
+```{r, include=FALSE}
+rtf2pdf("tlf/intro-ae3.rtf")
+```
+
 ```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
@@
 Default is `"c"` for center justification.
-To vary text justification by column, use character vector with length of vector equal to
+To vary text justification by column, use character vector with length of vector equals to
 number of columns displayed (e.g., `c("c", "l", "r")`).
@@
   write_rtf("tlf/intro-ae5.rtf")
+```
+
+```{r, include=FALSE}
+rtf2pdf("tlf/intro-ae5.rtf")
 ```
@@
 
+```{r, include=FALSE}
+rtf2pdf("tlf/intro-ae7.rtf")
+```
+
 ```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}

### 2022-02-15 preface.Rmd
Add install and build guide to readme

@@
 
-  Yilong Zhang <<yilong.zhang@merck.com>>, Nan Xiao, Keaven Anderson
+  Yilong Zhang <<yilong.zhang@merck.com>>, Nan Xiao, Keaven Anderson, Yalin Zhu
 
@@
 contributors <- read.csv("contributors.csv", stringsAsFactors = FALSE, na.strings = "")
-contributors$desc <- with(contributors, ifelse(is.na(username), name, paste0(name, " (\\@", username, ")")))
+contributors$desc <- with(contributors, ifelse(is.na(username), trimws(name), paste0(trimws(name), " (\\@", trimws(username), ")")))
 cat("  We are grateful for all the improvements brought by these contributors (in chronological order): ", sep = "")

### 2022-02-25 slides/r2rtf-function-summary.R
Deploying to gh-pages from @ elong0527/r4csr@73b1988f7cca381c8e03c30e6d6d970ae8155cf2 🚀

@@ -1,0 +1,203 @@
+# Generate an RTF table --------------------------------------------------------
+# Example from <https://merck.github.io/r2rtf/articles/example-sublineby-pageby-groupby.html>
+
+library(r2rtf)
+library(dplyr)
+library(pdftools)
+library(magick)
+
+data(r2rtf_adae)
+
+ae_t1 <- r2rtf_adae[200:260, ] %>%
+  mutate(
+    SUBLINEBY = paste0(
+      "Trial Number: ", STUDYID,
+      ", Site Number: ", SITEID
+    ),
+    SUBJLINE = paste0(
+      "Subject ID = ", USUBJID,
+      ", Gender = ", SEX,
+      ", Race = ", RACE,
+      ", AGE = ", AGE, " Years",
+      ", TRT = ", TRTA
+    ),
+    # create a subject line with participant's demographic information.
+    # this is for page_by argument in rtf_body function
+    AEDECD1 = tools::toTitleCase(AEDECOD), # propcase the AEDECOD
+    DUR = paste(ADURN, ADURU, sep = " ")
+  ) %>% # AE duration with unit
+  select(USUBJID, ASTDY, AEDECD1, DUR, AESEV, AESER, AEREL, AEACN, AEOUT, TRTA, SUBJLINE, SUBLINEBY) # display variable using this order
+
+ae_tbl <- ae_t1 %>%
+  # It is important to order variable properly.
+  arrange(SUBLINEBY, TRTA, SUBJLINE, USUBJID, ASTDY) %>%
+  rtf_page(orientation = "landscape", col_width = 9) %>%
+  rtf_page_header() %>%
+  rtf_page_footer(text = "CONFIDENTIAL") %>%
+  rtf_title(
+    "Listing of Subjects With Serious Adverse Events",
+    "ASaT"
+  ) %>%
+  rtf_colheader(
+    "Subject| Rel Day | Adverse | | | | |Action| |",
+    col_rel_width = c(2.5, 2, 4, 2, 3, 2, 3, 2, 5)
+  ) %>%
+  rtf_colheader(
+    "ID| of Onset | Event | Duration | Intensity | Serious | Related | Taken| Outcome",
+    border_top = "",
+    col_rel_width = c(2.5, 2, 4, 2, 3, 2, 3, 2, 5)
+  ) %>%
+  rtf_body(
+    col_rel_width = c(2.5, 2, 4, 2, 3, 2, 3, 2, 5, 1, 1, 1),
[…]

### 2022-02-28 slides/r4csr-gwu.Rmd
Re-render slides to use the latest function summary figure

@@
 author: "Yilong Zhang"
-date: "Jan 2022"
+date: "January 2022"
 output:
@@
 - R is widely used in clinical trial study design.
-  + Following @lachin1986evaluation
-  + [`gsDesign`](https://github.com/keaven/gsDesign): an R package for group sequential design under proportional hazards.
-  + [`simtrial`](https://github.com/Merck/simtrial), 
-  [`gsDesign2`](https://github.com/Merck/gsDesign2) and 
+  - Following @lachin1986evaluation
+  - [`gsDesign`](https://github.com/keaven/gsDesign): an R package for group sequential design under proportional hazards.
+  - [`simtrial`](https://github.com/Merck/simtrial),
+  [`gsDesign2`](https://github.com/Merck/gsDesign2), and
   [`gsdmvn`](https://github.com/Merck/gsdmvn): experimental R packages for group sequential design under non-proportional hazards
-  + Bookdown: <https://keaven.github.io/gsd-deming/>
+  - Bookdown: <https://keaven.github.io/gsd-deming/>
 
 - R is flexible for handling novel missing data approaches following estimand framework in ICH E9 (R1)
-  + Recurrent event data: @gao2017control @diao2020efficient 
-  + Robustness: @liu2021multiply
+  - Recurrent event data: @gao2017control @diao2020efficient
+  - Robustness: @liu2021multiply
 
 - R is used in Bayesian statistics.
-  + [stan](https://mc-stan.org/users/interfaces/rstan) and network meta-analysis for drug reimbursement analysis
+  - [stan](https://mc-stan.org/users/interfaces/rstan) and network meta-analysis for drug reimbursement analysis
 
 - R is widely used in visualization
-  + [SafetyGraphics](https://github.com/SafetyGraphics/safetyGraphics) 
-  + [forestly](https://elong0527.github.io/forestly/articles/example.html#construct-interactive-forest-plot-1): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_194/)
+  - [SafetyGraphics](https://github.com/SafetyGraphics/safetyGraphics)
+  - [forestly](https://elong0527.github.io/forestly/articles/example.html#construct-interactive-forest-plot-1): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_194/)
 
@@
 - Clinical study report is a key deliverable for clinical trials to regulatory agencies. (e.g., FDA, CFDA)
-  + [ICH E3 Structure and Content of Clinical Study Reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf)
+  - [ICH E3 Structure and Content of Clinical Study Reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf)
 
 - We try to fill in gaps to streamline workflow using R for clinical trial development:
-  + To develop, validate, and deliver analysis results.
-  + To submit analysis results to regulatory agencies in [eCTD format](https://en.wikipedia.org/wiki/Electronic_common_technical_document).
+  - To develop, validate, and deliver analysis results.
+  - To submit analysis results to regulatory agencies in [eCTD format](https://en.wikipedia.org/wiki/Electronic_common_technical_document).
 
 - Focus on table, listing, figure (TLFs) delivered in RTF/Microsoft Word format
-  + In the pharmaceutical industry, RTF/Microsoft Word play a central role in preparing clinical study reports.
-  + Different organization can have different table standards
+  - In the pharmaceutical industry, RTF/Microsoft Word play a central role in preparing clinical study reports.
+  - Different organization can have different table standards
 
-# R Consortium Pilot Submission
+# R Consortium pilot submission
 
-## Challenge and Assumption
+## Challenges and assumptions
 
-### Challenge
+### Challenges
 
@@
 
-### Assumption
+### Assumptions
 
@@
 - Open source pilot eCTD submission package to FDA
-  + Proprietary R package
-  + R scripts for analysis
-  + Analysis data reviewer guide (ADRG)
-  + etc. 
+  - Proprietary R package
+  - R scripts for analysis
+  - Analysis data reviewer guide (ADRG)
+  - etc.
 - Development: <https://github.com/RConsortium/submissions-pilot1>
@@
 - "Using FDA developed code, a statistical analyst was able to independently generate tables using the submitted data."
-- [original FDA response](https://github.com/RConsortium/submissions-wg/blob/main/Documents/Summary_R_Pilot_Submission.pdf)
+- [Original FDA response](https://github.com/RConsortium/submissions-wg/blob/main/Documents/Summary_R_Pilot_Submission.pdf)
 
@@
 - Pilot R Shiny app submission.
-- Pilot submission with advanced analysis methods (e.g. study design, missing data, Bayesian, etc).
+- Pilot submission with advanced analysis methods (e.g., study design, missing data, Bayesian, etc).
 
@@
 
-- Fixed R version: e.g. R4.1.0 
-- Fixed R package snapshot date: e.g. 2021-08-31 
-  + [RStudio package manager](https://packagemanager.rstudio.com/client/#/repos/1/overview)
-  + [MRAN](https://mran.microsoft.com/timemachine)
+- Fixed R version: e.g., R 4.1.0
+- Fixed R package snapshot date: e.g., 2021-08-31
+  - [RStudio Public Package Manager](https://packagemanager.rstudio.com/)
+  - [CRAN Time Machine by MRAN](https://mran.microsoft.com/timemachine)
 - Flexibility of input and output path:
-  + Define path as parameter. 
+  - Define path as parameter.
 - eCTD deliverables dry-run in Windows environment
@@
 
-```{r, out.width="80%", echo=FALSE, fig.align="center"}
+```{r, out.width="100%", echo=FALSE, fig.align="center"}
 knitr::include_graphics("function-summary.png")
@@
 and internal tools to define the mock-up table, listing and figure (TLFs) for statistical analysis of a clinical trial.
-  + More than 100 TLFs for efficacy and safety of a drug or vaccine.
+  - More than 100 TLFs for efficacy and safety of a drug or vaccine.
 
@@
 - R Validation Hub: <https://www.pharmar.org/>
-  + Focus on designing a framework that assesses the quality of an R package
-  + Presentations: <https://www.pharmar.org/present/>
+  - Focus on designing a framework that assesses the quality of an R package
+  - Presentations: <https://www.pharmar.org/present/>
 
 - R-based submission pilots to FDA:
-  + <https://rconsortium.github.io/r-submission-site/>
-  + Focus on improving practices of R-based clinical trial regulatory submission.
+  - <https://rconsortium.github.io/r-submission-site/>
+  - Focus on improving practices of R-based clinical trial regulatory submission.
 
 - R/Pharma: <https://rinpharma.com/>
-  + Annual conference focus on the use of R in the development of pharmaceuticals.
+  - Annual conference focus on the use of R in the development of pharmaceuticals.
 
 - R Consortium R Adoption Seminar Series
-  + <https://www.r-consortium.org/webinars>
+  - <https://www.r-consortium.org/webinars>
 
@@
 - Enhance compliance, reproducibility, traceability, and automation
-  + Automation of analysis, documentation, review, and testing
-  + Linkage among data, TLFs, and final reports
-  + R package qualification
+  - Automation of analysis, documentation, review, and testing
+  - Linkage among data, TLFs, and final reports
+  - R package qualification
 - Enable advanced study design and statistical methods
@@
 
-## Reference
+## References

### 2022-02-28 slides/china-r.Rmd
Deploying to gh-pages from @ elong0527/r4csr@4ca950c2f5252a5bdd26e7e5a7cd0318dadac192 🚀

@@
 - Clinical study report is a key deliverable for clinical trials to regulatory agencies. (e.g., FDA, CFDA)
-  + [ICH E3 Structure and Content of Clinical Study Reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf)
+  - [ICH E3 Structure and Content of Clinical Study Reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf)
 
 - We try to fill in gaps to streamline workflow using R:
-  + To develop, validate, deliver analysis results.
-  + To submit analysis results to regulatory agencies in [eCTD format](https://en.wikipedia.org/wiki/Electronic_common_technical_document).
+  - To develop, validate, deliver analysis results.
+  - To submit analysis results to regulatory agencies in [eCTD format](https://en.wikipedia.org/wiki/Electronic_common_technical_document).
 
 - Focus on table, listing, figure (TLFs) delivered in RTF/Microsoft Word format
-  + In the pharmaceutical industry, RTF/Microsoft Word play a central role in preparing clinical study reports.
-  + Different organization can have different table standards
+  - In the pharmaceutical industry, RTF/Microsoft Word play a central role in preparing clinical study reports.
+  - Different organization can have different table standards
 
@@
 - As a statistician, I use `tidyverse`, `r2rtf`, and internal tools to define the mock-up table, listing and figure (TLFs) for statistical analysis of a clinical trial.
-  + More than 100 TLFs for efficacy and safety of a drug or vaccine.
+  - More than 100 TLFs for efficacy and safety of a drug or vaccine.
 
@@
 - Contact us for R developer or Methodology Research Positions:
-  + Yilong Zhang: <https://elong0527.github.io/>
-  + Nan Xiao: <https://nanx.me/>
-  + Keaven Anderson: <https://www.linkedin.com/in/keaven-anderson-09aa7538/>
+  - Yilong Zhang: <https://elong0527.github.io/>
+  - Nan Xiao: <https://nanx.me/>
+  - Keaven Anderson: <https://www.linkedin.com/in/keaven-anderson-09aa7538/>
 
 - Multiple open positions (statisticians/programmers) in China, EU, and US
-  + US: <https://jobs.merck.com/bards>
-  + China/EU: <https://jobs.msd.com/cn/zh>
+  - US: <https://jobs.merck.com/bards>
+  - China/EU: <https://jobs.msd.com/cn/zh>
 

### 2022-02-28 slides/workshop-slides.Rmd
Deploying to gh-pages from @ elong0527/r4csr@4ca950c2f5252a5bdd26e7e5a7cd0318dadac192 🚀

@@
 
-```{r, message=FALSE}
+```{r, include=FALSE}
 library(dplyr)
@@
 - Delivering TLFs in CSR (Yilong Zhang)
-  + Session 1 (45 min)
-  + Break and/or Exercise (20 min)
-  + Session 2 (45 min)
-  + Break (10 min)
+  - Session 1 (45 min)
+  - Break and/or Exercise (20 min)
+  - Session 2 (45 min)
+  - Break (10 min)
 - Clinical trial project (Nan Xiao)
-  + Session 3 (30 min)
-  + Break (10 min)
+  - Session 3 (30 min)
+  - Break (10 min)
 - eCTD submission package (Nan Xiao)
-  + Session 4 (40 min)
-  + Q&A (10 min)
+  - Session 4 (40 min)
+  - Q&A (10 min)
 
@@
 
-## Training Objective
+## Training objective
 
@@
 - Team members from Merck & Co.
-  + [We are hiring!](https://jobs.merck.com/bards)
-  + Contact us if interested.
+  - [We are hiring!](https://jobs.merck.com/bards)
+  - Contact us if interested.
 
 - [Contributors](https://r4csr.org/) of the training materials.
-  + Please consider submitting issues or PR in the repo.
+  - Please consider submitting issues or PR in the repo.
 
 - [R Consortium](https://www.r-consortium.org/projects/isc-working-groups)
-  + R validation hub
-  + Submission working group
-  + R Tables for Regulatory Submission (RTRS) working group
-  + Please consider to join and contribute to a working group.
+  - R validation hub
+  - Submission working group
+  - R Tables for Regulatory Submission (RTRS) working group
+  - Please consider to join and contribute to a working group.
 
 - [ASA Princeton-Trenton Chapter](https://community.amstat.org/princetontrenton/home)
-  + Please consider to be a member of our chapter.
+  - Please consider to be a member of our chapter.
 
@@
 - Demo project:
-  + Analysis project: <https://github.com/elong0527/esubdemo>
-  + eCTD submission package: <https://github.com/elong0527/ectddemo>
+  - Analysis project: <https://github.com/elong0527/esubdemo>
+  - eCTD submission package: <https://github.com/elong0527/ectddemo>
 - R Consortium FDA submission pilot:
-  + <https://github.com/RConsortium/submissions-pilot1>
+  - <https://github.com/RConsortium/submissions-pilot1>
 
 - During the workshop, we will use `esubdemo` project.
-  + `esubdemo` is shared in RStudio Cloud (<https://rstudio.cloud/>)
-  + Project link will be shared in chat
-  + Post questions in group chat.
+  - `esubdemo` is shared in RStudio Cloud (<https://rstudio.cloud/>)
+  - Project link will be shared in chat
+  - Post questions in group chat.
 
@@
 
-## ICH E3 Guidance
+## ICH E3 guidance
 
@@
 
-# r2rtf Introduction
+# r2rtf introduction
 
@@
 - provide simple "verb" functions that correspond to each component of a table,
-  to help you translate a data frame to a table in RTF format.
-- enable pipes (`%>%`).
+  to help you translate a data frame to a table in RTF format;
+- enable pipes (`%>%`);
 - only focus on the **table format**.
-  + Data manipulation and analysis shall be handled by other R packages. (e.g. `tidyverse`)
+  - Data manipulation and analysis should be handled by other R packages. (e.g., `tidyverse`)
 
@@
 
-## Minimal Example
+## Minimal example
 
-`r2rtf` is designed to enable pipes (`%>%`)
+`r2rtf` is designed to enable pipes (`%>%`).
 
@@
 
-## Package Overview
+## Package overview
 
@@
 
-## Simple Example - Adverse Events
+## Simple example: adverse events
 
 `r2rtf` only focus on **table format**.
-Data manipulation and analysis shall be handled by other R packages. (e.g. `tidyverse`)
+Data manipulation and analysis should be handled by other R packages. (e.g., `tidyverse`)
 
@@
 
-## Function Summary
+## Function summary
 
@@
 
-## Function Illustration
+## Function illustration
 
-```{r, out.width="80%", echo=FALSE, fig.align="center"}
+```{r, out.width="100%", echo=FALSE, fig.align="center"}
 knitr::include_graphics("function-summary.png")
@@
 
-# Break and/or Exercise (20 min)
+# Break and/or exercise (20 min)
 
-# CSR Examples
+# CSR examples
 
-## Disposition Table
+## Disposition table
 
@@
 
-## Analysis Population
+## Analysis population
 
@@
 
-## Baseline Characteristics
+## Baseline characteristics
 
@@
 
-## Efficacy Table
+## Efficacy table
 
@@
 
-## AE Summary Table
+## AE Summary table
 
@@
 
-## Specific AE Table
+## Specific AE table
 
@@
 
-# Clinical Trial Project
+# Clinical trial project
 
@@
 
-## Folder Structure
+## Folder structure
 
@@
 
-## Project Management
+## Project management
 
@@
 
-# eCTD Submission
+# eCTD submission
 
@@
 
-## Submission Package
+## Submission package
 
@@
 
-## Running Environment
+## Running environment
 
@@
 
-## pkglite: Compact Package Representations
+## pkglite: compact package representations
 
@@
 
-## Pipe-Friendly Workflow
+## Pipe-friendly workflow
 
@@
 
-## File Specifications + File Collections
+## File specifications + file collections
 
@@
 
-```
+```text
 # Generated by pkglite: do not edit by hand

### 2022-03-18 images/pack.Rmd
Capture `pack()` output using asciicast

---
title: "Generate an asciicast SVG for pkglite::pack() output"
output: rmarkdown::html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "images/"
)
```

```{r echo=FALSE, results="hide"}
asciicast::init_knitr_engine(echo_input = FALSE)
```

```{r, include=FALSE}
options(asciicast_knitr_svg = TRUE)
```

```{asciicast, pack}
#' At: end
library("pkglite")

"esubdemo/" %>%
  collate(file_ectd(), file_auto("inst")) %>%
  pack(output = "r0pkgs.txt")
```

### 2022-03-18 submission-package.Rmd
fix some minor issues
Capture `pack()` output using asciicast

@@
 
-In eCTD deliverable, the analysis dataset and source code are saved
+In eCTD deliverable, the analysis datasets and source code are saved
 under the eCTD module 5 (clinical study reports) folder
@@
   pack(output = "r0pkgs.txt")
+```
+
+```{r, echo=FALSE, eval=FALSE}
+# Generated this asciicast SVG by running
+rmarkdown::render("images/pack.Rmd")
+
+patch_asciicast_theme <- function(x) {
+  x <- gsub("rgb\\(40,45,53\\)", replacement = "rgb\\(255,255,255\\)", x = x)
+  x <- gsub("rgb\\(185,192,203\\)", replacement = "rgb\\(0,0,0\\)", x = x)
+  x <- gsub("rgb\\(113,190,242\\)", replacement = "rgb\\(52,101,164\\)", x = x)
+  x <- gsub("rgb\\(168,204,140\\)", replacement = "rgb\\(78,154,6\\)", x = x)
+  x <- gsub("rgb\\(102,194,205\\)", replacement = "rgb\\(6,152,154\\)", x = x)
+  x
+}
+
+readLines("images/images/pack.svg", warn = FALSE) |>
+  patch_asciicast_theme() |>
+  writeLines(con = "images/images/pack.svg")
+
+# For best web browser compatibility:
+# Open pack.svg in web browser, print as PDF
+# pdfcrop pack.pdf
+# convert -density 300 pack-crop.pdf pack.png
 ```

### 2022-03-19 submission-environment.Rmd

@@
 ```{r, out.width="65%", echo=FALSE, fig.cap="cleanslate RStudio addin"}
+# Web browser DevTools -> Print as PDF ->
+# Edit PDF if necessary -> pdfcrop -> convert to PNG
 knitr::include_graphics("images/cleanslate-addin.png")

### 2022-03-19 submission-package.Rmd

--- 2021-11-28 submission-package.Rmd
@@
 ```{r, fig.cap="Preview the generated text file", out.width="100%", echo=FALSE}
+# Web browser DevTools -> Remove irrelevant elements ->
+# Print as PDF -> Edit PDF if necessary -> pdfcrop -> convert to PNG
 knitr::include_graphics("images/preview-txt.png")

### 2022-03-20 submission-package.Rmd

@@
 
-x <- available.packages(contriburl = contrib.url("https://mran.microsoft.com/snapshot/2021-08-06/"))
-df2$version <- x[match(df2$package, x[, "Package"]), "Version"]
+# # Commented out as the snapshot availability is not guaranteed
+# x <- available.packages(contriburl = contrib.url("https://mran.microsoft.com/snapshot/2021-08-06/"))
+# df2$version <- x[match(df2$package, x[, "Package"]), "Version"]
+df2$version <- c("0.2.0", "2.4.3", "1.0.7", "1.1.3", "1.6.2-1", "0.3.0")
 

### 2022-03-23 index.Rmd
Add events table

@@
 
-```{r, include=FALSE}
-source("common.R")
-```
-
 ```{r, eval = FALSE, include=FALSE}
@@
 
+## Events {-}
+
+```{css, echo=FALSE}
+table.table td:nth-child(1) { min-width: 27ch; }
+table.table td:nth-child(2) { min-width: 12ch; }
+table.table td:nth-child(3) { min-width: 10ch; }
+table.table tbody { border-bottom: 1px solid #B6BDC6; }
+```
+
+```{r, echo=FALSE}
+knitr::kable(read.table("events.tsv", stringsAsFactors = FALSE, header = TRUE, sep = "\t"))
+```
+
 ## License {-}

### 2022-03-24 index.Rmd

--- 2021-11-19 index.Rmd
@@
 
+## Events {-}
+
+```{css, echo=FALSE}
+table { font-variant-numeric: tabular-nums; }
+table.table td:nth-child(1) { min-width: 28ch; }
+table.table td:nth-child(2) { min-width: 14ch; }
+table.table td:nth-child(3) { min-width: 14ch; }
+table.table tbody { border-bottom: 1px solid #B6BDC6; }
+```
+
+```{r, echo=FALSE}
+knitr::kable(read.table("events.tsv", stringsAsFactors = FALSE, header = TRUE, sep = "\t"))
+```
+
 ## License {-}

### 2022-05-15 slides/r4csr-rstudio.Rmd
add r4csr-rstudio slides

--- 2022-02-28 slides/r4csr-gwu.Rmd
@@
 title: "R for Clinical Study Reports and Submission"
-subtitle: "The George Washington University Biostatistics Center"
+subtitle: "RStudio Pharma Meetup Series"
 author: "Yilong Zhang"
-date: "January 2022"
+date: "May 2022"
 output:
-  mkdocs::mk_ioslides_presentation:
-    mk_logo: "merck"
-    mk_col_title: "white"
-    mk_col_section: "teal"
-    mk_class: "public"
+  ioslides_presentation:
     widescreen: true
+    smaller: true
 bibliography: reference.bib
@@
 
+```{css, echo = FALSE}
+slides > slide.dark {
+  background: #59B2D1 !important;
+}
+
+.dark hgroup {
+  position: absolute;
+  left: 5%;
+  top: 40%;
+}
+
+slides > slide.title-slide {
+  background: #f2f2f2;
+}
+
+.title-slide hgroup {
+  position: absolute;
+  left: 5%;
+  top: 40%;
+}
+
+.title-slide hgroup h1 {
+  color: #404040;
+  font-size: 48px;
+}
+
+.title-slide hgroup h2 {
+  color: #404040;
+  font-size: 24px;
+}
+
+.title-slide hgroup p {
+  color: #404040;
+  font-size: 16px;
+}
+
+h1 {
+ color: #404040;
+}
+
+h2 {
+ color: #404040;
+}
+
+h3 {
+ color: #404040;
+}
+
+p {
+ color: #404040;
+}
+
+ul {
+ color: #404040;
+}
+
+li {
+ color: #404040;
+}
+
+a {
+    color: #59B2D1;
+    text-decoration: none;
+    border-bottom: none;
+}
+
+a > code {
+    color: #59B2D1;
+    text-decoration: none;
+    border-bottom: none;
+}
+
+slides > slide:not(.segue) > hgroup > h2 {
+    font-size: 34px;
+    line-height: 50px;
+    padding: 0px 0px 10px 0px;
+    border-bottom: 2px solid #59B2D1;
+}
+```
+
 ## Disclaimer
 
-- All opinions expressed are those of the presenter and not
-  Merck Sharp & Dohme Corp., a subsidiary of Merck & Co., Inc., Kenilworth, NJ, USA.
+- All opinions expressed are those of the presenter and not represent any organization.
 
@@
 - Clinical study report is a key deliverable for clinical trials to regulatory agencies. (e.g., FDA, CFDA)
-  - [ICH E3 Structure and Content of Clinical Study Reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf)
+  - [ICH E3 Structure and Content of Clinical Study Reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf).
 
@@
 
-- Focus on table, listing, figure (TLFs) delivered in RTF/Microsoft Word format
+- Focus on table, listing, figure (TLFs) delivered in RTF/Microsoft Word format.
   - In the pharmaceutical industry, RTF/Microsoft Word play a central role in preparing clinical study reports.
-  - Different organization can have different table standards
+  - Different organization can have different table standards.
 
 # R Consortium pilot submission
+
+## Working group
+
+- The R submission working group is a cross industry pharma working group focusing on improving practices of R-based clinical trial regulatory submission.
+  - [members and affiliations](https://rconsortium.github.io/submissions-wg/)
+- Mission: Easier R-based clinical trial regulatory submissions
+- [pilot 1 - common analyses](https://rconsortium.github.io/submissions-wg/pilot-overall.html#pilot-1---common-analyses)
+  - 2021.11   Initial submission to FDA. 
+  - 2022.12   Received FDA response 
+  - 2022.02   Revised submission to FDA
+  - 2022.03   Received final FDA response.
 
@@
 
+### Initial Submission
+
 - "Using R version 4.1.1, FDA was able to run the submitted code and confirm the applicant's tables and the submitted figure in report-tlf pdf file."
 - "Using FDA developed code, a statistical analyst was able to independently generate tables using the submitted data."
-- [Original FDA response](https://github.com/RConsortium/submissions-wg/blob/main/Documents/Summary_R_Pilot_Submission.pdf)
+- [Initial submission FDA response](https://github.com/RConsortium/submissions-wg/blob/main/Documents/Summary_R_Pilot_Submission.pdf)
+
+### Revised Submission
+
+- FDA agrees that the initial phase of the R Pilot submission has been completed.
+- [Revised  submission FDA response](https://github.com/RConsortium/submissions-wg/blob/main/Documents/Summary_R_Pilot_Submission2.pdf)
 
@@
 
-- Address minor issues reported by FDA reviewers.
-- Pilot R Shiny app submission.
+- Pilot R Shiny app submission to FDA.
+- Seek pilot submission opportunities to other regulator agencies. 
 - Pilot submission with advanced analysis methods (e.g., study design, missing data, Bayesian, etc).
@@
 
-## r2rtf: design
-
-`r2rtf` is designed to:
-
-- Generate highly customized tables
-- Limit package dependency
-- Target regulatory deliverable
-- Support pipes (`%>%`)
-
-## r2rtf: minimal example
-
-`r2rtf` is designed to be pipe-friendly (`%>%`)
-
-```{r, echo = TRUE, eval = FALSE}
-head(iris) %>%
-  rtf_body() %>%           # Step 1 Add table attributes
-  rtf_encode() %>%         # Step 2 Convert attributes to RTF encode
-  write_rtf("minimal.rtf") # Step 3 Write to a .rtf file
-```
-
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("minimal.pdf")
-```
-
-## r2rtf: function illustration
+## r2rtf: create TLFs in RTF format
 
@@
 
-Website: <https://merck.github.io/pkglite/>
-
 ## cleanslate: portable R environments
@@
 - Install a specific version of Rtools into the project folder 
-- (without administrator privileges)
+(without administrator privileges)
 
@@
 
-- As a statistician, I use `tidyverse`, `r2rtf`,
+- As statisticians, we use `tidyverse`, `r2rtf`,
 and internal tools to define the mock-up table, listing and figure (TLFs) for statistical analysis of a clinical trial.
@@
 
-- As a programmer, I use `tidyverse`, `r2rtf`, and internal tools to develop and/or validate analysis results based on mock-up TLFs.
+- As programmers, we use `tidyverse`, `r2rtf`, and internal tools to develop and/or validate analysis results based on mock-up TLFs.
 
-- As a statistican/programmer, I use `pkglite` and internal tools to prepare proprietary R packages and analysis R scripts for eCTD submission package.
+- As statisticans/programmers, we use `pkglite` and internal tools to prepare proprietary R packages and analysis R scripts for eCTD submission package.
 
-- As an internal/external reviewer, I use `cleanslate` to re-construct a portable environment (if required) to reproduce analysis results.
+- As an internal/external reviewers, we use `cleanslate` to re-construct a portable environment (if required) to reproduce analysis results.
 

### 2022-05-15 slides/render.R

@@ -1,3 +1,4 @@
 rmarkdown::render("slides/workshop-slides.Rmd")
 rmarkdown::render("slides/china-r.Rmd")
 rmarkdown::render("slides/r4csr-gwu.Rmd")
+rmarkdown::render("slides/r4csr-rstudio.Rmd")

### 2022-05-15 slides/r4csr-rstudio.Rmd
fix typo

@@
 
-- All opinions expressed are those of the presenter and not represent any organization.
+- All opinions expressed are those of the presenter and do not represent any organization.
 
@@
 
-- R is flexible for handling novel missing data approaches following estimand framework in ICH E9 (R1)
+- R is flexible for handling novel missing data approaches following the estimand framework in ICH E9 (R1)
   - Recurrent event data: @gao2017control @diao2020efficient
@@
   - In the pharmaceutical industry, RTF/Microsoft Word play a central role in preparing clinical study reports.
-  - Different organization can have different table standards.
+  - Different organizations can have different table standards.
 
@@
 
-- The R submission working group is a cross industry pharma working group focusing on improving practices of R-based clinical trial regulatory submission.
+- The R submission working group is a cross-industry pharma working group focusing on improving practices of R-based clinical trial regulatory submission.
   - [members and affiliations](https://rconsortium.github.io/submissions-wg/)
@@
 
-- How to submit internal developed (proprietary) R packages?
+- How to submit internally developed (proprietary) R packages?
 - How to follow [ICH and/or FDA guidances](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd)
@@
 
-- As statisticans/programmers, we use `pkglite` and internal tools to prepare proprietary R packages and analysis R scripts for eCTD submission package.
+- As statisticians/programmers, we use `pkglite` and internal tools to prepare proprietary R packages and analysis R scripts for eCTD submission packages.
 
@@
 - R/Pharma: <https://rinpharma.com/>
-  - Annual conference focus on the use of R in the development of pharmaceuticals.
+  - The annual conference focus on the use of R in the development of pharmaceuticals.
 

### 2022-07-12 preface.Rmd
Remove section numbering in preface
Update author email
update R package links

@@
 
-## Folder structure
+## Folder structure {.unnumbered}
 
@@
 
-## In this book
+## In this book {.unnumbered}
 
@@
 
-## Philosophy
+## Philosophy {.unnumbered}
 
 We share the same philosophy described in
-Section 1.1 of the [R Packages](https://r-pkgs.org/intro.html#intro-phil) book
+Section 1.1 of the [R Packages](https://r-pkgs.org/Introduction.html#intro-phil) book
 and quote here.
@@
 
-## Authors and contributors
+## Authors and contributors {.unnumbered}
 
@@
 
-  Yilong Zhang <<yilong.zhang@merck.com>>, Nan Xiao, Keaven Anderson, Yalin Zhu
+  Yilong Zhang <<elong0527@gmail.com>>, Nan Xiao, Keaven Anderson, Yalin Zhu
 

### 2022-07-12 project-folder.Rmd
update R package links

@@
 For example, the header below defines each variable in `fmt_est`.
-More details can be found in [Chapter 10 of the R Packages book](https://r-pkgs.org/man.html).
+More details can be found in [Chapter 16 of the R Packages book](https://r-pkgs.org/man.html).
 
@@
 Then `devtools::document()` is used to generate all the documentation files in the `man/` folder.
-More details can be found in [Chapter 10 of the R Packages book](https://r-pkgs.org/man.html).
+More details can be found in [Chapter 16 of the R Packages book](https://r-pkgs.org/man.html).
 
@@
 - [`devtools`](https://devtools.r-lib.org/): make package development easier.
-  + A good overview can be found in [Chapter 2 of the R Packages book](https://r-pkgs.org/whole-game.html).
+  + A good overview can be found in [Chapter 2 of the R Packages book](https://r-pkgs.org/Whole-game.html).
   + `devtools::load_all()`: load all functions in `R/` folder and running environment.
@@
 For an R package, there are already criteria to ensure R package integrity.
-More details can be found in [Chapter 19 of the R Packages book](https://r-pkgs.org/r-cmd-check.html).
+More details can be found in [Chapter 20 of the R Packages book](https://r-pkgs.org/check.html).

### 2022-07-12 project-management.Rmd
update R package links

@@
 The testthat R package can be used to streamline the validation process.
-More details of the `testthat` package for R package validation can be found in [Chapter 12 of the R package book](https://r-pkgs.org/tests.html).
+More details of the `testthat` package for R package validation can be found in [Chapter 12 of the R package book](https://r-pkgs.org/testing-basics.html).
 

### 2022-07-20 tlf-assemble.Rmd
Add TLF assemble chapter

# Assemble RTF TLFs {#assemble}

## Assemble TLFs to an RTF File

After TLFs are created, we may need to assemble them to one file to simplify review and delivery.
You can use `r2rtf::rtf_assemble()` to achieve the goal.

```{r, include=FALSE}
source("common.R")
```

```{r}
library(r2rtf)
```

We use the TLFs generated in the previous chapters as an example to illustrate the workflow.

### List RTF files to assemble

List all RTF files in the `outtable/` directory and/or the `outgraph/` directory.

The R code below lists all `.rtf` files in the output tables in the
`outtable/` directory and the output graphs in the `outgraph/` directory.

<!-- Note to authors: please modify the code below to find a better way to present this -->

```{r, eval=FALSE, echo=TRUE}
path <- list(
  outtable = "tlf/",
  outgraph = "tlf/"
)

# For a project the path has been pre-defined
rtf_table_path <- list.files(path$outtable, pattern = ".rtf")
rtf_graph_path <- list.files(path$outgraph, pattern = ".rtf")

dput(rtf_table_path)
#> c("intro-ae1.rtf", "intro-ae2.rtf", "intro-ae3.rtf", "intro-ae5.rtf",
#> "intro-ae7.rtf", "tbl_disp.rtf", "tbl_pop.rtf", "tlf_ae_summary.rtf",
#> "tlf_base.rtf", "tlf_eff.rtf", "tlf_eff1.rtf", "tlf_eff2.rtf",
#> "tlf_km.rtf", "tlf_spec_ae.rtf")

dput(rtf_graph_path)
#> c("intro-ae1.rtf", "intro-ae2.rtf", "intro-ae3.rtf", "intro-ae5.rtf",
#> "intro-ae7.rtf", "tbl_disp.rtf", "tbl_pop.rtf", "tlf_ae_summary.rtf",
#> "tlf_base.rtf", "tlf_eff.rtf", "tlf_eff1.rtf", "tlf_eff2.rtf",
#> "tlf_km.rtf", "tlf_spec_ae.rtf")
```

The complete list of `.rtf` files are printed in the Console.
You can then use this list as a starting point to arrange the order of all files.

An alternative way to get the RTF file list is to read from a spreadsheet
(e.g., using A&R grid) and work from there.

### Define file order and identify page orientation

Next, we should properly define the order of all TLF files,
and identify each TLF's page orientation (horizontal or vertical).

The R code below constructs an intermediate data frame `db` with three columns.
Each TLF file has its respective row to indicate

- `table`: Whether it is table or a plot, `TRUE` or `FALSE`
- `file`: File name
- `landscape`: Page orientation, `TRUE` or `FALSE`

<!-- Note to authors: please modify the code below following the last chunk, this is just a placeholder -->

```{r}
db <- data.frame(
  file = c(
    "intro-ae1.rtf",
    "intro-ae2.rtf",
    "intro-ae3.rtf",
    "intro-ae5.rtf",
    "intro-ae7.rtf",
    "tbl_disp.rtf",
    "tbl_pop.rtf",
    "tlf_ae_summary.rtf",
    "tlf_base.rtf",
    "tlf_eff.rtf",
    "tlf_eff1.rtf",
    "tlf_eff2.rtf",
    "tlf_km.rtf",
    "tlf_spec_ae.rtf"
  ),
  table = c(rep(TRUE, 3), FALSE, rep(TRUE, 6), FALSE, rep(TRUE, 2), FALSE),
  landscape = c(rep(FALSE, 13), rep(TRUE, 1)),
  stringsAsFactors = FALSE
)

db
```

<!-- Note to authors: the next section can probably be avoided if you capture both the file name and the full path in different columns in the data frame above, and consequently avoid the manual copy & paste. I guess one reason you did this is you had to convert the paths from Linux to Windows in the original Rmd, although that is not a generic flow. -->

### Construct full paths to RTF files

The R code below constructs file paths from two directories: `outtable` and `outgraph`.
You can extend the code to more than two directories.

```{r, eval=FALSE, echo=TRUE}
full_path <- ifelse(
  db$table,
  file.path(path$outtable, db$file),
  file.path(path$outgraph, db$file)
)
```

### Combine RTF files into one file

Combine RTF files into one file and saved in the `path$document` directory.

```{r, eval=FALSE, echo=TRUE}
r2rtf::rtf_assemble(
  input = full_path,
  output = file.path(path$document, "output.docx"),
  landscape = db$landscape
)
```

## Using toggle fields

The files are assembled using toggle fields.

- Use `Alt` + `F9` to **display** and **hide fields**.

```{r, echo=FALSE, out.width="99%", fig.cap="Using Alt + F9 to display fields", fig.align="center"}
knitr::include_graphics("images/rtf-alt-f9.png")
```

- To [update TLFs](https://support.microsoft.com/en-us/office/update-fields-7339a049-cb0d-4d5a-8679-97c20c643d4e), select field → right click → Update Field.

Below is how the file link looks like after the field got updated.

```{r, echo=FALSE, out.width="99%", fig.cap="Update fields", fig.align="center"}
knitr::include_graphics("images/rtf-update-field.png")
```

- Use `Alt` + `F9` to **display the tables**.
- Save the files.

Below is a snapshot of the final Word document output.

```{r, echo=FALSE, out.width="99%", fig.cap="Update fields", fig.align="center"}
knitr::include_graphics("images/rtf-after-update.png")
```

- We suggest to test one toggle field before updating all of them.
- If you modified content within a toggle field, remove all
  `\* MERGEFORMAT ` before updating toggle fields.
- If a document is ready for delivery to stakeholders, please
  [unlink toggle fields](https://success.powerdms.com/s/article/Removing-Fields-from-Word-Documents?language=en_US).

### 2022-08-12 tlf-assemble.Rmd
update tlf-assemble.Rmd

@@
-# Assemble RTF TLFs {#assemble}
-
-## Assemble TLFs to an RTF File
-
-After TLFs are created, we may need to assemble them to one file to simplify review and delivery.
-You can use `r2rtf::rtf_assemble()` to achieve the goal.
+# Assemble TLFs {#assemble}
 
@@
 
-We use the TLFs generated in the previous chapters as an example to illustrate the workflow.
+A CSR needs to include many TLFs. 
+After TLFs are created and saved into individual files, 
+we need to assemble them into one files in a pre-specified order.
 
-### List RTF files to assemble
+There are two general approach to achieve the goal.
 
-List all RTF files in the `outtable/` directory and/or the `outgraph/` directory.
+1. Combine RTF source code in individual files into one large RTF file. 
+1. Leverage Microsoft Word `Toggle Fields` feature to embed RTF files using 
+hyperlinks.
 
-The R code below lists all `.rtf` files in the output tables in the
-`outtable/` directory and the output graphs in the `outgraph/` directory.
-
-<!-- Note to authors: please modify the code below to find a better way to present this -->
-
-```{r, eval=FALSE, echo=TRUE}
-path <- list(
-  outtable = "tlf/",
-  outgraph = "tlf/"
-)
-
-# For a project the path has been pre-defined
-rtf_table_path <- list.files(path$outtable, pattern = ".rtf")
-rtf_graph_path <- list.files(path$outgraph, pattern = ".rtf")
-
-dput(rtf_table_path)
-#> c("intro-ae1.rtf", "intro-ae2.rtf", "intro-ae3.rtf", "intro-ae5.rtf",
-#> "intro-ae7.rtf", "tbl_disp.rtf", "tbl_pop.rtf", "tlf_ae_summary.rtf",
-#> "tlf_base.rtf", "tlf_eff.rtf", "tlf_eff1.rtf", "tlf_eff2.rtf",
-#> "tlf_km.rtf", "tlf_spec_ae.rtf")
-
-dput(rtf_graph_path)
-#> c("intro-ae1.rtf", "intro-ae2.rtf", "intro-ae3.rtf", "intro-ae5.rtf",
-#> "intro-ae7.rtf", "tbl_disp.rtf", "tbl_pop.rtf", "tlf_ae_summary.rtf",
-#> "tlf_base.rtf", "tlf_eff.rtf", "tlf_eff1.rtf", "tlf_eff2.rtf",
-#> "tlf_km.rtf", "tlf_spec_ae.rtf")
-```
-
-The complete list of `.rtf` files are printed in the Console.
-You can then use this list as a starting point to arrange the order of all files.
-
-An alternative way to get the RTF file list is to read from a spreadsheet
-(e.g., using A&R grid) and work from there.
-
-### Define file order and identify page orientation
-
-Next, we should properly define the order of all TLF files,
-and identify each TLF's page orientation (horizontal or vertical).
-
-The R code below constructs an intermediate data frame `db` with three columns.
-Each TLF file has its respective row to indicate
-
-- `table`: Whether it is table or a plot, `TRUE` or `FALSE`
-- `file`: File name
-- `landscape`: Page orientation, `TRUE` or `FALSE`
-
-<!-- Note to authors: please modify the code below following the last chunk, this is just a placeholder -->
+Let's illustrate the idea by using selected TLFs generated from previous chapters. 
+Here, we assume files are saved in the `tlf` folder.
 
 ```{r}
-db <- data.frame(
-  file = c(
-    "intro-ae1.rtf",
-    "intro-ae2.rtf",
-    "intro-ae3.rtf",
-    "intro-ae5.rtf",
-    "intro-ae7.rtf",
-    "tbl_disp.rtf",
-    "tbl_pop.rtf",
-    "tlf_ae_summary.rtf",
-    "tlf_base.rtf",
-    "tlf_eff.rtf",
-    "tlf_eff1.rtf",
-    "tlf_eff2.rtf",
-    "tlf_km.rtf",
-    "tlf_spec_ae.rtf"
-  ),
-  table = c(rep(TRUE, 3), FALSE, rep(TRUE, 6), FALSE, rep(TRUE, 2), FALSE),
-  landscape = c(rep(FALSE, 13), rep(TRUE, 1)),
-  stringsAsFactors = FALSE
-)
-
-db
-```
-
-<!-- Note to authors: the next section can probably be avoided if you capture both the file name and the full path in different columns in the data frame above, and consequently avoid the manual copy & paste.
-I guess one reason you did this is you had to convert the paths from Linux to Windows in the original Rmd, although that is not a generic flow.
--->
-
-### Construct full paths to RTF files
-
-The R code below constructs file paths from two directories: `outtable` and `outgraph`.
-You can extend the code to more than two directories.
-
-```{r, eval=FALSE, echo=TRUE}
-full_path <- ifelse(
-  db$table,
-  file.path(path$outtable, db$file),
-  file.path(path$outgraph, db$file)
+tlf_path <- c(
+  "tlf/tlf_disp.rtf",  # Disposition Table 
+  "tlf/tlf_eff.rtf",   # Efficacy Table
+  "tlf/tlf_km.rtf"     # KM Plot 
 )
@@
 
-### Combine RTF files into one file
+## Combine RTF Source Code
 
-Combine RTF files into one file and saved in the `path$document` directory.
+The `r2rtf::rtf_assemble()` function allow you to combine RTF 
+source code in individual files into one larger RTF file.
 
-```{r, eval=FALSE, echo=TRUE}
-r2rtf::rtf_assemble(
-  input = full_path,
-  output = file.path(path$document, "output.docx"),
-  landscape = db$landscape
-)
-```
+::: {.rmdnote}
+One limitation of combining RTF source code is that 
+we are not able to specify page orientation of each TLF 
+in the combined document. 
+:::
 
-## Using toggle fields
+## Using Toggle Fields
 
-The files are assembled using toggle fields.
+Microsoft Words uses toggle fields to embed files into one Word document. 
+The approach is a dynamic link between files by providing the file absolute path.
 
-- Use `Alt` + `F9` to **display** and **hide fields**.
+::: {.rmdnote}
+There is a small learning curve to learn how toggle fields work 
+in Microsoft Words. 
+After you become familiar with the workflow, 
+toggle fields can extend your capability to manage large amount of TLFs
+in RTF format.
+:::
+
+The `rtf_assemble()` function 
+allow you to create a word file with toggle fields as below.
+
+After open the generated `.docx` file, 
+you will see a blank file, because the file only contain fields
+with hyperlinks.
+
+By using `Alt` + `F9` to **display** the fields and you will 
+see information similar to the screenshot below.
 
@@
 
-- To [update TLFs](https://support.microsoft.com/en-us/office/update-fields-7339a049-cb0d-4d5a-8679-97c20c643d4e), select field → right click → Update Field.
+::: {.rmdnote}
+A typical error message is that system can not find the file 
+if you only provide relative path. 
+Please double check the correct absolute file path is in `INCLUDETEXT` field. 
+:::
 
-Below is how the file link looks like after the field got updated.
+To test the toggle field, you can right click one `INCLUDETEXT` filed and 
+click `Update Field`.
 
-```{r, echo=FALSE, out.width="99%", fig.cap="Update fields", fig.align="center"}
-knitr::include_graphics("images/rtf-update-field.png")
-```
-
-- Use `Alt` + `F9` to **display the tables**.
-- Save the files.
-
-Below is a snapshot of the final Word document output.
+If it works, you can see a table similar to the snapshot below
+by using `Alt + F9`. 
+It is a shortcut to change between results and field display mode.
 
@@
 
-- We suggest to test one toggle field before updating all of them.
-- If you modified content within a toggle field, remove all
-  `\* MERGEFORMAT ` before updating toggle fields.
-- If a document is ready for delivery to stakeholders, please
-  [unlink toggle fields](https://success.powerdms.com/s/article/Removing-Fields-from-Word-Documents?language=en_US).
+Now you can update all toggle fields to display all TLFs by 
+selecting all fields (`Ctrl + A`), then type `F9`. 
+We suggest to test one toggle field before updating all of them.
+
+As the `docx` file contain dynamic links, 
+you can keep updating the TLFs if you need to refresh content in 
+individual RTF files by selecting all fields (`Ctrl + A`), then type `F9`.
+
+::: {.rmdnote}
+If you modify table content in the combined `.docx` file, 
+you may get weird table layout if you update all fields within a toggle field. 
+To resolve the issue, please remove all `\* MERGEFORMAT` in the filed mode using `Alt + F9` before updating all toggle fields.
+:::
+
+After the combined TLF is ready for delivery, 
+you can also unlink toggle fields to save table contents, 
+because the absolute path may not work by others.  
+To unlink toggle fields, you can select all fields (`Ctrl + A`), 
+then type `Ctrl + Shift + F9`.

### 2022-08-12 tlf-assemble.Rmd
enable tmporary solution for `rtf_assemble`.

@@
 tlf_path <- c(
-  "tlf/tlf_disp.rtf",  # Disposition Table 
+  "tlf/tbl_disp.rtf",  # Disposition Table 
   "tlf/tlf_eff.rtf",   # Efficacy Table
@@
 
+```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("tlf/tlf-combine.pdf")
+```
+
 ## Combine RTF Source Code
 
-The `r2rtf::rtf_assemble()` function allow you to combine RTF 
+```{r, include=FALSE}
+#' Assemble RTF files into one single RTF file 
+#' 
+#' @param path a vector of RTF file path
+#' 
+#' @export
+rtf_assemble <- function(path){
+
+  rtf <- lapply(path, readLines)
+
+  n <- length(rtf)
+  start <- c(1, rep(2, n - 1))
+  end <- vapply(rtf, length, numeric(1))
+  end[-n] <- end[-n] - 1
+
+  for(i in 1:n){
+    rtf[[i]] <- rtf[[i]][start[i]:end[i]]
+    if(i < n) rtf[[i]] <- c(rtf[[i]], r2rtf:::as_rtf_new_page())
+  }
+  rtf <- do.call(c, rtf)
+
+  rtf
+
+}
+```
+
+The `rtf_assemble()` function allow you to combine RTF 
 source code in individual files into one larger RTF file.
@@
 :::
+
+```{r}
+rtf_assemble(tlf_path) |> 
+  r2rtf::write_rtf("tlf/rtf-combine.rtf")
+```
 

### 2022-08-12 tlf-assemble.Rmd
add ad-hoc functions.

@@
 
-```{r, include=FALSE}
+<details>
+<summary>`rtf_assemble1()` source code</summary>
+```{r}
 #' Assemble RTF files into one single RTF file 
@@
 #' @export
-rtf_assemble <- function(path){
+rtf_assemble1 <- function(path){
 
@@
 ```
+</details>
 
-The `rtf_assemble()` function allow you to combine RTF 
+The `rtf_assemble1()` function allow you to combine RTF 
 source code in individual files into one larger RTF file.
@@
 ```{r}
-rtf_assemble(tlf_path) |> 
+rtf_assemble1(tlf_path) |> 
   r2rtf::write_rtf("tlf/rtf-combine.rtf")
@@
 
-The `rtf_assemble()` function 
+<details>
+<summary>`rtf_assemble2()` source code</summary>
+```{r}
+#' Assemble RTF TLFs
+#'
+#' Add a set of RTF/TEXT fields into an rdocx object.
+#'
+#' @param input Character vector of file path.
+#' @param output Character string to the output file path.
+#' @param landscape Logical vector to determine whether to
+#' display files as portrait or landscape.
+#'
+#' @export
+rtf_assemble2 <- function(input, output, landscape = FALSE) {
+
+  input <- normalizePath(input)
+
+  if (!all(file.exists(input))) {
+    warning("Some files do not exist")
+  }
+
+  field <- ifelse(grepl("/", input),
+    paste0("INCLUDETEXT \"", gsub("/", "\\\\\\\\", input), "\""),
+    paste0("INCLUDETEXT \"", gsub("\\", "\\\\", input, fixed = "TRUE"), "\"")
+  )
+
+  if (length(landscape) == 1) {
+    landscape <- rep(landscape, length(field))
+  }
+
+  docx <- officer::read_docx()
+
+  for (i in seq_along(input)) {
+    docx <-
+      docx %>%
+      officer::body_add_fpar(
+        officer::fpar(
+          officer::ftext("Table "),
+          officer::run_word_field("SEQ Table \\* ARABIC"),
+          officer::run_linebreak(),
+          officer::run_word_field(field[i]),
+          officer::run_pagebreak()
+        )
+      )
+    if (landscape[i]) {
+      docx <- officer::body_end_section_landscape(docx)
+    } else {
+      docx <- officer::body_end_section_portrait(docx)
+    }
+  }
+
+  print(docx, target = output)
+
+  invisible(output)
+}
+```
+</details>
+
+The `rtf_assemble2()` function 
 allow you to create a word file with toggle fields as below.
+One benefit is to control the page direction of each TLF as below.
+
+```{r}
+rtf_assemble2(tlf_path, 
+              output = "tlf/rtf-combine-toggle.docx", 
+              landscape = c(FALSE, FALSE, TRUE))
+```
 

### 2022-08-12 tlf-assemble.Rmd
fix typo

@@
 
-A CSR needs to include many TLFs. 
 After TLFs are created and saved into individual files, 
-we need to assemble them into one files in a pre-specified order.
+we need to assemble them into one file in a pre-specified order.
 
-There are two general approach to achieve the goal.
+There are two general approaches to achieving the goal.
 
@@
 Let's illustrate the idea by using selected TLFs generated from previous chapters. 
-Here, we assume files are saved in the `tlf` folder.
+Here, we assume files are saved in the `tlf/` folder.
 
@@
 
-The `rtf_assemble1()` function allow you to combine RTF 
+The `rtf_assemble1()` function allows user to combine RTF 
 source code in individual files into one larger RTF file.
@@
 Microsoft Words uses toggle fields to embed files into one Word document. 
-The approach is a dynamic link between files by providing the file absolute path.
+The approach is a dynamic link between files by providing the absolute file path.
 
 ::: {.rmdnote}
-There is a small learning curve to learn how toggle fields work 
+There is a slight learning curve to learning how toggle fields work 
 in Microsoft Words. 
 After you become familiar with the workflow, 
-toggle fields can extend your capability to manage large amount of TLFs
+toggle fields can extend your capability to manage a large amount of TLFs
 in RTF format.
@@
 The `rtf_assemble2()` function 
-allow you to create a word file with toggle fields as below.
+allows you to create a word file with toggle fields as below.
 One benefit is to control the page direction of each TLF as below.
@@
 
-After open the generated `.docx` file, 
-you will see a blank file, because the file only contain fields
+After opening the generated `.docx` file, 
+you will see a blank file because the file only contains fields
 with hyperlinks.
 
-By using `Alt` + `F9` to **display** the fields and you will 
+By using `Alt + F9` to display the fields and you will 
 see information similar to the screenshot below.
@@
 A typical error message is that system can not find the file 
-if you only provide relative path. 
-Please double check the correct absolute file path is in `INCLUDETEXT` field. 
+if you only provide a relative path. 
+Please double-check that the correct absolute file path is in the `INCLUDETEXT` field. 
 :::
 
-To test the toggle field, you can right click one `INCLUDETEXT` filed and 
+To test the toggle field, you can right-click an `INCLUDETEXT` filed and 
 click `Update Field`.
@@
 selecting all fields (`Ctrl + A`), then type `F9`. 
-We suggest to test one toggle field before updating all of them.
+We suggest testing one toggle field before updating all of them.
 
@@
 If you modify table content in the combined `.docx` file, 
-you may get weird table layout if you update all fields within a toggle field. 
+you may get a weird table layout if you update all fields within a toggle field. 
 To resolve the issue, please remove all `\* MERGEFORMAT` in the filed mode using `Alt + F9` before updating all toggle fields.
@@
 you can also unlink toggle fields to save table contents, 
-because the absolute path may not work by others.  
+because the absolute path may not work for others.  
 To unlink toggle fields, you can select all fields (`Ctrl + A`), 

### 2022-08-12 index.Rmd

@@
 library(pkglite)
+library(officer)
 ```

### 2022-08-14 tlf-assemble.Rmd
Include rtf-combine.pdf instead of tlf-combine.pdf

@@
 tlf_path <- c(
-  "tlf/tbl_disp.rtf",  # Disposition Table 
-  "tlf/tlf_eff.rtf",   # Efficacy Table
-  "tlf/tlf_km.rtf"     # KM Plot 
+  "tlf/tbl_disp.rtf", # Disposition table
+  "tlf/tlf_eff.rtf",  # Efficacy table
+  "tlf/tlf_km.rtf"    # K-M plot
 )
@@
 ```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/tlf-combine.pdf")
+knitr::include_graphics("tlf/rtf-combine.pdf")
 ```
@@
 #' @export
-rtf_assemble1 <- function(path){
-
+rtf_assemble1 <- function(path) {
   rtf <- lapply(path, readLines)
@@
 
-  for(i in 1:n){
+  for (i in 1:n) {
     rtf[[i]] <- rtf[[i]][start[i]:end[i]]
-    if(i < n) rtf[[i]] <- c(rtf[[i]], r2rtf:::as_rtf_new_page())
+    if (i < n) rtf[[i]] <- c(rtf[[i]], r2rtf:::as_rtf_new_page())
   }
@@
   rtf
-
 }
@@
 rtf_assemble2 <- function(input, output, landscape = FALSE) {
-
   input <- normalizePath(input)
@@
 ```{r}
-rtf_assemble2(tlf_path, 
+rtf_assemble2(
+  tlf_path,
   output = "tlf/rtf-combine-toggle.docx",
-              landscape = c(FALSE, FALSE, TRUE))
+  landscape = c(FALSE, FALSE, TRUE)
+)
 ```

### 2022-08-17 slides/fda-workshop-slides.Rmd

--- 2022-02-28 slides/workshop-slides.Rmd
@@
 title: "R for Clinical Study Reports and Submission"
-subtitle: "ASA Princeton-Trenton Chapter"
-author: "Yilong Zhang, Nan Xiao"
-date: "December 2021"
+subtitle: "The ASA Biopharmaceutical Section Regulatory-Industry Statistics Workshop"
+author: "Yilong Zhang, Nan Xiao, Keaven Anderson"
+date: "September 2022"
 output:
-  mkdocs::mk_ioslides_presentation:
-    mk_logo: "invent"
-    mk_col_title: "white"
-    mk_col_section: "teal"
-    mk_class: "public"
+  ioslides_presentation:
     widescreen: true
+    smaller: true
+bibliography: reference.bib
 ---
@@
 knitr::opts_chunk$set(comment = "#>", echo = FALSE)
+```
+
+```{css, echo = FALSE}
+slides > slide.dark {
+  background: #59B2D1 !important;
+}
+
+.dark hgroup {
+  position: absolute;
+  left: 5%;
+  top: 40%;
+}
+
+slides > slide.title-slide {
+  background: #f2f2f2;
+}
+
+.title-slide hgroup {
+  position: absolute;
+  left: 5%;
+  top: 40%;
+}
+
+.title-slide hgroup h1 {
+  color: #404040;
+  font-size: 48px;
+}
+
+.title-slide hgroup h2 {
+  color: #404040;
+  font-size: 24px;
+}
+
+.title-slide hgroup p {
+  color: #404040;
+  font-size: 16px;
+}
+
+h1 {
+ color: #404040;
+}
+
+h2 {
+ color: #404040;
+}
+
+h3 {
+ color: #404040;
+}
+
+p {
+ color: #404040;
+}
+
+ul {
+ color: #404040;
+}
+
+li {
+ color: #404040;
+}
+
+a {
+    color: #59B2D1;
+    text-decoration: none;
+    border-bottom: none;
+}
+
+a > code {
+    color: #59B2D1;
+    text-decoration: none;
+    border-bottom: none;
+}
+
+slides > slide:not(.segue) > hgroup > h2 {
+    font-size: 34px;
+    line-height: 50px;
+    padding: 0px 0px 10px 0px;
+    border-bottom: 2px solid #59B2D1;
+}
 ```
@@
 
-- Delivering TLFs in CSR (Yilong Zhang)
+- Delivering TLFs in CSR 
   - Session 1 (45 min)
@@
   - Break (10 min)
-- Clinical trial project (Nan Xiao)
+- Clinical trial project 
   - Session 3 (30 min)
   - Break (10 min)
-- eCTD submission package (Nan Xiao)
+- eCTD submission package 
   - Session 4 (40 min)
@@
 
-- All opinions expressed are those of the presenter and not
-  Merck Sharp & Dohme Corp., a subsidiary of Merck & Co., Inc., Kenilworth, NJ, USA.
+- All opinions expressed are those of the presenter.
 
@@
 
-## Training objective
+## Training Objective
 
-- Learning how to create tables for clinical study reports.
+- Learning how to create TLFs for clinical study reports.
 - Learning how to organize a clinical development project.
@@
 
-- Team members from Merck & Co.
-  - [We are hiring!](https://jobs.merck.com/bards)
-  - Contact us if interested.
-
-- [Contributors](https://r4csr.org/) of the training materials.
-  - Please consider submitting issues or PR in the repo.
+- [Contributors](https://r4csr.org/preface.html#authors-and-contributors) of the training materials.
+  - Please consider submitting issues or PR in the [github repo](https://github.com/elong0527/r4csr).
 
@@
   - Please consider to join and contribute to a working group.
-
-- [ASA Princeton-Trenton Chapter](https://community.amstat.org/princetontrenton/home)
-  - Please consider to be a member of our chapter.
-
-<!--
-
-- R / Pharma organizers
-  + It is a fun and productive annual gathering.
-  + Please consider sharing stories and use cases to expand the community.
-
--->
 
@@
   - `esubdemo` is shared in RStudio Cloud (<https://rstudio.cloud/>)
-  - Project link will be shared in chat
-  - Post questions in group chat.
+  - Project link will be shared at the beginning of the course.
+
+## Motivation
+
+As an organization, we need to ensure compliance and reduce the risk of using R and R packages in regulatory deliverables.
+
+- R is widely used in clinical trial study design.
+  - Following @lachin1986evaluation
+  - [`gsDesign`](https://github.com/keaven/gsDesign): an R package for group sequential design under proportional hazards.
+  - [`simtrial`](https://github.com/Merck/simtrial),
+  [`gsDesign2`](https://github.com/Merck/gsDesign2), and
+  [`gsdmvn`](https://github.com/Merck/gsdmvn): experimental R packages for group sequential design under non-proportional hazards
+  - Bookdown: <https://keaven.github.io/gsd-deming/>
+
+- R is flexible for handling novel missing data approaches following the estimand framework in ICH E9 (R1)
+  - Recurrent event data: @gao2017control @diao2020efficient
+  - Robustness: @liu2021multiply
+
+- R is used in Bayesian statistics.
+  - [stan](https://mc-stan.org/users/interfaces/rstan) and network meta-analysis for drug reimbursement analysis
+
+- R is widely used in visualization
+  - [SafetyGraphics](https://github.com/SafetyGraphics/safetyGraphics)
+  - [forestly](https://elong0527.github.io/forestly/articles/example.html#construct-interactive-forest-plot-1): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_194/)
 
@@
 
-## ICH E3 guidance
+## ICH E3 Guidance
 
-The [ICH E3: structure and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) provide guidance to assist sponsors in the development of a CSR.
+The [ICH E3: structure and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) 
+provides guidance to assist sponsors in the development of a CSR.
 
@@
 
-- Public available CDISC pilot [study data located at CDISC Bitbucket repository](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).
+- Public available [CDISC pilot study data](https://github.com/elong0527/r4csr/tree/master/data-adam).
 
 - The dataset structure follows the [CDISC Analysis Data Model (ADaM)](https://www.cdisc.org/standards/foundational/adam).
-
-- Source data: <https://github.com/elong0527/r4csr/tree/master/data-adam>
 
@@
 
-# r2rtf introduction
+# r2rtf Introduction
 
@@
 
-## Minimal example
+## Minimal Example
 
@@
 
-## Package overview
+## Package Overview
 
@@
 
-## Simple example: adverse events
-
-`r2rtf` only focus on **table format**.
-Data manipulation and analysis should be handled by other R packages. (e.g., `tidyverse`)
-
-- Let's refer to examples in the [Section 1.4.2](https://r4csr.org/tlf-overview.html#r2rtf)
-
-## Function summary
+## Function Summary
 
@@
 
-## Function illustration
+## Function Illustration
 
@@
 
-# Break and/or exercise (20 min)
+# Break and/or Exercise (20 min)
 
-# CSR examples
+# CSR Examples
 
-## Disposition table
+## Disposition Table
 
@@
 
-## Analysis population
+## Analysis Population
 
@@
 
-## Baseline characteristics
+## Baseline Characteristics
 
@@
 
-## Efficacy table
+## Efficacy Table
 
@@
 
-## AE Summary table
+## AE Summary Table
 
@@
 
-## Specific AE table
+## Specific AE Table
 
@@
 
+## Assemble TLFs
+
+<https://r4csr.org/assemble.html>
+
 # Break (10 min)
 
-# Clinical trial project
+# Clinical Trial Project
 
@@
 
-## Folder structure
+## Folder Structure
 
@@
 
-## Project management
+## Project Management
 
@@
 
-# eCTD submission
+# eCTD Submission
 
-## Overview
+## Clarification from FDA
 
-<https://r4csr.org/submission-overview.html>
+- [FDA: Statistical Software Clarifying Statement](https://www.fda.gov/files/about%20fda/published/Statistical-Software-Clarifying-Statement-PDF.pdf)
 
-## Submission package
+> "FDA does not require use of any specific software for statistical analyses, and statistical software
+is not explicitly discussed in Title 21 of the Code of Federal Regulations [e.g., in 21CFR part
+11]. However, the software package(s) used for statistical analyses should be fully documented
+in the submission, including version and build identification."
+
+## R Consortium Pilot Submission Working group
+
+- The R submission working group is a cross-industry pharma working group focusing on improving practices of R-based clinical trial regulatory submission.
+  - [members and affiliations](https://rconsortium.github.io/submissions-wg/)
+- Mission: Easier R-based clinical trial regulatory submissions
+- [pilot 1 - common analyses](https://rconsortium.github.io/submissions-wg/pilot-overall.html#pilot-1---common-analyses)
+  - 2021.11   Initial submission to FDA. 
+  - 2022.12   Received FDA response 
+  - 2022.02   Revised submission to FDA
+  - 2022.03   Received final FDA response.
+
+## Challenges and Assumptions
+
+### Challenges
+
+- How to submit internally developed (proprietary) R packages?
+- How to follow [ICH and/or FDA guidances](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd)
+in preparing eCTD package?
+- How to enhance reproducibility from FDA reviewer's perspective?
+
+### Assumptions
+
+- Focus on analysis and reporting given available ADaM datasets.
+
+## Deliverables
+
+- Open source pilot eCTD submission package to FDA
+  - Proprietary R package
+  - R scripts for analysis
+  - Analysis data reviewer guide (ADRG)
+  - etc.
+- Development: <https://github.com/RConsortium/submissions-pilot1>
+- eCTD package: <https://github.com/RConsortium/submissions-pilot1-to-fda>
+- [R consortium announcement](https://www.r-consortium.org/blog/2021/12/08/successful-r-based-test-package-submitted-to-fda)
+
+## FDA Response
+
+### Initial Submission
+
+- "Using R version 4.1.1, FDA was able to run the submitted code and confirm the applicant's tables and the submitted figure in report-tlf pdf file."
+- "Using FDA developed code, a statistical analyst was able to independently generate tables using the submitted data."
+- [Initial submission FDA response](https://github.com/RConsortium/submissions-wg/blob/main/Documents/Summary_R_Pilot_Submission.pdf)
+
+### Revised Submission
+
+- FDA agrees that the initial phase of the R Pilot submission has been completed.
+- [Revised  submission FDA response](https://github.com/RConsortium/submissions-wg/blob/main/Documents/Summary_R_Pilot_Submission2.pdf)
+
+## Future Work
+
+- Pilot R Shiny app submission to FDA.
+- Seek pilot submission opportunities to other regulator agencies. 
+- Pilot submission with advanced analysis methods (e.g., study design, missing data, Bayesian, etc).
+
+## Submission Package
 
@@
 
-## Running environment
+## Running Environment
 
@@
 
-## pkglite: compact package representations
+## pkglite: Compact Package Representations
 
@@
 
-## Pipe-friendly workflow
+## Pipe-friendly Workflow
 
@@
 
-## File specifications + file collections
+## File Specifications + File Collections
 
@@
 # Q&A
+
+## References

### 2022-08-17 slides/render.R
Deploying to gh-pages from @ elong0527/r4csr@489af2dc28488b32c8cf9caad67ac30fc440b04f 🚀

@@ -2,3 +2,4 @@ rmarkdown::render("slides/workshop-slides.Rmd")
 rmarkdown::render("slides/china-r.Rmd")
 rmarkdown::render("slides/r4csr-gwu.Rmd")
 rmarkdown::render("slides/r4csr-rstudio.Rmd")
+rmarkdown::render("slides/fda-workshop-slides.Rmd")

### 2022-09-20 slides/fda-workshop-slides.Rmd
update project information

@@
   - `esubdemo` is shared in RStudio Cloud (<https://rstudio.cloud/>)
-  - Project link will be shared at the beginning of the course.
+  - [Project link](https://rstudio.cloud/spaces/183786/join?access_code=FsVLtAUPS9s8uSc38ybNDTr3LYgFWWVq2OQjl8L4).
 
@@
 - R is widely used in visualization
-  - [SafetyGraphics](https://github.com/SafetyGraphics/safetyGraphics)
-  - [forestly](https://elong0527.github.io/forestly/articles/example.html#construct-interactive-forest-plot-1): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_194/)
+  - [metalite.table1](https://elong0527.github.io/metalite.table1/articles/metalite-table1.html): Interactive demographic table.
+  - [forestly](https://elong0527.github.io/forestly/articles/forestly.html): Interactive forest plot for DMC safety monitoring in clinical trials
 

### 2022-09-20 slides/fda-workshop-slides.Rmd
Deploying to gh-pages from @ elong0527/r4csr@9eb966545461ba0193ea2284e5d1e59f0739f74a 🚀

@@
   - `esubdemo` is shared in RStudio Cloud (<https://rstudio.cloud/>)
-  - [Project link](https://rstudio.cloud/spaces/183786/join?access_code=FsVLtAUPS9s8uSc38ybNDTr3LYgFWWVq2OQjl8L4).
+  - You can also clone the `esubdemo` project to your local RStudio.
 

### 2022-10-24 slides/r4csr-psi.Rmd
add psi slides

--- 2022-05-15 slides/r4csr-rstudio.Rmd
@@
 ---
-title: "R for Clinical Study Reports and Submission"
-subtitle: "RStudio Pharma Meetup Series"
-author: "Yilong Zhang"
-date: "May 2022"
+title: "R submission Pilots to FDA"
+subtitle: "PSI Webinar Series: Showcasing R use in Pharma"
+author: "Yilong Zhang on Behalf of R Consortium “R Submission” Working Group"
+date: "Oct. 2022"
 output:
@@
 
-# R Consortium pilot submission
+# R Consortium Pilot Submission
 
-## Working group
+## The R Consortium R Submission Working Group
 
@@
   - [members and affiliations](https://rconsortium.github.io/submissions-wg/)
-- Mission: Easier R-based clinical trial regulatory submissions
+
+```{r, out.width = "60%", echo = FALSE, fig.align = "center"}
+knitr::include_graphics("submission-group-logo.png")
+```
+
+## Objective and Timeline
+
+- Objective: Create open source example for R-based clinical trial regulatory submissions
 - [pilot 1 - common analyses](https://rconsortium.github.io/submissions-wg/pilot-overall.html#pilot-1---common-analyses)
@@
 
-## Challenges and assumptions
+- [pilot 2 - shiny app](https://github.com/RConsortium/submissions-pilot2-to-fda)
+  - 2022.10   Public review of initial submission to FDA.
+
+- [pilot 3 - ADaM data](https://github.com/RConsortium/submissions-wg/blob/main/Documents/R%20Submission%20%20Pilot%201%20Extension%20Proposal.pdf)
+  - 2022.10   Proposal initiation.
+
+## Pilot 1 Challenges and Assumptions
 
@@
 
-## FDA response
+## FDA Response
 
@@
 
-## Future work
+## Future Work
 
-- Pilot R Shiny app submission to FDA.
 - Seek pilot submission opportunities to other regulator agencies. 
@@
 
-## Reproducibility spectrum
+## Reproducibility Spectrum
 
@@
 
-## Tools for reporting and submission
+## Tools for Submission
 
-Tools:
+- Tool: [`pkglite`](https://merck.github.io/pkglite/), represent and exchange R package source code as text files.
 
-- [`r2rtf`](https://merck.github.io/r2rtf/): create production-ready tables and figures in RTF format.
-- [`pkglite`](https://merck.github.io/pkglite/): represent and exchange R package source code as text files.
-- `cleanslate` (under internal validation): create portable R environments.
+- Reference: @zhao2022electronic
 
-Bookdown: <https://r4csr.org/>
+- Book: <https://r4csr.org/>
 
-## r2rtf: create TLFs in RTF format
-
-```{r, out.width="100%", echo=FALSE, fig.align="center"}
-knitr::include_graphics("function-summary.png")
-```
-
-## pkglite: compact package representations
+## pkglite: Compact Package Representations
 
@@
 
-## cleanslate: portable R environments
-
-- Create a project folder with specific context (`.Rproj`, `.Rprofile`, `.Renviron`)
-- Install a specific version of R into the project folder
-- Install a specific version of Rtools into the project folder 
-(without administrator privileges)
-
-```{r, eval=FALSE}
-library("cleanslate")
-
-"portable-project/" %>%
-  use_project(repo = "https://url/snapshot/2021-11-20/") %>%
-  use_rprofile() %>%
-  use_renviron() %>%
-  use_r_version(version = "4.1.1") %>%
-  use_rtools(version = "rtools40")
-```
-
-## Summary of workflow as user stories
-
-Within a regulatory R environment:
-
-- As statisticians, we use `tidyverse`, `r2rtf`,
-and internal tools to define the mock-up table, listing and figure (TLFs) for statistical analysis of a clinical trial.
-  - More than 100 TLFs for efficacy and safety of a drug or vaccine.
-
-- As programmers, we use `tidyverse`, `r2rtf`, and internal tools to develop and/or validate analysis results based on mock-up TLFs.
-
-- As statisticians/programmers, we use `pkglite` and internal tools to prepare proprietary R packages and analysis R scripts for eCTD submission packages.
-
-- As an internal/external reviewers, we use `cleanslate` to re-construct a portable environment (if required) to reproduce analysis results.
-
-More details: <https://r4csr.org/>
-
-## Folder structure
+## Folder Structure
 
@@
 
-## Cross-industry collaborations
+## Cross-industry Collaborations
 
@@
 
-## Future directions
+## Future Directions
 

### 2022-10-24 slides/r4csr-psi.Rmd
Deploying to gh-pages from @ elong0527/r4csr@3388db6577737ac10f48ae32ba016bcdf1ee42b6 🚀

@@
   - [`gsDesign`](https://github.com/keaven/gsDesign): an R package for group sequential design under proportional hazards.
-  - [`simtrial`](https://github.com/Merck/simtrial),
-  [`gsDesign2`](https://github.com/Merck/gsDesign2), and
-  [`gsdmvn`](https://github.com/Merck/gsdmvn): experimental R packages for group sequential design under non-proportional hazards
+  - [`gsDesign2`](https://github.com/Merck/gsDesign2): experimental R packages for group sequential design under non-proportional hazards
   - Bookdown: <https://keaven.github.io/gsd-deming/>

### 2022-10-25 slides/r4csr-psi.Rmd
Deploying to gh-pages from @ elong0527/r4csr@13301dc2ca02b158e52733a3457d7e6c7012ac18 🚀

@@
 - R is widely used in visualization
-  - [SafetyGraphics](https://github.com/SafetyGraphics/safetyGraphics)
+  - [metalite.table1](https://elong0527.github.io/metalite.table1/articles/metalite-table1.html): Interactive demographic table.
   - [forestly](https://elong0527.github.io/forestly/articles/example.html#construct-interactive-forest-plot-1): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_194/)

### 2022-12-07 tlf-ae-spec.Rmd
automate rtf to pdf conversion
TLF chapters PDF output patch - add `\mainmatter` and make `out.height` HTML-specific as LaTeX does not support px as a unit

@@
 we need to summarize number of participants for each specific AE in Section 12.2, Adverse Events (AEs).
+
+```{r, include=FALSE}
+source("common.R")
+```
 
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/tlf_spec_ae.pdf")
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, include=FALSE}
+rtf2pdf("tlf/tlf_spec_ae.rtf")
+```
+
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/tlf_spec_ae.pdf")

### 2022-12-07 tlf-overview.Rmd
minor updates
TLF chapters PDF output patch - add `\mainmatter` and make `out.height` HTML-specific as LaTeX does not support px as a unit

@@
+\mainmatter
+
 # (PART) Delivering TLFs in CSR {-}
@@
 Here we focus on `r2rtf` to illustrate the concept.  
-Reader is encouraged to explore other R packages to find the proper tools to fit your purpose. 
+Readers are encouraged to explore other R packages to find the proper tools to fit your purpose. 
 :::
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/intro-ae1.pdf")
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/intro-ae2.pdf")
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/intro-ae3.pdf")
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/intro-ae5.pdf")
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/intro-ae7.pdf")

### 2022-12-07 submission-package.Rmd

@@
 
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
+library("kableExtra")
+```
+
 ```{r, echo=FALSE}
-library("kableExtra")
-
 df1 <- data.frame(
@@
 names(df1) <- c("Program Name", "Output Table", "Title")
+```
 
+```{r, echo=FALSE}
+if (knitr::is_latex_output()) {
+  df1 %>% kbl(format = "latex")
+}
+
+if (knitr::is_html_output()) {
   df1 %>%
-  kbl(escape = FALSE) %>%
+    kbl(format = "html", escape = FALSE) %>%
     kable_classic(full_width = FALSE, html_font = "'Times New Roman', Times, serif", font_size = 18) %>%
@@
     row_spec(0, background = "#DFDFDF", bold = TRUE, extra_css = "border: 1px solid #000; text-align: center;")
+}
+```
 
+```{r, echo=FALSE}
 df2 <- data.frame(
@@
 )
+```
 
+```{r, echo=FALSE}
+if (knitr::is_latex_output()) {
+  df2 %>% kbl(format = "latex")
+}
+
+if (knitr::is_html_output()) {
   df2 %>%
-  kbl() %>%
+    kbl(format = "html") %>%
     kable_classic(full_width = FALSE, html_font = "'Times New Roman', Times, serif", font_size = 18) %>%
@@
     footnote(general = "R package versions based on CRAN on 2021-08-06", general_title = "")
+}
+```
 
+```{r, echo=FALSE}
 df3 <- data.frame(
@@
 )
+```
 
+```{r, echo=FALSE}
+if (knitr::is_latex_output()) {
+  df3 %>% kbl(format = "latex")
+}
+
+if (knitr::is_html_output()) {
   df3 %>%
-  kbl() %>%
+    kbl(format = "html") %>%
     kable_classic(full_width = FALSE, html_font = "'Times New Roman', Times, serif", font_size = 18) %>%
@@
     row_spec(0, background = "#DFDFDF", bold = TRUE, extra_css = "border: 1px solid #000; text-align: center;")
+}
 ```
@@
 )
+```
 
+```{r}
+if (knitr::is_latex_output()) {
+  df4 %>% kbl(format = "latex")
+}
+
+if (knitr::is_html_output()) {
   df4 %>%
-  kbl() %>%
+    kbl(format = "html") %>%
     kable_classic(full_width = FALSE, html_font = "'Times New Roman', Times, serif", font_size = 18) %>%
@@
 )
+```
 
+```{r, echo=FALSE}
+if (knitr::is_latex_output()) {
+  df5 %>% kbl(format = "latex")
+}
+
+if (knitr::is_html_output()) {
   df5 %>%
@@
     gsub("<thead>.*</thead>", "", .)
+}
 ```

### 2022-12-07 tlf-ae-summary.Rmd

@@
 we summarize number of participants that were included in each safety analysis in Section 12.2, Adverse Events (AEs).
+
+```{r, include=FALSE}
+source("common.R")
+```
 
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/tlf_ae_summary.pdf")
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, include=FALSE}
+rtf2pdf("tlf/tlf_ae_summary.rtf")
+```
+
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/tlf_ae_summary.pdf")

### 2022-12-07 tlf-assemble.Rmd
Update tlf-assemble.Rmd

@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/rtf-combine.pdf")
@@
 ::: {.rmdnote}
-There is a slight learning curve to learning how toggle fields work 
+There is a slight learning curve on how toggle fields work 
 in Microsoft Words. 

### 2022-12-07 tlf-baseline.Rmd
minor updates

@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/tlf_base.pdf")
@@
 space of text. They are helpful when you need to control the white space
-of a long phrase "AMERICAN INDIAN OR ALASKA NATIVE"
+of a long phrase, "AMERICAN INDIAN OR ALASKA NATIVE"
 in the table provides an example.

### 2022-12-07 tlf-disposition.Rmd

@@
 
+```{r, include=FALSE}
+source("common.R")
+```
+
 ```{r}
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/tbl_disp.pdf")
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, include=FALSE}
+rtf2pdf("tlf/tbl_disp.rtf")
+```
+
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/tbl_disp.pdf")

### 2022-12-07 tlf-efficacy-ancova.Rmd
rename file

--- 2021-11-15 tlf-efficacy.Rmd
@@
-# Efficacy
+# Efficacy table
 
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/tlf_eff.pdf")
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, include=FALSE}
+rtf2pdf("tlf/tlf_eff1.rtf")
+```
+
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/tlf_eff1.pdf")
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, include=FALSE}
+rtf2pdf("tlf/tlf_eff2.rtf")
+```
+
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/tlf_eff2.pdf")
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, include=FALSE}
+rtf2pdf("tlf/tlf_eff.rtf")
+```
+
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/tlf_eff.pdf")

### 2022-12-07 tlf-population.Rmd
source common.R in each Rmd to allow compile individually
fix some minor issues

@@
 analysis in Section 11.1, Data Sets Analysed.
+
+```{r, include=FALSE}
+source("common.R")
+```
 
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/tbl_pop.pdf")
@@
 simplicity and readability. More details about writing R functions can
-be found in the [Stat454
+be found in the [STAT 545
 course](https://stat545.com/functions-part1.html).
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, include=FALSE}
+rtf2pdf("tlf/tbl_pop.rtf")
+```
+
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/tbl_pop.pdf")

### 2022-12-08 project-folder.Rmd
Fix custom block annotation for pdf output

@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 Under the `vignettes/` folder, there are two folders: `adam/` and `tlf/`.
@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 This is the same level of reproducibility in most SAS environments:
@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 `.Rprofile` is only for project-specific R packages.
@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 RStudio Package Manager (RSPM) provides a solution to host both
@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 The `renv.lock` file and `renv/` folder are only for project-specific R package.
@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 A cloud-based R environment (e.g., RStudio Workbench)
@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 A container solution like Docker [@RJ-2020-007] could further enhance

### 2022-12-08 tlf-overview.Rmd
Fix custom block annotation for pdf output

@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 FDA's PDF specification is a general reference. Each organization can define
@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 There are other R packages to create TLFs in ASCII, RTF and Word format.

### 2022-12-08 submission-overview.Rmd

@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 "Sponsors should provide the software programs used to create all ADaM

### 2022-12-08 tlf-assemble.Rmd

@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 One limitation of combining RTF source code is that 
@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 There is a slight learning curve on how toggle fields work 
@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 A typical error message is that system can not find the file 
@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 If you modify table content in the combined `.docx` file, 

### 2022-12-08 tlf-efficacy-km.Rmd

@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 R packages such as
@@
 
-```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/tlf_km.pdf")

### 2022-12-18 tlf-assemble.Rmd
tlf-assemble: using r2rtf 0.4.0
- start to use `assemble_rtf` and `assemble_docx` to combine RTF files.

@@
 
-<details>
-<summary>`rtf_assemble1()` source code</summary>
-```{r}
-#' Assemble RTF files into one single RTF file
-#'
-#' @param path a vector of RTF file path
-#'
-#' @export
-rtf_assemble1 <- function(path) {
-  rtf <- lapply(path, readLines)
+> The code below requires r2rtf version >= 0.4.0.
 
-  n <- length(rtf)
-  start <- c(1, rep(2, n - 1))
-  end <- vapply(rtf, length, numeric(1))
-  end[-n] <- end[-n] - 1
-
-  for (i in 1:n) {
-    rtf[[i]] <- rtf[[i]][start[i]:end[i]]
-    if (i < n) rtf[[i]] <- c(rtf[[i]], r2rtf:::as_rtf_new_page())
-  }
-  rtf <- do.call(c, rtf)
-
-  rtf
-}
-```
-</details>
-
-The `rtf_assemble1()` function allows user to combine RTF 
+The `r2rtf::assemble_rtf()` function allows user to combine RTF 
 source code in individual files into one larger RTF file.
@@
 ```{r}
-rtf_assemble1(tlf_path) |>
-  r2rtf::write_rtf("tlf/rtf-combine.rtf")
+r2rtf::assemble_rtf(
+  input = tlf_path, 
+  output = "tlf/rtf-combine.rtf"
+)
 ```
@@
 
-<details>
-<summary>`rtf_assemble2()` source code</summary>
-```{r}
-#' Assemble RTF TLFs
-#'
-#' Add a set of RTF/TEXT fields into an rdocx object.
-#'
-#' @param input Character vector of file path.
-#' @param output Character string to the output file path.
-#' @param landscape Logical vector to determine whether to
-#' display files as portrait or landscape.
-#'
-#' @export
-rtf_assemble2 <- function(input, output, landscape = FALSE) {
-  input <- normalizePath(input)
-
-  if (!all(file.exists(input))) {
-    warning("Some files do not exist")
-  }
-
-  field <- ifelse(grepl("/", input),
-    paste0("INCLUDETEXT \"", gsub("/", "\\\\\\\\", input), "\""),
-    paste0("INCLUDETEXT \"", gsub("\\", "\\\\", input, fixed = "TRUE"), "\"")
-  )
-
-  if (length(landscape) == 1) {
-    landscape <- rep(landscape, length(field))
-  }
-
-  docx <- officer::read_docx()
-
-  for (i in seq_along(input)) {
-    docx <-
-      docx %>%
-      officer::body_add_fpar(
-        officer::fpar(
-          officer::ftext("Table "),
-          officer::run_word_field("SEQ Table \\* ARABIC"),
-          officer::run_linebreak(),
-          officer::run_word_field(field[i]),
-          officer::run_pagebreak()
-        )
-      )
-    if (landscape[i]) {
-      docx <- officer::body_end_section_landscape(docx)
-    } else {
-      docx <- officer::body_end_section_portrait(docx)
-    }
-  }
-
-  print(docx, target = output)
-
-  invisible(output)
-}
-```
-</details>
-
-The `rtf_assemble2()` function 
+The `assemble_docx()` function 
 allows you to create a word file with toggle fields as below.
@@
 ```{r}
-rtf_assemble2(
+r2rtf::assemble_docx(
   tlf_path,

### 2022-12-18 slides/r4csr-gwu.Rmd
update package name style (#91)
- avoid inline style for packge name

@@
 
-- [`r2rtf`](https://merck.github.io/r2rtf/): create production-ready tables and figures in RTF format.
+- [r2rtf](https://merck.github.io/r2rtf/): create production-ready tables and figures in RTF format.
 - [`pkglite`](https://merck.github.io/pkglite/): represent and exchange R package source code as text files.
@@
 
-`r2rtf` is designed to:
+r2rtf is designed to:
 
@@
 
-`r2rtf` is designed to be pipe-friendly (`%>%`)
+r2rtf is designed to be pipe-friendly (`%>%`)
 
@@
 
-- As a statistician, I use `tidyverse`, `r2rtf`,
+- As a statistician, I use tidyverse, r2rtf,
 and internal tools to define the mock-up table, listing and figure (TLFs) for statistical analysis of a clinical trial.
@@
 
-- As a programmer, I use `tidyverse`, `r2rtf`, and internal tools to develop and/or validate analysis results based on mock-up TLFs.
+- As a programmer, I use tidyverse, r2rtf, and internal tools to develop and/or validate analysis results based on mock-up TLFs.
 

### 2022-12-18 tlf-efficacy-ancova.Rmd
update package name style (#91)
- avoid inline style for packge name

@@
 
-The `emmeans` R package is used to obtain
+The emmeans R package is used to obtain
 within and between group least square (LS) mean
@@
 
-Then `r2rtf` is used to prepare the table format for `t1`.
+Then r2rtf is used to prepare the table format for `t1`.
 We also highlight how to handle special characters in this example.
@@
 For example, `{^a}` provides a superscript `a` for footnote notation.
-`r2rtf` also supports most LaTeX characters.
+r2rtf also supports most LaTeX characters.
 Examples can be found on the
-[`r2rtf` get started page](https://merck.github.io/r2rtf/articles/r2rtf.html#special-character).
+[r2rtf get started page](https://merck.github.io/r2rtf/articles/r2rtf.html#special-character).
 The `text_convert` argument in `r2rtf_*()` functions controls whether to convert special characters.
@@
 
-We also use `r2rtf` to prepare the table format for `t2`
+We also use r2rtf to prepare the table format for `t2`
 
@@
 
-Finally, we combine the two parts to get the final table using `r2rtf`.
+Finally, we combine the two parts to get the final table using r2rtf.
 This is achieved by providing a list of `t1_rtf` and `t2_rtf` as input for

### 2022-12-18 tlf-efficacy-km.Rmd
update package name style (#91)
- avoid inline style for packge name

@@
 
-The `survival` package is used to obtain the K-M estimate.
+The survival package is used to obtain the K-M estimate.
 
@@
 
-Now, we can use the `r2rtf` package to create a formatted RTF figure.
+Now, we can use the r2rtf package to create a formatted RTF figure.
 More details can be found on the [r2rtf website](https://merck.github.io/r2rtf/articles/example-figure.html).
@@
 - Step 3: Save figures into `png` files based on required analysis specification.
-- Step 4: Create RTF output using the `r2rtf` package.
+- Step 4: Create RTF output using the r2rtf package.

### 2022-12-19 project-folder.Rmd

@@
 For a clinical trial project, it is also important to provide proper documentation for those help functions.
-We use `roxygen2` to document functions.
+We use roxygen2 package to document functions.
 For example, the header below defines each variable in `fmt_est`.
@@
 
-The `roxygen2` documentation will be converted into
+The roxygen2 documentation will be converted into
 standard R documentation format, and saved as `.Rd` files in the `man/` folder.
@@
 
-The `man/` folder is used to save documentation automatically generated by `roxygen2`.
-A typical workflow is to add `roxygen2` documentation before each function in the `R/` folder.
+The `man/` folder is used to save documentation automatically generated by roxygen2.
+A typical workflow is to add roxygen2 documentation before each function in the `R/` folder.
 Then `devtools::document()` is used to generate all the documentation files in the `man/` folder.
@@
 including the control of the R version and the R package versions.
-By using the R package folder structure and proper tools (e.g., `renv`, `packrat`),
+By using the R package folder structure and proper tools (e.g., renv, packrat),
 we illustrate how to achieve reproducibility for R and R package versions.
@@
 First, we introduce the control of the R version.
-In the `esubdemo` project, a reproducible environment is created when you open
+In the esubdemo project, a reproducible environment is created when you open
 the `esubdemo.Rproj` from RStudio.
@@
 
-Secondly, we use `renv` to lock R package versions and save them in the `renv.lock` file.
-`renv` provides a robust and stable approach to managing R package versions
+Secondly, we use renv to lock R package versions and save them in the `renv.lock` file.
+renv provides a robust and stable approach to managing R package versions
 for project-specific R packages.
-An introduction of `renv` can be found on its
+An introduction of renv can be found on its
 [website](https://rstudio.github.io/renv/articles/renv.html).
@@
 
-The R code above in the `.Rprofile` initiates the `renv` running environment.
+The R code above in the `.Rprofile` initiates the renv running environment.
 As a user, you can use `renv::init()`, `renv::snapshot()`, and `renv::restore()`
@@
 
-In the analysis project, the `renv` package will
+In the analysis project, the renv package will
 
@@
 The `renv.lock` file and `renv/` folder are only for project-specific R package.
-A standard R package should not use `renv`.
+A standard R package should not use renv.
 :::
@@
 - Define a snapshot date in `inst/startup.R`.
-- Using `renv` to lock R versions within a project.
+- Using renv to lock R versions within a project.
 
@@
 
-- [`devtools`](https://devtools.r-lib.org/): make package development easier.
+- [devtools](https://devtools.r-lib.org/): make package development easier.
   + A good overview can be found in [Chapter 2 of the R Packages book](https://r-pkgs.org/Whole-game.html).
   + `devtools::load_all()`: load all functions in `R/` folder and running environment.
-  + `devtools::document()`: automatically create documentation using `roxygen2`.
+  + `devtools::document()`: automatically create documentation using roxygen2.
   + `devtools::check()`: automatically perform compliance check as an R package.
-  + `devtools::build_site()`: automatically run analysis scripts in batch and create a `pkgdown` website.
-- [`usethis`](https://usethis.r-lib.org/): automates repetitive tasks that arise during project setup and development.
-- [`testthat`](https://testthat.r-lib.org/): streamline testing code.
-  + A discussion of using the `testthat` for an A&R project can be found in (@madhutest).
-- [`pkgdown`](https://pkgdown.r-lib.org/): generate static HTML documentation website for an R package
+  + `devtools::build_site()`: automatically run analysis scripts in batch and create a pkgdown website.
+- [usethis](https://usethis.r-lib.org/): automates repetitive tasks that arise during project setup and development.
+- [testthat](https://testthat.r-lib.org/): streamline testing code.
+  + A discussion of using the testthat for an A&R project can be found in (@madhutest).
+- [pkgdown](https://pkgdown.r-lib.org/): generate static HTML documentation website for an R package
   + It also allows you to run all analysis code in batch.

### 2022-12-19 project-management.Rmd

--- 2022-01-29 project-management.Rmd
@@
 
-::: {.rmdnote}
+::: {.rmdnote data-latex=""}
 After project initiation, modifying `.Rprofile` will be a risk of reproducibility
@@
 The testthat R package can be used to streamline the validation process.
-More details of the `testthat` package for R package validation can be found in [Chapter 12 of the R package book](https://r-pkgs.org/tests.html).
+More details of the testthat package for R package validation can be found in [Chapter 12 of the R package book](https://r-pkgs.org/testing-basics.html).
 
@@
 
-The project lead can also use the R package `pkgdown` to build a complete website for a project-specific R package.
-The `pkgdown` website is a convenient way to run all analyses in batch and
+The project lead can also use the R package pkgdown to build a complete website for a project-specific R package.
+The pkgdown website is a convenient way to run all analyses in batch and
 integrate outputs in a website, which comprehensively covers
 project-specific R functions, TLF generation programs, outputs and validation tracking information, etc.
-For example, in the `esubdemo` project, we created the `pkgdown` website at <https://elong0527.github.io/esubdemo/>.
+For example, in the `esubdemo` project, we created the pkgdown website at <https://elong0527.github.io/esubdemo/>.
 
@@
 An organization can leverage CI/CD workflow to automatically enable those tasks,
-such as running testing cases and creating a `pkgdown` website.
+such as running testing cases and creating a pkgdown website.
 For example, in the `esubdemo` project, we set up

### 2022-12-19 slides/china-r.Rmd

@@
 
-- As a statistician, I use `tidyverse`, `r2rtf`, and internal tools to define the mock-up table, listing and figure (TLFs) for statistical analysis of a clinical trial.
+- As a statistician, I use tidyverse, r2rtf, and internal tools to define the mock-up table, listing and figure (TLFs) for statistical analysis of a clinical trial.
   - More than 100 TLFs for efficacy and safety of a drug or vaccine.
 
-- As a programmer, I use `tidyverse`, `r2rtf`, and internal tools to develop and/or validate analysis results based on mock-up TLFs.
+- As a programmer, I use tidyverse, r2rtf, and internal tools to develop and/or validate analysis results based on mock-up TLFs.
 

### 2022-12-19 slides/r4csr-rstudio.Rmd

@@
 
-- As statisticians, we use `tidyverse`, `r2rtf`,
+- As statisticians, we use tidyverse, r2rtf,
 and internal tools to define the mock-up table, listing and figure (TLFs) for statistical analysis of a clinical trial.
@@
 
-- As programmers, we use `tidyverse`, `r2rtf`, and internal tools to develop and/or validate analysis results based on mock-up TLFs.
+- As programmers, we use tidyverse, r2rtf, and internal tools to develop and/or validate analysis results based on mock-up TLFs.
 

### 2022-12-19 tlf-ae-spec.Rmd

@@
 To obtain the nested layout, we use the `page_by` argument in the
-`rtf_body` function. By defining `page_by="AESOC"`, `r2rtf` recognizes
+`rtf_body` function. By defining `page_by="AESOC"`, r2rtf recognizes
 the variable as a group indicator.
@@
 features can be found on the
-[`r2rtf` package website](https://merck.github.io/r2rtf/articles/example-sublineby-pageby-groupby.html).
+[r2rtf package website](https://merck.github.io/r2rtf/articles/example-sublineby-pageby-groupby.html).
 
@@
 - Step 6: Row-wise combine `t_pop` and `t_ae` into `tbl_ae_spec`.
-- Step 7: Format the output by using `r2rtf`.
+- Step 7: Format the output by using r2rtf.

### 2022-12-19 tlf-overview.Rmd

@@
 There are other R packages to create TLFs in ASCII, RTF and Word format.
-For example, `rtables`, `huxtable`, `pharmaRTF`, `gt`, `officer`, `flextable` etc.
-Here we focus on `r2rtf` to illustrate the concept.  
+For example, rtables, huxtable, pharmaRTF, gt, officer, flextable etc.
+Here we focus on r2rtf to illustrate the concept.  
 Readers are encouraged to explore other R packages to find the proper tools to fit your purpose. 
@@
 
-`tidyverse` is a collection of R packages to simplify the workflow to manipulate,
+tidyverse is a collection of R packages to simplify the workflow to manipulate,
 visualize and analyze data in R.
@@
 RStudio provided outstanding [cheatsheets](https://www.rstudio.com/resources/cheatsheets/)
-and [tutorials](https://github.com/rstudio-education/remaster-the-tidyverse) for `tidyverse`.
+and [tutorials](https://github.com/rstudio-education/remaster-the-tidyverse) for tidyverse.
 
-There are also books to introduce `tidyverse`.
-We assume the reader have experience in using `tidyverse` in this book.
+There are also books to introduce tidyverse.
+We assume the reader have experience in using tidyverse in this book.
 
@@
 
-`r2rtf` is an R package to create production-ready tables and figures in RTF format.
+r2rtf is an R package to create production-ready tables and figures in RTF format.
 This R package is designed to
@@
 - focus on the **table format** only.
-  Data manipulation and analysis shall be handled by other R packages (e.g., `tidyverse`).
+  Data manipulation and analysis shall be handled by other R packages (e.g., tidyverse).
 
@@
 
-We provide a brief introduction of `r2rtf` and show how to transfer
+We provide a brief introduction of r2rtf and show how to transfer
 data frames into table, listing, and figures (TLFs).
@@
 Other extended examples and features are covered on the
-[`r2rtf` package website](https://merck.github.io/r2rtf/articles/index.html).
+[r2rtf package website](https://merck.github.io/r2rtf/articles/index.html).
 
-To explore the basic RTF generation verbs in `r2rtf`,
-we will use the dataset `r2rtf_adae` saved in the `r2rtf` package.
+To explore the basic RTF generation verbs in r2rtf,
+we will use the dataset `r2rtf_adae` saved in the r2rtf package.
 This dataset contains adverse events (AEs) information from a clinical trial.
@@
 
-`dplyr` and `tidyr` packages within `tidyverse` are used
+dplyr and tidyr packages within tidyverse are used
 for data manipulation to create a data frame
@@
 
-`r2rtf` aims to provide one function for each type of table layout.
+r2rtf aims to provide one function for each type of table layout.
 Commonly used verbs include:
@@
 
-`r2rtf` supports 26 different border types. The details can be found on
+r2rtf supports 26 different border types. The details can be found on
 the [r2rtf package website](https://merck.github.io/r2rtf/articles/rtf-row.html#border-type).
@@
 
-In the `r2rtf` R package [get started](https://merck.github.io/r2rtf/articles/r2rtf.html) page,
+In the r2rtf R package [get started](https://merck.github.io/r2rtf/articles/r2rtf.html) page,
 there are more examples to illustrate how to customize

### 2023-01-13 slides/fda-workshop-slides.Rmd

@@
 
-- Data manipulation: `tidyverse`, `dplyr`, `tidyr`, `ggplot2` etc.
+- Data manipulation: tidyverse, dplyr, tidyr, ggplot2 etc.
 - ADaM data: `adsl`, `adae` etc.
@@
 
-- `tidyverse`: a collection of R packages to simplify the workflow to manipulate,
+- tidyverse: a collection of R packages to simplify the workflow to manipulate,
 visualize and analyze data in R.
 
-- `r2rtf`: an R package to create production-ready tables and figures in RTF format.
+- r2rtf: an R package to create production-ready tables and figures in RTF format.
 
@@
 
-- `r2rtf` is an R package to create production-ready tables and figures in RTF format.
+- r2rtf is an R package to create production-ready tables and figures in RTF format.
 
-`r2rtf` is designed to:
+r2rtf is designed to:
 
@@
 - only focus on the **table format**.
-  - Data manipulation and analysis should be handled by other R packages. (e.g., `tidyverse`)
+  - Data manipulation and analysis should be handled by other R packages. (e.g., tidyverse)
 
@@
 
-`r2rtf` is designed to enable pipes (`%>%`).
+r2rtf is designed to enable pipes (`%>%`).
 
@@
 
-`r2rtf` package provides the flexibility to customize table appearance for
+r2rtf package provides the flexibility to customize table appearance for
 
@@
 
-`r2rtf` package also provides the flexibility to convert figures in RTF format.
+r2rtf package also provides the flexibility to convert figures in RTF format.
 
@@
 
-`r2rtf` provides simple "verb" functions that correspond to each component of a table,
+r2rtf provides simple "verb" functions that correspond to each component of a table,
   to help you translate data frame to tables in RTF file.

### 2023-01-13 slides/workshop-slides.Rmd

@@
 
-`r2rtf` only focus on **table format**.
-Data manipulation and analysis should be handled by other R packages. (e.g., `tidyverse`)
+r2rtf only focus on **table format**.
+Data manipulation and analysis should be handled by other R packages. (e.g., tidyverse)
 

### 2023-01-13 tlf-population.Rmd

@@
 provide a simple (but not robust) wrapper function, `count_by()`, using
-the `dplyr` and `tidyr` package.
+the dplyr and tidyr package.
 
@@
 Now we combine individual rows into one table for reporting
-purpose. `tbl_pop` is used as input for `r2rtf` to create the final
+purpose. `tbl_pop` is used as input for r2rtf to create the final
 report.
@@
   `count_by()` function.
-- Step 3: Format the output from Step 2 using `r2rtf`.
+- Step 3: Format the output from Step 2 using r2rtf.

### 2023-01-13 tlf-assemble.Rmd
Polish text for TLF assemble chapter

@@
 1. Combine RTF source code in individual files into one large RTF file.
-1. Leverage Microsoft Word `Toggle Fields` feature to embed RTF files using 
+1. Leverage the `Toggle Fields` feature in Microsoft Word to embed RTF files using
 hyperlinks.
@@
 
-> The code below requires r2rtf version >= 0.4.0.
+::: {.rmdnote data-latex=""}
+The code below requires r2rtf version >= 1.0.0.
+:::
 
-The `r2rtf::assemble_rtf()` function allows user to combine RTF 
+The `r2rtf::assemble_rtf()` function allows the user to combine RTF
 source code in individual files into one larger RTF file.
@@
 One limitation of combining RTF source code is that
-we are not able to specify page orientation of each TLF 
+we are not able to specify the page orientation of each TLF
 in the combined document.
@@
 
-Microsoft Words uses toggle fields to embed files into one Word document. 
+Microsoft Word uses toggle fields to embed files into one Word document.
 The approach is a dynamic link between files by providing the absolute file path.
@@
 There is a slight learning curve on how toggle fields work
-in Microsoft Words. 
+in Microsoft Word.
 After you become familiar with the workflow,
-toggle fields can extend your capability to manage a large amount of TLFs
+toggle fields can extend your capability to manage a large number of TLFs
 in RTF format.
@@
 The `assemble_docx()` function
-allows you to create a word file with toggle fields as below.
+allows you to create a `.docx` file with toggle fields as below.
 One benefit is to control the page direction of each TLF as below.
@@
 
-By using `Alt + F9` to display the fields and you will 
+By using Alt + F9 to display the fields and you will
 see information similar to the screenshot below.
@@
 If it works, you can see a table similar to the snapshot below
-by using `Alt + F9`. 
+by using Alt + F9.
 It is a shortcut to change between results and field display mode.
@@
 Now you can update all toggle fields to display all TLFs by
-selecting all fields (`Ctrl + A`), then type `F9`. 
+selecting all fields (Ctrl + A), then press F9.
 We suggest testing one toggle field before updating all of them.
 
-As the `docx` file contain dynamic links, 
+As the `.docx` file contain dynamic links,
 you can keep updating the TLFs if you need to refresh content in
-individual RTF files by selecting all fields (`Ctrl + A`), then type `F9`.
+individual RTF files by selecting all fields (Ctrl + A), then press F9.
 
@@
 you may get a weird table layout if you update all fields within a toggle field.
-To resolve the issue, please remove all `\* MERGEFORMAT` in the filed mode using `Alt + F9` before updating all toggle fields.
+To resolve the issue, please remove all `\* MERGEFORMAT` in the filed mode using Alt + F9 before updating all toggle fields.
 :::
@@
 you can also unlink toggle fields to save table contents,
-because the absolute path may not work for others.  
-To unlink toggle fields, you can select all fields (`Ctrl + A`), 
-then type `Ctrl + Shift + F9`.
+because the absolute path may only work for some.
+To unlink toggle fields, you can select all fields (Ctrl + A),
+then press Ctrl + Shift + F9.

### 2023-01-14 check-url.R
Add URL checking code

@@ -1,0 +1,47 @@
+#' Check URLs in a bookdown project
+#'
+#' @param input Path to the bookdown project directory.
+#'
+#' @return URL checking results from `urlchecker::url_check()`
+#'
+#' @details This assumes there are no `.Rmd` and `.md` files with
+#' duplicated names in the bookdown project.
+#'
+#' @note The `tools::pkgVignettes()$docs` call in urlchecker
+#' requires two things (VignetteBuilder and VignetteEngine)
+#' to recognize `.Rmd` files as package vignettes.
+check_url <- function(input = ".") {
+  pkg_path <- tempfile()
+  dir.create(pkg_path)
+
+  # Create a DESCRIPTION file
+  write("VignetteBuilder: knitr", file = file.path(pkg_path, "DESCRIPTION"))
+
+  # Copy all .Rmd and .md files
+  rmd_path <- file.path(pkg_path, "vignettes")
+  dir.create(rmd_path)
+  file.copy(
+    list.files(path = input, pattern = "\\.Rmd$|\\.md$", recursive = TRUE),
+    to = rmd_path,
+    overwrite = TRUE
+  )
+  # Rename .md to .Rmd
+  md <- list.files(rmd_path, pattern = "\\.md$", full.names = TRUE)
+  if (length(md)) file.rename(md, to = paste0(tools::file_path_sans_ext(md), ".Rmd"))
+
+  # Make them look like vignettes
+  lapply(
+    list.files(rmd_path, full.names = TRUE),
+    function(x) {
+      write(
+        "---\nvignette: >\n  %\\VignetteEngine{knitr::rmarkdown}\n---",
+        file = x,
+        append = TRUE
+      )
+    }
+  )
+
+  urlchecker::url_check(pkg_path)
+}
+
+check_url()

### 2023-01-15 check-url.R
Remove file name assumptions for URL checking with flatten copy logic

@@ -1,47 +1,63 @@
+#' Flatten copy
+#'
+#' @param from Source directory path
+#' @param to Destination directory path
+#'
+#' @details Copy all `.Rmd` and `.md` files from source to destination,
+#' rename the `.md` files with an additional `.Rmd` extension, and get a
+#' flat destination directory structure with path-preserving file names.
+flatten_copy <- function(from, to) {
+  rmd <- list.files(from, pattern = "\\.Rmd$", recursive = TRUE, full.names = TRUE)
+  md <- list.files(from, pattern = "\\.md", recursive = TRUE, full.names = TRUE)
+
+  src <- c(rmd, md)
+  dst <- c(rmd, paste0(md, ".Rmd"))
+
+  # Remove starting `./` (if any)
+  dst <- gsub("^\\./", replacement = "", x = dst)
+  # Replace the forward slash in path with Unicode big solidus
+  dst <- gsub("/", replacement = "\u29F8", x = dst)
+
+  file.copy(src, to = file.path(to, dst))
+
+  invisible(to)
+}
+
 #' Check URLs in a bookdown project
 #'
 #' @param input Path to the bookdown project directory.
 #'
 #' @return URL checking results from `urlchecker::url_check()`
-#'
-#' @details This assumes there are no `.Rmd` and `.md` files with
-#' duplicated names in the bookdown project.
+#' for all `.Rmd` and `.md` files in the project.
 #'
 #' @note The `tools::pkgVignettes()$docs` call in urlchecker
-#' requires two things (VignetteBuilder and VignetteEngine)
+#' requires two things (`VignetteBuilder` and `VignetteEngine`)
 #' to recognize `.Rmd` files as package vignettes.
 check_url <- function(input = ".") {
-  pkg_path <- tempfile()
-  dir.create(pkg_path)
+  # Create a source package directory
+  pkg <- tempfile()
+  dir.create(pkg)
 
-  # Create a DESCRIPTION file
-  write("VignetteBuilder: knitr", file = file.path(pkg_path, "DESCRIPTION"))
+  # Create a minimal DESCRIPTION file
+  write("VignetteBuilder: knitr", file = file.path(pkg, "DESCRIPTION"))
 
-  # Copy all .Rmd and .md files
-  rmd_path <- file.path(pkg_path, "vignettes")
-  dir.create(rmd_path)
-  file.copy(
[…]

### 2023-01-15 check-url.R
Add URL checking support for Quarto projects

@@ -3,15 +3,15 @@
 #' @param from Source directory path
 #' @param to Destination directory path
 #'
-#' @details Copy all `.Rmd` and `.md` files from source to destination,
-#' rename the `.md` files with an additional `.Rmd` extension, and get a
-#' flat destination directory structure with path-preserving file names.
+#' @details Copy all `.Rmd`, `.qmd`, and `.md` files from source to destination,
+#' rename the `.qmd` and `.md` files with an additional `.Rmd` extension,
+#' and get a flat destination structure with path-preserving file names.
 flatten_copy <- function(from, to) {
   rmd <- list.files(from, pattern = "\\.Rmd$", recursive = TRUE, full.names = TRUE)
-  md <- list.files(from, pattern = "\\.md", recursive = TRUE, full.names = TRUE)
+  xmd <- list.files(from, pattern = "\\.qmd$|\\.md$", recursive = TRUE, full.names = TRUE)
 
-  src <- c(rmd, md)
-  dst <- c(rmd, paste0(md, ".Rmd"))
+  src <- c(rmd, xmd)
+  dst <- c(rmd, paste0(xmd, ".Rmd"))
 
   # Remove starting `./` (if any)
   dst <- gsub("^\\./", replacement = "", x = dst)
@@ -28,7 +28,7 @@ flatten_copy <- function(from, to) {
 #' @param input Path to the bookdown project directory.
 #'
 #' @return URL checking results from `urlchecker::url_check()`
-#' for all `.Rmd` and `.md` files in the project.
+#' for all `.Rmd`, `.qmd`, and `.md` files in the project.
 #'
 #' @note The `tools::pkgVignettes()$docs` call in urlchecker
 #' requires two things (`VignetteBuilder` and `VignetteEngine`)

### 2023-01-15 check-url.R
Make comments more generic to file types

@@ -41,12 +41,12 @@ check_url <- function(input = ".") {
   # Create a minimal DESCRIPTION file
   write("VignetteBuilder: knitr", file = file.path(pkg, "DESCRIPTION"))
 
-  # Flatten copy all .Rmd and .md files
+  # Flatten copy relevant files
   vig <- file.path(pkg, "vignettes")
   dir.create(vig)
   flatten_copy(input, vig)
 
-  # Make the .Rmd files look like vignettes
+  # Make the files look like vignettes
   lapply(
     list.files(vig, full.names = TRUE),
     function(x) {

### 2023-01-15 project-folder.Rmd
Fix broken URLs reported by URL checker

@@
 For R users, you already benefit from a well-defined and consistent folder structure.
-That is the [R package folder structure](https://github.com/rstudio/cheatsheets/raw/master/package-development.pdf).
+That is the [R package folder structure](https://github.com/rstudio/cheatsheets/blob/main/package-development.pdf).
 Every R package developer is required to follow the same convention to organize
@@
 This is the same level of reproducibility in most SAS environments:
-<https://support.sas.com/techsup/pcn/altopsys.html>
+<https://support.sas.com/en/technical-support/services-policies.html#altos>
 :::
@@
 # set up repository based on the snapshot date
-repos <- paste0("https://mran.microsoft.com/snapshot/", snapshot)
+repos <- paste0("https://mran.microsoft.com/snapshot/", snapshot, "/")
 
@@
 Current project R package repository:
-    https://mran.microsoft.com/snapshot/2021-08-06
+    https://mran.microsoft.com/snapshot/2021-08-06/
 ```

### 2023-01-15 slides/china-r.Rmd
Fix broken URLs reported by URL checker

@@
 
-- [FDA: Statistical Software Clarifying Statement](https://www.fda.gov/files/about%20fda/published/Statistical-Software-Clarifying-Statement-PDF.pdf)
+- [FDA: Statistical Software Clarifying Statement](https://www.fda.gov/media/161196/download)
 
@@
 We share the same philosophy described in
-[Section 1.1 of R Packages book](https://r-pkgs.org/intro.html#intro-phil) and quote here.
+[Section 1.1 of the R Packages book](https://r-pkgs.org/introduction.html#intro-phil) and quote here.
 
@@
 
-Training: [December 6th, Deming Conference 2021](https://demingconference.org/programs/2021-program/)
+Training: [December 6th, Deming Conference 2021](https://demingconference.org/wp-content/uploads/2021/07/2021-Deming-Conference-Printed-Program-v1.0.pdf)
 
@@
   - Nan Xiao: <https://nanx.me/>
-  - Keaven Anderson: <https://www.linkedin.com/in/keaven-anderson-09aa7538/>
+  - Keaven Anderson: <https://keaven.github.io/>
 

### 2023-01-15 slides/r4csr-gwu.Rmd
Fix broken URLs reported by URL checker

@@
   - [SafetyGraphics](https://github.com/SafetyGraphics/safetyGraphics)
-  - [forestly](https://elong0527.github.io/forestly/articles/example.html#construct-interactive-forest-plot-1): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_194/)
+  - [forestly](https://elong0527.github.io/forestly/articles/forestly.html): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_206/)
 

### 2023-01-15 slides/r4csr-psi.Rmd
Fix broken URLs reported by URL checker

@@
   - [metalite.table1](https://elong0527.github.io/metalite.table1/articles/metalite-table1.html): Interactive demographic table.
-  - [forestly](https://elong0527.github.io/forestly/articles/example.html#construct-interactive-forest-plot-1): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_194/)
+  - [forestly](https://elong0527.github.io/forestly/articles/example.html#construct-interactive-forest-plot-1): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_206/)
 

### 2023-01-15 slides/workshop-slides.Rmd
Fix broken URLs reported by URL checker

@@
 
-- [R Consortium](https://www.r-consortium.org/projects/isc-working-groups)
+- [R Consortium](https://www.r-consortium.org/all-projects/isc-working-groups)
   - R validation hub
@@
 
-- Public available CDISC pilot [study data located at CDISC Bitbucket repository](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).
+- Public available CDISC pilot [study data located at CDISC GitHub repository](https://github.com/cdisc-org/sdtm-adam-pilot-project/tree/master/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).
 
@@
 
-<https://r4csr.org/efficacy.html>
+<https://r4csr.org/efficacy-table.html>
 

### 2023-01-15 check-url.R
Move package constructing logic together

@@ -38,15 +38,15 @@ check_url <- function(input = ".") {
   pkg <- tempfile()
   dir.create(pkg)
 
-  # Create a minimal DESCRIPTION file
-  write("VignetteBuilder: knitr", file = file.path(pkg, "DESCRIPTION"))
-
   # Flatten copy relevant files
   vig <- file.path(pkg, "vignettes")
   dir.create(vig)
   flatten_copy(input, vig)
 
-  # Make the files look like vignettes
+  # Create a minimal DESCRIPTION file
+  write("VignetteBuilder: knitr", file = file.path(pkg, "DESCRIPTION"))
+
+  # Make the copied files look like vignettes
   lapply(
     list.files(vig, full.names = TRUE),
     function(x) {

### 2023-01-15 check-url.R
Update roxygen2 documentation

@@ -1,9 +1,12 @@
 #' Flatten copy
 #'
-#' @param from Source directory path
-#' @param to Destination directory path
+#' @param from Source directory path.
+#' @param to Destination directory path.
 #'
-#' @details Copy all `.Rmd`, `.qmd`, and `.md` files from source to destination,
+#' @return Destination directory path.
+#'
+#' @details
+#' Copy all `.Rmd`, `.qmd`, and `.md` files from source to destination,
 #' rename the `.qmd` and `.md` files with an additional `.Rmd` extension,
 #' and get a flat destination structure with path-preserving file names.
 flatten_copy <- function(from, to) {
@@ -23,15 +26,16 @@ flatten_copy <- function(from, to) {
   invisible(to)
 }
 
-#' Check URLs in a bookdown project
+#' Check URLs in an R Markdown or Quarto project
 #'
-#' @param input Path to the bookdown project directory.
+#' @param input Path to the project directory.
 #'
 #' @return URL checking results from `urlchecker::url_check()`
 #' for all `.Rmd`, `.qmd`, and `.md` files in the project.
 #'
-#' @note The `tools::pkgVignettes()$docs` call in urlchecker
-#' requires two things (`VignetteBuilder` and `VignetteEngine`)
+#' @details
+#' The `tools::pkgVignettes()$docs` call in urlchecker requires
+#' two core criteria (`VignetteBuilder` and `VignetteEngine`)
 #' to recognize `.Rmd` files as package vignettes.
 check_url <- function(input = ".") {
   # Create a source package directory

### 2023-01-16 preface.Rmd

@@
 We recommend using the
-[R package folder structure](https://github.com/rstudio/cheatsheets/raw/master/package-development.pdf)
+[R package folder structure](https://github.com/rstudio/cheatsheets/blob/main/package-development.pdf)
 to organize all the A&R-related source code and documentation for a clinical trial A&R project.
@@
 ```
+
+```{block2, type="flushright", html.tag="p"}
+Yilong Zhang, Nan Xiao, Keaven Anderson, Yalin Zhu
+
+December 2022
+```

### 2023-01-16 slides/r4csr-rstudio.Rmd

@@
   - [SafetyGraphics](https://github.com/SafetyGraphics/safetyGraphics)
-  - [forestly](https://elong0527.github.io/forestly/articles/example.html#construct-interactive-forest-plot-1): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_194/)
+  - [forestly](https://elong0527.github.io/forestly/articles/example.html#construct-interactive-forest-plot-1): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_206/)
 

### 2023-01-16 tlf-overview.Rmd

@@
 The Clinical Data Interchange Standards Consortium (CDISC) provides a
-[pilot project following ICH E3 guidance](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse).
+[pilot project following ICH E3 guidance](https://github.com/cdisc-org/sdtm-adam-pilot-project).
 
@@
 In the CDISC pilot project, an
-[example CSR](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/53-clin-stud-rep/535-rep-effic-safety-stud/5351-stud-rep-contr/cdiscpilot01/cdiscpilot01.pdf)
+[example CSR](https://github.com/cdisc-org/sdtm-adam-pilot-project/blob/master/updated-pilot-submission-package/900172/m5/53-clin-stud-rep/535-rep-effic-safety-stud/5351-stud-rep-contr/cdiscpilot01/cdiscpilot01.pdf)
 is also provided. If you are interested in more examples of clinical study reports,
@@
 We used publicly available CDISC pilot
-[study data located in the CDISC Bitbucket repository](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).
+[study data located in the CDISC GitHub repository](https://github.com/cdisc-org/sdtm-adam-pilot-project/tree/master/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).
 
@@
 Those R packages share
-[the tidy tools manifesto](https://cran.r-project.org/web/packages/tidyverse/vignettes/manifesto.html)
+[the tidy tools manifesto](https://tidyverse.tidyverse.org/articles/manifesto.html)
 and are easy to use for interactive data analysis.
 
-RStudio provided outstanding [cheatsheets](https://www.rstudio.com/resources/cheatsheets/)
+RStudio provided outstanding [cheatsheets](https://posit.co/resources/cheatsheets/)
 and [tutorials](https://github.com/rstudio-education/remaster-the-tidyverse) for tidyverse.

### 2023-05-22 project-folder.qmd
Migrate from bookdown to Quarto book (working HTML book)

--- 2023-01-15 project-folder.Rmd
@@
-# Project folder {#folder}
+# Project folder {#sec-folder}
 
@@
 
-## Consistency {#consistency}
+## Consistency {#sec-consistency}
 
@@
 
-::: {.rmdnote data-latex=""}
+::: {.callout-note}
 Under the `vignettes/` folder, there are two folders: `adam/` and `tlf/`.
@@
 
-## Reproducibility {#reproduce}
+## Reproducibility {#sec-reproduce}
 
@@
 
-::: {.rmdnote data-latex=""}
+::: {.callout-tip}
 This is the same level of reproducibility in most SAS environments:
@@
 
-::: {.rmdnote data-latex=""}
+::: {.callout-caution}
 `.Rprofile` is only for project-specific R packages.
@@
 
-::: {.rmdnote data-latex=""}
-RStudio Package Manager (RSPM) provides a solution to host both
-publicly available and internally developed R packages.
-However, the [public RSPM server](https://packagemanager.rstudio.com/client/#/repos/2/overview)
+::: {.callout-note}
+Posit Package Manager provides a solution to host both publicly available
+and internally developed R packages. However, the
+[Posit Public Package Manager](https://packagemanager.posit.co/)
 does not provide a daily snapshot as MRAN does.
@@
 
-::: {.rmdnote data-latex=""}
+::: {.callout-caution}
 The `renv.lock` file and `renv/` folder are only for project-specific R package.
@@
 
-::: {.rmdnote data-latex=""}
+::: {.callout-tip}
 A cloud-based R environment (e.g., RStudio Workbench)
@@
 
-::: {.rmdnote data-latex=""}
+::: {.callout-note}
 A container solution like Docker [@RJ-2020-007] could further enhance

### 2023-05-22 project-management.qmd
Migrate from bookdown to Quarto book (working HTML book)

--- 2022-12-19 project-management.Rmd
@@
-# Project management {#manage}
+# Project management {#sec-manage}
 
@@
 
-The project lead should also set up a project folder, as discussed in Chapter \@ref(folder).
+The project lead should also set up a project folder, as discussed in @sec-folder.
 The project initiation can be simplified by creating an
@@
 To enable reproducibility, the project leader should also review the startup file
-(i.e. `.Rprofile` discussed in Section \@ref(reproduce)) and define:
+(i.e. `.Rprofile` discussed in @sec-reproduce) and define:
 
@@
 
-::: {.rmdnote data-latex=""}
-After project initiation, modifying `.Rprofile` will be a risk of reproducibility
-and should be handled carefully if necessary.
+::: {.callout-caution}
+After project initiation, modifying `.Rprofile` will be a risk for
+reproducibility and should be handled carefully if necessary.
 :::
@@
 If there are any project-specific R functions that need to be developed,
-the R functions can be placed in the `R/` folder as discussed in Section \@ref(consistency).
+the R functions can be placed in the `R/` folder as discussed in
+@sec-consistency.
 

### 2023-05-22 project-overview.qmd
Migrate from bookdown to Quarto book (working HTML book)

--- 2021-11-08 project-overview.Rmd
@@
-# (PART) Clinical trial project {-}
-
 # Overview {#project-overview}
@@
 
-Chapter \@ref(folder) will discuss how to organize source code, documents, and
+@sec-folder will discuss how to organize source code, documents, and
 deliverables in an A&R clinical project.
@@
 
-Chapter \@ref(manage) will discuss a process or system development lifecycle
+@sec-manage will discuss a process or system development lifecycle
 to manage the A&R of a clinical project.

### 2023-05-22 submission-environment.qmd
Migrate from bookdown to Quarto book (working HTML book)

--- 2022-03-19 submission-environment.Rmd
@@
-# Running environment {#running-environment}
+# Running environment {#sec-running-environment}
 
@@
 
-`cleanslate` is an R package that offers a solution to
+cleanslate is an R package that offers a solution to
 create portable R environments.
 
-::: {.rmdnote}
-As of Q4 2021, the `cleanslate` package used in this chapter
+::: {.rmdnote data-latex=""}
+As of Q4 2021, the cleanslate package used in this chapter
 is still under active development and validation.
@@
 
-Install `cleanslate` from CRAN (once available):
+Install cleanslate from CRAN (once available):
 
@@
 
-The `cleanslate` package supports:
+The cleanslate package supports:
 
@@
 
-An essential feature of `cleanslate` is that it does **not** require
+An essential feature of cleanslate is that it does **not** require
 administrator privileges to run R and Rtools installers.
@@
 
-If you use `cleanslate`, remember to update the ADRG instructions for
+If you use cleanslate, remember to update the ADRG instructions for
 executing the analysis programs in R. Mostly, this can simplify
@@
 
-To make it convenient to use `cleanslate` in experiments,
-one can also use its RStudio addin. After `cleanslate`
+To make it convenient to use cleanslate in experiments,
+one can also use its RStudio addin. After cleanslate
 is installed, click `Addins` -> `cleanslate` -> `Create portable R environment`

### 2023-05-22 submission-overview.qmd
Migrate from bookdown to Quarto book (working HTML book)

--- 2021-11-08 submission-overview.Rmd
@@
-# (PART) eCTD submission {-}
-
 # Overview {#submission-overview}
@@
 
-> "Sponsors should provide the software programs used to create all ADaM
-datasets and **generate tables and figures associated with primary and
-secondary efficacy analyses**. Furthermore, sponsors should submit software
-programs used to generate additional information included in Section 14
-CLINICAL STUDIES of the Prescribing Information (PI)26 if applicable.
-**The specific software utilized should be specified in the ADRG**.
-The main purpose of requesting the submission of these programs
-**is to understand the process by which the variables for the respective
-analyses were created and to confirm the analysis algorithms**.
-Sponsors should submit software programs in **ASCII text format**;
-however, executable file extensions should not be used."
+> Sponsors should provide the software programs used to create all ADaM
+> datasets and **generate tables and figures associated with primary and
+> secondary efficacy analyses**. Furthermore, sponsors should submit software
+> programs used to generate additional information included in Section 14
+> CLINICAL STUDIES of the Prescribing Information (PI)26 if applicable.
+> **The specific software utilized should be specified in the ADRG**.
+> The main purpose of requesting the submission of these programs
+> **is to understand the process by which the variables for the respective
+> analyses were created and to confirm the analysis algorithms**.
+> Sponsors should submit software programs in **ASCII text format**;
+> however, executable file extensions should not be used.
 
-Chapter \@ref(submission-package) will focus on preparing
+@sec-submission-package will focus on preparing
 proprietary R packages and analysis code into proper
@@
 
-Chapter \@ref(running-environment) will discuss
+@sec-running-environment will discuss
 the recommendations to make the R code running environment

### 2023-05-22 submission-package.qmd
Migrate from bookdown to Quarto book (working HTML book)

--- 2022-12-07 submission-package.Rmd
@@
-# Submission package {#submission-package}
+# Submission package {#sec-submission-package}
 
@@
 
-This chapter uses `pkglite` to convert R source packages
+This chapter uses pkglite [@zhao2023electronic] to convert R source packages
 into text files and back.
@@
 with the primary R package(s) you want to submit,
-as `pkglite` supports packing multiple R packages into a single text file;
+as pkglite supports packing multiple R packages into a single text file;
 then restore and install them in the order they are packed.
@@
 Again, this can be checked in the project startup script,
-as discussed in section \@ref(reproduce).
+as discussed in @sec-reproduce.
 

### 2023-05-22 tlf-ae-specific.qmd
Migrate from bookdown to Quarto book (working HTML book)

--- 2022-12-19 tlf-ae-spec.Rmd
@@
 # Specific AE
+
+```{r, include=FALSE}
+source("_common.R")
+```
 
@@
 we need to summarize number of participants for each specific AE in Section 12.2, Adverse Events (AEs).
-
-```{r, include=FALSE}
-source("common.R")
-```
 
@@
 table. The percentage of participants for each AE can be
-calculated as shown in Chapter \@ref(aesummary).
+calculated as shown in @sec-aesummary.
 
@@
 ```{r}
+fmt_num <- function(x, digits, width = digits + 4) {
+  formatC(
+    x,
+    digits = digits,
+    format = "f",
+    width = width
+  )
+}
+```
+
+```{r}
 ana <- adae %>%
   mutate(
-    AESOC = toTitleCase(tolower(AESOC)),
-    AEDECOD = toTitleCase(tolower(AEDECOD))
+    AESOC = tools::toTitleCase(tolower(AESOC)),
+    AEDECOD = tools::toTitleCase(tolower(AEDECOD))
   )
@@
 We prepare reporting data for analysis population using code below:
+
+```{r}
+count_by <- function(data, # Input data set
+                     grp, # Group variable
+                     var, # Analysis variable
+                     var_label = var, # Analysis variable label
+                     id = "USUBJID") { # Subject ID variable
+  data <- data %>% rename(grp = !!grp, var = !!var, id = !!id)
+
+  left_join(
+    count(data, grp, var),
+    count(data, grp, name = "tot"),
+    by = "grp",
+  ) %>%
+    mutate(
+      pct = fmt_num(100 * n / tot, digits = 1),
+      n = fmt_num(n, digits = 0),
+      npct = paste0(n, " (", pct, ")")
+    ) %>%
+    pivot_wider(
+      id_cols = var,
+      names_from = grp,
+      values_from = c(n, pct, npct),
+      values_fill = list(n = "0", pct = fmt_num(0, digits = 0))
+    ) %>%
+    mutate(var_label = var_label)
+}
+```
 

### 2023-05-22 tlf-assemble.qmd
Migrate from bookdown to Quarto book (working HTML book)

--- 2023-01-13 tlf-assemble.Rmd
@@
 ```{r, include=FALSE}
-source("common.R")
+source("_common.R")
 ```
@@
 
-::: {.rmdnote data-latex=""}
+::: {.callout-note}
 The code below requires r2rtf version >= 1.0.0.
@@
 
-::: {.rmdnote data-latex=""}
+::: {.callout-caution}
 One limitation of combining RTF source code is that
@@
 
-::: {.rmdnote data-latex=""}
+::: {.callout-tip}
 There is a slight learning curve on how toggle fields work
@@
 
-::: {.rmdnote data-latex=""}
+::: {.callout-tip}
 A typical error message is that system can not find the file
@@
 
-::: {.rmdnote data-latex=""}
+::: {.callout-tip}
 If you modify table content in the combined `.docx` file,

### 2023-05-22 tlf-efficacy-ancova.qmd
Migrate from bookdown to Quarto book (working HTML book)

--- 2022-12-18 tlf-efficacy-ancova.Rmd
@@
 # Efficacy table
+
+```{r, include=FALSE}
+source("_common.R")
+```
 
@@
 in Section 11.4, Efficacy Results and Tabulations of Individual Participant.
-
-```{r, include=FALSE}
-source("common.R")
-```
 
@@
 To prepare the report, we create a few helper functions
-by using the `fmt_num()` function defined in Chapter \@ref(population).
+by using the `fmt_num()` function defined in @sec-population.
 
 - Format estimators
+
+```{r}
+fmt_num <- function(x, digits, width = digits + 4) {
+  formatC(
+    x,
+    digits = digits,
+    format = "f",
+    width = width
+  )
+}
+```
 
@@
   ) %>%
-  filter(str_detect(contrast, "- Placebo")) %>%
+  filter(stringr::str_detect(contrast, "- Placebo")) %>%
   select(contrast, ls, p)
@@
   ) %>%
-  rtf_colheader(paste(
+  rtf_colheader(
+    paste(
       "Treatment |",

### 2023-05-22 tlf-overview.qmd
Migrate from bookdown to Quarto book (working HTML book)

--- 2023-01-16 tlf-overview.Rmd
@@
-\mainmatter
-
-# (PART) Delivering TLFs in CSR {-}
-
 # Overview {#tlf-overview}
@@
 
-::: {.rmdnote data-latex=""}
+::: {.callout-note}
 FDA's PDF specification is a general reference. Each organization can define
@@
 
-::: {.rmdnote data-latex=""}
+::: {.callout-note}
 There are other R packages to create TLFs in ASCII, RTF and Word format.

### 2023-05-22 _common.R

@@ -1,3 +1,11 @@
+knitr::opts_chunk$set(
+  comment = "#>",
+  collapse = TRUE,
+  message = FALSE
+)
+
+options(dplyr.summarise.inform = FALSE)
+
 rtf2pdf <- function(input) {
   input <- normalizePath(input)
   x <- "export LD_LIBRARY_PATH=:/usr/lib/libreoffice/program:/usr/lib/x86_64-linux-gnu/"

### 2023-05-22 _utils.R

@@ -1,3 +1,14 @@
+# Utilities that can be ran manually
+
+# Convert RTF to PDF
+r2rtf:::rtf_convert_format(
+  list.files("tlf", pattern = ".rtf", full.names = TRUE),
+  output_dir = "tlf"
+)
+
+# Update manifest
+rsconnect::writeManifest(contentCategory = "site")
+
 #' Flatten copy
 #'
 #' @param from Source directory path.

### 2023-05-22 post-processing.R

@@ -1,3 +1,0 @@
-fs::dir_copy("tlf/", "_book/tlf", overwrite = TRUE)
-fs::dir_copy("slides/", "_book/slides", overwrite = TRUE)
-file.copy("static/CNAME", "_book", overwrite = TRUE)

### 2023-05-22 _utils.R
Remove `writeManifest()` call which is now obsolete

@@ -6,9 +6,6 @@ r2rtf:::rtf_convert_format(
   output_dir = "tlf"
 )
 
-# Update manifest
-rsconnect::writeManifest(contentCategory = "site")
-
 #' Flatten copy
 #'
 #' @param from Source directory path.

### 2023-05-23 index.qmd

`r if (knitr::is_latex_output()) '<!--'`

# Welcome {.unnumbered}

Welcome to R for Clinical Study Reports and Submission.
Clinical study reports (CSR) are crucial components in clinical trial development.
A CSR is an "integrated" full scientific report of an individual clinical trials.

The [ICH E3: Structure and Content of Clinical Study Reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf)
provides robust guidance to sponsors for developing a CSR.
This book demystifies the complexity by guiding you on how to use R for
streamlining the preparation of CSRs. Moreover, it walks you through the
process of preparing the submission to regulatory agencies. Regardless of your
expertise level in R programming, this book serves as an essential resource
in your statistical programming toolkit, enhancing efficiency and accuracy
in clinical trials reporting.

**This is a work-in-progress draft.**

## Events {.unnumbered}

```{r, echo=FALSE}
knitr::kable(read.table("events.tsv", stringsAsFactors = FALSE, header = TRUE, sep = "\t"))
```

## License {.unnumbered}

This book is licensed to you under the
[Creative Commons Attribution-NonCommercial-NoDerivatives 4.0 International License](https://creativecommons.org/licenses/by-nc-nd/4.0/).

`r if (knitr::is_latex_output()) '-->'`

### 2023-05-23 tlf-baseline.qmd

--- 2022-12-07 tlf-baseline.Rmd
@@
 # Baseline characteristics
+
+```{r, include=FALSE}
+source("_common.R")
+```
 
@@
 
-```{r, include=FALSE}
-source("common.R")
-```
-
 ```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
@@
 There are many R packages that can efficiently summarize baseline information.
-The [`table1`](https://github.com/benjaminrich/table1) R package is one of them.
+The [table1](https://github.com/benjaminrich/table1) R package is one of them.
 
@@
 For simplicity, we only analyze `SEX`, `AGE` and, `RACE` in this example
-using the `table1` R package. More details of the `table1` R package can
+using the table1 R package. More details of the table1 R package can
 be found in the package
@@
 
-The `table1` R package directly creates an HTML report.
+The table1 R package directly creates an HTML report.
 
@@
 recommended by regulatory agencies.
-`tbl_base` is used as input for `r2rtf` to create the final report.
+`tbl_base` is used as input for r2rtf to create the final report.
 
@@
 - Step 3: Transfer the output from Step 2 into a data frame that only contains ASCII characters.
-- Step 4: Define the format of the RTF table by using the R package `r2rtf`.
+- Step 4: Define the format of the RTF table by using the R package r2rtf.

### 2023-05-23 tlf-efficacy-km.qmd

--- 2022-12-18 tlf-efficacy-km.Rmd
@@
 # Efficacy figure
+
+```{r, include=FALSE}
+source("_common.R")
+```
 
@@
 
-::: {.rmdnote data-latex=""}
+::: {.callout-note}
 R packages such as

### 2023-05-23 tlf-population.qmd

--- 2023-01-13 tlf-population.Rmd
@@
-# Analysis population {#population}
+# Analysis population {#sec-population}
+
+```{r, include=FALSE}
+source("_common.R")
+```
 
@@
 analysis in Section 11.1, Data Sets Analysed.
-
-```{r, include=FALSE}
-source("common.R")
-```
 
@@
 - EFFFL: efficacy population flag
--   SAFFL: safty population flag
+- SAFFL: safety population flag
 
@@
 
-In Chapter \@ref(disposition), there are a few repeating steps to:
+In @sec-disposition, there are a few repeating steps to:
 
@@
 fmt_num <- function(x, digits, width = digits + 4) {
-  formatC(x,
+  formatC(
+    x,
     digits = digits,
@@
 For example, we can compare the two versions of code to format the
-percentage used in Chapter \@ref(disposition) and `fmt_num`.
+percentage used in @sec-disposition and `fmt_num()`.
 
@@
   rtf_title(
-    "Summary of Analysis Sets", ## "Participants Accounting in Analysis Population" is a weird title for me. Merck uses this table title for CSR? 
+    "Summary of Analysis Sets",
     "(All Participants Randomized)"

### 2023-05-25 preface.qmd
Migrate from bookdown to Quarto book (working HTML book)
Update authors in metadata and preface

--- 2023-01-16 preface.Rmd
@@
-# Preface {-}
+# Preface {.unnumbered}
 
@@
 
-  Yilong Zhang <<elong0527@gmail.com>>, Nan Xiao, Keaven Anderson, Yalin Zhu
+  [Yilong Zhang](https://elong0527.github.io/),
+  [Nan Xiao](https://nanx.me/),
+  [Keaven Anderson](https://keaven.github.io/),
+  Yalin Zhu
 
@@
 ```
-
-```{block2, type="flushright", html.tag="p"}
-Yilong Zhang, Nan Xiao, Keaven Anderson, Yalin Zhu
-
-December 2022
-```

### 2023-06-04 project-folder.qmd
RStudio -> Posit except RStudio IDE

@@
 [clone](https://happygitwithr.com/rstudio-git-github.html#clone-the-new-github-repository-to-your-computer-via-rstudio)
-the project using RStudio with
+the project using RStudio IDE with
 
@@
 In the esubdemo project, a reproducible environment is created when you open
-the `esubdemo.Rproj` from RStudio.
-When we open the `esubdemo` project, RStudio will execute the R code in `.Rprofile` automatically.
+the `esubdemo.Rproj` from RStudio IDE.
+When we open the `esubdemo` project, RStudio IDE will execute the R code in `.Rprofile` automatically.
 So we can use `.Rprofile` to set up a reproducible environment.
@@
 ::: {.callout-tip}
-A cloud-based R environment (e.g., RStudio Workbench)
+A cloud-based R environment (e.g., [Posit Workbench](https://posit.co/products/enterprise/workbench/))
 can enhance the reproducibility within an organization

### 2023-06-04 slides/fda-workshop-slides.Rmd
fix brocken link
fix broken link
RStudio -> Posit except RStudio IDE

@@
 - During the workshop, we will use `esubdemo` project.
-  - `esubdemo` is shared in RStudio Cloud (<https://rstudio.cloud/>)
-  - You can also clone the `esubdemo` project to your local RStudio.
+  - `esubdemo` is shared in Posit Cloud (<https://posit.cloud/>)
+  - You can also clone the `esubdemo` project to your local RStudio IDE.
 
@@
 
-<https://r4csr.org/disposition.html>
+<https://r4csr.org/tlf-disposition.html>
 
@@
 
-<https://r4csr.org/efficacy-table.html>
+<https://r4csr.org/tlf-efficacy-ancova.html>
 
@@
 
-<https://r4csr.org/specific-ae.html>
+<https://r4csr.org/tlf-ae-specific.html>
 
@@
 
-<https://r4csr.org/assemble.html>
+<https://r4csr.org/tlf-assemble.html>
 
@@
 
-<https://r4csr.org/folder.html>
+<https://r4csr.org/project-folder.html>
 

### 2023-06-04 slides/r4csr-gwu.Rmd
Built site for gh-pages
RStudio -> Posit except RStudio IDE

@@
 - Fixed R package snapshot date: e.g., 2021-08-31
-  - [RStudio Public Package Manager](https://packagemanager.rstudio.com/)
+  - [Posit Package Manager](https://packagemanager.posit.co/)
   - [CRAN Time Machine by MRAN](https://mran.microsoft.com/timemachine)
@@
 
-More details: <https://r4csr.org/folder.html>
+More details: <https://r4csr.org/project-folder.html>
 

### 2023-06-04 slides/workshop-slides.Rmd
fix broken link
RStudio -> Posit except RStudio IDE

@@
 - During the workshop, we will use `esubdemo` project.
-  - `esubdemo` is shared in RStudio Cloud (<https://rstudio.cloud/>)
+  - `esubdemo` is shared in Posit Cloud (<https://posit.cloud/>)
   - Project link will be shared in chat

### 2023-06-05 submission-environment.qmd
Add Quarto PDF output - magically working without tuning

@@
 
-::: {.rmdnote data-latex=""}
+::: {.callout-note}
 As of Q4 2021, the cleanslate package used in this chapter
@@
   to use the specified `repo` (an MRAN snapshot in this example),
-  and give instruction to set the R binary path in RStudio;
+  and give instruction to set the R binary path in RStudio IDE;
 - Create a `.Renviron` file under the project folder and set the
@@
 Go to the working directory created above, double click the .Rproj file
-to open the project in RStudio. Follow the instructions to select the
-project-specific R version, then restart RStudio. If successful,
+to open the project in RStudio IDE. Follow the instructions to select the
+project-specific R version, then restart RStudio IDE. If successful,
 the R version and package repo should be printed as defined above.
@@
 To make it convenient to use cleanslate in experiments,
-one can also use its RStudio addin. After cleanslate
+one can also use its RStudio IDE addin. After cleanslate
 is installed, click `Addins` -> `cleanslate` -> `Create portable R environment`
-in RStudio, or call `cleanslate:::create_env_addin()` to open it.
+in RStudio IDE, or call `cleanslate:::create_env_addin()` to open it.
 

### 2023-06-05 project-folder.qmd
Update MRAN snapshot links to Posit Public Package Manager snapshot links

@@
 # set up repository based on the snapshot date
-repos <- paste0("https://mran.microsoft.com/snapshot/", snapshot, "/")
+repos <- paste0("https://packagemanager.posit.co/cran/", snapshot)
 
@@
 We can also define the package repository to be a specific snapshot date.
-For example, we used Microsoft R Application Network (MRAN) to
+For example, we used Posit Public Package Manager to
 define the snapshot date to be `2021-08-06`.
@@
 
-In other words, all R packages installed in this RStudio project
+In other words, all R packages installed in this R project
 are based on the frozen R version at the snapshot date.
-Here it is `2021-08-06` by using the MRAN server.
+Here it is `2021-08-06` by using the Posit Package Manager.
 
-Below information will be displayed after a new R session is opened.
+The information below will be displayed after a new R session is opened.
 
@@
 Current project R package repository:
-    https://mran.microsoft.com/snapshot/2021-08-06/
+    https://packagemanager.posit.co/cran/2021-08-06
 ```
@@
 ::: {.callout-note}
-Posit Package Manager provides a solution to host both publicly available
-and internally developed R packages. However, the
 [Posit Public Package Manager](https://packagemanager.posit.co/)
-does not provide a daily snapshot as MRAN does.
+hosts daily CRAN snapshots for Mondays to Fridays of the week.
+Posit Package Manager, when deployed internally within an organization,
+provides a solution to host both publicly available and internally
+developed R packages.
 :::

### 2023-06-05 submission-package.qmd
Update MRAN snapshot links to Posit Public Package Manager snapshot links

@@
 # # Commented out as the snapshot availability is not guaranteed
-# x <- available.packages(contriburl = contrib.url("https://mran.microsoft.com/snapshot/2021-08-06/"))
+# x <- available.packages(contriburl = contrib.url("https://packagemanager.posit.co/cran/2021-08-06"))
 # df2$version <- x[match(df2$package, x[, "Package"]), "Version"]
@@
 
-options(repos = "https://mran.microsoft.com/snapshot/2021-08-06/")
+options(repos = "https://packagemanager.posit.co/cran/2021-08-06")
 

### 2023-06-05 tlf-overview.qmd
RStudio -> Posit

@@
 
-RStudio provided outstanding [cheatsheets](https://posit.co/resources/cheatsheets/)
+Posit provided outstanding [cheatsheets](https://posit.co/resources/cheatsheets/)
 and [tutorials](https://github.com/rstudio-education/remaster-the-tidyverse) for tidyverse.

### 2023-06-05 submission-environment.qmd
Update remaining MRAN mentions

@@
   from = "https://cran.r-project.org/",
-  repo = "https://mran.microsoft.com/snapshot/2021-08-06/"
+  repo = "https://packagemanager.posit.co/cran/2021-08-06"
 )
@@
 - Create a `.Rprofile` file under the project folder, set `options(repos)`
-  to use the specified `repo` (an MRAN snapshot in this example),
+  to use the specified `repo` (a Posit Public Package Manager snapshot in this example),
   and give instruction to set the R binary path in RStudio IDE;

### 2023-06-06 slides/china-r.Rmd
fix brocken link
Built site for gh-pages

@@
 
-Training: [December 6th, Deming Conference 2021](https://demingconference.org/wp-content/uploads/2021/07/2021-Deming-Conference-Printed-Program-v1.0.pdf)
+Training: [December 6th, Deming Conference 2021](https://web.archive.org/web/20220531042701/https://demingconference.org/wp-content/uploads/2021/11/2021-Deming-Conference-Virtual-Printed-Program-v1.1.pdf)
 
@@
 
-More details: <https://r4csr.org/manage.html>
+More details: <https://r4csr.org/project-management.html>
 

### 2023-06-06 slides/fda-workshop-slides.Rmd
Built site for gh-pages

@@
 
-<https://r4csr.org/population.html>
+<https://r4csr.org/tlf-population.html>
 
@@
 
-<https://r4csr.org/baseline-characteristics.html>
+<https://r4csr.org/tlf-baseline.html>
 
@@
 
-<https://r4csr.org/aesummary.html>
+<https://r4csr.org/tlf-ae-summary.html>
 
@@
 
-<https://r4csr.org/manage.html>
+<https://r4csr.org/project-management.html>
 
@@
 - Mission: Easier R-based clinical trial regulatory submissions
-- [pilot 1 - common analyses](https://rconsortium.github.io/submissions-wg/pilot-overall.html#pilot-1---common-analyses)
+- [pilot 1 - common analyses](https://rconsortium.github.io/submissions-wg/pilot1.html)
   - 2021.11   Initial submission to FDA. 
@@
 - "Using FDA developed code, a statistical analyst was able to independently generate tables using the submitted data."
-- [Initial submission FDA response](https://github.com/RConsortium/submissions-wg/blob/main/Documents/Summary_R_Pilot_Submission.pdf)
+- [Initial submission FDA response](https://github.com/RConsortium/submissions-wg/blob/main/_Documents/Summary_R_Pilot_Submission.pdf)
 
@@
 - FDA agrees that the initial phase of the R Pilot submission has been completed.
-- [Revised  submission FDA response](https://github.com/RConsortium/submissions-wg/blob/main/Documents/Summary_R_Pilot_Submission2.pdf)
+- [Revised  submission FDA response](https://github.com/RConsortium/submissions-wg/blob/main/_Documents/Summary_R_Pilot_Submission2.pdf)
 
@@
 
-<https://r4csr.org/running-environment.html>
+<https://r4csr.org/submission-environment.html>
 

### 2023-06-06 slides/r4csr-gwu.Rmd
Built site for gh-pages

@@
 - "Using FDA developed code, a statistical analyst was able to independently generate tables using the submitted data."
-- [Original FDA response](https://github.com/RConsortium/submissions-wg/blob/main/Documents/Summary_R_Pilot_Submission.pdf)
+- [Original FDA response](https://github.com/RConsortium/submissions-wg/blob/main/_Documents/Summary_R_Pilot_Submission.pdf)
 

### 2023-06-06 slides/r4csr-psi.Rmd
Built site for gh-pages

@@
 - Objective: Create open source example for R-based clinical trial regulatory submissions
-- [pilot 1 - common analyses](https://rconsortium.github.io/submissions-wg/pilot-overall.html#pilot-1---common-analyses)
+- [pilot 1 - common analyses](https://rconsortium.github.io/submissions-wg/pilot1.html)
   - 2021.11   Initial submission to FDA. 
@@
 
-- [pilot 3 - ADaM data](https://github.com/RConsortium/submissions-wg/blob/main/Documents/R%20Submission%20%20Pilot%201%20Extension%20Proposal.pdf)
+- [pilot 3 - ADaM data](https://github.com/RConsortium/submissions-wg/blob/main/_Documents/R%20Submission%20%20Pilot%201%20Extension%20Proposal.pdf)
   - 2022.10   Proposal initiation.

### 2023-07-09 index.qmd
update index.qmd

@@
 The [ICH E3: Structure and Content of Clinical Study Reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf)
-provides robust guidance to sponsors for developing a CSR.
-This book demystifies the complexity by guiding you on how to use R for
-streamlining the preparation of CSRs. Moreover, it walks you through the
-process of preparing the submission to regulatory agencies. Regardless of your
-expertise level in R programming, this book serves as an essential resource
-in your statistical programming toolkit, enhancing efficiency and accuracy
-in clinical trials reporting.
+offer comprehensive instructions to sponsors on the creation of a CSR.
+This book is a clear and straightforward guide on using R to streamline the process of preparing CSRs.
+Additionally, it provides detailed guidance on the submission process to regulatory agencies.
+Whether you are a beginner or an experienced R programmer, this book is an indispensable asset in your clinical reporting arsenal.
 
@@
 ```{r, echo=FALSE}
-knitr::kable(read.table("events.tsv", stringsAsFactors = FALSE, header = TRUE, sep = "\t"))
+knitr::kable(
+  read.table("events.tsv", 
+             stringsAsFactors = FALSE, 
+             header = TRUE, 
+             sep = "\t")
+  )
 ```

### 2023-07-09 preface.qmd
- update preface.qmd - update R package reference

--- 2022-07-12 preface.Rmd
@@
 
-In clinical trial development, source code needs to be developed and maintained to generate and deliver
-Study Data Tabulation Model (SDTM), Analysis Dataset Model (ADaM) datasets and tables, listings, and figures (TLFs).
-A typical example is a Phase 3 clinical trial where hundreds of TLFs are required for submission.
-Considering the number of programs required for such an effort,
-a consistent and well-defined folder structure is crucial in
-managing a clinical trial analysis and reporting (A&R) project.
+In the development of clinical trials, it is necessary to create and manage source code for generating and delivering Study Data Tabulation Model (SDTM), Analysis Dataset Model (ADaM) datasets, as well as tables, listings, and figures (TLFs).
+This is particularly evident in Phase 3 trials, where numerous TLFs are needed for submission.
+To effectively handle the large number of programs involved in such endeavors, it is essential to establish a consistent and well-defined folder structure for managing the analysis and reporting (A&R) project of a clinical trial.
 
-We recommend using the
-[R package folder structure](https://github.com/rstudio/cheatsheets/raw/master/package-development.pdf)
-to organize all the A&R-related source code and documentation for a clinical trial A&R project.
-The R package folder structure is well defined and widely used in the R community through repositories (e.g., CRAN).
+To streamline the organization of source code and documentation for a clinical trial A&R project, we suggest employing the R package folder structure.
+This folder structure is extensively utilized within the R community and is well-defined, often found in repositories like CRAN.
+By adopting this structure, you can benefit from a standardized and widely accepted framework for managing your A&R-related materials in an efficient and accessible manner.
 
-The consistent approach of using the R package folder structure
-simplifies the communication for all developers within and across organizations.
+Using the R package folder structure provides a consistent approach that simplifies communication among developers, both within and across organizations.
 
-- For a new R developer, it is an essential step to develop R packages
-  when you want to share your work with others.
-  You will learn one folder structure widely used in the R community
-  with outstanding tutorials and tools for free.
+- For newcomers to R development, creating R packages is an essential step when sharing their work with others.
+The R community offers a widely adopted folder structure accompanied by excellent tutorials and free tools.
 - For an experienced R developer, there is a minimal learning curve.
-- For an organization, it simplifies the process, tool, template,
-  and training development, because a unified folder structure is used
-  to develop and maintain standard tool and analysis projects.
+- For an organization, adopting the R package folder structure simplifies the development of processes, tools, templates, and training.
+It enables the use of a unified folder structure for building and maintaining standardized tool and analysis projects.
 
-The workflow around an R package can also improve
-the traceability and reproducibility of an analysis project [@marwick2018packaging].
+The workflow around an R package can also improve the traceability and reproducibility of an analysis project [@marwick2018packaging].
 
-We will revisit folder structure when we discuss project management for a clinical trial project.
+We will revisit the folder structure topic when discussing project management for a clinical trial project.
 
-In addition, R package folder structure is also recommended
-to develop Shiny apps as discussed
-in Chapter 20 of the [Mastering Shiny](https://mastering-shiny.org/scaling-packaging.html) book
-and the [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/golem.html) book.
+Additionally, the R package folder structure is also recommended for developing Shiny apps, as discussed in Chapter 20 of the [Mastering Shiny](https://mastering-shiny.org/scaling-packaging.html) book and the [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/golem.html) book.
 
@@
 
-This book is an intermediate-level book by
-assuming readers have R programming and clinical development knowledge.
-The assumption of each part is as below:
+This book is designed for intermediate-level readers who possess knowledge in both R programming and clinical development.
+Each part of the book makes certain assumptions about the readers' background:
 
-- Part 1, "Delivering TLFs in CSR" provides general information with examples
-  to create tables, listings, and figures.
-  In this part, we assume readers are individual contributors to a clinical project
-  with some experience in R programming.
-  We expect readers are familiar with data manipulation in R.
-  Some good references include [Hands-On Programming with R](https://rstudio-education.github.io/hopr/),
+- Part 1, titled "Delivering TLFs in CSR," provides general information and examples on creating tables, listings, and figures.
+It assumes that readers are individual contributors to a clinical project with prior experience in R programming.
+Familiarity with data manipulation in R is expected.
+Some recommended references for this part include [Hands-On Programming with R](https://rstudio-education.github.io/hopr/),
   [R for Data Science](https://r4ds.had.co.nz/) and
@@
 
-- Part 2, "Clinical trial project" provides general information
-  with examples to manage a clinical trial A&R project.
-  In this part, we assume a reader is a project lead with experience in R package development.
-  Some good references include [R Packages](https://r-pkgs.org/) and the
+- Part 2, titled "Clinical trial project," provides general information and examples on managing a clinical trial A&R project.
+It assumes that readers are project leads who have experience in R package development.
+Recommended references for this part include [R Packages](https://r-pkgs.org/) and the
   [tidyverse style guide](https://style.tidyverse.org/).
 
-- Part 3, "eCTD submission package" provides general information in preparing
-  submission packages related to clinical study report (CSR) in
-  electronic Common Technical Document (eCTD).
-  In this part, we assume a reader is a project lead of a clinical project
-  with experience in R package development and submission.
+- Part 3, titled "eCTD submission package," provides general information on preparing submission packages related to the CSR in the electronic Common Technical Document (eCTD) format.
+It assumes that readers are project leads of clinical projects who possess experience in R package development and submission.
 
@@
 We share the same philosophy described in
-Section 1.1 of the [R Packages](https://r-pkgs.org/Introduction.html#intro-phil) book
-and quote here.
+the introduction of the [R Packages](https://r-pkgs.org/Introduction.html#intro-phil) book [@wickham2023r]
+, which we quote below:
 
@@
 - "Do as little as possible by hand. Do as much as possible with functions."
-- "The goal is to spend your time thinking about what you want to do
-rather than thinking about the minutiae of the package structure."
 
@@
 
-The document is maintained by a community.
-While reading the document, you can be a contributor as well.
-The quality of this document relies on you.
+This document is a collaborative effort maintained by a community.
+As you read through it, you also have the opportunity to contribute and enhance its quality.
+Your input and involvement play a vital role in shaping the excellence of this document.
 
-- Authors: contributed the majority of content to at least one chapter.
+- Authors: made significant contributions to at least one chapter, constituting the majority of the content.
 
-  Yilong Zhang <<elong0527@gmail.com>>, Nan Xiao, Keaven Anderson, Yalin Zhu
+  [Yilong Zhang](https://elong0527.github.io/),
+  [Nan Xiao](https://nanx.me/),
+  [Keaven Anderson](https://keaven.github.io/),
+  [Yalin Zhu](https://yalin.netlify.app/)
 

### 2023-07-09 tlf-overview.qmd
update tlf-overview.qmd

@@
 
-In clinical trials, a critical step is to submit trial results to regulatory agencies.
-[Electronic Common Technical Document (eCTD)](https://en.wikipedia.org/wiki/Electronic_common_technical_document)
-has become a worldwide regulatory submission standard format.
-For example, the United States Food and Drug Administration (US FDA) requires
-new drug applications and biologics license applications
-[must be submitted using the eCTD format](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd).
-The Clinical Data Interchange Standards Consortium (CDISC) provides a
-[pilot project following ICH E3 guidance](https://github.com/cdisc-org/sdtm-adam-pilot-project).
+Submitting clinical trial results to regulatory agencies is a crucial aspect of clinical development. 
+The [Electronic Common Technical Document (eCTD)](https://en.wikipedia.org/wiki/Electronic_common_technical_document) has emerged as the global standard format for regulatory submissions.
+For instance, the United States Food and Drug Administration (US FDA) [mandates the use of eCTD]((https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd)) for new drug applications and biologics license applications.
 
-Within eCTD, clinical study reports (CSRs) are located at module 5.
-[ICH E3 guidance](https://www.ich.org/page/efficacy-guidelines) provides
-a compilation of the structure and content of clinical study reports.
+A CSR provides comprehensive information about the methods and results of an individual clinical study.
+To support the statistical analysis, numerous tables, listings, and figures are included within the main text and appendices.
+As part of the CDISC pilot project, an [example CSR](https://github.com/cdisc-org/sdtm-adam-pilot-project/blob/master/updated-pilot-submission-package/900172/m5/53-clin-stud-rep/535-rep-effic-safety-stud/5351-stud-rep-contr/cdiscpilot01/cdiscpilot01.pdf) is also available for reference.
+If you seek additional examples of CSR, you can visit the clinical data website of the [European Medicines Agency (EMA) clinical data website](https://clinicaldata.ema.europa.eu/web/cdp/home).
 
-A typical CSR contains full details on the methods and results of an individual clinical study.
-In support of the statistical analysis, a large number of tables, listings, and figures are
-incorporated into the main text and appendices.
-In the CDISC pilot project, an
-[example CSR](https://github.com/cdisc-org/sdtm-adam-pilot-project/blob/master/updated-pilot-submission-package/900172/m5/53-clin-stud-rep/535-rep-effic-safety-stud/5351-stud-rep-contr/cdiscpilot01/cdiscpilot01.pdf)
-is also provided. If you are interested in more examples of clinical study reports,
-you can go to the [European Medicines Agency (EMA) clinical data website](https://clinicaldata.ema.europa.eu/web/cdp/home).
+The creation of CSR is a collaborative effort that involves various professionals such as clinicians, medical writers, statisticians, statistical programmers. 
+In this context, we will focus on the specific deliverables provided by statisticians and statistical programmers.
 
-Building CSRs is teamwork between clinicians, medical writers, statisticians, statistical programmers,
-and other relevant specialists such as experts on biomarkers.
-Here, we focus on the work and deliverables completed by statisticians and statistical programmers.
-In an organization, they commonly work together to
-define, develop, validate and deliver tables, listings, and figures (TLFs) required for a CSR to
-summarize the efficacy and/or safety of the pharmaceutical product.
-Microsoft Word is widely used to prepare CSR in the pharmaceutical industry.
-Therefore, `.rtf`, `.doc`, `.docx` are commonly used formats in their deliverables.
+Within an organization, these professionals typically collaborate to define, develop, validate, and deliver the necessary tables, listings, and figures (TLFs) for a CSR.
+These TLFs serve to summarize the efficacy and/or safety of the pharmaceutical product under study.
+In the pharmaceutical industry, Microsoft Word is widely utilized for CSR preparation.
+As a result, the deliverables from statisticians and statistical programmers are commonly provided in formats such as `.rtf`, `.doc`, `.docx` to align with industry standards and requirements.
 
-In this chapter, our focus is to illustrate how to create tables, listings, and figures (TLFs) in RTF format
-that is commonly used in a CSR. The examples are in compliance with the
-[FDA's Portable Document Format (PDF) Specifications](https://www.fda.gov/media/76797/download).
+Our focus is to demonstrate the process of generating TLFs in RTF format, which is commonly employed in CSRs. The examples provided in this chapter adhere to the 
+[ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf) and the [FDA's PDF Specifications](https://www.fda.gov/media/76797/download).
 
@@
 more specific TLF format requirements that can be different from the examples in this book.
+The FDA's PDF specification serves as a general reference for formatting requirements.
+Each organization has the flexibility to define its own specific requirements for TLFs. 
+These specific format requirements may differ from the examples provided in this book.
+It is advisable to consult and adhere to the guidelines and specifications set by your respective organization when preparing TLFs for submission.
 :::
 
-## Structure and content
-
-In the rest of this chapter, we are following the
-[ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf)
-on the structure and content of clinical study reports.
-
-In a CSR, most of TLFs are located in
+By following the ICH E3 guidance, most of TLFs in a CSR are located at
 
@@
 
-We used publicly available CDISC pilot
-[study data located in the CDISC GitHub repository](https://github.com/cdisc-org/sdtm-adam-pilot-project/tree/master/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).
-
-For simplicity, we have downloaded all these datasets into the `data-adam/`
-folder of this project and converted them from the `.xpt` format to
-the `.sas7bdat` format.
-
 The dataset structure follows
@@
 
+In this project, we used publicly available CDISC pilot study data, which is accessible through the [CDISC GitHub repository](https://github.com/cdisc-org/sdtm-adam-pilot-project/tree/master/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).
+
+To streamline the process, we have downloaded all the datasets from the repository and stored them in the [`data-adam/` folder](https://github.com/elong0527/r4csr/tree/main/data-adam) within this project.
+Additionally, we converted these datasets from the `.xpt` format to the `.sas7bdat` format for ease of use and compatibility.
+The dataset structure adheres to the CDISC [Analysis Data Model (ADaM)](https://www.cdisc.org/standards/foundational/adam) standard.
+
 ## Tools
 
-In this part, we mainly use the R packages below to illustrate
-how to deliver TLFs in a CSR.
+To exemplify the generation of TLFs in RTF format, we rely on the functionality provided by two R packages:
 
-- [tidyverse](https://www.tidyverse.org/): prepare datasets ready for reporting.
-- [r2rtf](https://merck.github.io/r2rtf/): create RTF outputs
+- [tidyverse](https://www.tidyverse.org/): preparation of datasets in a format suitable for reporting purposes.
+The tidyverse package offers a comprehensive suite of tools and functions for data manipulation and transformation, ensuring that the data is structured appropriately.
+- [r2rtf](https://merck.github.io/r2rtf/): creation RTF files. The r2rtf package offers functions specifically designed for generating RTF files, allowing us to produce TLFs in the desired format.
 
 ::: {.callout-note}
-There are other R packages to create TLFs in ASCII, RTF and Word format.
-For example, rtables, huxtable, pharmaRTF, gt, officer, flextable etc.
-Here we focus on r2rtf to illustrate the concept.
-Readers are encouraged to explore other R packages to find the proper tools to fit your purpose.
+There are indeed several other R packages available that can assist in creating TLFs in ASCII, RTF, and Word formats such as rtables, huxtable, pharmaRTF, gt, officer, and flextable.
+However, in this particular context, we will concentrate on demonstrating the concept using the r2rtf package.
+It is highly recommended for readers to explore and experiment with various R packages to identify the most suitable tools that align with their specific needs and objectives. 
 :::
@@
 
-tidyverse is a collection of R packages to simplify the workflow to manipulate,
-visualize and analyze data in R.
-Those R packages share
-[the tidy tools manifesto](https://tidyverse.tidyverse.org/articles/manifesto.html)
-and are easy to use for interactive data analysis.
+The tidyverse is a comprehensive collection of R packages that aim to simplify the workflow of manipulating, visualizing, and analyzing data in R.
+These packages adhere to the principles outlined in the [the tidy tools manifesto](https://tidyverse.tidyverse.org/articles/manifesto.html) and offer user-friendly interfaces for interactive data analysis.
 
-Posit provided outstanding [cheatsheets](https://posit.co/resources/cheatsheets/)
-and [tutorials](https://github.com/rstudio-education/remaster-the-tidyverse) for tidyverse.
+The creators of the tidyverse, Posit, have provided exceptional [cheatsheets](https://posit.co/resources/cheatsheets/)
+and [tutorials](https://github.com/rstudio-education/remaster-the-tidyverse) that serve as valuable resources for learning and mastering the functionalities of these packages.
 
-There are also books to introduce tidyverse.
-We assume the reader have experience in using tidyverse in this book.
+Furthermore, there are several books available that serve as introductions to the tidyverse. 
+For example:
 
@@
 
+::: {.callout-note}
+In this book, we assume that the reader already has experience in using the tidyverse.
+This prior knowledge and familiarity with the tidyverse tools enable a more efficient and focused exploration of the concepts presented throughout the book.
+:::
+
 ### r2rtf
 
-r2rtf is an R package to create production-ready tables and figures in RTF format.
-This R package is designed to
+r2rtf is an R package specifically designed to create production-ready tables and figures in RTF format.
 
@@
   to help you translate a data frame to a table in an RTF file;
-- enable pipes (`%>%`);
-- focus on the **table format** only.
-  Data manipulation and analysis shall be handled by other R packages (e.g., tidyverse).
+- enable pipes (`|>`);
+- focus on the **table format** only. Data manipulation and analysis tasks can be handled by other R packages like the tidyverse
 
-Before creating an RTF table, we need to
+Before generating an RTF table using r2rtf, there are a few steps to follow:
 
-- figure out the table layout;
-- split the layout into small tasks in the form of a computer program;
-- execute the program.
+- Determine the desired layout of the table.
+- Break down the layout into smaller tasks, which can be programmed.
+- Execute the program to generate the table.
 
@@
 
-Other extended examples and features are covered on the
+We provide a concise introduction to r2rtf and demonstrate how to convert data frames into TLFs. 
+For more comprehensive examples and additional features, we encourage readers to explore the 
 [r2rtf package website](https://merck.github.io/r2rtf/articles/index.html).
 
-To explore the basic RTF generation verbs in r2rtf,
-we will use the dataset `r2rtf_adae` saved in the r2rtf package.
-This dataset contains adverse events (AEs) information from a clinical trial.
+To illustrate the basic usage of the r2rtf package, we will work with the "r2rtf_adae" dataset, available within the r2rtf package. 
+This dataset contains information on adverse events (AEs) from a clinical trial, 
+which will serve as a practical example for generating RTF tables using r2rtf.
 
-We will begin by loading the packages:
+To begin, let's load the required packages:
 
-```{r}
-library(dplyr) # Manipulate data
-library(tidyr) # Manipulate data
+```{r, message=FALSE}
+library(tidyverse) # Manipulate data
 library(r2rtf) # Reporting in RTF format
@@
 
-Below is the meaning of relevant variables.
-More information can be found on the help page of the dataset (`?r2rtf_adae`)
+In this example, we will focus on three variables from the `r2rtf_adae` dataset:
 
-In this example, we consider three variables:
+- `USUBJID`: unique subject identifier.
+- `TRTA`: actual treatment group.
+- `AEDECOD`: dictionary-derived derm.
 
-- USUBJID: Unique Subject Identifier
-- TRTA: Actual Treatment
-- AEDECOD: Dictionary-Derived Term
+::: {.callout-note}
+Additional information about these variables can be found on the help page of the dataset, 
+which can be accessed by using the command `?r2rtf_adae` in R.
+:::
 
 ```{r}
-r2rtf_adae %>%
-  select(USUBJID, TRTA, AEDECOD) %>%
+r2rtf_adae |>
+  select(USUBJID, TRTA, AEDECOD) |>
   head(4)
@@
 
-dplyr and tidyr packages within tidyverse are used
-for data manipulation to create a data frame
-that contains all the information we want to add in an RTF table.
+To manipulate the data and create a data frame containing the necessary information for the RTF table, we can use the dplyr and tidyr packages within the tidyverse.
 
@@
 
-Now we have a dataset `tbl` in preparing the final RTF table.
-
-r2rtf aims to provide one function for each type of table layout.
-Commonly used verbs include:
+Having prepared the dataset `tbl`, 
+we can now proceed with constructing the final RTF table using the r2rtf package. 
+The r2rtf package has various functions, each designed for a specific type of table layout. Some commonly used verbs include:
 
@@
 
-All these verbs are designed to enable the usage of pipes (`%>%`).
-A full list of all functions can be found in the
-[r2rtf package function reference manual](https://merck.github.io/r2rtf/reference/index.html).
+Functions provided by the r2rtf package are designed to work seamlessly with the pipe operator (|>).
+This allows for a more concise and readable code structure, enhancing the efficiency of table creation in RTF format.
+A full list of functions in the r2rtf package can be found in the
+[package reference page](https://merck.github.io/r2rtf/reference/index.html).
 
-A minimal example below illustrates how to combine verbs using pipes to create an RTF table.
+Here is a minimal example that demonstrates how to combine functions using pipes to create an RTF table
 
@@
 ```{r}
-head(tbl) %>%
-  rtf_body() %>% # Step 1 Add table attributes
-  rtf_encode() %>% # Step 2 Convert attributes to RTF encode
-  write_rtf("tlf/intro-ae1.rtf") # Step 3 Write to a .rtf file
-```
-
-```{r, include=FALSE}
-rtf2pdf("tlf/intro-ae1.rtf")
+head(tbl) |>
+  rtf_body() |> 
+  rtf_encode() |> 
+  write_rtf("tlf/intro-ae1.rtf") 
 ```
@@
 ```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
+rtf2pdf("tlf/intro-ae1.rtf")
 knitr::include_graphics("tlf/intro-ae1.pdf")
@@
 
-If we want to adjust the width of each column to
-provide more space to the first column,
-this can be achieved by updating the `col_rel_width` argument
-in the `rtf_body()` function.
+In the previous example, we may want to add more column space to the first column.
+We can achieve the goal by updating the `col_rel_width argument` in the rtf_body() function.
 
-In this example, the input of `col_rel_width` is a vector
-with the same length for the number of columns.
-This argument defines the relative width of each column
-within a pre-defined total column width.
+In the example below, the `col_rel_width` argument expects a vector with the same length as the number of columns in the table `tbl`.
+This vector defines the relative width of each column within a predetermined total column width.
+Here, the relative width is defined as `3:2:2:2` that allow us to allocate more space to specific columns.
 
-In this example, the defined relative width is `3:2:2:2`.
-Only the ratio of `col_rel_width` is used.
-Therefore it is equivalent to use `col_rel_width = c(6, 4, 4, 4)`
-or `col_rel_width = c(1.5, 1, 1, 1)`.
+::: {.callout-note}
+Only the ratio of the `col_rel_width` values is considered. 
+Therefore, using `col_rel_width = c(6, 4, 4, 4)` or `col_rel_width = c(1.5, 1, 1, 1)` would yield equivalent results, as they maintain the same ratio.
+:::
 
 ```{r}
-head(tbl) %>%
-  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
-  # define relative width
-  rtf_encode() %>%
+head(tbl) |>
+  rtf_body(col_rel_width = c(3, 2, 2, 2)) |>
+  rtf_encode() |>
   write_rtf("tlf/intro-ae2.rtf")
@@
 
-```{r, include=FALSE}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 rtf2pdf("tlf/intro-ae2.rtf")
-```
-
-```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/intro-ae2.pdf")
@@
 
-In the previous example, we found the issue of a misaligned column header.
-We can fix the issue by using the `rtf_colheader()` function.
+In the previous example, we encountered a misalignment issue with the column header. 
+To address this, we can use the `rtf_colheader()` 
+function to adjust column header width and provide more informative column headers.
 
-In `rtf_colheader()`, the `colheader` argument is used to provide the content of the column header.
-We use `"|"` to separate the columns.
-
-In the example below, `"Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose"`
-define a column header with 4 columns.
+Within the `rtf_colheader()` function, the `colheader` argument is used to specify the content of the column header. 
+The columns are separated using the "|" symbol.
+In the following example, we define the column header as `"Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose"`, representing the four columns in the table:
 
 ```{r}
-head(tbl) %>%
+head(tbl) |>
   rtf_colheader(
@@
     col_rel_width = c(3, 2, 2, 2)
-  ) %>%
-  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
+  ) |>
+  rtf_body(col_rel_width = c(3, 2, 2, 2)) |>
   rtf_encode() %>%
@@
 
-```{r, include=FALSE}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 rtf2pdf("tlf/intro-ae3.rtf")
+knitr::include_graphics("tlf/intro-ae3.pdf")
+```
+
+In `rtf_body()` and `rtf_colheader()`, 
+the `text_justification` argument is used to align text within the generated RTF table. 
+The default value is `"c"`, representing center justification. 
+However, you can customize the text justification by column using a character vector 
+with a length equal to the number of displayed columns.
+Here is a table displaying the possible inputs for the `text_justification` argument:
+
+```{r, echo = FALSE}
+r2rtf:::justification() |> 
+  select(1:2) |>
+  knitr::kable()
+```
+
+Below is an example to make the first column left-aligned and the rest columns center-aligned.
+
+```{r}
+head(tbl) |>
+  rtf_body(text_justification = c("l", "c", "c", "c")) |>
+  rtf_encode() |>
+  write_rtf("tlf/intro-ae5.rtf")
 ```
@@
 ```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
-knitr::include_graphics("tlf/intro-ae3.pdf")
-```
-
-In `rtf_*()` functions such as `rtf_body()`, `rtf_footnote()`,
-the `text_justification` argument is used to align text.
-Default is `"c"` for center justification.
-To vary text justification by column, use character vector with length of vector equals to
-number of columns displayed (e.g., `c("c", "l", "r")`).
-
-All possible inputs can be found in the table below.
-
-```{r}
-r2rtf:::justification()
-```
-
-Below is an example to make the first column left-aligned and center-aligned for the rest.
-
-```{r}
-head(tbl) %>%
-  rtf_body(text_justification = c("l", "c", "c", "c")) %>%
-  rtf_encode() %>%
-  write_rtf("tlf/intro-ae5.rtf")
-```
-
-```{r, include=FALSE}
 rtf2pdf("tlf/intro-ae5.rtf")
-```
-
-```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/intro-ae5.pdf")
@@
 
-In `rtf_*()` functions such as `rtf_body()`, `rtf_footnote()`, etc.,
-`border_left`, `border_right`, `border_top`, and `border_bottom` control cell borders.
+The `border_left`, `border_right`, `border_top`, and `border_bottom` arguments in the `rtf_body()` and `rtf_colheader()` functions are used to control the cell borders in the RTF table.
+If we want to remove the top border of `"Adverse Events"` in the header,
+we can change the default value `"single"` to `""` in the `border_top` argument.
+Below is an example to demonstrate the possibility of adding multiple column headers
+with proper border lines.
 
-If we want to remove the top border of `"Adverse Events"` in the header,
-we can change the default value `"single"` to `""` in the `border_top` argument, as shown below.
-
-r2rtf supports 26 different border types. The details can be found on
-the [r2rtf package website](https://merck.github.io/r2rtf/articles/rtf-row.html#border-type).
-
-In this example, we also demonstrate the possibility of adding multiple column headers.
+::: {.callout-note}
+ the r2rtf package supports 26 different border types, each offering unique border styles.
+For more details and examples regarding these border types, you can refer to the [r2rtf package website](https://merck.github.io/r2rtf/articles/rtf-row.html#border-type.
+:::
 
@@
     col_rel_width = c(3, 6)
-  ) %>%
+  ) |>
   rtf_colheader(
@@
     col_rel_width = c(3, 2, 2, 2)
-  ) %>%
+  ) |>
   rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
-  rtf_encode() %>%
+  rtf_encode() |>
   write_rtf("tlf/intro-ae7.rtf")
@@
 
-```{r, include=FALSE}
+```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 rtf2pdf("tlf/intro-ae7.rtf")
-```
-
-```{r, out.width = "100%", out.height = if (knitr::is_html_output()) "400px", echo = FALSE, fig.align = "center"}
 knitr::include_graphics("tlf/intro-ae7.pdf")
@@
 
-In the r2rtf R package [get started](https://merck.github.io/r2rtf/articles/r2rtf.html) page,
-there are more examples to illustrate how to customize
+The r2rtf R [package website](https://merck.github.io/r2rtf/articles/index.html) provides additional examples that demonstrate how to customize various aspects of the generated RTF tables.
+These examples cover topics such as customizing the title, subtitle, footnote, data source, and handling special characters within the table content.
 
-- title, subtitle
-- footnote, data source
-- special character
-- etc.
-
-Those features will be introduced when we first use them in the rest of the chapters.
+In the upcoming chapters of this book, we will introduce and explore these features as they become relevant to the specific use cases and scenarios discussed.
+By following along with the chapters, readers will gradually learn how to leverage these features to customize and enhance their RTF tables in real examples.

### 2023-07-09 preface.qmd
Break lines to 80 characters each and style code chunks

@@
 
-In the development of clinical trials, it is necessary to create and manage source code for generating and delivering Study Data Tabulation Model (SDTM), Analysis Dataset Model (ADaM) datasets, as well as tables, listings, and figures (TLFs).
-This is particularly evident in Phase 3 trials, where numerous TLFs are needed for submission.
-To effectively handle the large number of programs involved in such endeavors, it is essential to establish a consistent and well-defined folder structure for managing the analysis and reporting (A&R) project of a clinical trial.
+In the development of clinical trials, it is necessary to create and manage
+source code for generating and delivering Study Data Tabulation Model (SDTM),
+Analysis Dataset Model (ADaM) datasets, as well as tables, listings, and
+figures (TLFs). This is particularly evident in Phase 3 trials, where numerous
+TLFs are needed for submission. To effectively handle the large number of
+programs involved in such endeavors, it is essential to establish a consistent
+and well-defined folder structure for managing the analysis and reporting (A&R)
+project of a clinical trial.
 
-To streamline the organization of source code and documentation for a clinical trial A&R project, we suggest employing the R package folder structure.
-This folder structure is extensively utilized within the R community and is well-defined, often found in repositories like CRAN.
-By adopting this structure, you can benefit from a standardized and widely accepted framework for managing your A&R-related materials in an efficient and accessible manner.
+To streamline the organization of source code and documentation for a
+clinical trial A&R project, we suggest employing the R package folder structure.
+This folder structure is extensively utilized within the R community and is
+well-defined, often found in repositories like CRAN. By adopting this structure,
+you can benefit from a standardized and widely accepted framework for managing
+your A&R-related materials in an efficient and accessible manner.
 
-Using the R package folder structure provides a consistent approach that simplifies communication among developers, both within and across organizations.
+Using the R package folder structure provides a consistent approach that
+simplifies communication among developers, both within and across organizations.
 
-- For newcomers to R development, creating R packages is an essential step when sharing their work with others.
-The R community offers a widely adopted folder structure accompanied by excellent tutorials and free tools.
+- For newcomers to R development, creating R packages is an essential step
+  when sharing their work with others. The R community offers a widely adopted
+  folder structure accompanied by excellent tutorials and free tools.
 - For an experienced R developer, there is a minimal learning curve.
-- For an organization, adopting the R package folder structure simplifies the development of processes, tools, templates, and training.
-It enables the use of a unified folder structure for building and maintaining standardized tool and analysis projects.
+- For an organization, adopting the R package folder structure simplifies
+  the development of processes, tools, templates, and training.
+  It enables the use of a unified folder structure for building and
+  maintaining standardized tool and analysis projects.
 
-The workflow around an R package can also improve the traceability and reproducibility of an analysis project [@marwick2018packaging].
+The workflow around an R package can also improve the traceability and
+reproducibility of an analysis project [@marwick2018packaging].
 
-We will revisit the folder structure topic when discussing project management for a clinical trial project.
+We will revisit the folder structure topic when discussing project management
+for a clinical trial project.
 
-Additionally, the R package folder structure is also recommended for developing Shiny apps, as discussed in Chapter 20 of the [Mastering Shiny](https://mastering-shiny.org/scaling-packaging.html) book and the [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/golem.html) book.
+Additionally, the R package folder structure is also recommended for developing
+Shiny apps, as discussed in Chapter 20 of the
+[Mastering Shiny](https://mastering-shiny.org/scaling-packaging.html) book
+and the [Engineering Production-Grade Shiny Apps](https://engineering-shiny.org/golem.html) book.
 
@@
 
-This book is designed for intermediate-level readers who possess knowledge in both R programming and clinical development.
-Each part of the book makes certain assumptions about the readers' background:
+This book is designed for intermediate-level readers who possess knowledge
+in both R programming and clinical development. Each part of the book makes
+certain assumptions about the readers' background:
 
-- Part 1, titled "Delivering TLFs in CSR," provides general information and examples on creating tables, listings, and figures.
-It assumes that readers are individual contributors to a clinical project with prior experience in R programming.
-Familiarity with data manipulation in R is expected.
-Some recommended references for this part include [Hands-On Programming with R](https://rstudio-education.github.io/hopr/),
-  [R for Data Science](https://r4ds.had.co.nz/) and
-  [Data Manipulation with R](https://www.amazon.com/Data-Manipulation-R-Use/dp/0387747303).
+- Part 1, titled "Delivering TLFs in CSR", provides general information and
+  examples on creating tables, listings, and figures. It assumes that readers
+  are individual contributors to a clinical project with prior experience in
+  R programming. Familiarity with data manipulation in R is expected.
+  Some recommended references for this part include
+  [Hands-On Programming with R](https://rstudio-education.github.io/hopr/),
+  [R for Data Science](https://r4ds.had.co.nz/), and
+  [Data Manipulation with R](https://doi.org/10.1007/978-0-387-74731-6).
 
-- Part 2, titled "Clinical trial project," provides general information and examples on managing a clinical trial A&R project.
-It assumes that readers are project leads who have experience in R package development.
-Recommended references for this part include [R Packages](https://r-pkgs.org/) and the
+- Part 2, titled "Clinical trial project", provides general information and
+  examples on managing a clinical trial A&R project. It assumes that readers
+  are project leads who have experience in R package development.
+  Recommended references for this part include
+  [R Packages](https://r-pkgs.org/) and the
   [tidyverse style guide](https://style.tidyverse.org/).
 
-- Part 3, titled "eCTD submission package," provides general information on preparing submission packages related to the CSR in the electronic Common Technical Document (eCTD) format.
-It assumes that readers are project leads of clinical projects who possess experience in R package development and submission.
+- Part 3, titled "eCTD submission package", provides general information on
+  preparing submission packages related to the CSR in the
+  electronic Common Technical Document (eCTD) format.
+  It assumes that readers are project leads of clinical projects who possess
+  experience in R package development and submission.
 
@@
 
-We share the same philosophy described in
-the introduction of the [R Packages](https://r-pkgs.org/Introduction.html#intro-phil) book [@wickham2023r]
-, which we quote below:
+We share the same philosophy described in the introduction of the
+[R Packages](https://r-pkgs.org/introduction.html#sec-intro-phil)
+book [@wickham2023r], which we quote below:
 
@@
 This document is a collaborative effort maintained by a community.
-As you read through it, you also have the opportunity to contribute and enhance its quality.
-Your input and involvement play a vital role in shaping the excellence of this document.
+As you read through it, you also have the opportunity to contribute
+and enhance its quality. Your input and involvement play a vital role
+in shaping the excellence of this document.
 
-- Authors: made significant contributions to at least one chapter, constituting the majority of the content.
+- Authors: made significant contributions to at least one chapter,
+  constituting the majority of the content.
 
@@
 
-- Contributors: contributed at least one commit to the [source code](https://github.com/elong0527/r4csr).
+- Contributors: contributed at least one commit to the
+  [source code](https://github.com/elong0527/r4csr).
 

### 2023-07-09 index.qmd

@@
 The [ICH E3: Structure and Content of Clinical Study Reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf)
-offer comprehensive instructions to sponsors on the creation of a CSR.
-This book is a clear and straightforward guide on using R to streamline the process of preparing CSRs.
-Additionally, it provides detailed guidance on the submission process to regulatory agencies.
-Whether you are a beginner or an experienced R programmer, this book is an indispensable asset in your clinical reporting arsenal.
+offers comprehensive instructions to sponsors on the creation of a CSR.
+This book is a clear and straightforward guide on using R to streamline the
+process of preparing CSRs. Additionally, it provides detailed guidance on the
+submission process to regulatory agencies. Whether you are a beginner or an
+experienced R programmer, this book is an indispensable asset in your
+clinical reporting toolkit.
 
@@
     header = TRUE,
-             sep = "\t")
+    sep = "\t"
+  )
 )

### 2023-07-09 tlf-overview.qmd

@@
 
-Submitting clinical trial results to regulatory agencies is a crucial aspect of clinical development. 
-The [Electronic Common Technical Document (eCTD)](https://en.wikipedia.org/wiki/Electronic_common_technical_document) has emerged as the global standard format for regulatory submissions.
-For instance, the United States Food and Drug Administration (US FDA) [mandates the use of eCTD]((https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd)) for new drug applications and biologics license applications.
+Submitting clinical trial results to regulatory agencies is a crucial aspect
+of clinical development.
+The [Electronic Common Technical Document (eCTD)](https://en.wikipedia.org/wiki/Electronic_common_technical_document)
+has emerged as the global standard format for regulatory submissions.
+For instance, the United States Food and Drug Administration (US FDA)
+[mandates the use of eCTD]((https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd))
+for new drug applications and biologics license applications.
 
-A CSR provides comprehensive information about the methods and results of an individual clinical study.
-To support the statistical analysis, numerous tables, listings, and figures are included within the main text and appendices.
-As part of the CDISC pilot project, an [example CSR](https://github.com/cdisc-org/sdtm-adam-pilot-project/blob/master/updated-pilot-submission-package/900172/m5/53-clin-stud-rep/535-rep-effic-safety-stud/5351-stud-rep-contr/cdiscpilot01/cdiscpilot01.pdf) is also available for reference.
-If you seek additional examples of CSR, you can visit the clinical data website of the [European Medicines Agency (EMA) clinical data website](https://clinicaldata.ema.europa.eu/web/cdp/home).
+A CSR provides comprehensive information about the methods and results of an
+individual clinical study. To support the statistical analysis, numerous tables,
+listings, and figures are included within the main text and appendices.
+As part of the CDISC pilot project, an
+[example CSR](https://github.com/cdisc-org/sdtm-adam-pilot-project/blob/master/updated-pilot-submission-package/900172/m5/53-clin-stud-rep/535-rep-effic-safety-stud/5351-stud-rep-contr/cdiscpilot01/cdiscpilot01.pdf)
+is also available for reference. If you seek additional examples of CSR,
+you can visit the clinical data website of the
+[European Medicines Agency (EMA) clinical data website](https://clinicaldata.ema.europa.eu/web/cdp/home).
 
-The creation of CSR is a collaborative effort that involves various professionals such as clinicians, medical writers, statisticians, statistical programmers. 
-In this context, we will focus on the specific deliverables provided by statisticians and statistical programmers.
+The creation of CSR is a collaborative effort that involves various
+professionals such as clinicians, medical writers, statisticians,
+statistical programmers. In this context, we will focus on the specific
+deliverables provided by statisticians and statistical programmers.
 
-Within an organization, these professionals typically collaborate to define, develop, validate, and deliver the necessary tables, listings, and figures (TLFs) for a CSR.
-These TLFs serve to summarize the efficacy and/or safety of the pharmaceutical product under study.
-In the pharmaceutical industry, Microsoft Word is widely utilized for CSR preparation.
-As a result, the deliverables from statisticians and statistical programmers are commonly provided in formats such as `.rtf`, `.doc`, `.docx` to align with industry standards and requirements.
+Within an organization, these professionals typically collaborate to define,
+develop, validate, and deliver the necessary tables, listings,
+and figures (TLFs) for a CSR. These TLFs serve to summarize the efficacy
+and/or safety of the pharmaceutical product under study.
+In the pharmaceutical industry, Microsoft Word is widely utilized for
+CSR preparation. As a result, the deliverables from statisticians and
+statistical programmers are commonly provided in formats such as
+`.rtf`, `.doc`, `.docx` to align with industry standards and requirements.
 
-Our focus is to demonstrate the process of generating TLFs in RTF format, which is commonly employed in CSRs. The examples provided in this chapter adhere to the 
-[ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf) and the [FDA's PDF Specifications](https://www.fda.gov/media/76797/download).
+Our focus is to demonstrate the process of generating TLFs in RTF format,
+which is commonly employed in CSRs. The examples provided in this chapter
+adhere to the
+[ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf)
+and the [FDA's PDF Specifications](https://www.fda.gov/media/76797/download).
 
@@
 FDA's PDF specification is a general reference. Each organization can define
-more specific TLF format requirements that can be different from the examples in this book.
-The FDA's PDF specification serves as a general reference for formatting requirements.
-Each organization has the flexibility to define its own specific requirements for TLFs. 
-These specific format requirements may differ from the examples provided in this book.
-It is advisable to consult and adhere to the guidelines and specifications set by your respective organization when preparing TLFs for submission.
+more specific TLF format requirements that can be different from the examples
+in this book. The FDA's PDF specification serves as a general reference for
+formatting requirements. Each organization has the flexibility to define its
+own specific requirements for TLFs. These specific format requirements may
+differ from the examples provided in this book. It is advisable to consult
+and adhere to the guidelines and specifications set by your respective
+organization when preparing TLFs for submission.
 :::
@@
 
-In this project, we used publicly available CDISC pilot study data, which is accessible through the [CDISC GitHub repository](https://github.com/cdisc-org/sdtm-adam-pilot-project/tree/master/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).
+In this project, we used publicly available CDISC pilot study data, which is
+accessible through the
+[CDISC GitHub repository](https://github.com/cdisc-org/sdtm-adam-pilot-project/tree/master/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).
 
-To streamline the process, we have downloaded all the datasets from the repository and stored them in the [`data-adam/` folder](https://github.com/elong0527/r4csr/tree/main/data-adam) within this project.
-Additionally, we converted these datasets from the `.xpt` format to the `.sas7bdat` format for ease of use and compatibility.
-The dataset structure adheres to the CDISC [Analysis Data Model (ADaM)](https://www.cdisc.org/standards/foundational/adam) standard.
+To streamline the process, we have downloaded all the datasets from the
+repository and stored them in the
+[`data-adam/` folder](https://github.com/elong0527/r4csr/tree/main/data-adam)
+within this project. Additionally, we converted these datasets from the
+`.xpt` format to the `.sas7bdat` format for ease of use and compatibility.
+The dataset structure adheres to the CDISC
+[Analysis Data Model (ADaM)](https://www.cdisc.org/standards/foundational/adam)
+standard.
 
@@
 
-To exemplify the generation of TLFs in RTF format, we rely on the functionality provided by two R packages:
+To exemplify the generation of TLFs in RTF format, we rely on the functionality
+provided by two R packages:
 
-- [tidyverse](https://www.tidyverse.org/): preparation of datasets in a format suitable for reporting purposes.
-The tidyverse package offers a comprehensive suite of tools and functions for data manipulation and transformation, ensuring that the data is structured appropriately.
-- [r2rtf](https://merck.github.io/r2rtf/): creation RTF files. The r2rtf package offers functions specifically designed for generating RTF files, allowing us to produce TLFs in the desired format.
+- [tidyverse](https://www.tidyverse.org/): preparation of datasets in a format
+  suitable for reporting purposes. The tidyverse package offers a comprehensive
+  suite of tools and functions for data manipulation and transformation,
+  ensuring that the data is structured appropriately.
+- [r2rtf](https://merck.github.io/r2rtf/): creation RTF files.
+  The r2rtf package offers functions specifically designed for generating
+  RTF files, allowing us to produce TLFs in the desired format.
 
 ::: {.callout-note}
-There are indeed several other R packages available that can assist in creating TLFs in ASCII, RTF, and Word formats such as rtables, huxtable, pharmaRTF, gt, officer, and flextable.
-However, in this particular context, we will concentrate on demonstrating the concept using the r2rtf package.
-It is highly recommended for readers to explore and experiment with various R packages to identify the most suitable tools that align with their specific needs and objectives. 
+There are indeed several other R packages available that can assist in
+creating TLFs in ASCII, RTF, and Word formats such as rtables, huxtable,
+pharmaRTF, gt, officer, and flextable. However, in this particular context,
+we will concentrate on demonstrating the concept using the r2rtf package.
+It is highly recommended for readers to explore and experiment with various
+R packages to identify the most suitable tools that align with their
+specific needs and objectives.
 :::
@@
 
-The tidyverse is a comprehensive collection of R packages that aim to simplify the workflow of manipulating, visualizing, and analyzing data in R.
-These packages adhere to the principles outlined in the [the tidy tools manifesto](https://tidyverse.tidyverse.org/articles/manifesto.html) and offer user-friendly interfaces for interactive data analysis.
+The tidyverse is a comprehensive collection of R packages that aim to
+simplify the workflow of manipulating, visualizing, and analyzing data in R.
+These packages adhere to the principles outlined in the
+[the tidy tools manifesto](https://tidyverse.tidyverse.org/articles/manifesto.html)
+and offer user-friendly interfaces for interactive data analysis.
 
-The creators of the tidyverse, Posit, have provided exceptional [cheatsheets](https://posit.co/resources/cheatsheets/)
-and [tutorials](https://github.com/rstudio-education/remaster-the-tidyverse) that serve as valuable resources for learning and mastering the functionalities of these packages.
+The creators of the tidyverse, Posit, have provided exceptional
+[cheatsheets](https://posit.co/resources/cheatsheets/)
+and [tutorials](https://github.com/rstudio-education/remaster-the-tidyverse)
+that serve as valuable resources for learning and mastering the
+functionalities of these packages.
 
-Furthermore, there are several books available that serve as introductions to the tidyverse. 
-For example:
+Furthermore, there are several books available that serve as introductions
+to the tidyverse. For example:
 
@@
 ::: {.callout-note}
-In this book, we assume that the reader already has experience in using the tidyverse.
-This prior knowledge and familiarity with the tidyverse tools enable a more efficient and focused exploration of the concepts presented throughout the book.
+In this book, we assume that the reader already has experience in using the
+tidyverse. This prior knowledge and familiarity with the tidyverse tools
+enable a more efficient and focused exploration of the concepts presented
+throughout the book.
 :::
@@
 
-r2rtf is an R package specifically designed to create production-ready tables and figures in RTF format.
+r2rtf is an R package specifically designed to create production-ready
+tables and figures in RTF format.
 
-- provide simple "verb" functions that correspond to each component of a table,
-  to help you translate a data frame to a table in an RTF file;
-- enable pipes (`|>`);
-- focus on the **table format** only. Data manipulation and analysis tasks can be handled by other R packages like the tidyverse
+- Provide simple "verb" functions that correspond to each component of a table,
+  to help you translate a data frame to a table in an RTF file.
+- Enable pipes (`|>`).
+- Focus on the **table format** only. Data manipulation and analysis tasks
+  can be handled by other R packages like the tidyverse.
 
@@
 
-We provide a concise introduction to r2rtf and demonstrate how to convert data frames into TLFs. 
-For more comprehensive examples and additional features, we encourage readers to explore the 
+We provide a concise introduction to r2rtf and demonstrate how to convert
+data frames into TLFs. For more comprehensive examples and additional features,
+we encourage readers to explore the
 [r2rtf package website](https://merck.github.io/r2rtf/articles/index.html).
 
-To illustrate the basic usage of the r2rtf package, we will work with the "r2rtf_adae" dataset, available within the r2rtf package. 
+To illustrate the basic usage of the r2rtf package, we will work with the
+"r2rtf_adae" dataset, available within the r2rtf package.
 This dataset contains information on adverse events (AEs) from a clinical trial,
@@
 ::: {.callout-note}
-Additional information about these variables can be found on the help page of the dataset, 
-which can be accessed by using the command `?r2rtf_adae` in R.
+Additional information about these variables can be found on the help page
+of the dataset, which can be accessed by using the command `?r2rtf_adae` in R.
 :::
@@
 
-To manipulate the data and create a data frame containing the necessary information for the RTF table, we can use the dplyr and tidyr packages within the tidyverse.
+To manipulate the data and create a data frame containing the necessary
+information for the RTF table, we can use the dplyr and tidyr packages
+within the tidyverse.
 
@@
 
-Having prepared the dataset `tbl`, 
-we can now proceed with constructing the final RTF table using the r2rtf package. 
-The r2rtf package has various functions, each designed for a specific type of table layout. Some commonly used verbs include:
+Having prepared the dataset `tbl`, we can now proceed with constructing the
+final RTF table using the r2rtf package.
+The r2rtf package has various functions, each designed for a specific type
+of table layout. Some commonly used verbs include:
 
@@
 
-Functions provided by the r2rtf package are designed to work seamlessly with the pipe operator (|>).
-This allows for a more concise and readable code structure, enhancing the efficiency of table creation in RTF format.
+Functions provided by the r2rtf package are designed to work seamlessly
+with the pipe operator (`|>`). This allows for a more concise and readable
+code structure, enhancing the efficiency of table creation in RTF format.
 A full list of functions in the r2rtf package can be found in the
@@
 
-Here is a minimal example that demonstrates how to combine functions using pipes to create an RTF table
+Here is a minimal example that demonstrates how to combine functions using
+pipes to create an RTF table.
 
@@
 - `rtf_encode()` transfers table layout information into RTF syntax.
-- `write_rtf()` save RTF encoding into a file with file extension `.rtf`
+- `write_rtf()` save RTF encoding into a file with file extension `.rtf`.
 
@@
 
-In the previous example, we may want to add more column space to the first column.
-We can achieve the goal by updating the `col_rel_width argument` in the rtf_body() function.
+In the previous example, we may want to add more column space to the first
+column. We can achieve the goal by updating the `col_rel_width argument`
+in the `rtf_body()` function.
 
-In the example below, the `col_rel_width` argument expects a vector with the same length as the number of columns in the table `tbl`.
-This vector defines the relative width of each column within a predetermined total column width.
-Here, the relative width is defined as `3:2:2:2` that allow us to allocate more space to specific columns.
+In the example below, the `col_rel_width` argument expects a vector with
+the same length as the number of columns in the table `tbl`.
+This vector defines the relative width of each column within a
+predetermined total column width.
+Here, the relative width is defined as `3:2:2:2` that allow us to
+allocate more space to specific columns.
 
@@
 Only the ratio of the `col_rel_width` values is considered.
-Therefore, using `col_rel_width = c(6, 4, 4, 4)` or `col_rel_width = c(1.5, 1, 1, 1)` would yield equivalent results, as they maintain the same ratio.
+Therefore, using `col_rel_width = c(6, 4, 4, 4)` or
+`col_rel_width = c(1.5, 1, 1, 1)` would yield equivalent results,
+as they maintain the same ratio.
 :::
@@
 
-In the previous example, we encountered a misalignment issue with the column header. 
-To address this, we can use the `rtf_colheader()` 
-function to adjust column header width and provide more informative column headers.
+In the previous example, we encountered a misalignment issue with the
+column header. To address this, we can use the `rtf_colheader()`
+function to adjust column header width and provide more informative
+column headers.
 
-Within the `rtf_colheader()` function, the `colheader` argument is used to specify the content of the column header. 
-The columns are separated using the "|" symbol.
-In the following example, we define the column header as `"Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose"`, representing the four columns in the table:
+Within the `rtf_colheader()` function, the `colheader` argument is used to
+specify the content of the column header.
+The columns are separated using the `|` symbol.
+In the following example, we define the column header as
+`"Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose"`,
+representing the four columns in the table:
 
@@
 
-In `rtf_body()` and `rtf_colheader()`, 
-the `text_justification` argument is used to align text within the generated RTF table. 
+In `rtf_body()` and `rtf_colheader()`, the `text_justification` argument is
+used to align text within the generated RTF table.
 The default value is `"c"`, representing center justification.
-However, you can customize the text justification by column using a character vector 
-with a length equal to the number of displayed columns.
-Here is a table displaying the possible inputs for the `text_justification` argument:
+However, you can customize the text justification by column using a character
+vector with a length equal to the number of displayed columns.
+Here is a table displaying the possible inputs for the `text_justification`
+argument:
 
@@
 
-Below is an example to make the first column left-aligned and the rest columns center-aligned.
+Below is an example to make the first column left-aligned and the rest columns
+center-aligned.
 
@@
 
-The `border_left`, `border_right`, `border_top`, and `border_bottom` arguments in the `rtf_body()` and `rtf_colheader()` functions are used to control the cell borders in the RTF table.
+The `border_left`, `border_right`, `border_top`, and `border_bottom` arguments
+in the `rtf_body()` and `rtf_colheader()` functions are used to control the
+cell borders in the RTF table.
 If we want to remove the top border of `"Adverse Events"` in the header,
 we can change the default value `"single"` to `""` in the `border_top` argument.
-Below is an example to demonstrate the possibility of adding multiple column headers
-with proper border lines.
+Below is an example to demonstrate the possibility of adding multiple column
+headers with proper border lines.
 
 ::: {.callout-note}
- the r2rtf package supports 26 different border types, each offering unique border styles.
-For more details and examples regarding these border types, you can refer to the [r2rtf package website](https://merck.github.io/r2rtf/articles/rtf-row.html#border-type.
+ the r2rtf package supports 26 different border types, each offering unique
+ border styles. For more details and examples regarding these border types,
+ you can refer to the
+ [r2rtf package website](https://merck.github.io/r2rtf/articles/rtf-row.html#border-type).
 :::
@@
 
-The r2rtf R [package website](https://merck.github.io/r2rtf/articles/index.html) provides additional examples that demonstrate how to customize various aspects of the generated RTF tables.
-These examples cover topics such as customizing the title, subtitle, footnote, data source, and handling special characters within the table content.
+The r2rtf R [package website](https://merck.github.io/r2rtf/articles/index.html)
+provides additional examples that demonstrate how to customize various aspects
+of the generated RTF tables. These examples cover topics such as customizing
+the title, subtitle, footnote, data source, and handling special characters
+within the table content.
 
-In the upcoming chapters of this book, we will introduce and explore these features as they become relevant to the specific use cases and scenarios discussed.
-By following along with the chapters, readers will gradually learn how to leverage these features to customize and enhance their RTF tables in real examples.
+In the upcoming chapters of this book, we will introduce and explore these
+features as they become relevant to the specific use cases and scenarios
+discussed. By following along with the chapters, readers will gradually
+learn how to leverage these features to customize and enhance their RTF
+tables in real examples.

### 2023-07-16 tlf-disposition.qmd
Migrate from bookdown to Quarto book (working HTML book)
update tlf-disposition.qmd

--- 2022-12-07 tlf-disposition.Rmd
@@
-# Disposition {#disposition}
+# Disposition {#sec-disposition}
+
+```{r, include=FALSE}
+source("_common.R")
+```
 
@@
 
-The disposition of participants table reports the numbers of participants who were randomized,
-and who entered and completed each phase of the study.
-In addition, the reasons for all post-randomization discontinuations,
-grouped by treatment and by major reason (lost to follow-up, adverse event, poor compliance, etc.) are reported.
-
-```{r, include=FALSE}
-source("common.R")
-```
+The participant disposition table provides an overview of the study's participant numbers, 
+detailing those who were randomized and their progression through each phase. 
+It also presents the reasons for discontinuation after randomization, 
+categorized by treatment and major factors such as 
+lost to follow-up, adverse events, poor compliance, and others.
 
@@
 
-In this chapter, we show how to create a typical disposition table.
+This chapter provides a step-by-step guide on creating a simple disposition table in RTF format.
+The expected output of the table is as follows:
 
@@
 
-The first step is to read in the relevant datasets into R.
-For a disposition table, all the required information is saved in a Subject-level Analysis Dataset (ADSL).
-This dataset is provided in `sas7bdat` format, which is a SAS data format currently used in many clinical trial analysis and reporting.
-The `haven` package is able to read the dataset, while maintaining its attributes (e.g., variable labels).
+## Input data
+
+The initial step involves reading the necessary datasets into R. 
+In the case of a disposition table, 
+the required information is typically stored in a Subject-level Analysis Dataset (ADSL). 
+The dataset provided is in `sas7bdat` format, commonly used in clinical trial analysis and reporting. 
+To read the dataset while preserving its attributes, such as variable labels, 
+we can use the `read_sas` function in haven package.
 
@@
 
-The following variables are used in the preparation of a simplified disposition of participants table:
+To prepare a simplified disposition of participants table using the `adsl` dataset, the following variables are utilized:
 
-- USUBJID: unique subject identifier
-- TRT01P: planned treatment
-- TRT01PN: planned treatment numeric encoding
-- DISCONFL: discontinued from study flag
-- DCREASCD: discontinued from study reason coded
+- `USUBJID`: unique subject identifier
+- `TRT01P`: planned treatment
+- `TRT01PN`: planned treatment numeric encoding
+- `DISCONFL`: discontinued from study flag
+- `DCREASCD`: discontinued from study reason
 
@@
 
-In the code below, we calculate the number of participants in the analysis population by treatment arms.
+## Output data
+
+The objective of this section is to create a dataset suitable for generating the expected report. 
+This process will consist of several steps to generate each component of the reporting dataset.
+
+we can calculate the number of participants 
+in the analysis population for each treatment arm by using following code.
 
@@
 ```
+
+Next, we calculate the number of participants who have been discontinued from the study.
 
@@
 
-In the code below, we calculate the number and percentage of
-participants who completed/discontinued the study for different reasons by treatment arms.
+Additionally, we calculate the number and percentage of participants
+who completed or discontinued the study, categorized by treatment arms and different reasons.
 
@@
 
-In the code below, we calculate the number and percentage of participants who complete the study by treatment arms.
-We split `n_reason` because we want to customize the row order of the table.
+In the following code, we compute the number and percentage of participants 
+who successfully complete the study for each treatment arm. 
+The splitting of `n_reason` is done to allow for customizing the row order within the resulting table.
 
@@
 
-In the code below, we calculate the numbers and percentages of participants who discontinued the study for different reasons by treatment arms.
-For display purpose, `paste0("    ", row)` is used to add leading spaces to produce indentation in the final report.
+The following code calculates the numbers and percentages of participants 
+who discontinued the study for various reasons, categorized by treatment arms. 
+To achieve indentation in the final report, 
+the code `paste0("    ", row)` is employed to add leading spaces.
 
@@
 
-Now we combine individual rows into one table for reporting purpose.
-`tbl_disp` is used as input for `r2rtf` to create final report.
+To summarize the data manipulation steps, 
+we consolidate individual rows into a single table for reporting purposes. 
+This table, named `tbl_disp`, 
+serves as the input for generating the final report using the `r2rtf` package.
 
@@
 
-In the below code, formatting of the final table is defined.
-Items that were not discussed in the previous sections, are highlighted below.
+## Reporting
 
-The `rtf_title` defines table title.
-We can provide a vector for the title argument.
-Each value is a separate line.
-The format can also be controlled by providing a vector input in text format.
+By using `tbl_disp` as input, we create generate the disposition table using following code.
+The `r2rtf` elements that were not covered in the previous sections are highlighted.
+Here, the `rtf_title` function is utilized to define the title of the table. 
+For the `title` argument, 
+you can provide a vector where each value represents a separate line.
 
@@
 ```
-
-The procedure to generate a disposition table can be summarized as follows:
-
-- Step 1: Read subject level data (i.e., `adsl`) into R.
-- Step 2: Count participants in the analysis population and name the dataset `n_rand`.
-- Step 3: Calculate the number and percentage of participants who discontinued the study by treatment arm, and name the dataset `n_disc`.
-- Step 4: Calculate the numbers and percentages of participants who discontinued the study for different reasons by treatment arm,
-  and name the dataset `n_reason`.
-- Step 5: Calculate the number and percentage of participants who completed the study by treatment arm, and name the dataset `n_complete`.
-- Step 6: Bind `n_rand`, `n_disc`, `n_reason`, and `n_complete` by row.
-- Step 7: Write the final table to RTF

### 2023-07-16 tlf-disposition.qmd
update `tlf-population.qmd`

@@
 
-To prepare a simplified disposition of participants table using the `adsl` dataset, the following variables are utilized:
+To prepare a simplified disposition of participants table using the `adsl` dataset, the following variables are used:
 
@@
 
-By using `tbl_disp` as input, we create generate the disposition table using following code.
+By using `tbl_disp` as input, we can generate the disposition table using following code.
 The `r2rtf` elements that were not covered in the previous sections are highlighted.

### 2023-07-16 tlf-population.qmd
update `tlf-population.qmd`

@@
 
-In this chapter, we illustrate how to create a summary table for
-the analysis population for a study.
+This chapter provides a step-by-step guide on creating a simple table for
+the analysis population of a study in RTF format.
+The expected output of the table is as follows:
 
@@
 
-The first step is to read relevant datasets into R. For the analysis
-population table, all the required information is saved in the ADSL
-dataset. We can use the `haven` package to read the dataset.
+## Input data
+
+The initial step involves reading the ADSL datasets into R.
 
@@
 
-We illustrate how to prepare a report data for a simplified analysis
-population table using variables below:
+the following variables are utilized:
 
@@
 
-Before we write the analysis code, let's discuss the possibility of reusing R
-code by writing helper functions.
+Before we writing the analysis code, 
+let's explore the possibility of maximizing 
+code reuse in R by creating helper functions.
 
-As discussed in [R for data
+::: {.callout-note}
+As emphasized in the book [R for data
 science](https://r4ds.had.co.nz/functions.html#when-should-you-write-a-function),
@@
 pasted a block of code more than twice".
+:::
 
-In @sec-disposition, there are a few repeating steps to:
+In @sec-disposition, there are a few repetitive steps involved:
 
-- Format the percentages using the `formatC()` function.
-- Calculate the numbers and percentages by treatment arm.
+- Formatting the percentages using the `formatC()` function.
+- Calculating  the numbers and percentages by treatment arm.
 
@@
 
-To format numbers and percentages, we create a function called
-`fmt_num()`. It is a very simple function wrapping `formatC()`.
+To address this, we introduce two ad-hoc functions that 
+will assist us in creating tables throughout the remainder of this book.
+
+The first function, named `fmt_num()`, is a straightforward implementation that 
+wraps `formatC()` to format numbers and percentages.
 
@@
 
-The main reason to create the `fmt_num()` function is to enhance the
-readability of the analysis code.
+One purpose of creating the `fmt_num()` function 
+is to improve the readability of the analysis code.
 
-For example, we can compare the two versions of code to format the
-percentage used in @sec-disposition and `fmt_num()`.
+To illustrate this, let's compare two versions of code for formatting percentages: 
+the one used in @sec-disposition and the alternative approach using fmt_num().
 
@@
 
-The function can be enhanced in multiple ways, but here we only focus on
-simplicity and readability. More details about writing R functions can
-be found in the [STAT 545
+To compute the numbers and percentages of participants in different groups,
+we offer a straightforward (albeit not robust) 
+wrapper function called `count_by()`. 
+This function depends on the dplyr and tidyr packages.
+
+::: {.callout-note}
+While there are several ways to improve the function, 
+our current focus is on enhancing simplicity and readability. 
+A good tutorial can be found from the  [STAT 545
 course](https://stat545.com/functions-part1.html).
+:::
 
@@
 
-By using the `count_by()` function, we can simplify the analysis code as
-below.
+The `count_by()` function help us simplify the analysis code:
 
@@
 
-## Analysis code
+## Output data
 
-With the helper function `count_by`, we can easily prepare a report
-dataset as
+The objective of this section is to create a dataset suitable for generating the expected report. 
+This process will consist of several steps to generate each component of the reporting dataset.
+
+With the aid of the helper function `count_by()`, 
+preparing a report dataset becomes a straightforward task.
 
@@
 
-Now we combine individual rows into one table for reporting
-purpose. `tbl_pop` is used as input for r2rtf to create the final
-report.
-
 ```{r}
@@
 
-We define the format of the output using code below.
+## Reporting
+
+By using `tbl_pop` as input, we can generate the analysis population table using following code.
 
@@
 ```
-
-The procedure to generate an analysis population table can be summarized
-as follows:
-
-- Step 1: Read data (i.e., `adsl`) into R.
-- Step 2: Bind the counts/percentages of the ITT population, the
-  efficacy population, and the safety population by row using the
-  `count_by()` function.
-- Step 3: Format the output from Step 2 using r2rtf.

### 2023-07-16 tlf-baseline.qmd
improve `tlf-baseline.qmd`

@@
 
-In this chapter, we illustrate how to create a simplified
-baseline characteristics table for a study.
+This chapter provides a step-by-step guide on creating a simple 
+baseline characteristics table in RTF format.
+The expected output of the table is as follows:
 
@@
 ```{r}
-library(table1)
-library(r2rtf)
 library(haven)
@@
 library(tools)
+library(table1)
+library(r2rtf)
 ```
 
-As in previous chapters, we first read the `adsl` dataset that contains all
-the required information for the baseline characteristics table.
+## Input data
+
+We begin by reading the `adsl` dataset, 
+which comprises all the necessary information for generating the baseline characteristics table.
 
@@
 
-For simplicity, we only analyze `SEX`, `AGE` and, `RACE` in this example
-using the table1 R package. More details of the table1 R package can
+For simplicity, we only analyze `SEX`, `AGE` and, `RACE` in this example.
+
+## Output data
+
+The table1 R package directly creates an HTML report for the baseline characteristic table.
+More details of the table1 R package can
 be found in the package
 [vignettes](https://benjaminrich.github.io/table1/vignettes/table1-examples.html).
-
-The table1 R package directly creates an HTML report.
 
@@
 
-The code below transfer the output into a dataframe
-that only contains ASCII characters
-recommended by regulatory agencies.
-`tbl_base` is used as input for r2rtf to create the final report.
+let's transfer the output into a dataframe that 
+exclusively consists of ASCII characters, as recommended by regulatory agencies.
 
@@
 
-We define the format of the output. We highlight items that
-are not discussed in previous discussion.
+## Reporting
 
-`text_indent_first` and `text_indent_left` are used to control the indent
-space of text. They are helpful when you need to control the white space
-of a long phrase, "AMERICAN INDIAN OR ALASKA NATIVE"
-in the table provides an example.
+By using `tbl_base` as the input, we can generate the baseline characteristics table.
+
+In the code below, `text_indent_first` and `text_indent_left`
+are used to control the indentation space of the text. 
+These parameters come in handy when you need to manage the whitespace within a lengthy phrase.
 
@@
 ```
-
-In conclusion, the procedure to generate demographic and baseline characteristics table is summarized as follows:
-
-- Step 1: Read the data set.
-- Step 2: Use `table1::table1()` to get the baseline characteristics table.
-- Step 3: Transfer the output from Step 2 into a data frame that only contains ASCII characters.
-- Step 4: Define the format of the RTF table by using the R package r2rtf.

### 2023-07-16 tlf-baseline.qmd
add package use case

@@
 ```{r}
-library(haven)
-library(dplyr)
-library(tidyr)
-library(stringr)
-library(tools)
-library(table1)
-library(r2rtf)
+library(haven)   # Read SAS data
+library(dplyr)   # Manipulate data
+library(tidyr)   # Manipulate data
+library(stringr) # Manipulate string
+library(tools)   # Manipulate string
+library(table1)  # Baseline characteristics table
+library(r2rtf)   # Reporting in RTF format
 ```

### 2023-07-29 slides/r4csr-gwu.Rmd
Update or remove obsolete URLs

@@
   - [SafetyGraphics](https://github.com/SafetyGraphics/safetyGraphics)
-  - [forestly](https://elong0527.github.io/forestly/articles/forestly.html): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_206/)
+  - [forestly](https://merck.github.io/forestly/articles/forestly.html): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_206/)
 
@@
   - [Posit Package Manager](https://packagemanager.posit.co/)
-  - [CRAN Time Machine by MRAN](https://mran.microsoft.com/timemachine)
+  - CRAN Time Machine by MRAN
 - Flexibility of input and output path:

### 2023-07-29 slides/r4csr-psi.Rmd
Update or remove obsolete URLs

@@
   - [metalite.table1](https://elong0527.github.io/metalite.table1/articles/metalite-table1.html): Interactive demographic table.
-  - [forestly](https://elong0527.github.io/forestly/articles/example.html#construct-interactive-forest-plot-1): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_206/)
+  - [forestly](https://merck.github.io/forestly/articles/forestly.html): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_206/)
 

### 2023-07-29 slides/r4csr-rstudio.Rmd
Built site for gh-pages
Update or remove obsolete URLs

@@
   - [SafetyGraphics](https://github.com/SafetyGraphics/safetyGraphics)
-  - [forestly](https://elong0527.github.io/forestly/articles/example.html#construct-interactive-forest-plot-1): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_206/)
+  - [forestly](https://merck.github.io/forestly/articles/forestly.html): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_206/)
 

### 2023-07-29 tlf-overview.qmd

@@
 For instance, the United States Food and Drug Administration (US FDA)
-[mandates the use of eCTD]((https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd))
+[mandates the use of eCTD](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd)
 for new drug applications and biologics license applications.

### 2024-02-20 slides/china-r.Rmd
Update R submission site URL

@@
 - R-based submission pilots to FDA:
-  - <https://rconsortium.github.io/r-submission-site/>
+  - <https://rconsortium.github.io/submissions-wg/>
   - Focus on improving practices of R-based clinical trial regulatory submission.

### 2024-02-22 _common.R
Customize data frame and tibble printing methods

@@ -13,3 +13,25 @@ rtf2pdf <- function(input) {
   z <- paste(x, y, sep = " && ")
   if (Sys.getenv("GITHUB_ACTIONS") != "") system(z) else invisible(NULL)
 }
+
+# Customize data frame and tibble printing methods.
+# See <https://github.com/elong0527/r4csr/issues/115> for details.
+knit_print.data.frame <- function(x, ...) {
+  paste(capture.output(base::print.data.frame(head(x, 4))), collapse = "\n")
+}
+
+registerS3method(
+  "knit_print", "data.frame", knit_print.data.frame,
+  envir = asNamespace("knitr")
+)
+
+options(pillar.advice = FALSE)
+
+knit_print.tbl_df <- function(x, ...) {
+  paste(capture.output(print(x, n = 4)), collapse = "\n")
+}
+
+registerS3method(
+  "knit_print", "tbl_df", knit_print.tbl_df,
+  envir = asNamespace("knitr")
+)

### 2024-02-22 _common.R
Add "x more rows" footer for printing data frames

@@ -16,8 +16,15 @@ rtf2pdf <- function(input) {
 
 # Customize data frame and tibble printing methods.
 # See <https://github.com/elong0527/r4csr/issues/115> for details.
+more_rows <- function(x, n) if (nrow(x) <= n) NULL else paste0("# ℹ ", nrow(x) - n, " more rows")
 knit_print.data.frame <- function(x, ...) {
-  paste(capture.output(base::print.data.frame(head(x, 4))), collapse = "\n")
+  paste(
+    c(
+      capture.output(base::print.data.frame(head(x, 4))),
+      more_rows(x, 4)
+    ),
+    collapse = "\n"
+  )
 }
 
 registerS3method(

### 2024-02-23 tlf-overview.qmd

@@
 ```{r}
-r2rtf_adae |>
-  select(USUBJID, TRTA, AEDECOD) |>
-  head(4)
+r2rtf_adae |> select(USUBJID, TRTA, AEDECOD)
 ```
@@
 
-tbl %>% head(4)
+tbl
 ```
@@
 ```{r}
-head(tbl) |>
+tbl |>
+  head() |>
   rtf_body() |>
@@
 ```{r}
-head(tbl) |>
+tbl |>
+  head() |>
   rtf_body(col_rel_width = c(3, 2, 2, 2)) |>
@@
 ```{r}
-head(tbl) |>
+tbl |>
+  head() |>
   rtf_colheader(
@@
 ```{r}
-head(tbl) |>
+tbl |>
+  head() |>
   rtf_body(text_justification = c("l", "c", "c", "c")) |>
@@
 ```{r}
-head(tbl) %>%
+tbl |>
+  head() |>
   rtf_colheader(

### 2024-04-07 tex/before_body.tex
Change PDF output template and fix preview images

\null\vfill
\begin{flushleft}
\thispagestyle{empty}
\textit{R for Clinical Study Reports and Submission}

© Placeholder Name, Inc.

ISBN-1234567891234

\noindent Lorem ipsum dolor sit amet, consectetur adipiscing elit.
Nulla et elementum libero. In hac habitasse platea dictumst.
Vestibulum ante ipsum primis in faucibus orci luctus et ultrices
posuere cubilia Curae; Donec sed odio dui. Nullam quis risus eget
urna mollis ornare vel eu leo. Pellentesque habitant morbi tristique
senectus et netus et malesuada fames ac turpis egestas.
Curabitur blandit tempus porttitor.
Integer posuere erat a ante venenatis dapibus posuere velit aliquet.
\end{flushleft}

\frontmatter

% Dedication
\begin{center}
    \thispagestyle{empty}
    \vspace*{\fill}
    \huge{\textit{Placeholder for dedication text}}
    \vspace*{\fill}
\end{center}

\mainmatter
